### 2017-03-15 main.tex
A starting point

\documentclass[letterpaper]{article}
\usepackage{proceed2e}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,epsfig,booktabs}

% Set the typeface to Times Roman
\usepackage{times}

\title{Quotient Normalized Maximum Likelihood Criterion for Learning Bayesian Network Structures}

\author{} % LEAVE BLANK FOR ORIGINAL SUBMISSION.
          % UAI  reviewing is double-blind.

% The author names and affiliations should appear only in the accepted paper.
%
%\author{ {\bf Harry Q.~Bovik\thanks{Footnote for author to give an
%alternate address.}} \\
%Computer Science Dept. \\
%Cranberry University\\
%Pittsburgh, PA 15213 \\
%\And
%{\bf Coauthor}  \\
%Affiliation          \\
%Address \\
%\And
%{\bf Coauthor}   \\
%Affiliation \\
%Address    \\
%(if needed)\\
%}

% Example definitions.
% --------------------
%\def\x{{\mathbf x}}
%\def\L{{\cal L}}

\hyphenation{col-umns}
\hyphenation{Bayes-ian}
\hyphenation{oth-er}

\begin{document}

\maketitle

\begin{abstract}
Learning the dependency structure of a multivariate distribution from
the observational data is an important task since it allows us to
speculate about the underlying causal mechanisms that induce
dependencies. This learning task can be approached as a model
selection problem, but recent studies have revealed that the popular
Bayesian model selection criterion is not satisfactory. We review some
of the information theoretic alternatives and introduce a new one
called a quotient normalized maximum likelihood criterion.
\end{abstract}

\section{INTRODUCTION}
\label{sec:intro}
Bayesian networks~\cite{Pear88} are popular models for presenting
multivariate statistical dependencies that may have been induced by
underlying causal mechanisms.  Techniques for learning the structure
of Bayesian networks from the observational data has therefore been
used for many tasks such as discovering cell signaling pathways from
the protein activity data~\cite{bn4sigpath02}, revealing the business
process structures~\cite{bn4bpmining} from the transaction logs and
modeling brain region connectivity using fMRI
data~\cite{bn4brainconnect}.

Learning the structure of statistical dependencies can be seen as a
model selection task where each model is a different hypothesis about
the conditional dependencies between sets of variables. Traditional
model selection criteria such as the Akaike information
criterion~\cite{Akai73} and the Bayesian information
criterion~\cite{Schw78} have also been used for the task, but recent
comparisons have not been favorable for these criteria in terms of
structural stability and/or predictive
performance~\cite{cosco.pgm08a}.  The most popular criterion has been
the marginal likelihood (usually called BDeu for reasons explained
later), but recent studies~\cite{cosco.uai07,Steck08} have found this
criterion to be very sensitive to hyper parameters and to yield
undesirably complex models for small sample sizes.

The information theoretic normalized maximum likelihood (NML)
criterion~\cite{Shta87,Riss96a} would otherwise be an ideal candidate
for a good criterion.  but its exact calculation is likely to be
prohibitively demanding. In 2008, Silander et al. introduced a
hyper-parameter free, NML inspired criterion called a factorized NML
(fNML)~\cite{cosco.pgm08a} that was shown to yield good predictive
models without sensitivity problems.  However, from the structure
learning point of view, the fNML still sometimes appears to yield
overly complex models. In this paper we introduce another NML related
criterion, a quotient NML (qNML) that yields simpler models without
sacrificing predictive accuracy. Furthermore, unlike the fNML, the
qNML gives equal scores to the structures that encode the same
independence statements. Like other common model selection criteria,
the qNML is also consistent.

We will next briefly introduce Bayesian networks and then review the
BDeu and fNML criteria and introduce the qNML criterion.  We will also
summarize the results for 20 datasets to back up our claim of qNML
yielding parsimonious models with good predictive capabilities.

\section{BAYESIAN NETWORKS}
\label{sec:bns}

Bayesian networks are a general way to describe the dependencies
between the components of an $n$\nobreakdash-dimensional discrete data
vector $X=(X_{1},\ldots,X_{n})$ in which the component $X_{i}$ may
take any of the discrete values in a set $\{1,\ldots,r_{i}\}$.
Despite denoting the values with small integers, the model will treat
the components of $X$ categorical.

\subsection{Likelihood}
\label{ssec:likelihood}

Bayesian network $B=(G,\theta)$ defines a probability distribution for
$X$. The component $G$ defines the structure of the model as a
directed acyclic graph that has exactly one node for each component of
$X$. The structure $G=(G_{1},\ldots,G_{n})$ defines for each
variable/node $X_{i}$ its (possibly empty) parent set $G_{i}$, i.e.,
the nodes from which there are a directed edges to the variable
$X_{i}$.

Given a realization $x$ of $X$, we denote the sub\nobreakdash-vector
of $x$ that consists of the values of the parents of $X_{i}$ in $x$ as
$G_{i}(x)$. It is customary to enumerate all the possible
sub\nobreakdash-vectors $G_{i}(X)$ from $1$ to $q_{i}=\prod_{h\in
  G_{i}}r_{h}.$ In case $G_{i}$ is empty, we define $q_{i}=1$ and
$P(G_{i}(x)=1)=1$ for all vectors $x$.

For each variable $X_{i}$ there is a $q_{i}\times r_{i}$ table
$\theta_{i}$ of parameters whose $j^{\textnormal{th}}$ row and
$k^{\textnormal{th}}$ column defines the conditional probability
$P(X_{i}=k\mid G_{i}(X)=j;\theta)=\theta_{ijk}$.  With structure $G$
and parameters $\theta$, we can now express the likelihood function of
the model as
\begin{equation}
P(x|G,\theta)=\prod_{i=1}^{n}P(x_{i}\mid G_{i}(x);\theta_{i})=\prod_{i=1}^{n}\theta_{iG_{i}(x)x_{i}}.
\end{equation}

\subsection{Bayesian structure learning}

Bayesian learning of Bayesian network structures is based on the
posterior probability $P(G|D,\alpha)$, where $\alpha$ denotes the
hyper\nobreakdash-parameters for the model parameters $\theta$, and
the $D$ is a collection of $N$ $n$\nobreakdash-dimensional i.i.d. data
vectors collected to a $N\times n$ design matrix.  We use structure
$G$ to extend the common matrix indexing notation $D[i,j]$, and allow
more general row selectors.  Notably the notation $D[G_i=j,i]$ is used
to denote those rows of the column $i$ in which the parents $G_i$
contain the value configuration number $j\in\{1,\ldots,q_i\}$.

It is common to assume the uniform prior for structures, in which case
the objective function for structure learning is reduced to the
marginal likelihood $P(D|G,\alpha)$.  If the model parameters
$\theta_{ij}$ are further assumed to be independently Dirichlet
distributed only depending on $i$ and $G_{i}$ and the data $D$ is
assumed to have no missing values, the marginal likelihood can be
decomposed as
\begin{eqnarray}
\label{eqn:bayesmix}
P(D|G,\alpha) & = & \prod_{i=1}^{n}\prod_{j=1}^{q_i}\int
P(D[G_i=j,i]|\alpha)\\ & = & \prod_{i=1}^{n}\prod_{j=1}^{q_i}\int
P([G_i=j,i]|\theta)P(\theta|\alpha) d\theta.\nonumber
\end{eqnarray}
In coding terms this means that each data column $D[\cdot,i]$ is first partitioned based on the
values in columns $G_i$, and each part is then coded using a Bayesian mixture that can be expressed
in a closed form~\cite{Bunt91, Heck95}.

\subsection {Problems, solutions and problems}

Finding the satisfactory Dirichlet hyperparameters for the Bayesian
mixture above has, however, turned out to be problematic. Early on,
one of the desiderata for a good model selection criterion was that it
would yield equal scores for essentially equivalent
models~\cite{Verm90}.  For example, the score for the structure
$X_1\rightarrow X_2$ should be the same as the score for the model
$X_2 \rightarrow X_1$ since they both correspond to the hypothesis
that variables $X_1$ and $X_2$ are statistically dependent on each
other.  It can be shown~\cite{Heck95} that to achieve this, not all
the hyperparameters $\alpha$ are possible and for practical reasons
Buntine~\cite{Bunt91} suggested a so called BDeu score with just one
hyperparameter $\alpha\in R_{++}$ so that $\theta_{ij\cdot}\sim
Dir(\frac{\alpha}{q_i r_i},\ldots,\frac{\alpha}{q_i r_i})$.  However,
it soon turned out that BDeu score was very sensitive to the selection
of this hyperparameter~\cite{cosco.uai07} and that for small sample
sizes this method detects spurious correlations~\cite{Steck08} leading
to models with suspiciously many parameters.

A natural solution to avoid parameter sensitivity of BDeu would be to
use a normalized maximum likelihood (NML)
criterion~\cite{Shta87,Riss96a}, i.e., to find the structure $G$ that
maximizes
\begin{equation}
P_{NML}(D;G)=\frac{P(D|\hat\theta(D;G))}{\sum_{D'}{P(D'|\hat\theta(D';G))}},
\end{equation}
where $\hat\theta$ denotes the (easy to find) maximum likelihood
parameters and the sum in the denominator goes over all the possible
$N\times n$ data matrices. While it is easy to see that this criterion
satisfies the requirement of giving equal scores to equal structures,
the normalizing constant renders the computation
infeasible. Consequently, Silander et al.~\cite{cosco.pgm08a}
suggested solving the parameter sensitivity problem by adopting the
NML-code to the column partitions, i.e., changing the Bayesian mixture
in equation~(\ref{eqn:bayesmix}) to
\begin{equation}
P^1_{NML}(D[G_i=j,i];G)=\frac{P(D|\hat\theta(D[G_i=j,i];G))}{\sum_{D'}{P(D';\hat\theta(D'|G))}},
\end{equation}
where $D'\in{\{1,\ldots,r_i\}}^{|D[G_i=j,i]|}$.  The logarithm of the
denominator is often called a regret, since it indicates the extra
code length needed compared to the code length obtained using the (a
priori unknown) maximum likelihood parameters. The regret for
$P^1_{NML}$ depends only on the length $N$ of the categorical data
vector with $r$ different categorical values.  While the naive
computation of the regret is still prohibitive, it can be approximated
efficiently using a so called Szpankowski
approximation~\cite{cosco.aistat03}:
\begin{eqnarray}
\label{eqn:szp}
\lefteqn{reg(N,r) \approx \frac{\sqrt{2} r \Gamma{\left(\frac{r}{2} \right)}}{3 \sqrt{N} \Gamma{\left(\frac{r-1}{2}  \right)}}} \\
 &&+ \left(\frac{r-1}{2}\right) \log{\left (\frac{N}{2} \right )}- \log \Gamma{\left(\frac{r}{2} \right)} + \frac{1}{2} \log{\left (\pi \right )}\nonumber\\
 &&+ \frac{r^{2} \Gamma^{2}{\left(\frac{r}{2} \right)}}{9N \Gamma^{2}{\left(\frac{r-1}{2}\right)}} + \frac{2r^3-3r^2-2r+3}{36N}\nonumber.
\end{eqnarray}
fNML solves the parameter sensitivity problem and yields predictive
models superior to BDeu.  However, the criterion does not satisfy the
property of giving the same score for the models that correspond to
the same dependence statements. Furthermore, the learned structures
still sometimes appear surprisingly complicated.

\section{QUOTIENT NML SCORE}

We will now introduce a quotient normalized maximum likelihood (qNML)
criterion for learning Bayesian network structures.  While equally
efficient to compute than BDeu and fNML, it is free from
hyperparameters, and it can be proven to give equal scores to
equivalent models. Furthermore, it coincides with the actual NML score
for exponentially many models. In our empirical tests it produces
models featuring good predictive performance with significantly
simpler structures than BDeu and fNML.

Like BDeu and fNML, qNML can be expressed as a product of $n$ terms,
one for each variable, but unlike the other two, it is not based on
further partitioning the corresponding data column
\begin{eqnarray}
s^{qNML}(D;G) & := & \sum_{i=1}^n s^{qNML}_i(D;G)\\
 & := & \sum_{i=1}^n \frac{P^1_{NML}(D[\cdot,(i,G_i)];G)}{P^1_{NML}(D[\cdot,G_i];G)}.\nonumber
\end{eqnarray}
The trick here is to model a subset of columns as though there were no
conditional independencies among the corresponding variables $S
\subset X$.  In this case, we can collapse the $\prod_{X_i\in S} r_i$
value configurations and consider them as values of a single variable
with $\prod_{X_i\in S} r_i$ different values which can then be modeled
with a one-dimensional $P^1_{NML}$ code.  The $s^{qNML}$ does not
necessarily define a distribution for $D$, but it is easy to verify
that it coincides with the $P_{NML}(D;G)$ for all the networks that
are composed of fully connected components.  The number of such
networks equals the number of nonempty partitions of a set of $n$
elements, i.e., the $n^\text{th}$ Bell number.

\subsection {On implementation}

Collapsing many variables into a single one causes $r$ in the
equation~(\ref{eqn:szp}) to be exponentially large (in our experiments
around $2^{20}$).  This causes the $\Gamma^2$ terms to grow very large
calling for care when computing the ratios of such huge numbers. More
specifically, scoring the $i^\text{th}$ data column, i.e., computing
$s^{qNML}_i$, requires the difference $dreg(N,q_i,r_i)=reg(N,q_i) -
reg(N,r_i q_i)$, where $q$ denotes the number of values in the
collapsed parent variable.  A numerically stable approximation is:
\begin{eqnarray}
\lefteqn{dreg(N,q,r)\approx\frac{1-r^2}{18 N} q^{2}} \\
&&+ \frac{\left(q - 1\right) \sqrt{4 q - 2} + \left(1 - q r\right) \sqrt{4 q r - 2}}{6 \sqrt{N}} \nonumber\\
&&+\frac{q\left(3 N \left(r \log{r}  - \left(r - 1\right) \left(\log{N} + 1\right)\right) + r\right)}{6 N} \nonumber\\
&&+ \frac{1}{12}(\log{(r+\frac{1-r}{q+1})}- 7\log{r}).\nonumber
\end{eqnarray}

\section{EXPERIMENTAL RESULTS}

To empirically compare the model selection criteria we took 20 UCI
data sets~\cite{Lichman:2013} and run 1000 train and test experiments
for 20 different sample sizes for all of them. The training was
conducted using a dynamic programming based exact structure
learning\cite{cosco.uai06} which limited the number $n$ of variables
to less than 20.

Figure~\ref{fig:bcmean} shows an example of the behavior of the
studied selection criteria.  For small sample sizes fNML and qNML
yield equally accurate predictions while BDeu requires more samples to
obtain competitive performance. The curves are mean curves from 1000
independent train and test splits.

\begin{figure}
\centering
\includegraphics[width=8cm,height=5cm]{breast_cancer_mean.pdf}
\caption{Predictive log-loss in a breast cancer data as a function of
  sample size for different model selection criteria.}
\label{fig:bcmean}
\end{figure}

Figure~\ref{fig:bcnpmean} shows how fNML still sometimes behaves
strangely in terms of model complexity here measured by the number of
parameters in the model.  qNML, instead, appears to yield more
parsimonious models.

\begin{figure}
\centering
\includegraphics[width=8cm,height=5cm]{breast_cancer_np_mean.pdf}
\caption{Number of parameters in a breast cancer model as a function
  of sample size for different model selection criteria.}
\label{fig:bcnpmean}
\end{figure}

\begin{table}
\centering
\begin{tabular}{crrrr}
\toprule
           Data &     N &   BDeu &            fNML &            qNML \\
\midrule
           Iris &    16 &   3.59 &            3.47 &   \textbf{3.43} \\
  PostOperative &    20 &  10.24 &   \textbf{8.27} &            8.32 \\
           Wine &    27 &  18.11 &           12.68 &  \textbf{12.51} \\
 HeartHungarian &    45 &  10.15 &            9.10 &   \textbf{8.90} \\
   BreastCancer &    45 &  12.01 &  \textbf{10.57} &           10.66 \\
 HeartCleveland &    48 &  16.36 &           12.81 &  \textbf{12.54} \\
          Ecoli &    51 &   5.77 &   \textbf{5.30} &            5.37 \\
          Liver &    54 &   4.07 &            3.97 &   \textbf{3.91} \\
          Glass &    55 &   7.16 &   \textbf{6.22} &            6.28 \\
        Thyroid &    55 &   2.83 &   \textbf{2.75} &            2.77 \\
   HeartStatlog &    56 &  14.15 &           11.92 &  \textbf{11.60} \\
      TicTacToe &    96 &  10.99 &  \textbf{10.73} &           11.05 \\
        Balance &    96 &   7.48 &   \textbf{7.39} &            7.78 \\
    BcWisconsin &   105 &   5.81 &            5.12 &   \textbf{5.08} \\
       Diabetes &   117 &   4.95 &            4.95 &   \textbf{4.89} \\
          Yeast &   225 &   5.54 &   \textbf{5.40} &            5.42 \\
        Abalone &   418 &   3.93 &   \textbf{3.88} &            3.96 \\
     PageBlocks &   548 &   2.36 &   \textbf{2.32} &            2.38 \\
        Shuttle &  2900 &   1.70 &   \textbf{1.69} &            1.72 \\
          Adult &  4885 &  10.22 &  \textbf{10.08} &           10.12 \\
\bottomrule
\end{tabular}
\caption{Predictive log losses for small sample sizes for different model selection criteria in 20 different datasets.}
\label{tbl:preds}
	\end{table}

Table~\ref{tbl:preds} features predictive losses for the small sample
sizes.  (The results for large sample sizes converge for all sensible
criteria.)  As we can see, fNML still usually (12/20) performs best
but qNML is not much worse while BDeu may sometimes perform much
worse.

Looking at the number of parameters for the same datasets and sample
sizes (Table~\ref{tbl:nofparams}) reveals that qNML usually (18/20)
yields simplest models, and sometimes, like in the BreastCancer data,
the differences are of orders of magnitude.

\begin{table}
\centering
\begin{tabular}{crrrr}
\toprule
           Data &     N &          BDeu &  fNML &          qNML \\
\midrule
           Iris &    16 &            36 &    33 &   \textbf{28} \\
  PostOperative &    20 &          1304 &   464 &  \textbf{115} \\
           Wine &    27 &         11559 &   780 &  \textbf{214} \\
 HeartHungarian &    45 &          4706 &   838 &  \textbf{100} \\
   BreastCancer &    45 &         26774 &  2710 &  \textbf{246} \\
 HeartCleveland &    48 &         29346 &  1407 &  \textbf{204} \\
          Ecoli &    51 &           113 &   167 &   \textbf{80} \\
          Liver &    54 &            40 &    60 &   \textbf{24} \\
          Glass &    55 &          1488 &   572 &  \textbf{100} \\
        Thyroid &    55 &            37 &    70 &   \textbf{28} \\
   HeartStatlog &    56 &         18414 &  1099 &  \textbf{159} \\
      TicTacToe &    96 &         13645 &  2438 &  \textbf{696} \\
        Balance &    96 &   \textbf{20} &   237 &           168 \\
    BcWisconsin &   105 &          5921 &   816 &   \textbf{88} \\
       Diabetes &   117 &            35 &   216 &   \textbf{34} \\
          Yeast &   225 &   \textbf{78} &   307 &            80 \\
        Abalone &   418 &            90 &   156 &   \textbf{62} \\
     PageBlocks &   548 &           699 &   377 &   \textbf{56} \\
        Shuttle &  2900 &           380 &   594 &  \textbf{111} \\
          Adult &  4885 &  \textbf{599} &  1431 &           876 \\
\bottomrule
\end{tabular}
\caption{Number of model parameters for small sample sizes for different model selection criteria in 20 different datasets.}
\label{tbl:nofparams}
\end{table}

\section{CONCLUSION}

We have presented qNML, a new model selection criterion for learning
structures of Bayesian networks.  While being competitive in
predictive terms, it often yields significantly simpler models than
other common model selection criteria.  The computational cost of qNML
equals the cost of earlier criteria.  The criterion also gives equal
scores for models that encode same independence hypotheses about the
joint probability distribution.  qNML also coincides with NML
criterion for exponentially many models.

\subsubsection*{Acknowledgements}

Use unnumbered third level headings for the acknowledgements title.
All acknowledgements go at the end of the paper.

%%\subsubsection*{References}

\bibliographystyle{apalike}
\bibliography{cosco}

\end{document}

### 2017-03-15 qNMLisEQV.tex
Added a proof (separate file) for score equivalence.

%% LyX 2.2.2 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{babel}

\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}

\begin{document}

\title{qNML is score equivalent}

\maketitle

Chickering showed that any equivalent structure can be reached from
another by a series of arc reversal operations without leaving the
equivalence class~\cite{Chick95}. This means that in order to show
that equivalent structures leads to the same qNML score, we only need
to prove that this is the case for equivalent structures that differ
by a single arc reversal.

The network without any arcs is the sole member of its equivalence
class. All the other networks have at least one arc. We will first
presents some lemmas that characterize the relations of the parent sets
before and after the arc reversal. Let $G$ be a network structure
and $G'$ an equivalent network structure after the arc from $A$
to $B$ has been reversed to point from $B$ to $A$. We
will denote the parent sets of $A$ and $B$ in $G$ by $G_{A}$
and $G_{B}$, and the parent sets of $A$ and $B$ in $G'$ by $G'_{A}$
and $G'_{B}$. Note that not all the arc reversals lead to the equivalent
structures. By saying that $G$ and $G'$ are equivalent we imply
that our arc reversal does not destroy existing V-structures or create
new ones. The crucial observation is that if reversing the arc that
goes from $A$ to $B$ does not create or destroy V-structures, it
must be that the parents of $B$ other than $A$ are exactly the same
as parents of $A$ (statement 3 below).

\begin{lemma}\label{thm:sameparents}
\item $G_{B}=G_{A}\cup\{A\}$
\end{lemma}

\begin{proof}
We show the direction $G_{A}\cup\{A\}\subset G_{B}$ 
by contradiction. First of all, a parent of $A$ cannot be a child
of $B$ or otherwise there would be a loop in $G.$ If $A$ had a
parent $Z$ that is not a parent of $B$, then $B$ and $Z$ were
not adjacent and the reversal of the arc would create a new V-structure
$(B,A,Z)$. Since this was forbidden by the equivalence of $G$ and
$G'$, $Z$ must also be a parent of $B$.

Also, $B$ cannot have parents other than $A$ that are not also parents
of $A$, i.e., $G_{B}\subset G_{A}\cup{A}$ . If there were such
a parent $Z$ it should be anyway adjacent to $A$, otherwise the
V-structure $(A,B,Z)$ would be destroyed in the reversal. Since the
$Z$ was not a parent of $A$, the only possibility for adjacency
is that $Z$ were a child of $A$. However, in this case arc reversal
would lead to a loop $B,A,Z,B$.
\end{proof}

We wil next list some simple consequences of the equivalence preserving
arc reversal:

\begin{lemma}
\label{thm:reveqs}
\hangindent\leftmargini
\hspace{1mm}
\begin{enumerate}
\item $G'_{B}=G_{B}\setminus\{A\}$
\item $G'_{A}=G_{A}\cup\{B\}$ 
\item $G'_{B}=G_{A}$
\item $\{B\}\cup G_{B}=\{B\}\cup G_{A}\cup\{A\}=G'_{A}\cup\{A\}$
\item $G'_{A}=G_{A}\cup\{B\}=G'_{B}\cup\{B\}$
\end{enumerate}
\end{lemma}

\begin{proof}

The statements 1 and 2 are trivial since they just state the fact
that an arc reversal removes $A$ from the parents of $B$ and adds
$B$ to the parents of $A$.

The
statement 3 follows directly by using the lemma \ref{thm:sameparents}
to the statement 1. Combining these equalities we can generate more of them 
such as statements 4 and 5. 
\end{proof}

Let us take structures $G$ and $G'$. Since we can assume the data to be fixed we lighten
up the notations and write 
$P^1_{NML}(i,G_i) := P^1_{NML}(D[\cdot,(i,G_i)];G)$ and
$P^1_{NML}(G_i) := P^1_{NML}(D[\cdot,G_i];G)$.

\begin{theorem}
$  s^{qNML}(D;G)=s^{qNML}(D;G').$
\end{theorem}

\begin{proof}

  Now the scores for structures can be decomposed as the
  $s^{qNML}(D;G)=\sum_{i=1}^{n}s_i^{qNML}(D;G)$ and 
  $s^{qNML}(D;G')=\sum_{i=1}^{n}s_i^{qNML}(D;G')$.

Since only the terms corresponding to the variables $A$ and $B$
in these sums are different, it is enough to show that

$$
s_A^{qNML}(D;G)+s_B^{qNML}(D;G) = s_A^{qNML}(D;G')+s_B^{qNML}(D;G')
$$
Now

\begin{align*}
s_A^{qNML}(D;G)+s_B^{qNML}(D;G)& =\log\frac{P^1_{NML}(A,G_{A})}{P^1_{NML}(G_{A})}\frac{P^1_{NML}(B,G_{B})}{P^1_{NML}(G_{B})}\\
 & =\log 1\cdot\frac{P^1_{NML}(B,G_{B})}{P^1_{NML}(G_{A})}\\
 & =\log \frac{P^1_{NML}(B,G'_{B})}{P^1_{NML}(G'_{A})}\frac{P^1_{NML}(A,G'_{A})}{P^1_{NML}(G'_{B})}\\
 & =s_A^{qNML}(D;G')+s_B^{qNML}(D;G').
\end{align*}
The second equation follows from the lemma \ref{thm:sameparents}, and the third from
the statements 5 and 4 of the lemma \ref{thm:reveqs}.
\end{proof}

\bibliographystyle{apalike}
\bibliography{cosco}

\end{document}

### 2017-03-16 main.tex
Corrected a typo
Added proof of score equivalence.

@@
 \usepackage[margin=1in]{geometry}
-\usepackage{amsmath,epsfig,booktabs}
+\usepackage{amsmath,amsthm,epsfig,booktabs}
+
+\newtheorem{lemma}{Lemma}
+\newtheorem{theorem}{Theorem}
 
@@
 \label{eqn:szp}
-\lefteqn{reg(N,r) \approx \frac{\sqrt{2} r \Gamma{\left(\frac{r}{2} \right)}}{3 \sqrt{N} \Gamma{\left(\frac{r-1}{2}  \right)}}} \\
- &&+ \left(\frac{r-1}{2}\right) \log{\left (\frac{N}{2} \right )}- \log \Gamma{\left(\frac{r}{2} \right)} + \frac{1}{2} \log{\left (\pi \right )}\nonumber\\
- &&+ \frac{r^{2} \Gamma^{2}{\left(\frac{r}{2} \right)}}{9N \Gamma^{2}{\left(\frac{r-1}{2}\right)}} + \frac{2r^3-3r^2-2r+3}{36N}\nonumber.
+\lefteqn{reg(N,r) \approx \frac{\sqrt{2} r \Gamma{\left(\frac{r}{2} \right)}}
+                               {3 \sqrt{N} \Gamma{\left(\frac{r-1}{2}  \right)}}} \\
+&&+ \left(\frac{r-1}{2}\right) \log{\left (\frac{N}{2} \right )}
+- \log \Gamma{\left(\frac{r}{2} \right)} + \frac{1}{2} \log{\left (\pi \right )}\nonumber\\
+&&- \frac{r^{2} \Gamma^{2}{\left(\frac{r }{2} \right)}}
+         {9N \Gamma^{2}{\left(\frac{r-1}{2}\right)}}
++ \frac{2r^3-3r^2-2r+3}{36N}\nonumber.
 \end{eqnarray}
@@
 s^{qNML}(D;G) & := & \sum_{i=1}^n s^{qNML}_i(D;G)\\
- & := & \sum_{i=1}^n \frac{P^1_{NML}(D[\cdot,(i,G_i)];G)}{P^1_{NML}(D[\cdot,G_i];G)}.\nonumber
+& := & \sum_{i=1}^n \log \frac{P^1_{NML}(D[\cdot,(i,G_i)];G)}
+                             {P^1_{NML}(D[\cdot,G_i];G)}.\nonumber
 \end{eqnarray}
@@
 elements, i.e., the $n^\text{th}$ Bell number.
+
+\subsection {qNML is score equivalent}
+
+The qNML yields equal scores for network structures that encode same
+sets of independencies. Verma and Pearl~\cite{Verm90} showed that the
+equivalent networks are exactly those who a) are same when directed
+arcs are substitued by undirected ones and b) who have same
+V-structures, i.e.  the variable triplets $(A,B,C)$ where both $A$ and
+$B$ are parents of $C$, but there is no arc between $A$ and $B$ (in
+neither direction).  Later, Chickering~\cite{Chick95} showed that all
+the equivalent network structures, and only them, can be reached from
+each other by reversing, one by one, so called covered arcs, i.e. the
+arcs from node $A$ to $B$, for which $B$'s parents other than $A$ are
+exactly the $A$'s parents ($G_B=\{A\}\cup G_A$).
+
+We will next state this as a
+theorem and sketch a proof for it. A more detailed proof appears in the
+suplementary material.
+\begin{theorem}
+  \label{thm:scoreqv}
+  Let $G$ and $G'$ be two Bayesian network structures that differ only
+  by a single covered arc reversal, i.e., the arc from $A$ to $B$ in $G$
+  has been reversed in $G'$ to point from $B$ to $A$, then
+  $$s^{qNML}(D;G)=s^{qNML}(D;G').$$
+\end{theorem}
+\begin{proof}
+  Now the scores for structures can be decomposed as the
+  $s^{qNML}(D;G)=\sum_{i=1}^{n}s_i^{qNML}(D;G)$ and
+  $s^{qNML}(D;G')=\sum_{i=1}^{n}s_i^{qNML}(D;G')$.  Since only the
+  terms corresponding to the variables $A$ and $B$ in these sums are
+  different, it is enough to show that sum of these two terms are
+  equal for $G$ and $G'$. Since we can assume the data to be fixed we
+  lighten up the notations and write
+  $P^1_{NML}(i,G_i) := P^1_{NML}(D[\cdot,(i,G_i)];G)$ and
+  $P^1_{NML}(G_i)   := P^1_{NML}(D[\cdot,G_i];G)$.
+  \begin{eqnarray}
+    \lefteqn{s_A^{qNML}(D;G)+s_B^{qNML}(D;G)} \nonumber\\
+    && =\log\frac{P^1_{NML}(A,G_{A})}{P^1_{NML}(G_{A})}
+            \frac{P^1_{NML}(B,G_{B})}{P^1_{NML}(G_{B})}\nonumber\\
+    && =\log 1\cdot\frac{P^1_{NML}(B,G_{B})}{P^1_{NML}(G_{A})}\nonumber\\
+    && =\log \frac{P^1_{NML}(B,G'_{B})}{P^1_{NML}(G'_{A})}
+             \frac{P^1_{NML}(A,G'_{A})}{P^1_{NML}(G'_{B})}\nonumber\\
+ && =s_A^{qNML}(D;G')+s_B^{qNML}(D;G'),\nonumber
+\end{eqnarray}
+  due to equations
+  $\{A\}\cup G_A  = G_B$,
+  $\{B\}\cup G'_B = G'_A$,
+  $\{B\}\cup G_B  = \{A\} \cup G'_A$,
+  and $G_A        = G'_B$.
+\end{proof}
+
+\subsection{qNML is consistent}
 

### 2017-03-16 abstract.tex
Splitting main to sections

--- 2017-03-15 main.tex
@@
-\documentclass[letterpaper]{article}
-\usepackage{proceed2e}
-\usepackage[margin=1in]{geometry}
-\usepackage{amsmath,epsfig,booktabs}
-
-% Set the typeface to Times Roman
-\usepackage{times}
-
-\title{Quotient Normalized Maximum Likelihood Criterion for Learning Bayesian Network Structures}
-
-\author{} % LEAVE BLANK FOR ORIGINAL SUBMISSION.
-          % UAI  reviewing is double-blind.
-
-% The author names and affiliations should appear only in the accepted paper.
-%
-%\author{ {\bf Harry Q.~Bovik\thanks{Footnote for author to give an
-%alternate address.}} \\
-%Computer Science Dept. \\
-%Cranberry University\\
-%Pittsburgh, PA 15213 \\
-%\And
-%{\bf Coauthor}  \\
-%Affiliation          \\
-%Address \\
-%\And
-%{\bf Coauthor}   \\
-%Affiliation \\
-%Address    \\
-%(if needed)\\
-%}
-
-% Example definitions.
-% --------------------
-%\def\x{{\mathbf x}}
-%\def\L{{\cal L}}
-
-\hyphenation{col-umns}
-\hyphenation{Bayes-ian}
-\hyphenation{oth-er}
-
-\begin{document}
-
-\maketitle
-
 \begin{abstract}
@@
 \end{abstract}
-
-\section{INTRODUCTION}
-\label{sec:intro}
-Bayesian networks~\cite{Pear88} are popular models for presenting
-multivariate statistical dependencies that may have been induced by
-underlying causal mechanisms.  Techniques for learning the structure
-of Bayesian networks from the observational data has therefore been
-used for many tasks such as discovering cell signaling pathways from
-the protein activity data~\cite{bn4sigpath02}, revealing the business
-process structures~\cite{bn4bpmining} from the transaction logs and
-modeling brain region connectivity using fMRI
-data~\cite{bn4brainconnect}.
-
-Learning the structure of statistical dependencies can be seen as a
-model selection task where each model is a different hypothesis about
-the conditional dependencies between sets of variables. Traditional
-model selection criteria such as the Akaike information
-criterion~\cite{Akai73} and the Bayesian information
-criterion~\cite{Schw78} have also been used for the task, but recent
-comparisons have not been favorable for these criteria in terms of
-structural stability and/or predictive
-performance~\cite{cosco.pgm08a}.  The most popular criterion has been
-the marginal likelihood (usually called BDeu for reasons explained
-later), but recent studies~\cite{cosco.uai07,Steck08} have found this
-criterion to be very sensitive to hyper parameters and to yield
-undesirably complex models for small sample sizes.
-
-The information theoretic normalized maximum likelihood (NML)
-criterion~\cite{Shta87,Riss96a} would otherwise be an ideal candidate
-for a good criterion.  but its exact calculation is likely to be
-prohibitively demanding. In 2008, Silander et al. introduced a
-hyper-parameter free, NML inspired criterion called a factorized NML
-(fNML)~\cite{cosco.pgm08a} that was shown to yield good predictive
-models without sensitivity problems.  However, from the structure
-learning point of view, the fNML still sometimes appears to yield
-overly complex models. In this paper we introduce another NML related
-criterion, a quotient NML (qNML) that yields simpler models without
-sacrificing predictive accuracy. Furthermore, unlike the fNML, the
-qNML gives equal scores to the structures that encode the same
-independence statements. Like other common model selection criteria,
-the qNML is also consistent.
-
-We will next briefly introduce Bayesian networks and then review the
-BDeu and fNML criteria and introduce the qNML criterion.  We will also
-summarize the results for 20 datasets to back up our claim of qNML
-yielding parsimonious models with good predictive capabilities.
-
-\section{BAYESIAN NETWORKS}
-\label{sec:bns}
-
-Bayesian networks are a general way to describe the dependencies
-between the components of an $n$\nobreakdash-dimensional discrete data
-vector $X=(X_{1},\ldots,X_{n})$ in which the component $X_{i}$ may
-take any of the discrete values in a set $\{1,\ldots,r_{i}\}$.
-Despite denoting the values with small integers, the model will treat
-the components of $X$ categorical.
-
-\subsection{Likelihood}
-\label{ssec:likelihood}
-
-Bayesian network $B=(G,\theta)$ defines a probability distribution for
-$X$. The component $G$ defines the structure of the model as a
-directed acyclic graph that has exactly one node for each component of
-$X$. The structure $G=(G_{1},\ldots,G_{n})$ defines for each
-variable/node $X_{i}$ its (possibly empty) parent set $G_{i}$, i.e.,
-the nodes from which there are a directed edges to the variable
-$X_{i}$.
-
-Given a realization $x$ of $X$, we denote the sub\nobreakdash-vector
-of $x$ that consists of the values of the parents of $X_{i}$ in $x$ as
-$G_{i}(x)$. It is customary to enumerate all the possible
-sub\nobreakdash-vectors $G_{i}(X)$ from $1$ to $q_{i}=\prod_{h\in
-  G_{i}}r_{h}.$ In case $G_{i}$ is empty, we define $q_{i}=1$ and
-$P(G_{i}(x)=1)=1$ for all vectors $x$.
-
-For each variable $X_{i}$ there is a $q_{i}\times r_{i}$ table
-$\theta_{i}$ of parameters whose $j^{\textnormal{th}}$ row and
-$k^{\textnormal{th}}$ column defines the conditional probability
-$P(X_{i}=k\mid G_{i}(X)=j;\theta)=\theta_{ijk}$.  With structure $G$
-and parameters $\theta$, we can now express the likelihood function of
-the model as
-\begin{equation}
-P(x|G,\theta)=\prod_{i=1}^{n}P(x_{i}\mid G_{i}(x);\theta_{i})=\prod_{i=1}^{n}\theta_{iG_{i}(x)x_{i}}.
-\end{equation}
-
-\subsection{Bayesian structure learning}
-
-Bayesian learning of Bayesian network structures is based on the
-posterior probability $P(G|D,\alpha)$, where $\alpha$ denotes the
-hyper\nobreakdash-parameters for the model parameters $\theta$, and
-the $D$ is a collection of $N$ $n$\nobreakdash-dimensional i.i.d. data
-vectors collected to a $N\times n$ design matrix.  We use structure
-$G$ to extend the common matrix indexing notation $D[i,j]$, and allow
-more general row selectors.  Notably the notation $D[G_i=j,i]$ is used
-to denote those rows of the column $i$ in which the parents $G_i$
-contain the value configuration number $j\in\{1,\ldots,q_i\}$.
-
-It is common to assume the uniform prior for structures, in which case
-the objective function for structure learning is reduced to the
-marginal likelihood $P(D|G,\alpha)$.  If the model parameters
-$\theta_{ij}$ are further assumed to be independently Dirichlet
-distributed only depending on $i$ and $G_{i}$ and the data $D$ is
-assumed to have no missing values, the marginal likelihood can be
-decomposed as
-\begin{eqnarray}
-\label{eqn:bayesmix}
-P(D|G,\alpha) & = & \prod_{i=1}^{n}\prod_{j=1}^{q_i}\int
-P(D[G_i=j,i]|\alpha)\\ & = & \prod_{i=1}^{n}\prod_{j=1}^{q_i}\int
-P([G_i=j,i]|\theta)P(\theta|\alpha) d\theta.\nonumber
-\end{eqnarray}
-In coding terms this means that each data column $D[\cdot,i]$ is first partitioned based on the
-values in columns $G_i$, and each part is then coded using a Bayesian mixture that can be expressed
-in a closed form~\cite{Bunt91, Heck95}.
-
-\subsection {Problems, solutions and problems}
-
-Finding the satisfactory Dirichlet hyperparameters for the Bayesian
-mixture above has, however, turned out to be problematic. Early on,
-one of the desiderata for a good model selection criterion was that it
-would yield equal scores for essentially equivalent
-models~\cite{Verm90}.  For example, the score for the structure
-$X_1\rightarrow X_2$ should be the same as the score for the model
-$X_2 \rightarrow X_1$ since they both correspond to the hypothesis
-that variables $X_1$ and $X_2$ are statistically dependent on each
-other.  It can be shown~\cite{Heck95} that to achieve this, not all
-the hyperparameters $\alpha$ are possible and for practical reasons
-Buntine~\cite{Bunt91} suggested a so called BDeu score with just one
-hyperparameter $\alpha\in R_{++}$ so that $\theta_{ij\cdot}\sim
-Dir(\frac{\alpha}{q_i r_i},\ldots,\frac{\alpha}{q_i r_i})$.  However,
-it soon turned out that BDeu score was very sensitive to the selection
-of this hyperparameter~\cite{cosco.uai07} and that for small sample
-sizes this method detects spurious correlations~\cite{Steck08} leading
-to models with suspiciously many parameters.
-
-A natural solution to avoid parameter sensitivity of BDeu would be to
-use a normalized maximum likelihood (NML)
-criterion~\cite{Shta87,Riss96a}, i.e., to find the structure $G$ that
-maximizes
-\begin{equation}
-P_{NML}(D;G)=\frac{P(D|\hat\theta(D;G))}{\sum_{D'}{P(D'|\hat\theta(D';G))}},
-\end{equation}
-where $\hat\theta$ denotes the (easy to find) maximum likelihood
-parameters and the sum in the denominator goes over all the possible
-$N\times n$ data matrices. While it is easy to see that this criterion
-satisfies the requirement of giving equal scores to equal structures,
-the normalizing constant renders the computation
-infeasible. Consequently, Silander et al.~\cite{cosco.pgm08a}
-suggested solving the parameter sensitivity problem by adopting the
-NML-code to the column partitions, i.e., changing the Bayesian mixture
-in equation~(\ref{eqn:bayesmix}) to
-\begin{equation}
-P^1_{NML}(D[G_i=j,i];G)=\frac{P(D|\hat\theta(D[G_i=j,i];G))}{\sum_{D'}{P(D';\hat\theta(D'|G))}},
-\end{equation}
-where $D'\in{\{1,\ldots,r_i\}}^{|D[G_i=j,i]|}$.  The logarithm of the
-denominator is often called a regret, since it indicates the extra
-code length needed compared to the code length obtained using the (a
-priori unknown) maximum likelihood parameters. The regret for
-$P^1_{NML}$ depends only on the length $N$ of the categorical data
-vector with $r$ different categorical values.  While the naive
-computation of the regret is still prohibitive, it can be approximated
-efficiently using a so called Szpankowski
-approximation~\cite{cosco.aistat03}:
-\begin{eqnarray}
-\label{eqn:szp}
-\lefteqn{reg(N,r) \approx \frac{\sqrt{2} r \Gamma{\left(\frac{r}{2} \right)}}{3 \sqrt{N} \Gamma{\left(\frac{r-1}{2}  \right)}}} \\
- &&+ \left(\frac{r-1}{2}\right) \log{\left (\frac{N}{2} \right )}- \log \Gamma{\left(\frac{r}{2} \right)} + \frac{1}{2} \log{\left (\pi \right )}\nonumber\\
- &&+ \frac{r^{2} \Gamma^{2}{\left(\frac{r}{2} \right)}}{9N \Gamma^{2}{\left(\frac{r-1}{2}\right)}} + \frac{2r^3-3r^2-2r+3}{36N}\nonumber.
-\end{eqnarray}
-fNML solves the parameter sensitivity problem and yields predictive
-models superior to BDeu.  However, the criterion does not satisfy the
-property of giving the same score for the models that correspond to
-the same dependence statements. Furthermore, the learned structures
-still sometimes appear surprisingly complicated.
-
-\section{QUOTIENT NML SCORE}
-
-We will now introduce a quotient normalized maximum likelihood (qNML)
-criterion for learning Bayesian network structures.  While equally
-efficient to compute than BDeu and fNML, it is free from
-hyperparameters, and it can be proven to give equal scores to
-equivalent models. Furthermore, it coincides with the actual NML score
-for exponentially many models. In our empirical tests it produces
-models featuring good predictive performance with significantly
-simpler structures than BDeu and fNML.
-
-Like BDeu and fNML, qNML can be expressed as a product of $n$ terms,
-one for each variable, but unlike the other two, it is not based on
-further partitioning the corresponding data column
-\begin{eqnarray}
-s^{qNML}(D;G) & := & \sum_{i=1}^n s^{qNML}_i(D;G)\\
- & := & \sum_{i=1}^n \frac{P^1_{NML}(D[\cdot,(i,G_i)];G)}{P^1_{NML}(D[\cdot,G_i];G)}.\nonumber
-\end{eqnarray}
-The trick here is to model a subset of columns as though there were no
-conditional independencies among the corresponding variables $S
-\subset X$.  In this case, we can collapse the $\prod_{X_i\in S} r_i$
-value configurations and consider them as values of a single variable
-with $\prod_{X_i\in S} r_i$ different values which can then be modeled
-with a one-dimensional $P^1_{NML}$ code.  The $s^{qNML}$ does not
-necessarily define a distribution for $D$, but it is easy to verify
-that it coincides with the $P_{NML}(D;G)$ for all the networks that
-are composed of fully connected components.  The number of such
-networks equals the number of nonempty partitions of a set of $n$
-elements, i.e., the $n^\text{th}$ Bell number.
-
-\subsection {On implementation}
-
-Collapsing many variables into a single one causes $r$ in the
-equation~(\ref{eqn:szp}) to be exponentially large (in our experiments
-around $2^{20}$).  This causes the $\Gamma^2$ terms to grow very large
-calling for care when computing the ratios of such huge numbers. More
-specifically, scoring the $i^\text{th}$ data column, i.e., computing
-$s^{qNML}_i$, requires the difference $dreg(N,q_i,r_i)=reg(N,q_i) -
-reg(N,r_i q_i)$, where $q$ denotes the number of values in the
-collapsed parent variable.  A numerically stable approximation is:
-\begin{eqnarray}
-\lefteqn{dreg(N,q,r)\approx\frac{1-r^2}{18 N} q^{2}} \\
-&&+ \frac{\left(q - 1\right) \sqrt{4 q - 2} + \left(1 - q r\right) \sqrt{4 q r - 2}}{6 \sqrt{N}} \nonumber\\
-&&+\frac{q\left(3 N \left(r \log{r}  - \left(r - 1\right) \left(\log{N} + 1\right)\right) + r\right)}{6 N} \nonumber\\
-&&+ \frac{1}{12}(\log{(r+\frac{1-r}{q+1})}- 7\log{r}).\nonumber
-\end{eqnarray}
-
-\section{EXPERIMENTAL RESULTS}
-
-To empirically compare the model selection criteria we took 20 UCI
-data sets~\cite{Lichman:2013} and run 1000 train and test experiments
-for 20 different sample sizes for all of them. The training was
-conducted using a dynamic programming based exact structure
-learning\cite{cosco.uai06} which limited the number $n$ of variables
-to less than 20.
-
-Figure~\ref{fig:bcmean} shows an example of the behavior of the
-studied selection criteria.  For small sample sizes fNML and qNML
-yield equally accurate predictions while BDeu requires more samples to
-obtain competitive performance. The curves are mean curves from 1000
-independent train and test splits.
-
-\begin{figure}
-\centering
-\includegraphics[width=8cm,height=5cm]{breast_cancer_mean.pdf}
-\caption{Predictive log-loss in a breast cancer data as a function of
-  sample size for different model selection criteria.}
-\label{fig:bcmean}
-\end{figure}
-
-Figure~\ref{fig:bcnpmean} shows how fNML still sometimes behaves
-strangely in terms of model complexity here measured by the number of
-parameters in the model.  qNML, instead, appears to yield more
-parsimonious models.
-
-\begin{figure}
-\centering
-\includegraphics[width=8cm,height=5cm]{breast_cancer_np_mean.pdf}
-\caption{Number of parameters in a breast cancer model as a function
-  of sample size for different model selection criteria.}
-\label{fig:bcnpmean}
-\end{figure}
-
-\begin{table}
-\centering
-\begin{tabular}{crrrr}
-\toprule
-           Data &     N &   BDeu &            fNML &            qNML \\
-\midrule
-           Iris &    16 &   3.59 &            3.47 &   \textbf{3.43} \\
-  PostOperative &    20 &  10.24 &   \textbf{8.27} &            8.32 \\
-           Wine &    27 &  18.11 &           12.68 &  \textbf{12.51} \\
- HeartHungarian &    45 &  10.15 &            9.10 &   \textbf{8.90} \\
-   BreastCancer &    45 &  12.01 &  \textbf{10.57} &           10.66 \\
- HeartCleveland &    48 &  16.36 &           12.81 &  \textbf{12.54} \\
-          Ecoli &    51 &   5.77 &   \textbf{5.30} &            5.37 \\
-          Liver &    54 &   4.07 &            3.97 &   \textbf{3.91} \\
-          Glass &    55 &   7.16 &   \textbf{6.22} &            6.28 \\
-        Thyroid &    55 &   2.83 &   \textbf{2.75} &            2.77 \\
-   HeartStatlog &    56 &  14.15 &           11.92 &  \textbf{11.60} \\
-      TicTacToe &    96 &  10.99 &  \textbf{10.73} &           11.05 \\
-        Balance &    96 &   7.48 &   \textbf{7.39} &            7.78 \\
-    BcWisconsin &   105 &   5.81 &            5.12 &   \textbf{5.08} \\
-       Diabetes &   117 &   4.95 &            4.95 &   \textbf{4.89} \\
-          Yeast &   225 &   5.54 &   \textbf{5.40} &            5.42 \\
-        Abalone &   418 &   3.93 &   \textbf{3.88} &            3.96 \\
-     PageBlocks &   548 &   2.36 &   \textbf{2.32} &            2.38 \\
-        Shuttle &  2900 &   1.70 &   \textbf{1.69} &            1.72 \\
-          Adult &  4885 &  10.22 &  \textbf{10.08} &           10.12 \\
-\bottomrule
-\end{tabular}
-\caption{Predictive log losses for small sample sizes for different model selection criteria in 20 different datasets.}
-\label{tbl:preds}
-	\end{table}
-
-Table~\ref{tbl:preds} features predictive losses for the small sample
-sizes.  (The results for large sample sizes converge for all sensible
-criteria.)  As we can see, fNML still usually (12/20) performs best
-but qNML is not much worse while BDeu may sometimes perform much
-worse.
-
-Looking at the number of parameters for the same datasets and sample
-sizes (Table~\ref{tbl:nofparams}) reveals that qNML usually (18/20)
-yields simplest models, and sometimes, like in the BreastCancer data,
-the differences are of orders of magnitude.
-
-\begin{table}
-\centering
-\begin{tabular}{crrrr}
-\toprule
-           Data &     N &          BDeu &  fNML &          qNML \\
-\midrule
-           Iris &    16 &            36 &    33 &   \textbf{28} \\
-  PostOperative &    20 &          1304 &   464 &  \textbf{115} \\
-           Wine &    27 &         11559 &   780 &  \textbf{214} \\
- HeartHungarian &    45 &          4706 &   838 &  \textbf{100} \\
-   BreastCancer &    45 &         26774 &  2710 &  \textbf{246} \\
- HeartCleveland &    48 &         29346 &  1407 &  \textbf{204} \\
-          Ecoli &    51 &           113 &   167 &   \textbf{80} \\
-          Liver &    54 &            40 &    60 &   \textbf{24} \\
-          Glass &    55 &          1488 &   572 &  \textbf{100} \\
-        Thyroid &    55 &            37 &    70 &   \textbf{28} \\
-   HeartStatlog &    56 &         18414 &  1099 &  \textbf{159} \\
-      TicTacToe &    96 &         13645 &  2438 &  \textbf{696} \\
-        Balance &    96 &   \textbf{20} &   237 &           168 \\
-    BcWisconsin &   105 &          5921 &   816 &   \textbf{88} \\
-       Diabetes &   117 &            35 &   216 &   \textbf{34} \\
-          Yeast &   225 &   \textbf{78} &   307 &            80 \\
-        Abalone &   418 &            90 &   156 &   \textbf{62} \\
-     PageBlocks &   548 &           699 &   377 &   \textbf{56} \\
-        Shuttle &  2900 &           380 &   594 &  \textbf{111} \\
-          Adult &  4885 &  \textbf{599} &  1431 &           876 \\
-\bottomrule
-\end{tabular}
-\caption{Number of model parameters for small sample sizes for different model selection criteria in 20 different datasets.}
-\label{tbl:nofparams}
-\end{table}
-
-\section{CONCLUSION}
-
-We have presented qNML, a new model selection criterion for learning
-structures of Bayesian networks.  While being competitive in
-predictive terms, it often yields significantly simpler models than
-other common model selection criteria.  The computational cost of qNML
-equals the cost of earlier criteria.  The criterion also gives equal
-scores for models that encode same independence hypotheses about the
-joint probability distribution.  qNML also coincides with NML
-criterion for exponentially many models.
-
-\subsubsection*{Acknowledgements}
-
-Use unnumbered third level headings for the acknowledgements title.
-All acknowledgements go at the end of the paper.
-
-%%\subsubsection*{References}
-
-\bibliographystyle{apalike}
-\bibliography{cosco}
-
-\end{document}

### 2017-03-16 bn.tex
Splitting main to sections

--- 2017-03-15 main.tex
@@
-\documentclass[letterpaper]{article}
-\usepackage{proceed2e}
-\usepackage[margin=1in]{geometry}
-\usepackage{amsmath,epsfig,booktabs}
-
-% Set the typeface to Times Roman
-\usepackage{times}
-
-\title{Quotient Normalized Maximum Likelihood Criterion for Learning Bayesian Network Structures}
-
-\author{} % LEAVE BLANK FOR ORIGINAL SUBMISSION.
-          % UAI  reviewing is double-blind.
-
-% The author names and affiliations should appear only in the accepted paper.
-%
-%\author{ {\bf Harry Q.~Bovik\thanks{Footnote for author to give an
-%alternate address.}} \\
-%Computer Science Dept. \\
-%Cranberry University\\
-%Pittsburgh, PA 15213 \\
-%\And
-%{\bf Coauthor}  \\
-%Affiliation          \\
-%Address \\
-%\And
-%{\bf Coauthor}   \\
-%Affiliation \\
-%Address    \\
-%(if needed)\\
-%}
-
-% Example definitions.
-% --------------------
-%\def\x{{\mathbf x}}
-%\def\L{{\cal L}}
-
-\hyphenation{col-umns}
-\hyphenation{Bayes-ian}
-\hyphenation{oth-er}
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-Learning the dependency structure of a multivariate distribution from
-the observational data is an important task since it allows us to
-speculate about the underlying causal mechanisms that induce
-dependencies. This learning task can be approached as a model
-selection problem, but recent studies have revealed that the popular
-Bayesian model selection criterion is not satisfactory. We review some
-of the information theoretic alternatives and introduce a new one
-called a quotient normalized maximum likelihood criterion.
-\end{abstract}
-
-\section{INTRODUCTION}
-\label{sec:intro}
-Bayesian networks~\cite{Pear88} are popular models for presenting
-multivariate statistical dependencies that may have been induced by
-underlying causal mechanisms.  Techniques for learning the structure
-of Bayesian networks from the observational data has therefore been
-used for many tasks such as discovering cell signaling pathways from
-the protein activity data~\cite{bn4sigpath02}, revealing the business
-process structures~\cite{bn4bpmining} from the transaction logs and
-modeling brain region connectivity using fMRI
-data~\cite{bn4brainconnect}.
-
-Learning the structure of statistical dependencies can be seen as a
-model selection task where each model is a different hypothesis about
-the conditional dependencies between sets of variables. Traditional
-model selection criteria such as the Akaike information
-criterion~\cite{Akai73} and the Bayesian information
-criterion~\cite{Schw78} have also been used for the task, but recent
-comparisons have not been favorable for these criteria in terms of
-structural stability and/or predictive
-performance~\cite{cosco.pgm08a}.  The most popular criterion has been
-the marginal likelihood (usually called BDeu for reasons explained
-later), but recent studies~\cite{cosco.uai07,Steck08} have found this
-criterion to be very sensitive to hyper parameters and to yield
-undesirably complex models for small sample sizes.
-
-The information theoretic normalized maximum likelihood (NML)
-criterion~\cite{Shta87,Riss96a} would otherwise be an ideal candidate
-for a good criterion.  but its exact calculation is likely to be
-prohibitively demanding. In 2008, Silander et al. introduced a
-hyper-parameter free, NML inspired criterion called a factorized NML
-(fNML)~\cite{cosco.pgm08a} that was shown to yield good predictive
-models without sensitivity problems.  However, from the structure
-learning point of view, the fNML still sometimes appears to yield
-overly complex models. In this paper we introduce another NML related
-criterion, a quotient NML (qNML) that yields simpler models without
-sacrificing predictive accuracy. Furthermore, unlike the fNML, the
-qNML gives equal scores to the structures that encode the same
-independence statements. Like other common model selection criteria,
-the qNML is also consistent.
-
-We will next briefly introduce Bayesian networks and then review the
-BDeu and fNML criteria and introduce the qNML criterion.  We will also
-summarize the results for 20 datasets to back up our claim of qNML
-yielding parsimonious models with good predictive capabilities.
-
 \section{BAYESIAN NETWORKS}
@@
 still sometimes appear surprisingly complicated.
-
-\section{QUOTIENT NML SCORE}
-
-We will now introduce a quotient normalized maximum likelihood (qNML)
-criterion for learning Bayesian network structures.  While equally
-efficient to compute than BDeu and fNML, it is free from
-hyperparameters, and it can be proven to give equal scores to
-equivalent models. Furthermore, it coincides with the actual NML score
-for exponentially many models. In our empirical tests it produces
-models featuring good predictive performance with significantly
-simpler structures than BDeu and fNML.
-
-Like BDeu and fNML, qNML can be expressed as a product of $n$ terms,
-one for each variable, but unlike the other two, it is not based on
-further partitioning the corresponding data column
-\begin{eqnarray}
-s^{qNML}(D;G) & := & \sum_{i=1}^n s^{qNML}_i(D;G)\\
- & := & \sum_{i=1}^n \frac{P^1_{NML}(D[\cdot,(i,G_i)];G)}{P^1_{NML}(D[\cdot,G_i];G)}.\nonumber
-\end{eqnarray}
-The trick here is to model a subset of columns as though there were no
-conditional independencies among the corresponding variables $S
-\subset X$.  In this case, we can collapse the $\prod_{X_i\in S} r_i$
-value configurations and consider them as values of a single variable
-with $\prod_{X_i\in S} r_i$ different values which can then be modeled
-with a one-dimensional $P^1_{NML}$ code.  The $s^{qNML}$ does not
-necessarily define a distribution for $D$, but it is easy to verify
-that it coincides with the $P_{NML}(D;G)$ for all the networks that
-are composed of fully connected components.  The number of such
-networks equals the number of nonempty partitions of a set of $n$
-elements, i.e., the $n^\text{th}$ Bell number.
-
-\subsection {On implementation}
-
-Collapsing many variables into a single one causes $r$ in the
-equation~(\ref{eqn:szp}) to be exponentially large (in our experiments
-around $2^{20}$).  This causes the $\Gamma^2$ terms to grow very large
-calling for care when computing the ratios of such huge numbers. More
-specifically, scoring the $i^\text{th}$ data column, i.e., computing
-$s^{qNML}_i$, requires the difference $dreg(N,q_i,r_i)=reg(N,q_i) -
-reg(N,r_i q_i)$, where $q$ denotes the number of values in the
-collapsed parent variable.  A numerically stable approximation is:
-\begin{eqnarray}
-\lefteqn{dreg(N,q,r)\approx\frac{1-r^2}{18 N} q^{2}} \\
-&&+ \frac{\left(q - 1\right) \sqrt{4 q - 2} + \left(1 - q r\right) \sqrt{4 q r - 2}}{6 \sqrt{N}} \nonumber\\
-&&+\frac{q\left(3 N \left(r \log{r}  - \left(r - 1\right) \left(\log{N} + 1\right)\right) + r\right)}{6 N} \nonumber\\
-&&+ \frac{1}{12}(\log{(r+\frac{1-r}{q+1})}- 7\log{r}).\nonumber
-\end{eqnarray}
-
-\section{EXPERIMENTAL RESULTS}
-
-To empirically compare the model selection criteria we took 20 UCI
-data sets~\cite{Lichman:2013} and run 1000 train and test experiments
-for 20 different sample sizes for all of them. The training was
-conducted using a dynamic programming based exact structure
-learning\cite{cosco.uai06} which limited the number $n$ of variables
-to less than 20.
-
-Figure~\ref{fig:bcmean} shows an example of the behavior of the
-studied selection criteria.  For small sample sizes fNML and qNML
-yield equally accurate predictions while BDeu requires more samples to
-obtain competitive performance. The curves are mean curves from 1000
-independent train and test splits.
-
-\begin{figure}
-\centering
-\includegraphics[width=8cm,height=5cm]{breast_cancer_mean.pdf}
-\caption{Predictive log-loss in a breast cancer data as a function of
-  sample size for different model selection criteria.}
-\label{fig:bcmean}
-\end{figure}
-
-Figure~\ref{fig:bcnpmean} shows how fNML still sometimes behaves
-strangely in terms of model complexity here measured by the number of
-parameters in the model.  qNML, instead, appears to yield more
-parsimonious models.
-
-\begin{figure}
-\centering
-\includegraphics[width=8cm,height=5cm]{breast_cancer_np_mean.pdf}
-\caption{Number of parameters in a breast cancer model as a function
-  of sample size for different model selection criteria.}
-\label{fig:bcnpmean}
-\end{figure}
-
-\begin{table}
-\centering
-\begin{tabular}{crrrr}
-\toprule
-           Data &     N &   BDeu &            fNML &            qNML \\
-\midrule
-           Iris &    16 &   3.59 &            3.47 &   \textbf{3.43} \\
-  PostOperative &    20 &  10.24 &   \textbf{8.27} &            8.32 \\
-           Wine &    27 &  18.11 &           12.68 &  \textbf{12.51} \\
- HeartHungarian &    45 &  10.15 &            9.10 &   \textbf{8.90} \\
-   BreastCancer &    45 &  12.01 &  \textbf{10.57} &           10.66 \\
- HeartCleveland &    48 &  16.36 &           12.81 &  \textbf{12.54} \\
-          Ecoli &    51 &   5.77 &   \textbf{5.30} &            5.37 \\
-          Liver &    54 &   4.07 &            3.97 &   \textbf{3.91} \\
-          Glass &    55 &   7.16 &   \textbf{6.22} &            6.28 \\
-        Thyroid &    55 &   2.83 &   \textbf{2.75} &            2.77 \\
-   HeartStatlog &    56 &  14.15 &           11.92 &  \textbf{11.60} \\
-      TicTacToe &    96 &  10.99 &  \textbf{10.73} &           11.05 \\
-        Balance &    96 &   7.48 &   \textbf{7.39} &            7.78 \\
-    BcWisconsin &   105 &   5.81 &            5.12 &   \textbf{5.08} \\
-       Diabetes &   117 &   4.95 &            4.95 &   \textbf{4.89} \\
-          Yeast &   225 &   5.54 &   \textbf{5.40} &            5.42 \\
-        Abalone &   418 &   3.93 &   \textbf{3.88} &            3.96 \\
-     PageBlocks &   548 &   2.36 &   \textbf{2.32} &            2.38 \\
-        Shuttle &  2900 &   1.70 &   \textbf{1.69} &            1.72 \\
-          Adult &  4885 &  10.22 &  \textbf{10.08} &           10.12 \\
-\bottomrule
-\end{tabular}
-\caption{Predictive log losses for small sample sizes for different model selection criteria in 20 different datasets.}
-\label{tbl:preds}
-	\end{table}
-
-Table~\ref{tbl:preds} features predictive losses for the small sample
-sizes.  (The results for large sample sizes converge for all sensible
-criteria.)  As we can see, fNML still usually (12/20) performs best
-but qNML is not much worse while BDeu may sometimes perform much
-worse.
-
-Looking at the number of parameters for the same datasets and sample
-sizes (Table~\ref{tbl:nofparams}) reveals that qNML usually (18/20)
-yields simplest models, and sometimes, like in the BreastCancer data,
-the differences are of orders of magnitude.
-
-\begin{table}
-\centering
-\begin{tabular}{crrrr}
-\toprule
-           Data &     N &          BDeu &  fNML &          qNML \\
-\midrule
-           Iris &    16 &            36 &    33 &   \textbf{28} \\
-  PostOperative &    20 &          1304 &   464 &  \textbf{115} \\
-           Wine &    27 &         11559 &   780 &  \textbf{214} \\
- HeartHungarian &    45 &          4706 &   838 &  \textbf{100} \\
-   BreastCancer &    45 &         26774 &  2710 &  \textbf{246} \\
- HeartCleveland &    48 &         29346 &  1407 &  \textbf{204} \\
-          Ecoli &    51 &           113 &   167 &   \textbf{80} \\
-          Liver &    54 &            40 &    60 &   \textbf{24} \\
-          Glass &    55 &          1488 &   572 &  \textbf{100} \\
-        Thyroid &    55 &            37 &    70 &   \textbf{28} \\
-   HeartStatlog &    56 &         18414 &  1099 &  \textbf{159} \\
-      TicTacToe &    96 &         13645 &  2438 &  \textbf{696} \\
-        Balance &    96 &   \textbf{20} &   237 &           168 \\
-    BcWisconsin &   105 &          5921 &   816 &   \textbf{88} \\
-       Diabetes &   117 &            35 &   216 &   \textbf{34} \\
-          Yeast &   225 &   \textbf{78} &   307 &            80 \\
-        Abalone &   418 &            90 &   156 &   \textbf{62} \\
-     PageBlocks &   548 &           699 &   377 &   \textbf{56} \\
-        Shuttle &  2900 &           380 &   594 &  \textbf{111} \\
-          Adult &  4885 &  \textbf{599} &  1431 &           876 \\
-\bottomrule
-\end{tabular}
-\caption{Number of model parameters for small sample sizes for different model selection criteria in 20 different datasets.}
-\label{tbl:nofparams}
-\end{table}
-
-\section{CONCLUSION}
-
-We have presented qNML, a new model selection criterion for learning
-structures of Bayesian networks.  While being competitive in
-predictive terms, it often yields significantly simpler models than
-other common model selection criteria.  The computational cost of qNML
-equals the cost of earlier criteria.  The criterion also gives equal
-scores for models that encode same independence hypotheses about the
-joint probability distribution.  qNML also coincides with NML
-criterion for exponentially many models.
-
-\subsubsection*{Acknowledgements}
-
-Use unnumbered third level headings for the acknowledgements title.
-All acknowledgements go at the end of the paper.
-
-%%\subsubsection*{References}
-
-\bibliographystyle{apalike}
-\bibliography{cosco}
-
-\end{document}

### 2017-03-16 conclusion.tex
Splitting main to sections

--- 2017-03-15 main.tex
@@
-\documentclass[letterpaper]{article}
-\usepackage{proceed2e}
-\usepackage[margin=1in]{geometry}
-\usepackage{amsmath,epsfig,booktabs}
-
-% Set the typeface to Times Roman
-\usepackage{times}
-
-\title{Quotient Normalized Maximum Likelihood Criterion for Learning Bayesian Network Structures}
-
-\author{} % LEAVE BLANK FOR ORIGINAL SUBMISSION.
-          % UAI  reviewing is double-blind.
-
-% The author names and affiliations should appear only in the accepted paper.
-%
-%\author{ {\bf Harry Q.~Bovik\thanks{Footnote for author to give an
-%alternate address.}} \\
-%Computer Science Dept. \\
-%Cranberry University\\
-%Pittsburgh, PA 15213 \\
-%\And
-%{\bf Coauthor}  \\
-%Affiliation          \\
-%Address \\
-%\And
-%{\bf Coauthor}   \\
-%Affiliation \\
-%Address    \\
-%(if needed)\\
-%}
-
-% Example definitions.
-% --------------------
-%\def\x{{\mathbf x}}
-%\def\L{{\cal L}}
-
-\hyphenation{col-umns}
-\hyphenation{Bayes-ian}
-\hyphenation{oth-er}
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-Learning the dependency structure of a multivariate distribution from
-the observational data is an important task since it allows us to
-speculate about the underlying causal mechanisms that induce
-dependencies. This learning task can be approached as a model
-selection problem, but recent studies have revealed that the popular
-Bayesian model selection criterion is not satisfactory. We review some
-of the information theoretic alternatives and introduce a new one
-called a quotient normalized maximum likelihood criterion.
-\end{abstract}
-
-\section{INTRODUCTION}
-\label{sec:intro}
-Bayesian networks~\cite{Pear88} are popular models for presenting
-multivariate statistical dependencies that may have been induced by
-underlying causal mechanisms.  Techniques for learning the structure
-of Bayesian networks from the observational data has therefore been
-used for many tasks such as discovering cell signaling pathways from
-the protein activity data~\cite{bn4sigpath02}, revealing the business
-process structures~\cite{bn4bpmining} from the transaction logs and
-modeling brain region connectivity using fMRI
-data~\cite{bn4brainconnect}.
-
-Learning the structure of statistical dependencies can be seen as a
-model selection task where each model is a different hypothesis about
-the conditional dependencies between sets of variables. Traditional
-model selection criteria such as the Akaike information
-criterion~\cite{Akai73} and the Bayesian information
-criterion~\cite{Schw78} have also been used for the task, but recent
-comparisons have not been favorable for these criteria in terms of
-structural stability and/or predictive
-performance~\cite{cosco.pgm08a}.  The most popular criterion has been
-the marginal likelihood (usually called BDeu for reasons explained
-later), but recent studies~\cite{cosco.uai07,Steck08} have found this
-criterion to be very sensitive to hyper parameters and to yield
-undesirably complex models for small sample sizes.
-
-The information theoretic normalized maximum likelihood (NML)
-criterion~\cite{Shta87,Riss96a} would otherwise be an ideal candidate
-for a good criterion.  but its exact calculation is likely to be
-prohibitively demanding. In 2008, Silander et al. introduced a
-hyper-parameter free, NML inspired criterion called a factorized NML
-(fNML)~\cite{cosco.pgm08a} that was shown to yield good predictive
-models without sensitivity problems.  However, from the structure
-learning point of view, the fNML still sometimes appears to yield
-overly complex models. In this paper we introduce another NML related
-criterion, a quotient NML (qNML) that yields simpler models without
-sacrificing predictive accuracy. Furthermore, unlike the fNML, the
-qNML gives equal scores to the structures that encode the same
-independence statements. Like other common model selection criteria,
-the qNML is also consistent.
-
-We will next briefly introduce Bayesian networks and then review the
-BDeu and fNML criteria and introduce the qNML criterion.  We will also
-summarize the results for 20 datasets to back up our claim of qNML
-yielding parsimonious models with good predictive capabilities.
-
-\section{BAYESIAN NETWORKS}
-\label{sec:bns}
-
-Bayesian networks are a general way to describe the dependencies
-between the components of an $n$\nobreakdash-dimensional discrete data
-vector $X=(X_{1},\ldots,X_{n})$ in which the component $X_{i}$ may
-take any of the discrete values in a set $\{1,\ldots,r_{i}\}$.
-Despite denoting the values with small integers, the model will treat
-the components of $X$ categorical.
-
-\subsection{Likelihood}
-\label{ssec:likelihood}
-
-Bayesian network $B=(G,\theta)$ defines a probability distribution for
-$X$. The component $G$ defines the structure of the model as a
-directed acyclic graph that has exactly one node for each component of
-$X$. The structure $G=(G_{1},\ldots,G_{n})$ defines for each
-variable/node $X_{i}$ its (possibly empty) parent set $G_{i}$, i.e.,
-the nodes from which there are a directed edges to the variable
-$X_{i}$.
-
-Given a realization $x$ of $X$, we denote the sub\nobreakdash-vector
-of $x$ that consists of the values of the parents of $X_{i}$ in $x$ as
-$G_{i}(x)$. It is customary to enumerate all the possible
-sub\nobreakdash-vectors $G_{i}(X)$ from $1$ to $q_{i}=\prod_{h\in
-  G_{i}}r_{h}.$ In case $G_{i}$ is empty, we define $q_{i}=1$ and
-$P(G_{i}(x)=1)=1$ for all vectors $x$.
-
-For each variable $X_{i}$ there is a $q_{i}\times r_{i}$ table
-$\theta_{i}$ of parameters whose $j^{\textnormal{th}}$ row and
-$k^{\textnormal{th}}$ column defines the conditional probability
-$P(X_{i}=k\mid G_{i}(X)=j;\theta)=\theta_{ijk}$.  With structure $G$
-and parameters $\theta$, we can now express the likelihood function of
-the model as
-\begin{equation}
-P(x|G,\theta)=\prod_{i=1}^{n}P(x_{i}\mid G_{i}(x);\theta_{i})=\prod_{i=1}^{n}\theta_{iG_{i}(x)x_{i}}.
-\end{equation}
-
-\subsection{Bayesian structure learning}
-
-Bayesian learning of Bayesian network structures is based on the
-posterior probability $P(G|D,\alpha)$, where $\alpha$ denotes the
-hyper\nobreakdash-parameters for the model parameters $\theta$, and
-the $D$ is a collection of $N$ $n$\nobreakdash-dimensional i.i.d. data
-vectors collected to a $N\times n$ design matrix.  We use structure
-$G$ to extend the common matrix indexing notation $D[i,j]$, and allow
-more general row selectors.  Notably the notation $D[G_i=j,i]$ is used
-to denote those rows of the column $i$ in which the parents $G_i$
-contain the value configuration number $j\in\{1,\ldots,q_i\}$.
-
-It is common to assume the uniform prior for structures, in which case
-the objective function for structure learning is reduced to the
-marginal likelihood $P(D|G,\alpha)$.  If the model parameters
-$\theta_{ij}$ are further assumed to be independently Dirichlet
-distributed only depending on $i$ and $G_{i}$ and the data $D$ is
-assumed to have no missing values, the marginal likelihood can be
-decomposed as
-\begin{eqnarray}
-\label{eqn:bayesmix}
-P(D|G,\alpha) & = & \prod_{i=1}^{n}\prod_{j=1}^{q_i}\int
-P(D[G_i=j,i]|\alpha)\\ & = & \prod_{i=1}^{n}\prod_{j=1}^{q_i}\int
-P([G_i=j,i]|\theta)P(\theta|\alpha) d\theta.\nonumber
-\end{eqnarray}
-In coding terms this means that each data column $D[\cdot,i]$ is first partitioned based on the
-values in columns $G_i$, and each part is then coded using a Bayesian mixture that can be expressed
-in a closed form~\cite{Bunt91, Heck95}.
-
-\subsection {Problems, solutions and problems}
-
-Finding the satisfactory Dirichlet hyperparameters for the Bayesian
-mixture above has, however, turned out to be problematic. Early on,
-one of the desiderata for a good model selection criterion was that it
-would yield equal scores for essentially equivalent
-models~\cite{Verm90}.  For example, the score for the structure
-$X_1\rightarrow X_2$ should be the same as the score for the model
-$X_2 \rightarrow X_1$ since they both correspond to the hypothesis
-that variables $X_1$ and $X_2$ are statistically dependent on each
-other.  It can be shown~\cite{Heck95} that to achieve this, not all
-the hyperparameters $\alpha$ are possible and for practical reasons
-Buntine~\cite{Bunt91} suggested a so called BDeu score with just one
-hyperparameter $\alpha\in R_{++}$ so that $\theta_{ij\cdot}\sim
-Dir(\frac{\alpha}{q_i r_i},\ldots,\frac{\alpha}{q_i r_i})$.  However,
-it soon turned out that BDeu score was very sensitive to the selection
-of this hyperparameter~\cite{cosco.uai07} and that for small sample
-sizes this method detects spurious correlations~\cite{Steck08} leading
-to models with suspiciously many parameters.
-
-A natural solution to avoid parameter sensitivity of BDeu would be to
-use a normalized maximum likelihood (NML)
-criterion~\cite{Shta87,Riss96a}, i.e., to find the structure $G$ that
-maximizes
-\begin{equation}
-P_{NML}(D;G)=\frac{P(D|\hat\theta(D;G))}{\sum_{D'}{P(D'|\hat\theta(D';G))}},
-\end{equation}
-where $\hat\theta$ denotes the (easy to find) maximum likelihood
-parameters and the sum in the denominator goes over all the possible
-$N\times n$ data matrices. While it is easy to see that this criterion
-satisfies the requirement of giving equal scores to equal structures,
-the normalizing constant renders the computation
-infeasible. Consequently, Silander et al.~\cite{cosco.pgm08a}
-suggested solving the parameter sensitivity problem by adopting the
-NML-code to the column partitions, i.e., changing the Bayesian mixture
-in equation~(\ref{eqn:bayesmix}) to
-\begin{equation}
-P^1_{NML}(D[G_i=j,i];G)=\frac{P(D|\hat\theta(D[G_i=j,i];G))}{\sum_{D'}{P(D';\hat\theta(D'|G))}},
-\end{equation}
-where $D'\in{\{1,\ldots,r_i\}}^{|D[G_i=j,i]|}$.  The logarithm of the
-denominator is often called a regret, since it indicates the extra
-code length needed compared to the code length obtained using the (a
-priori unknown) maximum likelihood parameters. The regret for
-$P^1_{NML}$ depends only on the length $N$ of the categorical data
-vector with $r$ different categorical values.  While the naive
-computation of the regret is still prohibitive, it can be approximated
-efficiently using a so called Szpankowski
-approximation~\cite{cosco.aistat03}:
-\begin{eqnarray}
-\label{eqn:szp}
-\lefteqn{reg(N,r) \approx \frac{\sqrt{2} r \Gamma{\left(\frac{r}{2} \right)}}{3 \sqrt{N} \Gamma{\left(\frac{r-1}{2}  \right)}}} \\
- &&+ \left(\frac{r-1}{2}\right) \log{\left (\frac{N}{2} \right )}- \log \Gamma{\left(\frac{r}{2} \right)} + \frac{1}{2} \log{\left (\pi \right )}\nonumber\\
- &&+ \frac{r^{2} \Gamma^{2}{\left(\frac{r}{2} \right)}}{9N \Gamma^{2}{\left(\frac{r-1}{2}\right)}} + \frac{2r^3-3r^2-2r+3}{36N}\nonumber.
-\end{eqnarray}
-fNML solves the parameter sensitivity problem and yields predictive
-models superior to BDeu.  However, the criterion does not satisfy the
-property of giving the same score for the models that correspond to
-the same dependence statements. Furthermore, the learned structures
-still sometimes appear surprisingly complicated.
-
-\section{QUOTIENT NML SCORE}
-
-We will now introduce a quotient normalized maximum likelihood (qNML)
-criterion for learning Bayesian network structures.  While equally
-efficient to compute than BDeu and fNML, it is free from
-hyperparameters, and it can be proven to give equal scores to
-equivalent models. Furthermore, it coincides with the actual NML score
-for exponentially many models. In our empirical tests it produces
-models featuring good predictive performance with significantly
-simpler structures than BDeu and fNML.
-
-Like BDeu and fNML, qNML can be expressed as a product of $n$ terms,
-one for each variable, but unlike the other two, it is not based on
-further partitioning the corresponding data column
-\begin{eqnarray}
-s^{qNML}(D;G) & := & \sum_{i=1}^n s^{qNML}_i(D;G)\\
- & := & \sum_{i=1}^n \frac{P^1_{NML}(D[\cdot,(i,G_i)];G)}{P^1_{NML}(D[\cdot,G_i];G)}.\nonumber
-\end{eqnarray}
-The trick here is to model a subset of columns as though there were no
-conditional independencies among the corresponding variables $S
-\subset X$.  In this case, we can collapse the $\prod_{X_i\in S} r_i$
-value configurations and consider them as values of a single variable
-with $\prod_{X_i\in S} r_i$ different values which can then be modeled
-with a one-dimensional $P^1_{NML}$ code.  The $s^{qNML}$ does not
-necessarily define a distribution for $D$, but it is easy to verify
-that it coincides with the $P_{NML}(D;G)$ for all the networks that
-are composed of fully connected components.  The number of such
-networks equals the number of nonempty partitions of a set of $n$
-elements, i.e., the $n^\text{th}$ Bell number.
-
-\subsection {On implementation}
-
-Collapsing many variables into a single one causes $r$ in the
-equation~(\ref{eqn:szp}) to be exponentially large (in our experiments
-around $2^{20}$).  This causes the $\Gamma^2$ terms to grow very large
-calling for care when computing the ratios of such huge numbers. More
-specifically, scoring the $i^\text{th}$ data column, i.e., computing
-$s^{qNML}_i$, requires the difference $dreg(N,q_i,r_i)=reg(N,q_i) -
-reg(N,r_i q_i)$, where $q$ denotes the number of values in the
-collapsed parent variable.  A numerically stable approximation is:
-\begin{eqnarray}
-\lefteqn{dreg(N,q,r)\approx\frac{1-r^2}{18 N} q^{2}} \\
-&&+ \frac{\left(q - 1\right) \sqrt{4 q - 2} + \left(1 - q r\right) \sqrt{4 q r - 2}}{6 \sqrt{N}} \nonumber\\
-&&+\frac{q\left(3 N \left(r \log{r}  - \left(r - 1\right) \left(\log{N} + 1\right)\right) + r\right)}{6 N} \nonumber\\
-&&+ \frac{1}{12}(\log{(r+\frac{1-r}{q+1})}- 7\log{r}).\nonumber
-\end{eqnarray}
-
-\section{EXPERIMENTAL RESULTS}
-
-To empirically compare the model selection criteria we took 20 UCI
-data sets~\cite{Lichman:2013} and run 1000 train and test experiments
-for 20 different sample sizes for all of them. The training was
-conducted using a dynamic programming based exact structure
-learning\cite{cosco.uai06} which limited the number $n$ of variables
-to less than 20.
-
-Figure~\ref{fig:bcmean} shows an example of the behavior of the
-studied selection criteria.  For small sample sizes fNML and qNML
-yield equally accurate predictions while BDeu requires more samples to
-obtain competitive performance. The curves are mean curves from 1000
-independent train and test splits.
-
-\begin{figure}
-\centering
-\includegraphics[width=8cm,height=5cm]{breast_cancer_mean.pdf}
-\caption{Predictive log-loss in a breast cancer data as a function of
-  sample size for different model selection criteria.}
-\label{fig:bcmean}
-\end{figure}
-
-Figure~\ref{fig:bcnpmean} shows how fNML still sometimes behaves
-strangely in terms of model complexity here measured by the number of
-parameters in the model.  qNML, instead, appears to yield more
-parsimonious models.
-
-\begin{figure}
-\centering
-\includegraphics[width=8cm,height=5cm]{breast_cancer_np_mean.pdf}
-\caption{Number of parameters in a breast cancer model as a function
-  of sample size for different model selection criteria.}
-\label{fig:bcnpmean}
-\end{figure}
-
-\begin{table}
-\centering
-\begin{tabular}{crrrr}
-\toprule
-           Data &     N &   BDeu &            fNML &            qNML \\
-\midrule
-           Iris &    16 &   3.59 &            3.47 &   \textbf{3.43} \\
-  PostOperative &    20 &  10.24 &   \textbf{8.27} &            8.32 \\
-           Wine &    27 &  18.11 &           12.68 &  \textbf{12.51} \\
- HeartHungarian &    45 &  10.15 &            9.10 &   \textbf{8.90} \\
-   BreastCancer &    45 &  12.01 &  \textbf{10.57} &           10.66 \\
- HeartCleveland &    48 &  16.36 &           12.81 &  \textbf{12.54} \\
-          Ecoli &    51 &   5.77 &   \textbf{5.30} &            5.37 \\
-          Liver &    54 &   4.07 &            3.97 &   \textbf{3.91} \\
-          Glass &    55 &   7.16 &   \textbf{6.22} &            6.28 \\
-        Thyroid &    55 &   2.83 &   \textbf{2.75} &            2.77 \\
-   HeartStatlog &    56 &  14.15 &           11.92 &  \textbf{11.60} \\
-      TicTacToe &    96 &  10.99 &  \textbf{10.73} &           11.05 \\
-        Balance &    96 &   7.48 &   \textbf{7.39} &            7.78 \\
-    BcWisconsin &   105 &   5.81 &            5.12 &   \textbf{5.08} \\
-       Diabetes &   117 &   4.95 &            4.95 &   \textbf{4.89} \\
-          Yeast &   225 &   5.54 &   \textbf{5.40} &            5.42 \\
-        Abalone &   418 &   3.93 &   \textbf{3.88} &            3.96 \\
-     PageBlocks &   548 &   2.36 &   \textbf{2.32} &            2.38 \\
-        Shuttle &  2900 &   1.70 &   \textbf{1.69} &            1.72 \\
-          Adult &  4885 &  10.22 &  \textbf{10.08} &           10.12 \\
-\bottomrule
-\end{tabular}
-\caption{Predictive log losses for small sample sizes for different model selection criteria in 20 different datasets.}
-\label{tbl:preds}
-	\end{table}
-
-Table~\ref{tbl:preds} features predictive losses for the small sample
-sizes.  (The results for large sample sizes converge for all sensible
-criteria.)  As we can see, fNML still usually (12/20) performs best
-but qNML is not much worse while BDeu may sometimes perform much
-worse.
-
-Looking at the number of parameters for the same datasets and sample
-sizes (Table~\ref{tbl:nofparams}) reveals that qNML usually (18/20)
-yields simplest models, and sometimes, like in the BreastCancer data,
-the differences are of orders of magnitude.
-
-\begin{table}
-\centering
-\begin{tabular}{crrrr}
-\toprule
-           Data &     N &          BDeu &  fNML &          qNML \\
-\midrule
-           Iris &    16 &            36 &    33 &   \textbf{28} \\
-  PostOperative &    20 &          1304 &   464 &  \textbf{115} \\
-           Wine &    27 &         11559 &   780 &  \textbf{214} \\
- HeartHungarian &    45 &          4706 &   838 &  \textbf{100} \\
-   BreastCancer &    45 &         26774 &  2710 &  \textbf{246} \\
- HeartCleveland &    48 &         29346 &  1407 &  \textbf{204} \\
-          Ecoli &    51 &           113 &   167 &   \textbf{80} \\
-          Liver &    54 &            40 &    60 &   \textbf{24} \\
-          Glass &    55 &          1488 &   572 &  \textbf{100} \\
-        Thyroid &    55 &            37 &    70 &   \textbf{28} \\
-   HeartStatlog &    56 &         18414 &  1099 &  \textbf{159} \\
-      TicTacToe &    96 &         13645 &  2438 &  \textbf{696} \\
-        Balance &    96 &   \textbf{20} &   237 &           168 \\
-    BcWisconsin &   105 &          5921 &   816 &   \textbf{88} \\
-       Diabetes &   117 &            35 &   216 &   \textbf{34} \\
-          Yeast &   225 &   \textbf{78} &   307 &            80 \\
-        Abalone &   418 &            90 &   156 &   \textbf{62} \\
-     PageBlocks &   548 &           699 &   377 &   \textbf{56} \\
-        Shuttle &  2900 &           380 &   594 &  \textbf{111} \\
-          Adult &  4885 &  \textbf{599} &  1431 &           876 \\
-\bottomrule
-\end{tabular}
-\caption{Number of model parameters for small sample sizes for different model selection criteria in 20 different datasets.}
-\label{tbl:nofparams}
-\end{table}
-
 \section{CONCLUSION}
@@
 criterion for exponentially many models.
-
-\subsubsection*{Acknowledgements}
-
-Use unnumbered third level headings for the acknowledgements title.
-All acknowledgements go at the end of the paper.
-
-%%\subsubsection*{References}
-
-\bibliographystyle{apalike}
-\bibliography{cosco}
-
-\end{document}

### 2017-03-16 experiments.tex
Splitting main to sections

--- 2017-03-15 main.tex
@@
-\documentclass[letterpaper]{article}
-\usepackage{proceed2e}
-\usepackage[margin=1in]{geometry}
-\usepackage{amsmath,epsfig,booktabs}
-
-% Set the typeface to Times Roman
-\usepackage{times}
-
-\title{Quotient Normalized Maximum Likelihood Criterion for Learning Bayesian Network Structures}
-
-\author{} % LEAVE BLANK FOR ORIGINAL SUBMISSION.
-          % UAI  reviewing is double-blind.
-
-% The author names and affiliations should appear only in the accepted paper.
-%
-%\author{ {\bf Harry Q.~Bovik\thanks{Footnote for author to give an
-%alternate address.}} \\
-%Computer Science Dept. \\
-%Cranberry University\\
-%Pittsburgh, PA 15213 \\
-%\And
-%{\bf Coauthor}  \\
-%Affiliation          \\
-%Address \\
-%\And
-%{\bf Coauthor}   \\
-%Affiliation \\
-%Address    \\
-%(if needed)\\
-%}
-
-% Example definitions.
-% --------------------
-%\def\x{{\mathbf x}}
-%\def\L{{\cal L}}
-
-\hyphenation{col-umns}
-\hyphenation{Bayes-ian}
-\hyphenation{oth-er}
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-Learning the dependency structure of a multivariate distribution from
-the observational data is an important task since it allows us to
-speculate about the underlying causal mechanisms that induce
-dependencies. This learning task can be approached as a model
-selection problem, but recent studies have revealed that the popular
-Bayesian model selection criterion is not satisfactory. We review some
-of the information theoretic alternatives and introduce a new one
-called a quotient normalized maximum likelihood criterion.
-\end{abstract}
-
-\section{INTRODUCTION}
-\label{sec:intro}
-Bayesian networks~\cite{Pear88} are popular models for presenting
-multivariate statistical dependencies that may have been induced by
-underlying causal mechanisms.  Techniques for learning the structure
-of Bayesian networks from the observational data has therefore been
-used for many tasks such as discovering cell signaling pathways from
-the protein activity data~\cite{bn4sigpath02}, revealing the business
-process structures~\cite{bn4bpmining} from the transaction logs and
-modeling brain region connectivity using fMRI
-data~\cite{bn4brainconnect}.
-
-Learning the structure of statistical dependencies can be seen as a
-model selection task where each model is a different hypothesis about
-the conditional dependencies between sets of variables. Traditional
-model selection criteria such as the Akaike information
-criterion~\cite{Akai73} and the Bayesian information
-criterion~\cite{Schw78} have also been used for the task, but recent
-comparisons have not been favorable for these criteria in terms of
-structural stability and/or predictive
-performance~\cite{cosco.pgm08a}.  The most popular criterion has been
-the marginal likelihood (usually called BDeu for reasons explained
-later), but recent studies~\cite{cosco.uai07,Steck08} have found this
-criterion to be very sensitive to hyper parameters and to yield
-undesirably complex models for small sample sizes.
-
-The information theoretic normalized maximum likelihood (NML)
-criterion~\cite{Shta87,Riss96a} would otherwise be an ideal candidate
-for a good criterion.  but its exact calculation is likely to be
-prohibitively demanding. In 2008, Silander et al. introduced a
-hyper-parameter free, NML inspired criterion called a factorized NML
-(fNML)~\cite{cosco.pgm08a} that was shown to yield good predictive
-models without sensitivity problems.  However, from the structure
-learning point of view, the fNML still sometimes appears to yield
-overly complex models. In this paper we introduce another NML related
-criterion, a quotient NML (qNML) that yields simpler models without
-sacrificing predictive accuracy. Furthermore, unlike the fNML, the
-qNML gives equal scores to the structures that encode the same
-independence statements. Like other common model selection criteria,
-the qNML is also consistent.
-
-We will next briefly introduce Bayesian networks and then review the
-BDeu and fNML criteria and introduce the qNML criterion.  We will also
-summarize the results for 20 datasets to back up our claim of qNML
-yielding parsimonious models with good predictive capabilities.
-
-\section{BAYESIAN NETWORKS}
-\label{sec:bns}
-
-Bayesian networks are a general way to describe the dependencies
-between the components of an $n$\nobreakdash-dimensional discrete data
-vector $X=(X_{1},\ldots,X_{n})$ in which the component $X_{i}$ may
-take any of the discrete values in a set $\{1,\ldots,r_{i}\}$.
-Despite denoting the values with small integers, the model will treat
-the components of $X$ categorical.
-
-\subsection{Likelihood}
-\label{ssec:likelihood}
-
-Bayesian network $B=(G,\theta)$ defines a probability distribution for
-$X$. The component $G$ defines the structure of the model as a
-directed acyclic graph that has exactly one node for each component of
-$X$. The structure $G=(G_{1},\ldots,G_{n})$ defines for each
-variable/node $X_{i}$ its (possibly empty) parent set $G_{i}$, i.e.,
-the nodes from which there are a directed edges to the variable
-$X_{i}$.
-
-Given a realization $x$ of $X$, we denote the sub\nobreakdash-vector
-of $x$ that consists of the values of the parents of $X_{i}$ in $x$ as
-$G_{i}(x)$. It is customary to enumerate all the possible
-sub\nobreakdash-vectors $G_{i}(X)$ from $1$ to $q_{i}=\prod_{h\in
-  G_{i}}r_{h}.$ In case $G_{i}$ is empty, we define $q_{i}=1$ and
-$P(G_{i}(x)=1)=1$ for all vectors $x$.
-
-For each variable $X_{i}$ there is a $q_{i}\times r_{i}$ table
-$\theta_{i}$ of parameters whose $j^{\textnormal{th}}$ row and
-$k^{\textnormal{th}}$ column defines the conditional probability
-$P(X_{i}=k\mid G_{i}(X)=j;\theta)=\theta_{ijk}$.  With structure $G$
-and parameters $\theta$, we can now express the likelihood function of
-the model as
-\begin{equation}
-P(x|G,\theta)=\prod_{i=1}^{n}P(x_{i}\mid G_{i}(x);\theta_{i})=\prod_{i=1}^{n}\theta_{iG_{i}(x)x_{i}}.
-\end{equation}
-
-\subsection{Bayesian structure learning}
-
-Bayesian learning of Bayesian network structures is based on the
-posterior probability $P(G|D,\alpha)$, where $\alpha$ denotes the
-hyper\nobreakdash-parameters for the model parameters $\theta$, and
-the $D$ is a collection of $N$ $n$\nobreakdash-dimensional i.i.d. data
-vectors collected to a $N\times n$ design matrix.  We use structure
-$G$ to extend the common matrix indexing notation $D[i,j]$, and allow
-more general row selectors.  Notably the notation $D[G_i=j,i]$ is used
-to denote those rows of the column $i$ in which the parents $G_i$
-contain the value configuration number $j\in\{1,\ldots,q_i\}$.
-
-It is common to assume the uniform prior for structures, in which case
-the objective function for structure learning is reduced to the
-marginal likelihood $P(D|G,\alpha)$.  If the model parameters
-$\theta_{ij}$ are further assumed to be independently Dirichlet
-distributed only depending on $i$ and $G_{i}$ and the data $D$ is
-assumed to have no missing values, the marginal likelihood can be
-decomposed as
-\begin{eqnarray}
-\label{eqn:bayesmix}
-P(D|G,\alpha) & = & \prod_{i=1}^{n}\prod_{j=1}^{q_i}\int
-P(D[G_i=j,i]|\alpha)\\ & = & \prod_{i=1}^{n}\prod_{j=1}^{q_i}\int
-P([G_i=j,i]|\theta)P(\theta|\alpha) d\theta.\nonumber
-\end{eqnarray}
-In coding terms this means that each data column $D[\cdot,i]$ is first partitioned based on the
-values in columns $G_i$, and each part is then coded using a Bayesian mixture that can be expressed
-in a closed form~\cite{Bunt91, Heck95}.
-
-\subsection {Problems, solutions and problems}
-
-Finding the satisfactory Dirichlet hyperparameters for the Bayesian
-mixture above has, however, turned out to be problematic. Early on,
-one of the desiderata for a good model selection criterion was that it
-would yield equal scores for essentially equivalent
-models~\cite{Verm90}.  For example, the score for the structure
-$X_1\rightarrow X_2$ should be the same as the score for the model
-$X_2 \rightarrow X_1$ since they both correspond to the hypothesis
-that variables $X_1$ and $X_2$ are statistically dependent on each
-other.  It can be shown~\cite{Heck95} that to achieve this, not all
-the hyperparameters $\alpha$ are possible and for practical reasons
-Buntine~\cite{Bunt91} suggested a so called BDeu score with just one
-hyperparameter $\alpha\in R_{++}$ so that $\theta_{ij\cdot}\sim
-Dir(\frac{\alpha}{q_i r_i},\ldots,\frac{\alpha}{q_i r_i})$.  However,
-it soon turned out that BDeu score was very sensitive to the selection
-of this hyperparameter~\cite{cosco.uai07} and that for small sample
-sizes this method detects spurious correlations~\cite{Steck08} leading
-to models with suspiciously many parameters.
-
-A natural solution to avoid parameter sensitivity of BDeu would be to
-use a normalized maximum likelihood (NML)
-criterion~\cite{Shta87,Riss96a}, i.e., to find the structure $G$ that
-maximizes
-\begin{equation}
-P_{NML}(D;G)=\frac{P(D|\hat\theta(D;G))}{\sum_{D'}{P(D'|\hat\theta(D';G))}},
-\end{equation}
-where $\hat\theta$ denotes the (easy to find) maximum likelihood
-parameters and the sum in the denominator goes over all the possible
-$N\times n$ data matrices. While it is easy to see that this criterion
-satisfies the requirement of giving equal scores to equal structures,
-the normalizing constant renders the computation
-infeasible. Consequently, Silander et al.~\cite{cosco.pgm08a}
-suggested solving the parameter sensitivity problem by adopting the
-NML-code to the column partitions, i.e., changing the Bayesian mixture
-in equation~(\ref{eqn:bayesmix}) to
-\begin{equation}
-P^1_{NML}(D[G_i=j,i];G)=\frac{P(D|\hat\theta(D[G_i=j,i];G))}{\sum_{D'}{P(D';\hat\theta(D'|G))}},
-\end{equation}
-where $D'\in{\{1,\ldots,r_i\}}^{|D[G_i=j,i]|}$.  The logarithm of the
-denominator is often called a regret, since it indicates the extra
-code length needed compared to the code length obtained using the (a
-priori unknown) maximum likelihood parameters. The regret for
-$P^1_{NML}$ depends only on the length $N$ of the categorical data
-vector with $r$ different categorical values.  While the naive
-computation of the regret is still prohibitive, it can be approximated
-efficiently using a so called Szpankowski
-approximation~\cite{cosco.aistat03}:
-\begin{eqnarray}
-\label{eqn:szp}
-\lefteqn{reg(N,r) \approx \frac{\sqrt{2} r \Gamma{\left(\frac{r}{2} \right)}}{3 \sqrt{N} \Gamma{\left(\frac{r-1}{2}  \right)}}} \\
- &&+ \left(\frac{r-1}{2}\right) \log{\left (\frac{N}{2} \right )}- \log \Gamma{\left(\frac{r}{2} \right)} + \frac{1}{2} \log{\left (\pi \right )}\nonumber\\
- &&+ \frac{r^{2} \Gamma^{2}{\left(\frac{r}{2} \right)}}{9N \Gamma^{2}{\left(\frac{r-1}{2}\right)}} + \frac{2r^3-3r^2-2r+3}{36N}\nonumber.
-\end{eqnarray}
-fNML solves the parameter sensitivity problem and yields predictive
-models superior to BDeu.  However, the criterion does not satisfy the
-property of giving the same score for the models that correspond to
-the same dependence statements. Furthermore, the learned structures
-still sometimes appear surprisingly complicated.
-
-\section{QUOTIENT NML SCORE}
-
-We will now introduce a quotient normalized maximum likelihood (qNML)
-criterion for learning Bayesian network structures.  While equally
-efficient to compute than BDeu and fNML, it is free from
-hyperparameters, and it can be proven to give equal scores to
-equivalent models. Furthermore, it coincides with the actual NML score
-for exponentially many models. In our empirical tests it produces
-models featuring good predictive performance with significantly
-simpler structures than BDeu and fNML.
-
-Like BDeu and fNML, qNML can be expressed as a product of $n$ terms,
-one for each variable, but unlike the other two, it is not based on
-further partitioning the corresponding data column
-\begin{eqnarray}
-s^{qNML}(D;G) & := & \sum_{i=1}^n s^{qNML}_i(D;G)\\
- & := & \sum_{i=1}^n \frac{P^1_{NML}(D[\cdot,(i,G_i)];G)}{P^1_{NML}(D[\cdot,G_i];G)}.\nonumber
-\end{eqnarray}
-The trick here is to model a subset of columns as though there were no
-conditional independencies among the corresponding variables $S
-\subset X$.  In this case, we can collapse the $\prod_{X_i\in S} r_i$
-value configurations and consider them as values of a single variable
-with $\prod_{X_i\in S} r_i$ different values which can then be modeled
-with a one-dimensional $P^1_{NML}$ code.  The $s^{qNML}$ does not
-necessarily define a distribution for $D$, but it is easy to verify
-that it coincides with the $P_{NML}(D;G)$ for all the networks that
-are composed of fully connected components.  The number of such
-networks equals the number of nonempty partitions of a set of $n$
-elements, i.e., the $n^\text{th}$ Bell number.
-
-\subsection {On implementation}
-
-Collapsing many variables into a single one causes $r$ in the
-equation~(\ref{eqn:szp}) to be exponentially large (in our experiments
-around $2^{20}$).  This causes the $\Gamma^2$ terms to grow very large
-calling for care when computing the ratios of such huge numbers. More
-specifically, scoring the $i^\text{th}$ data column, i.e., computing
-$s^{qNML}_i$, requires the difference $dreg(N,q_i,r_i)=reg(N,q_i) -
-reg(N,r_i q_i)$, where $q$ denotes the number of values in the
-collapsed parent variable.  A numerically stable approximation is:
-\begin{eqnarray}
-\lefteqn{dreg(N,q,r)\approx\frac{1-r^2}{18 N} q^{2}} \\
-&&+ \frac{\left(q - 1\right) \sqrt{4 q - 2} + \left(1 - q r\right) \sqrt{4 q r - 2}}{6 \sqrt{N}} \nonumber\\
-&&+\frac{q\left(3 N \left(r \log{r}  - \left(r - 1\right) \left(\log{N} + 1\right)\right) + r\right)}{6 N} \nonumber\\
-&&+ \frac{1}{12}(\log{(r+\frac{1-r}{q+1})}- 7\log{r}).\nonumber
-\end{eqnarray}
-
 \section{EXPERIMENTAL RESULTS}
@@
 \end{table}
-
-\section{CONCLUSION}
-
-We have presented qNML, a new model selection criterion for learning
-structures of Bayesian networks.  While being competitive in
-predictive terms, it often yields significantly simpler models than
-other common model selection criteria.  The computational cost of qNML
-equals the cost of earlier criteria.  The criterion also gives equal
-scores for models that encode same independence hypotheses about the
-joint probability distribution.  qNML also coincides with NML
-criterion for exponentially many models.
-
-\subsubsection*{Acknowledgements}
-
-Use unnumbered third level headings for the acknowledgements title.
-All acknowledgements go at the end of the paper.
-
-%%\subsubsection*{References}
-
-\bibliographystyle{apalike}
-\bibliography{cosco}
-
-\end{document}

### 2017-03-16 intro.tex
Splitting main to sections

--- 2017-03-15 main.tex
@@
-\documentclass[letterpaper]{article}
-\usepackage{proceed2e}
-\usepackage[margin=1in]{geometry}
-\usepackage{amsmath,epsfig,booktabs}
-
-% Set the typeface to Times Roman
-\usepackage{times}
-
-\title{Quotient Normalized Maximum Likelihood Criterion for Learning Bayesian Network Structures}
-
-\author{} % LEAVE BLANK FOR ORIGINAL SUBMISSION.
-          % UAI  reviewing is double-blind.
-
-% The author names and affiliations should appear only in the accepted paper.
-%
-%\author{ {\bf Harry Q.~Bovik\thanks{Footnote for author to give an
-%alternate address.}} \\
-%Computer Science Dept. \\
-%Cranberry University\\
-%Pittsburgh, PA 15213 \\
-%\And
-%{\bf Coauthor}  \\
-%Affiliation          \\
-%Address \\
-%\And
-%{\bf Coauthor}   \\
-%Affiliation \\
-%Address    \\
-%(if needed)\\
-%}
-
-% Example definitions.
-% --------------------
-%\def\x{{\mathbf x}}
-%\def\L{{\cal L}}
-
-\hyphenation{col-umns}
-\hyphenation{Bayes-ian}
-\hyphenation{oth-er}
-
-\begin{document}
-
-\maketitle
-
-\begin{abstract}
-Learning the dependency structure of a multivariate distribution from
-the observational data is an important task since it allows us to
-speculate about the underlying causal mechanisms that induce
-dependencies. This learning task can be approached as a model
-selection problem, but recent studies have revealed that the popular
-Bayesian model selection criterion is not satisfactory. We review some
-of the information theoretic alternatives and introduce a new one
-called a quotient normalized maximum likelihood criterion.
-\end{abstract}
-
 \section{INTRODUCTION}
@@
 yielding parsimonious models with good predictive capabilities.
-
-\section{BAYESIAN NETWORKS}
-\label{sec:bns}
-
-Bayesian networks are a general way to describe the dependencies
-between the components of an $n$\nobreakdash-dimensional discrete data
-vector $X=(X_{1},\ldots,X_{n})$ in which the component $X_{i}$ may
-take any of the discrete values in a set $\{1,\ldots,r_{i}\}$.
-Despite denoting the values with small integers, the model will treat
-the components of $X$ categorical.
-
-\subsection{Likelihood}
-\label{ssec:likelihood}
-
-Bayesian network $B=(G,\theta)$ defines a probability distribution for
-$X$. The component $G$ defines the structure of the model as a
-directed acyclic graph that has exactly one node for each component of
-$X$. The structure $G=(G_{1},\ldots,G_{n})$ defines for each
-variable/node $X_{i}$ its (possibly empty) parent set $G_{i}$, i.e.,
-the nodes from which there are a directed edges to the variable
-$X_{i}$.
-
-Given a realization $x$ of $X$, we denote the sub\nobreakdash-vector
-of $x$ that consists of the values of the parents of $X_{i}$ in $x$ as
-$G_{i}(x)$. It is customary to enumerate all the possible
-sub\nobreakdash-vectors $G_{i}(X)$ from $1$ to $q_{i}=\prod_{h\in
-  G_{i}}r_{h}.$ In case $G_{i}$ is empty, we define $q_{i}=1$ and
-$P(G_{i}(x)=1)=1$ for all vectors $x$.
-
-For each variable $X_{i}$ there is a $q_{i}\times r_{i}$ table
-$\theta_{i}$ of parameters whose $j^{\textnormal{th}}$ row and
-$k^{\textnormal{th}}$ column defines the conditional probability
-$P(X_{i}=k\mid G_{i}(X)=j;\theta)=\theta_{ijk}$.  With structure $G$
-and parameters $\theta$, we can now express the likelihood function of
-the model as
-\begin{equation}
-P(x|G,\theta)=\prod_{i=1}^{n}P(x_{i}\mid G_{i}(x);\theta_{i})=\prod_{i=1}^{n}\theta_{iG_{i}(x)x_{i}}.
-\end{equation}
-
-\subsection{Bayesian structure learning}
-
-Bayesian learning of Bayesian network structures is based on the
-posterior probability $P(G|D,\alpha)$, where $\alpha$ denotes the
-hyper\nobreakdash-parameters for the model parameters $\theta$, and
-the $D$ is a collection of $N$ $n$\nobreakdash-dimensional i.i.d. data
-vectors collected to a $N\times n$ design matrix.  We use structure
-$G$ to extend the common matrix indexing notation $D[i,j]$, and allow
-more general row selectors.  Notably the notation $D[G_i=j,i]$ is used
-to denote those rows of the column $i$ in which the parents $G_i$
-contain the value configuration number $j\in\{1,\ldots,q_i\}$.
-
-It is common to assume the uniform prior for structures, in which case
-the objective function for structure learning is reduced to the
-marginal likelihood $P(D|G,\alpha)$.  If the model parameters
-$\theta_{ij}$ are further assumed to be independently Dirichlet
-distributed only depending on $i$ and $G_{i}$ and the data $D$ is
-assumed to have no missing values, the marginal likelihood can be
-decomposed as
-\begin{eqnarray}
-\label{eqn:bayesmix}
-P(D|G,\alpha) & = & \prod_{i=1}^{n}\prod_{j=1}^{q_i}\int
-P(D[G_i=j,i]|\alpha)\\ & = & \prod_{i=1}^{n}\prod_{j=1}^{q_i}\int
-P([G_i=j,i]|\theta)P(\theta|\alpha) d\theta.\nonumber
-\end{eqnarray}
-In coding terms this means that each data column $D[\cdot,i]$ is first partitioned based on the
-values in columns $G_i$, and each part is then coded using a Bayesian mixture that can be expressed
-in a closed form~\cite{Bunt91, Heck95}.
-
-\subsection {Problems, solutions and problems}
-
-Finding the satisfactory Dirichlet hyperparameters for the Bayesian
-mixture above has, however, turned out to be problematic. Early on,
-one of the desiderata for a good model selection criterion was that it
-would yield equal scores for essentially equivalent
-models~\cite{Verm90}.  For example, the score for the structure
-$X_1\rightarrow X_2$ should be the same as the score for the model
-$X_2 \rightarrow X_1$ since they both correspond to the hypothesis
-that variables $X_1$ and $X_2$ are statistically dependent on each
-other.  It can be shown~\cite{Heck95} that to achieve this, not all
-the hyperparameters $\alpha$ are possible and for practical reasons
-Buntine~\cite{Bunt91} suggested a so called BDeu score with just one
-hyperparameter $\alpha\in R_{++}$ so that $\theta_{ij\cdot}\sim
-Dir(\frac{\alpha}{q_i r_i},\ldots,\frac{\alpha}{q_i r_i})$.  However,
-it soon turned out that BDeu score was very sensitive to the selection
-of this hyperparameter~\cite{cosco.uai07} and that for small sample
-sizes this method detects spurious correlations~\cite{Steck08} leading
-to models with suspiciously many parameters.
-
-A natural solution to avoid parameter sensitivity of BDeu would be to
-use a normalized maximum likelihood (NML)
-criterion~\cite{Shta87,Riss96a}, i.e., to find the structure $G$ that
-maximizes
-\begin{equation}
-P_{NML}(D;G)=\frac{P(D|\hat\theta(D;G))}{\sum_{D'}{P(D'|\hat\theta(D';G))}},
-\end{equation}
-where $\hat\theta$ denotes the (easy to find) maximum likelihood
-parameters and the sum in the denominator goes over all the possible
-$N\times n$ data matrices. While it is easy to see that this criterion
-satisfies the requirement of giving equal scores to equal structures,
-the normalizing constant renders the computation
-infeasible. Consequently, Silander et al.~\cite{cosco.pgm08a}
-suggested solving the parameter sensitivity problem by adopting the
-NML-code to the column partitions, i.e., changing the Bayesian mixture
-in equation~(\ref{eqn:bayesmix}) to
-\begin{equation}
-P^1_{NML}(D[G_i=j,i];G)=\frac{P(D|\hat\theta(D[G_i=j,i];G))}{\sum_{D'}{P(D';\hat\theta(D'|G))}},
-\end{equation}
-where $D'\in{\{1,\ldots,r_i\}}^{|D[G_i=j,i]|}$.  The logarithm of the
-denominator is often called a regret, since it indicates the extra
-code length needed compared to the code length obtained using the (a
-priori unknown) maximum likelihood parameters. The regret for
-$P^1_{NML}$ depends only on the length $N$ of the categorical data
-vector with $r$ different categorical values.  While the naive
-computation of the regret is still prohibitive, it can be approximated
-efficiently using a so called Szpankowski
-approximation~\cite{cosco.aistat03}:
-\begin{eqnarray}
-\label{eqn:szp}
-\lefteqn{reg(N,r) \approx \frac{\sqrt{2} r \Gamma{\left(\frac{r}{2} \right)}}{3 \sqrt{N} \Gamma{\left(\frac{r-1}{2}  \right)}}} \\
- &&+ \left(\frac{r-1}{2}\right) \log{\left (\frac{N}{2} \right )}- \log \Gamma{\left(\frac{r}{2} \right)} + \frac{1}{2} \log{\left (\pi \right )}\nonumber\\
- &&+ \frac{r^{2} \Gamma^{2}{\left(\frac{r}{2} \right)}}{9N \Gamma^{2}{\left(\frac{r-1}{2}\right)}} + \frac{2r^3-3r^2-2r+3}{36N}\nonumber.
-\end{eqnarray}
-fNML solves the parameter sensitivity problem and yields predictive
-models superior to BDeu.  However, the criterion does not satisfy the
-property of giving the same score for the models that correspond to
-the same dependence statements. Furthermore, the learned structures
-still sometimes appear surprisingly complicated.
-
-\section{QUOTIENT NML SCORE}
-
-We will now introduce a quotient normalized maximum likelihood (qNML)
-criterion for learning Bayesian network structures.  While equally
-efficient to compute than BDeu and fNML, it is free from
-hyperparameters, and it can be proven to give equal scores to
-equivalent models. Furthermore, it coincides with the actual NML score
-for exponentially many models. In our empirical tests it produces
-models featuring good predictive performance with significantly
-simpler structures than BDeu and fNML.
-
-Like BDeu and fNML, qNML can be expressed as a product of $n$ terms,
-one for each variable, but unlike the other two, it is not based on
-further partitioning the corresponding data column
-\begin{eqnarray}
-s^{qNML}(D;G) & := & \sum_{i=1}^n s^{qNML}_i(D;G)\\
- & := & \sum_{i=1}^n \frac{P^1_{NML}(D[\cdot,(i,G_i)];G)}{P^1_{NML}(D[\cdot,G_i];G)}.\nonumber
-\end{eqnarray}
-The trick here is to model a subset of columns as though there were no
-conditional independencies among the corresponding variables $S
-\subset X$.  In this case, we can collapse the $\prod_{X_i\in S} r_i$
-value configurations and consider them as values of a single variable
-with $\prod_{X_i\in S} r_i$ different values which can then be modeled
-with a one-dimensional $P^1_{NML}$ code.  The $s^{qNML}$ does not
-necessarily define a distribution for $D$, but it is easy to verify
-that it coincides with the $P_{NML}(D;G)$ for all the networks that
-are composed of fully connected components.  The number of such
-networks equals the number of nonempty partitions of a set of $n$
-elements, i.e., the $n^\text{th}$ Bell number.
-
-\subsection {On implementation}
-
-Collapsing many variables into a single one causes $r$ in the
-equation~(\ref{eqn:szp}) to be exponentially large (in our experiments
-around $2^{20}$).  This causes the $\Gamma^2$ terms to grow very large
-calling for care when computing the ratios of such huge numbers. More
-specifically, scoring the $i^\text{th}$ data column, i.e., computing
-$s^{qNML}_i$, requires the difference $dreg(N,q_i,r_i)=reg(N,q_i) -
-reg(N,r_i q_i)$, where $q$ denotes the number of values in the
-collapsed parent variable.  A numerically stable approximation is:
-\begin{eqnarray}
-\lefteqn{dreg(N,q,r)\approx\frac{1-r^2}{18 N} q^{2}} \\
-&&+ \frac{\left(q - 1\right) \sqrt{4 q - 2} + \left(1 - q r\right) \sqrt{4 q r - 2}}{6 \sqrt{N}} \nonumber\\
-&&+\frac{q\left(3 N \left(r \log{r}  - \left(r - 1\right) \left(\log{N} + 1\right)\right) + r\right)}{6 N} \nonumber\\
-&&+ \frac{1}{12}(\log{(r+\frac{1-r}{q+1})}- 7\log{r}).\nonumber
-\end{eqnarray}
-
-\section{EXPERIMENTAL RESULTS}
-
-To empirically compare the model selection criteria we took 20 UCI
-data sets~\cite{Lichman:2013} and run 1000 train and test experiments
-for 20 different sample sizes for all of them. The training was
-conducted using a dynamic programming based exact structure
-learning\cite{cosco.uai06} which limited the number $n$ of variables
-to less than 20.
-
-Figure~\ref{fig:bcmean} shows an example of the behavior of the
-studied selection criteria.  For small sample sizes fNML and qNML
-yield equally accurate predictions while BDeu requires more samples to
-obtain competitive performance. The curves are mean curves from 1000
-independent train and test splits.
-
-\begin{figure}
-\centering
-\includegraphics[width=8cm,height=5cm]{breast_cancer_mean.pdf}
-\caption{Predictive log-loss in a breast cancer data as a function of
-  sample size for different model selection criteria.}
-\label{fig:bcmean}
-\end{figure}
-
-Figure~\ref{fig:bcnpmean} shows how fNML still sometimes behaves
-strangely in terms of model complexity here measured by the number of
-parameters in the model.  qNML, instead, appears to yield more
-parsimonious models.
-
-\begin{figure}
-\centering
-\includegraphics[width=8cm,height=5cm]{breast_cancer_np_mean.pdf}
-\caption{Number of parameters in a breast cancer model as a function
-  of sample size for different model selection criteria.}
-\label{fig:bcnpmean}
-\end{figure}
-
-\begin{table}
-\centering
-\begin{tabular}{crrrr}
-\toprule
-           Data &     N &   BDeu &            fNML &            qNML \\
-\midrule
-           Iris &    16 &   3.59 &            3.47 &   \textbf{3.43} \\
-  PostOperative &    20 &  10.24 &   \textbf{8.27} &            8.32 \\
-           Wine &    27 &  18.11 &           12.68 &  \textbf{12.51} \\
- HeartHungarian &    45 &  10.15 &            9.10 &   \textbf{8.90} \\
-   BreastCancer &    45 &  12.01 &  \textbf{10.57} &           10.66 \\
- HeartCleveland &    48 &  16.36 &           12.81 &  \textbf{12.54} \\
-          Ecoli &    51 &   5.77 &   \textbf{5.30} &            5.37 \\
-          Liver &    54 &   4.07 &            3.97 &   \textbf{3.91} \\
-          Glass &    55 &   7.16 &   \textbf{6.22} &            6.28 \\
-        Thyroid &    55 &   2.83 &   \textbf{2.75} &            2.77 \\
-   HeartStatlog &    56 &  14.15 &           11.92 &  \textbf{11.60} \\
-      TicTacToe &    96 &  10.99 &  \textbf{10.73} &           11.05 \\
-        Balance &    96 &   7.48 &   \textbf{7.39} &            7.78 \\
-    BcWisconsin &   105 &   5.81 &            5.12 &   \textbf{5.08} \\
-       Diabetes &   117 &   4.95 &            4.95 &   \textbf{4.89} \\
-          Yeast &   225 &   5.54 &   \textbf{5.40} &            5.42 \\
-        Abalone &   418 &   3.93 &   \textbf{3.88} &            3.96 \\
-     PageBlocks &   548 &   2.36 &   \textbf{2.32} &            2.38 \\
-        Shuttle &  2900 &   1.70 &   \textbf{1.69} &            1.72 \\
-          Adult &  4885 &  10.22 &  \textbf{10.08} &           10.12 \\
-\bottomrule
-\end{tabular}
-\caption{Predictive log losses for small sample sizes for different model selection criteria in 20 different datasets.}
-\label{tbl:preds}
-	\end{table}
-
-Table~\ref{tbl:preds} features predictive losses for the small sample
-sizes.  (The results for large sample sizes converge for all sensible
-criteria.)  As we can see, fNML still usually (12/20) performs best
-but qNML is not much worse while BDeu may sometimes perform much
-worse.
-
-Looking at the number of parameters for the same datasets and sample
-sizes (Table~\ref{tbl:nofparams}) reveals that qNML usually (18/20)
-yields simplest models, and sometimes, like in the BreastCancer data,
-the differences are of orders of magnitude.
-
-\begin{table}
-\centering
-\begin{tabular}{crrrr}
-\toprule
-           Data &     N &          BDeu &  fNML &          qNML \\
-\midrule
-           Iris &    16 &            36 &    33 &   \textbf{28} \\
-  PostOperative &    20 &          1304 &   464 &  \textbf{115} \\
-           Wine &    27 &         11559 &   780 &  \textbf{214} \\
- HeartHungarian &    45 &          4706 &   838 &  \textbf{100} \\
-   BreastCancer &    45 &         26774 &  2710 &  \textbf{246} \\
- HeartCleveland &    48 &         29346 &  1407 &  \textbf{204} \\
-          Ecoli &    51 &           113 &   167 &   \textbf{80} \\
-          Liver &    54 &            40 &    60 &   \textbf{24} \\
-          Glass &    55 &          1488 &   572 &  \textbf{100} \\
-        Thyroid &    55 &            37 &    70 &   \textbf{28} \\
-   HeartStatlog &    56 &         18414 &  1099 &  \textbf{159} \\
-      TicTacToe &    96 &         13645 &  2438 &  \textbf{696} \\
-        Balance &    96 &   \textbf{20} &   237 &           168 \\
-    BcWisconsin &   105 &          5921 &   816 &   \textbf{88} \\
-       Diabetes &   117 &            35 &   216 &   \textbf{34} \\
-          Yeast &   225 &   \textbf{78} &   307 &            80 \\
-        Abalone &   418 &            90 &   156 &   \textbf{62} \\
-     PageBlocks &   548 &           699 &   377 &   \textbf{56} \\
-        Shuttle &  2900 &           380 &   594 &  \textbf{111} \\
-          Adult &  4885 &  \textbf{599} &  1431 &           876 \\
-\bottomrule
-\end{tabular}
-\caption{Number of model parameters for small sample sizes for different model selection criteria in 20 different datasets.}
-\label{tbl:nofparams}
-\end{table}
-
-\section{CONCLUSION}
-
-We have presented qNML, a new model selection criterion for learning
-structures of Bayesian networks.  While being competitive in
-predictive terms, it often yields significantly simpler models than
-other common model selection criteria.  The computational cost of qNML
-equals the cost of earlier criteria.  The criterion also gives equal
-scores for models that encode same independence hypotheses about the
-joint probability distribution.  qNML also coincides with NML
-criterion for exponentially many models.
-
-\subsubsection*{Acknowledgements}
-
-Use unnumbered third level headings for the acknowledgements title.
-All acknowledgements go at the end of the paper.
-
-%%\subsubsection*{References}
-
-\bibliographystyle{apalike}
-\bibliography{cosco}
-
-\end{document}

### 2017-03-16 main.tex
Splitting main to sections

--- 2017-03-15 main.tex
@@
 
-% Example definitions.
-% --------------------
-%\def\x{{\mathbf x}}
-%\def\L{{\cal L}}
-
 \hyphenation{col-umns}
@@
 
-\begin{abstract}
-Learning the dependency structure of a multivariate distribution from
-the observational data is an important task since it allows us to
-speculate about the underlying causal mechanisms that induce
-dependencies. This learning task can be approached as a model
-selection problem, but recent studies have revealed that the popular
-Bayesian model selection criterion is not satisfactory. We review some
-of the information theoretic alternatives and introduce a new one
-called a quotient normalized maximum likelihood criterion.
-\end{abstract}
-
-\section{INTRODUCTION}
-\label{sec:intro}
-Bayesian networks~\cite{Pear88} are popular models for presenting
-multivariate statistical dependencies that may have been induced by
-underlying causal mechanisms.  Techniques for learning the structure
-of Bayesian networks from the observational data has therefore been
-used for many tasks such as discovering cell signaling pathways from
-the protein activity data~\cite{bn4sigpath02}, revealing the business
-process structures~\cite{bn4bpmining} from the transaction logs and
-modeling brain region connectivity using fMRI
-data~\cite{bn4brainconnect}.
-
-Learning the structure of statistical dependencies can be seen as a
-model selection task where each model is a different hypothesis about
-the conditional dependencies between sets of variables. Traditional
-model selection criteria such as the Akaike information
-criterion~\cite{Akai73} and the Bayesian information
-criterion~\cite{Schw78} have also been used for the task, but recent
-comparisons have not been favorable for these criteria in terms of
-structural stability and/or predictive
-performance~\cite{cosco.pgm08a}.  The most popular criterion has been
-the marginal likelihood (usually called BDeu for reasons explained
-later), but recent studies~\cite{cosco.uai07,Steck08} have found this
-criterion to be very sensitive to hyper parameters and to yield
-undesirably complex models for small sample sizes.
-
-The information theoretic normalized maximum likelihood (NML)
-criterion~\cite{Shta87,Riss96a} would otherwise be an ideal candidate
-for a good criterion.  but its exact calculation is likely to be
-prohibitively demanding. In 2008, Silander et al. introduced a
-hyper-parameter free, NML inspired criterion called a factorized NML
-(fNML)~\cite{cosco.pgm08a} that was shown to yield good predictive
-models without sensitivity problems.  However, from the structure
-learning point of view, the fNML still sometimes appears to yield
-overly complex models. In this paper we introduce another NML related
-criterion, a quotient NML (qNML) that yields simpler models without
-sacrificing predictive accuracy. Furthermore, unlike the fNML, the
-qNML gives equal scores to the structures that encode the same
-independence statements. Like other common model selection criteria,
-the qNML is also consistent.
-
-We will next briefly introduce Bayesian networks and then review the
-BDeu and fNML criteria and introduce the qNML criterion.  We will also
-summarize the results for 20 datasets to back up our claim of qNML
-yielding parsimonious models with good predictive capabilities.
-
-\section{BAYESIAN NETWORKS}
-\label{sec:bns}
-
-Bayesian networks are a general way to describe the dependencies
-between the components of an $n$\nobreakdash-dimensional discrete data
-vector $X=(X_{1},\ldots,X_{n})$ in which the component $X_{i}$ may
-take any of the discrete values in a set $\{1,\ldots,r_{i}\}$.
-Despite denoting the values with small integers, the model will treat
-the components of $X$ categorical.
-
-\subsection{Likelihood}
-\label{ssec:likelihood}
-
-Bayesian network $B=(G,\theta)$ defines a probability distribution for
-$X$. The component $G$ defines the structure of the model as a
-directed acyclic graph that has exactly one node for each component of
-$X$. The structure $G=(G_{1},\ldots,G_{n})$ defines for each
-variable/node $X_{i}$ its (possibly empty) parent set $G_{i}$, i.e.,
-the nodes from which there are a directed edges to the variable
-$X_{i}$.
-
-Given a realization $x$ of $X$, we denote the sub\nobreakdash-vector
-of $x$ that consists of the values of the parents of $X_{i}$ in $x$ as
-$G_{i}(x)$. It is customary to enumerate all the possible
-sub\nobreakdash-vectors $G_{i}(X)$ from $1$ to $q_{i}=\prod_{h\in
-  G_{i}}r_{h}.$ In case $G_{i}$ is empty, we define $q_{i}=1$ and
-$P(G_{i}(x)=1)=1$ for all vectors $x$.
-
-For each variable $X_{i}$ there is a $q_{i}\times r_{i}$ table
-$\theta_{i}$ of parameters whose $j^{\textnormal{th}}$ row and
-$k^{\textnormal{th}}$ column defines the conditional probability
-$P(X_{i}=k\mid G_{i}(X)=j;\theta)=\theta_{ijk}$.  With structure $G$
-and parameters $\theta$, we can now express the likelihood function of
-the model as
-\begin{equation}
-P(x|G,\theta)=\prod_{i=1}^{n}P(x_{i}\mid G_{i}(x);\theta_{i})=\prod_{i=1}^{n}\theta_{iG_{i}(x)x_{i}}.
-\end{equation}
-
-\subsection{Bayesian structure learning}
-
-Bayesian learning of Bayesian network structures is based on the
-posterior probability $P(G|D,\alpha)$, where $\alpha$ denotes the
-hyper\nobreakdash-parameters for the model parameters $\theta$, and
-the $D$ is a collection of $N$ $n$\nobreakdash-dimensional i.i.d. data
-vectors collected to a $N\times n$ design matrix.  We use structure
-$G$ to extend the common matrix indexing notation $D[i,j]$, and allow
-more general row selectors.  Notably the notation $D[G_i=j,i]$ is used
-to denote those rows of the column $i$ in which the parents $G_i$
-contain the value configuration number $j\in\{1,\ldots,q_i\}$.
-
-It is common to assume the uniform prior for structures, in which case
-the objective function for structure learning is reduced to the
-marginal likelihood $P(D|G,\alpha)$.  If the model parameters
-$\theta_{ij}$ are further assumed to be independently Dirichlet
-distributed only depending on $i$ and $G_{i}$ and the data $D$ is
-assumed to have no missing values, the marginal likelihood can be
-decomposed as
-\begin{eqnarray}
-\label{eqn:bayesmix}
-P(D|G,\alpha) & = & \prod_{i=1}^{n}\prod_{j=1}^{q_i}\int
-P(D[G_i=j,i]|\alpha)\\ & = & \prod_{i=1}^{n}\prod_{j=1}^{q_i}\int
-P([G_i=j,i]|\theta)P(\theta|\alpha) d\theta.\nonumber
-\end{eqnarray}
-In coding terms this means that each data column $D[\cdot,i]$ is first partitioned based on the
-values in columns $G_i$, and each part is then coded using a Bayesian mixture that can be expressed
-in a closed form~\cite{Bunt91, Heck95}.
-
-\subsection {Problems, solutions and problems}
-
-Finding the satisfactory Dirichlet hyperparameters for the Bayesian
-mixture above has, however, turned out to be problematic. Early on,
-one of the desiderata for a good model selection criterion was that it
-would yield equal scores for essentially equivalent
-models~\cite{Verm90}.  For example, the score for the structure
-$X_1\rightarrow X_2$ should be the same as the score for the model
-$X_2 \rightarrow X_1$ since they both correspond to the hypothesis
-that variables $X_1$ and $X_2$ are statistically dependent on each
-other.  It can be shown~\cite{Heck95} that to achieve this, not all
-the hyperparameters $\alpha$ are possible and for practical reasons
-Buntine~\cite{Bunt91} suggested a so called BDeu score with just one
-hyperparameter $\alpha\in R_{++}$ so that $\theta_{ij\cdot}\sim
-Dir(\frac{\alpha}{q_i r_i},\ldots,\frac{\alpha}{q_i r_i})$.  However,
-it soon turned out that BDeu score was very sensitive to the selection
-of this hyperparameter~\cite{cosco.uai07} and that for small sample
-sizes this method detects spurious correlations~\cite{Steck08} leading
-to models with suspiciously many parameters.
-
-A natural solution to avoid parameter sensitivity of BDeu would be to
-use a normalized maximum likelihood (NML)
-criterion~\cite{Shta87,Riss96a}, i.e., to find the structure $G$ that
-maximizes
-\begin{equation}
-P_{NML}(D;G)=\frac{P(D|\hat\theta(D;G))}{\sum_{D'}{P(D'|\hat\theta(D';G))}},
-\end{equation}
-where $\hat\theta$ denotes the (easy to find) maximum likelihood
-parameters and the sum in the denominator goes over all the possible
-$N\times n$ data matrices. While it is easy to see that this criterion
-satisfies the requirement of giving equal scores to equal structures,
-the normalizing constant renders the computation
-infeasible. Consequently, Silander et al.~\cite{cosco.pgm08a}
-suggested solving the parameter sensitivity problem by adopting the
-NML-code to the column partitions, i.e., changing the Bayesian mixture
-in equation~(\ref{eqn:bayesmix}) to
-\begin{equation}
-P^1_{NML}(D[G_i=j,i];G)=\frac{P(D|\hat\theta(D[G_i=j,i];G))}{\sum_{D'}{P(D';\hat\theta(D'|G))}},
-\end{equation}
-where $D'\in{\{1,\ldots,r_i\}}^{|D[G_i=j,i]|}$.  The logarithm of the
-denominator is often called a regret, since it indicates the extra
-code length needed compared to the code length obtained using the (a
-priori unknown) maximum likelihood parameters. The regret for
-$P^1_{NML}$ depends only on the length $N$ of the categorical data
-vector with $r$ different categorical values.  While the naive
-computation of the regret is still prohibitive, it can be approximated
-efficiently using a so called Szpankowski
-approximation~\cite{cosco.aistat03}:
-\begin{eqnarray}
-\label{eqn:szp}
-\lefteqn{reg(N,r) \approx \frac{\sqrt{2} r \Gamma{\left(\frac{r}{2} \right)}}{3 \sqrt{N} \Gamma{\left(\frac{r-1}{2}  \right)}}} \\
- &&+ \left(\frac{r-1}{2}\right) \log{\left (\frac{N}{2} \right )}- \log \Gamma{\left(\frac{r}{2} \right)} + \frac{1}{2} \log{\left (\pi \right )}\nonumber\\
- &&+ \frac{r^{2} \Gamma^{2}{\left(\frac{r}{2} \right)}}{9N \Gamma^{2}{\left(\frac{r-1}{2}\right)}} + \frac{2r^3-3r^2-2r+3}{36N}\nonumber.
-\end{eqnarray}
-fNML solves the parameter sensitivity problem and yields predictive
-models superior to BDeu.  However, the criterion does not satisfy the
-property of giving the same score for the models that correspond to
-the same dependence statements. Furthermore, the learned structures
-still sometimes appear surprisingly complicated.
-
-\section{QUOTIENT NML SCORE}
-
-We will now introduce a quotient normalized maximum likelihood (qNML)
-criterion for learning Bayesian network structures.  While equally
-efficient to compute than BDeu and fNML, it is free from
-hyperparameters, and it can be proven to give equal scores to
-equivalent models. Furthermore, it coincides with the actual NML score
-for exponentially many models. In our empirical tests it produces
-models featuring good predictive performance with significantly
-simpler structures than BDeu and fNML.
-
-Like BDeu and fNML, qNML can be expressed as a product of $n$ terms,
-one for each variable, but unlike the other two, it is not based on
-further partitioning the corresponding data column
-\begin{eqnarray}
-s^{qNML}(D;G) & := & \sum_{i=1}^n s^{qNML}_i(D;G)\\
- & := & \sum_{i=1}^n \frac{P^1_{NML}(D[\cdot,(i,G_i)];G)}{P^1_{NML}(D[\cdot,G_i];G)}.\nonumber
-\end{eqnarray}
-The trick here is to model a subset of columns as though there were no
-conditional independencies among the corresponding variables $S
-\subset X$.  In this case, we can collapse the $\prod_{X_i\in S} r_i$
-value configurations and consider them as values of a single variable
-with $\prod_{X_i\in S} r_i$ different values which can then be modeled
-with a one-dimensional $P^1_{NML}$ code.  The $s^{qNML}$ does not
-necessarily define a distribution for $D$, but it is easy to verify
-that it coincides with the $P_{NML}(D;G)$ for all the networks that
-are composed of fully connected components.  The number of such
-networks equals the number of nonempty partitions of a set of $n$
-elements, i.e., the $n^\text{th}$ Bell number.
-
-\subsection {On implementation}
-
-Collapsing many variables into a single one causes $r$ in the
-equation~(\ref{eqn:szp}) to be exponentially large (in our experiments
-around $2^{20}$).  This causes the $\Gamma^2$ terms to grow very large
-calling for care when computing the ratios of such huge numbers. More
-specifically, scoring the $i^\text{th}$ data column, i.e., computing
-$s^{qNML}_i$, requires the difference $dreg(N,q_i,r_i)=reg(N,q_i) -
-reg(N,r_i q_i)$, where $q$ denotes the number of values in the
-collapsed parent variable.  A numerically stable approximation is:
-\begin{eqnarray}
-\lefteqn{dreg(N,q,r)\approx\frac{1-r^2}{18 N} q^{2}} \\
-&&+ \frac{\left(q - 1\right) \sqrt{4 q - 2} + \left(1 - q r\right) \sqrt{4 q r - 2}}{6 \sqrt{N}} \nonumber\\
-&&+\frac{q\left(3 N \left(r \log{r}  - \left(r - 1\right) \left(\log{N} + 1\right)\right) + r\right)}{6 N} \nonumber\\
-&&+ \frac{1}{12}(\log{(r+\frac{1-r}{q+1})}- 7\log{r}).\nonumber
-\end{eqnarray}
-
-\section{EXPERIMENTAL RESULTS}
-
-To empirically compare the model selection criteria we took 20 UCI
-data sets~\cite{Lichman:2013} and run 1000 train and test experiments
-for 20 different sample sizes for all of them. The training was
-conducted using a dynamic programming based exact structure
-learning\cite{cosco.uai06} which limited the number $n$ of variables
-to less than 20.
-
-Figure~\ref{fig:bcmean} shows an example of the behavior of the
-studied selection criteria.  For small sample sizes fNML and qNML
-yield equally accurate predictions while BDeu requires more samples to
-obtain competitive performance. The curves are mean curves from 1000
-independent train and test splits.
-
-\begin{figure}
-\centering
-\includegraphics[width=8cm,height=5cm]{breast_cancer_mean.pdf}
-\caption{Predictive log-loss in a breast cancer data as a function of
-  sample size for different model selection criteria.}
-\label{fig:bcmean}
-\end{figure}
-
-Figure~\ref{fig:bcnpmean} shows how fNML still sometimes behaves
-strangely in terms of model complexity here measured by the number of
-parameters in the model.  qNML, instead, appears to yield more
-parsimonious models.
-
-\begin{figure}
-\centering
-\includegraphics[width=8cm,height=5cm]{breast_cancer_np_mean.pdf}
-\caption{Number of parameters in a breast cancer model as a function
-  of sample size for different model selection criteria.}
-\label{fig:bcnpmean}
-\end{figure}
-
-\begin{table}
-\centering
-\begin{tabular}{crrrr}
-\toprule
-           Data &     N &   BDeu &            fNML &            qNML \\
-\midrule
-           Iris &    16 &   3.59 &            3.47 &   \textbf{3.43} \\
-  PostOperative &    20 &  10.24 &   \textbf{8.27} &            8.32 \\
-           Wine &    27 &  18.11 &           12.68 &  \textbf{12.51} \\
- HeartHungarian &    45 &  10.15 &            9.10 &   \textbf{8.90} \\
-   BreastCancer &    45 &  12.01 &  \textbf{10.57} &           10.66 \\
- HeartCleveland &    48 &  16.36 &           12.81 &  \textbf{12.54} \\
-          Ecoli &    51 &   5.77 &   \textbf{5.30} &            5.37 \\
-          Liver &    54 &   4.07 &            3.97 &   \textbf{3.91} \\
-          Glass &    55 &   7.16 &   \textbf{6.22} &            6.28 \\
-        Thyroid &    55 &   2.83 &   \textbf{2.75} &            2.77 \\
-   HeartStatlog &    56 &  14.15 &           11.92 &  \textbf{11.60} \\
-      TicTacToe &    96 &  10.99 &  \textbf{10.73} &           11.05 \\
-        Balance &    96 &   7.48 &   \textbf{7.39} &            7.78 \\
-    BcWisconsin &   105 &   5.81 &            5.12 &   \textbf{5.08} \\
-       Diabetes &   117 &   4.95 &            4.95 &   \textbf{4.89} \\
-          Yeast &   225 &   5.54 &   \textbf{5.40} &            5.42 \\
-        Abalone &   418 &   3.93 &   \textbf{3.88} &            3.96 \\
-     PageBlocks &   548 &   2.36 &   \textbf{2.32} &            2.38 \\
-        Shuttle &  2900 &   1.70 &   \textbf{1.69} &            1.72 \\
-          Adult &  4885 &  10.22 &  \textbf{10.08} &           10.12 \\
-\bottomrule
-\end{tabular}
-\caption{Predictive log losses for small sample sizes for different model selection criteria in 20 different datasets.}
-\label{tbl:preds}
-	\end{table}
-
-Table~\ref{tbl:preds} features predictive losses for the small sample
-sizes.  (The results for large sample sizes converge for all sensible
-criteria.)  As we can see, fNML still usually (12/20) performs best
-but qNML is not much worse while BDeu may sometimes perform much
-worse.
-
-Looking at the number of parameters for the same datasets and sample
-sizes (Table~\ref{tbl:nofparams}) reveals that qNML usually (18/20)
-yields simplest models, and sometimes, like in the BreastCancer data,
-the differences are of orders of magnitude.
-
-\begin{table}
-\centering
-\begin{tabular}{crrrr}
-\toprule
-           Data &     N &          BDeu &  fNML &          qNML \\
-\midrule
-           Iris &    16 &            36 &    33 &   \textbf{28} \\
-  PostOperative &    20 &          1304 &   464 &  \textbf{115} \\
-           Wine &    27 &         11559 &   780 &  \textbf{214} \\
- HeartHungarian &    45 &          4706 &   838 &  \textbf{100} \\
-   BreastCancer &    45 &         26774 &  2710 &  \textbf{246} \\
- HeartCleveland &    48 &         29346 &  1407 &  \textbf{204} \\
-          Ecoli &    51 &           113 &   167 &   \textbf{80} \\
-          Liver &    54 &            40 &    60 &   \textbf{24} \\
-          Glass &    55 &          1488 &   572 &  \textbf{100} \\
-        Thyroid &    55 &            37 &    70 &   \textbf{28} \\
-   HeartStatlog &    56 &         18414 &  1099 &  \textbf{159} \\
-      TicTacToe &    96 &         13645 &  2438 &  \textbf{696} \\
-        Balance &    96 &   \textbf{20} &   237 &           168 \\
-    BcWisconsin &   105 &          5921 &   816 &   \textbf{88} \\
-       Diabetes &   117 &            35 &   216 &   \textbf{34} \\
-          Yeast &   225 &   \textbf{78} &   307 &            80 \\
-        Abalone &   418 &            90 &   156 &   \textbf{62} \\
-     PageBlocks &   548 &           699 &   377 &   \textbf{56} \\
-        Shuttle &  2900 &           380 &   594 &  \textbf{111} \\
-          Adult &  4885 &  \textbf{599} &  1431 &           876 \\
-\bottomrule
-\end{tabular}
-\caption{Number of model parameters for small sample sizes for different model selection criteria in 20 different datasets.}
-\label{tbl:nofparams}
-\end{table}
-
-\section{CONCLUSION}
-
-We have presented qNML, a new model selection criterion for learning
-structures of Bayesian networks.  While being competitive in
-predictive terms, it often yields significantly simpler models than
-other common model selection criteria.  The computational cost of qNML
-equals the cost of earlier criteria.  The criterion also gives equal
-scores for models that encode same independence hypotheses about the
-joint probability distribution.  qNML also coincides with NML
-criterion for exponentially many models.
+\input{abstract.tex}
+\input{intro.tex}
+\input{bn.tex} 
+\input{qnml.tex}
+\input{experiments.tex}
+\input{conclusion}
 

### 2017-03-16 qnml.tex
Splitting main to sections

\section{QUOTIENT NML SCORE}

We will now introduce a quotient normalized maximum likelihood (qNML)
criterion for learning Bayesian network structures.  While equally
efficient to compute than BDeu and fNML, it is free from
hyperparameters, and it can be proven to give equal scores to
equivalent models. Furthermore, it coincides with the actual NML score
for exponentially many models. In our empirical tests it produces
models featuring good predictive performance with significantly
simpler structures than BDeu and fNML.

Like BDeu and fNML, qNML can be expressed as a product of $n$ terms,
one for each variable, but unlike the other two, it is not based on
further partitioning the corresponding data column
\begin{eqnarray}
s^{qNML}(D;G) & := & \sum_{i=1}^n s^{qNML}_i(D;G)\\
& := & \sum_{i=1}^n \log \frac{P^1_{NML}(D[\cdot,(i,G_i)];G)}
                             {P^1_{NML}(D[\cdot,G_i];G)}.\nonumber
\end{eqnarray}
The trick here is to model a subset of columns as though there were no
conditional independencies among the corresponding variables $S
\subset X$.  In this case, we can collapse the $\prod_{X_i\in S} r_i$
value configurations and consider them as values of a single variable
with $\prod_{X_i\in S} r_i$ different values which can then be modeled
with a one-dimensional $P^1_{NML}$ code.  The $s^{qNML}$ does not
necessarily define a distribution for $D$, but it is easy to verify
that it coincides with the $P_{NML}(D;G)$ for all the networks that
are composed of fully connected components.  The number of such
networks equals the number of nonempty partitions of a set of $n$
elements, i.e., the $n^\text{th}$ Bell number.

\subsection {qNML is score equivalent}

The qNML yields equal scores for network structures that encode same
sets of independencies. Verma and Pearl~\cite{Verm90} showed that the
equivalent networks are exactly those who a) are same when directed
arcs are substitued by undirected ones and b) who have same
V-structures, i.e.  the variable triplets $(A,B,C)$ where both $A$ and
$B$ are parents of $C$, but there is no arc between $A$ and $B$ (in
neither direction).  Later, Chickering~\cite{Chick95} showed that all
the equivalent network structures, and only them, can be reached from
each other by reversing, one by one, so called covered arcs, i.e. the
arcs from node $A$ to $B$, for which $B$'s parents other than $A$ are
exactly the $A$'s parents ($G_B=\{A\}\cup G_A$).

We will next state this as a
theorem and sketch a proof for it. A more detailed proof appears in the
suplementary material.
\begin{theorem}
  \label{thm:scoreqv}
  Let $G$ and $G'$ be two Bayesian network structures that differ only
  by a single covered arc reversal, i.e., the arc from $A$ to $B$ in $G$
  has been reversed in $G'$ to point from $B$ to $A$, then
  $$s^{qNML}(D;G)=s^{qNML}(D;G').$$
\end{theorem}
\begin{proof}
  Now the scores for structures can be decomposed as the
  $s^{qNML}(D;G)=\sum_{i=1}^{n}s_i^{qNML}(D;G)$ and
  $s^{qNML}(D;G')=\sum_{i=1}^{n}s_i^{qNML}(D;G')$.  Since only the
  terms corresponding to the variables $A$ and $B$ in these sums are
  different, it is enough to show that sum of these two terms are
  equal for $G$ and $G'$. Since we can assume the data to be fixed we
  lighten up the notations and write
  $P^1_{NML}(i,G_i) := P^1_{NML}(D[\cdot,(i,G_i)];G)$ and
  $P^1_{NML}(G_i)   := P^1_{NML}(D[\cdot,G_i];G)$.
  \begin{eqnarray}
    \lefteqn{s_A^{qNML}(D;G)+s_B^{qNML}(D;G)} \nonumber\\
    && =\log\frac{P^1_{NML}(A,G_{A})}{P^1_{NML}(G_{A})}
            \frac{P^1_{NML}(B,G_{B})}{P^1_{NML}(G_{B})}\nonumber\\
    && =\log 1\cdot\frac{P^1_{NML}(B,G_{B})}{P^1_{NML}(G_{A})}\nonumber\\
    && =\log \frac{P^1_{NML}(B,G'_{B})}{P^1_{NML}(G'_{A})}
             \frac{P^1_{NML}(A,G'_{A})}{P^1_{NML}(G'_{B})}\nonumber\\
 && =s_A^{qNML}(D;G')+s_B^{qNML}(D;G'),\nonumber
\end{eqnarray}
  due to equations
  $\{A\}\cup G_A  = G_B$,
  $\{B\}\cup G'_B = G'_A$,
  $\{B\}\cup G_B  = \{A\} \cup G'_A$,
  and $G_A        = G'_B$.
\end{proof}

\subsection{qNML is consistent}

\subsection {On implementation}

Collapsing many variables into a single one causes $r$ in the
equation~(\ref{eqn:szp}) to be exponentially large (in our experiments
around $2^{20}$).  This causes the $\Gamma^2$ terms to grow very large
calling for care when computing the ratios of such huge numbers. More
specifically, scoring the $i^\text{th}$ data column, i.e., computing
$s^{qNML}_i$, requires the difference $dreg(N,q_i,r_i)=reg(N,q_i) -
reg(N,r_i q_i)$, where $q$ denotes the number of values in the
collapsed parent variable.  A numerically stable approximation is:
\begin{eqnarray}
\lefteqn{dreg(N,q,r)\approx\frac{1-r^2}{18 N} q^{2}} \\
&&+ \frac{\left(q - 1\right) \sqrt{4 q - 2} + \left(1 - q r\right) \sqrt{4 q r - 2}}{6 \sqrt{N}} \nonumber\\
&&+\frac{q\left(3 N \left(r \log{r}  - \left(r - 1\right) \left(\log{N} + 1\right)\right) + r\right)}{6 N} \nonumber\\
&&+ \frac{1}{12}(\log{(r+\frac{1-r}{q+1})}- 7\log{r}).\nonumber
\end{eqnarray}

### 2017-03-16 qnml.tex
consistency proof added

@@
 
-\subsection {On implementation}
+One important property possessed by nearly every model selection criterion is consistency.
+In our context, consistency means that given a data matrix with $N$ samples coming from a distribution faithful to some DAG $G$, the qNML will give the highest score to true graph $G$ with a probability tending to one as $N$ increases.
+We will show this by first proving that qNML is asymptotically equivalent to widely used BIC criterion which is known to be consistent \cite{Schw78, Haug88}.
+
+The BIC can be written as
+\begin{equation}
+\textnormal{BIC}(D,G) = \sum_{i = 1}^n \log P(D_i \ | \ \hat{\theta}_{i | G_i} ) - \frac{q_i(r_i - 1)}{2} \log N,
+\end{equation}where $D_i = D[\cdot, i]$ stands for the data on variable $i$, $\hat{\theta}_{i | G_i}$ denotes the maximum likelihood parameters of the conditional distribution of variable $i$ given its parents in $G$.
+Since the both BIC and qNML scores are decomposable, we can focus on studying the local scores.
+Using the definition of qNML, we have
+\begin{align}\label{qnmlDef2}
+s^{qNML}_i(D;G) &= \log \frac{P(D_{i, G_i} \ | \ \hat{\theta}_{i, G_i} )}{P(D_{G_i} \ | \ \hat{\theta}_{G_i} )} \notag \\
+& -(reg(N,q_i r_i) - reg(N,q_i)).
+\end{align}
+
+Now, we observe that the first term in (\ref{qnmlDef2}) can written as
+$$
+\log \frac{P(D_{i, G_i} \ | \ \hat{\theta}_{i, G_i} )}{P(D_{G_i} \ | \ \hat{\theta}_{G_i} )} = \log P(D_i \ | \ \hat{\theta}_{i | G_i} )
+$$ which simply shows that the goodness-of-the-fit terms agree between the criteria.
+Next, we consider the difference of regrets in (\ref{qnmlDef2}) which corresponds to the penalty term of BIC.
+The following theorem states that these two are asymptotically equal.
+
+\begin{theorem}
+Let $r_i$ and $q_i$ denote the number of possible values for variable $X_i$ and the possible configurations of parents $X_{G_i}$, respectively. As $N \to \infty$,
+$$reg(N,q_i r_i) - reg(N,q_i) = \frac{q_i(r_i - 1)}{2}\log N + O(1).$$
+\end{theorem}   
+\begin{proof}
+By considering only the leading term in (\ref{eqn:szp}), the regret for a single multinomial variable with $m$ categories can be written asymptotically as
+$$
+reg(N,m) = \frac{m-1}{2}\log N + O(1).
+$$ Using this, we have
+\begin{align*}
+reg(N,q_i r_i)& -  reg(N,q_i) \\ 
+&= \frac{q_ir_i-1}{2}\log N-\frac{q_i-1}{2}\log N + O(1) \\
+&= \frac{q_ir_i-1-q_i + 1}{2}\log N + O(1) \\
+&= \frac{q_i(r_i - 1)}{2} \log N + O(1).
+\end{align*}
+\end{proof}
+This suffices in order to show that qNML is asymptotically equivalent to BIC and, hence, consistent.
+The outline of this proof follows the similar pattern as found in \cite{SilanderIJAR10} where the consistency of fNML was proved.
+
+\subsection{On implementation}
 

### 2017-03-20 MLterms.tex
MLterms equal tex added

\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{bm}
\usepackage[round]{natbib}
\usepackage{bm}
\newcommand{\Le}{\left(}
\newcommand{\Ri}{\right)}
\newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{lemma}{Lemma}

\begin{document}

\begin{lemma}
$$
\log \frac{P(D_{i, G_i} \ | \ \hat{\theta}_{i, G_i} )}{P(D_{G_i} \ | \ \hat{\theta}_{G_i} )} = \log P(D_i \ | \ \hat{\theta}_{i | G_i} )
$$
\end{lemma}
\begin{proof}
We can write the terms in the left hand side of the equation as
$$
P(D_{i, G_i} \ | \ \hat{\theta}_{i, G_i}) = \prod_{j,k} \Le \frac{N_{ijk}}{N}  \Ri^{N_{ijk}}
$$and
$$
P(D_{G_i} \ | \ \hat{\theta}_{G_i} ) =  \prod_{j} \Le \frac{N_{ij}}{N}  \Ri^{N_{ij}}.
$$Here, $N_{ijk}$ denotes the number of times we observe $X_i$ taking value $k$ when its parents are in $j$:th configuration in our data matrix $D$. Also, $N_{ij} = \sum_k N_{ijk}$ (and $\sum_{k,j}N_{ijk} = N$ for all $i$).

Now,
\begin{align*}
\frac{P(D_{i, G_i} \ | \ \hat{\theta}_{i, G_i} )}{P(D_{G_i} \ | \ \hat{\theta}_{G_i} )} &= \frac{ \prod_{j,k} \Le \frac{N_{ijk}}{N}  \Ri^{N_{ijk}}}{\prod_{j} \Le \frac{N_{ij}}{N}  \Ri^{N_{ij}}} \\
&= \frac{ \prod_{j,k} \Le \frac{N_{ijk}}{N}  \Ri^{N_{ijk}}}{\prod_{j} \Le \frac{N_{ij}}{N}  \Ri^{\sum_k N_{ijk}}} \\
&= \frac{ \prod_{j,k} \Le \frac{N_{ijk}}{N}  \Ri^{N_{ijk}}}{\prod_{j}\prod_{k} \Le \frac{N_{ij}}{N}  \Ri^{N_{ijk}}} \\ &= \frac{ \prod_{j,k} \Le \frac{N_{ijk}}{N}  \Ri^{N_{ijk}}}{\prod_{j,k} \Le \frac{N_{ij}}{N}  \Ri^{N_{ijk}}} \\
&= \prod_{j,k} \Le \frac{N_{ijk}}{N_{ij}} \Ri^{N_{ijk}} \\
&= P(D_i \ | \ \hat{\theta}_{i | G_i} ).
\end{align*} 
\end{proof}
\end{document}

### 2017-03-23 qnml.tex
edited qNML-section

@@
 \begin{eqnarray}
+\label{eqn:qnmldef}
 s^{qNML}(D;G) & := & \sum_{i=1}^n s^{qNML}_i(D;G)\\
@@
 equivalent networks are exactly those who a) are same when directed
-arcs are substitued by undirected ones and b) who have same
+arcs are substituted by undirected ones and b) who have same
 V-structures, i.e.  the variable triplets $(A,B,C)$ where both $A$ and
@@
 theorem and sketch a proof for it. A more detailed proof appears in the
-suplementary material.
+supplementary material.
 \begin{theorem}
@@
 
-One important property possessed by nearly every model selection criterion is consistency.
-In our context, consistency means that given a data matrix with $N$ samples coming from a distribution faithful to some DAG $G$, the qNML will give the highest score to true graph $G$ with a probability tending to one as $N$ increases.
-We will show this by first proving that qNML is asymptotically equivalent to widely used BIC criterion which is known to be consistent \cite{Schw78, Haug88}.
+One important property possessed by nearly every model selection
+criterion is consistency. In our context, consistency means that given
+a data matrix with $N$ samples coming from a distribution faithful to
+some DAG $G$, the qNML will give the highest score to true graph $G$
+with a probability tending to one as $N$ increases. We will show this
+by first proving that qNML is asymptotically equivalent to widely used
+BIC criterion which is known to be consistent \cite{Schw78, Haug88}.
 
@@
 \textnormal{BIC}(D,G) = \sum_{i = 1}^n \log P(D_i \ | \ \hat{\theta}_{i | G_i} ) - \frac{q_i(r_i - 1)}{2} \log N,
-\end{equation}where $D_i = D[\cdot, i]$ stands for the data on variable $i$, $\hat{\theta}_{i | G_i}$ denotes the maximum likelihood parameters of the conditional distribution of variable $i$ given its parents in $G$.
-Since the both BIC and qNML scores are decomposable, we can focus on studying the local scores.
-Using the definition of qNML, we have
+\end{equation}
+where $D_i = D[\cdot, i]$ stands for the data on variable $i$,
+$\hat{\theta}_{i | G_i}$ denotes the maximum likelihood parameters of
+the conditional distribution of variable $i$ given its parents in
+$G$. Since the both BIC and qNML scores are decomposable, we can focus
+on studying the local scores. Using the definition of qNML, we have
 \begin{align}\label{qnmlDef2}
-s^{qNML}_i(D;G) &= \log \frac{P(D_{i, G_i} \ | \ \hat{\theta}_{i, G_i} )}{P(D_{G_i} \ | \ \hat{\theta}_{G_i} )} \notag \\
-& -(reg(N,q_i r_i) - reg(N,q_i)).
+s^{qNML}_i(D;G) &= \log \frac{P(D_{i, G_i} \ | \ \hat{\theta}_{i, G_i}
+  )}{P(D_{G_i} \ | \ \hat{\theta}_{G_i} )} \notag \\ & -(reg(N,q_i
+r_i) - reg(N,q_i)).
 \end{align}
@@
 \log \frac{P(D_{i, G_i} \ | \ \hat{\theta}_{i, G_i} )}{P(D_{G_i} \ | \ \hat{\theta}_{G_i} )} = \log P(D_i \ | \ \hat{\theta}_{i | G_i} )
-$$ which simply shows that the goodness-of-the-fit terms agree between the criteria.
-Next, we consider the difference of regrets in (\ref{qnmlDef2}) which corresponds to the penalty term of BIC.
-The following theorem states that these two are asymptotically equal.
+$$ which simply shows that the goodness-of-the-fit terms agree between
+the criteria. Next, we consider the difference of regrets in
+(\ref{qnmlDef2}) which corresponds to the penalty term of BIC. The
+following theorem states that these two are asymptotically equal.
 
 \begin{theorem}
-Let $r_i$ and $q_i$ denote the number of possible values for variable $X_i$ and the possible configurations of parents $X_{G_i}$, respectively. As $N \to \infty$,
+Let $r_i$ and $q_i$ denote the number of possible values for variable
+$X_i$ and the possible configurations of parents $X_{G_i}$,
+respectively. As $N \to \infty$,
 $$reg(N,q_i r_i) - reg(N,q_i) = \frac{q_i(r_i - 1)}{2}\log N + O(1).$$
@@
 \begin{proof}
-By considering only the leading term in (\ref{eqn:szp}), the regret for a single multinomial variable with $m$ categories can be written asymptotically as
+By considering only the leading term in (\ref{eqn:szp}), the regret
+for a single multinomial variable with $m$ categories can be written
+asymptotically as
 $$
@@
 \end{proof}
-This suffices in order to show that qNML is asymptotically equivalent to BIC and, hence, consistent.
-The outline of this proof follows the similar pattern as found in \cite{SilanderIJAR10} where the consistency of fNML was proved.
+This suffices in order to show that qNML is asymptotically equivalent
+to BIC and, hence, consistent. The outline of this proof follows the
+similar pattern as found in \cite{SilanderIJAR10} where the
+consistency of fNML was proved.
+
+\subsection{qNML equals NML for exponentially many models}
+The fNML criterion equals NML criterion just for the Bayesian network
+with no arcs.  We will next show that the qNML equals NML when the
+connected components of the network are tournaments (complete directed
+acyclic subgraphs of G).
+
+We first note that the NML-criterion for the Bayesian networks
+decomposes by the connected components. Namely, if $G$ consists of $C$
+connected components $(G^1,\ldots,G^C)$ with variable sets $(V^1,\ldots,V^C)$,
+the maximum likelihood for the data $D$ decomposes, and we can write
+\begin{eqnarray}
+  \lefteqn{P_{NML}(D;G)} \nonumber \\
+  && = \frac{P(D;\hat\theta(D),G)}
+            {\sum_{D_{V_1}}\ldots \sum_{D_{V_C}}\prod_{c=1}^C P(D_{V^c};\hat\theta(D_{V^c}),G)} \nonumber \\
+            && = \frac{\prod_{c=1}^C P(D_{V^c};\hat\theta(D_{V^c}),G)} 
+            {\prod_{c=1}^C\sum_D^cP(D_{V^c};\hat\theta(D_{V^c}),G)}.\nonumber \\
+            && = \prod_{c=1}^CP_{NML}(D_{V^c};G).
+\end{eqnarray}
+
+Clearly, the qNML score also decomposes by the connected components,
+so it remains to show that if the (sub)network $G$ is a tournament,
+then for any data $D$, $s^{qNML}(D;G)=\log P_{NML}(D;G)$.  Due to the
+score equivalence of both the NML criterion and the qNML criterion, we
+may pick such a tournament $G$ that linear ordering defined by $G$
+matches the ordering of the data columns, i.e., $i<j$ implies $G_i
+\subset G_j$. Now from the definition~(\ref{eqn:qnmldef}) of the qNML
+criterion we see that for the tournament $G$, the sum telescopes
+leaving us with $s^{qNML}(D;G) = \log P^1_{NML}(D[\cdot,G];G)$, thus
+it is enough to show that $P^1_{NML}(D;G)=P_{NML}(D;G)$.
+This follows, since for any data vector $x$ in data $D$, the
+$P^1(x;\hat\theta(D),G) = P(x;\hat\theta(D),G)$,
+where we have denoted by $P^1$ the model that takes $n$-dimensional
+vectors to be values of the single (collapsed) categorical variable.
+Denoting prefixes of data vector $x$ by $x^{:i}$, and number of times
+such a prefix appears in a $N\times n$ datamatrix $D$ by $n_D(x^{:i})$,
+($n_D(x^0)=N)$, we have
+\begin{eqnarray}
+  P(D;\hat\theta(D),G) &=& \prod_{j=1}^{N}P(d_j;\theta(D),G) \nonumber \\
+  &=& \prod_{j=1}^{N} \prod_{i=1}^{n} \frac{n_D(d_j^{:i})}{n_D(d_j^{:i-1})} \nonumber\\
+  &=& \prod_{j=1}^{N} \frac{n_D(d_j^{:n})}{N} \nonumber \\
+  &=& P^1(D;\hat\theta^1(D),G).
+\end{eqnarray}
+Since both $P^1_{NML}$ and $P_{NML}$ are defined in terms of the
+maximum likelihood probabilities, the equality above implies the equality
+of these distributions.
+
+Equality established, we would like to still state the number of
+different networks connected components of which are tournaments.  We
+start by generating all the $p(n)$ integer partitions i.e. ways to
+partition $n$ labelled into parts with different size-profiles.  For
+example, for $n=4$, we have $p(4)=5$ partition profiles $[[4], [3, 1],
+  [2, 2], [2, 1, 1], [1, 1, 1, 1]]$.  Each of these partition size
+profiles corresponds to many different networks, apart from the last
+one that corresponds just to the empty network.  We count number of
+networks for one such partition size profile (and then later sum these
+counts up).  For any such partition profile $(p_1,\ldots,p_k)$ we can
+(think we can) count the ways we can pick the nodes to different parts
+and then order each part. This leads to the product $n\choose
+p_1$$p_1$!${n-p_1}\choose p_2$$p_2!$${n-p_1-p2}\choose
+p_3$$p3!$$\ldots$${n-\sum_{j=1}^{k-1}p_j}\choose p_k$$p_k!$. However,
+the order of different parts of the same size does not matter, so for
+all groups of parts having the same size, we have to divide the
+product above by the factorial of the size of such group. Notice also,
+that the product above telescopes, leaving us a formula for OEIS
+sequence A000262 as described by Thomas Wieder:
+
+With $p(n) =$ the number of integer partitions of $n$, 
+     $d(i)$ = the number of different parts of the $i$-th partition of $n$, $m(i,j) =$ multiplicity of the $j$-th part of the $i$-th partition of $n$,  one has:
+
+$$
+a(n) = \sum_{i=1}^{p(n)} \frac{n!}{\prod_{j=1}^{d(i)} m(i,j)!}.
+$$
+
+HERE SOME TEXT ABOUT GROWTH RATE
 

### 2017-03-26 qnml.tex
consistency proof reformulated a bit

@@
 
-The BIC can be written as
-\begin{equation}
+The BIC criterion can be written as
+\begin{equation}\label{BIC}
 \textnormal{BIC}(D,G) = \sum_{i = 1}^n \log P(D_i \ | \ \hat{\theta}_{i | G_i} ) - \frac{q_i(r_i - 1)}{2} \log N,
@@
 the conditional distribution of variable $i$ given its parents in
-$G$. Since the both BIC and qNML scores are decomposable, we can focus
-on studying the local scores. Using the definition of qNML, we have
+$G$.
+
+Since the both BIC and qNML scores are decomposable, we can focus
+on studying the local scores. We will next show that, asymptotically, the local qNML score equals the local BIC score. This is formulated in the following theorem:
+
+\begin{theorem}\label{consistency}
+Let $r_i$ and $q_i$ denote the number of possible values for variable
+$X_i$ and the possible configurations of parents $X_{G_i}$,
+respectively. As $N \to \infty$
+$$
+s^{qNML}_i(D;G) =  \log P(D_i \ | \ \hat{\theta}_{i | G_i} )  - \frac{q_i(r_i - 1)}{2} \log N.
+$$
+\end{theorem}
+
+In order to prove this, we start with the definition of qNML and write
 \begin{align}\label{qnmlDef2}
@@
 
-Now, we observe that the first term in (\ref{qnmlDef2}) can written as
+By comparing the equations (\ref{BIC}) and (\ref{qnmlDef2}), we see that proving our clam boils down to showing two things: 1) the terms involving the maximized likelihoods are equal and 2) the penalty terms are asymptotically equivalent.
+We will formulate these into two lemmas.
+
+\begin{lemma}\label{MLLemma} The maximized likelihood terms in equations (\ref{BIC}) and (\ref{qnmlDef2}) are equal:    
 $$
-\log \frac{P(D_{i, G_i} \ | \ \hat{\theta}_{i, G_i} )}{P(D_{G_i} \ | \ \hat{\theta}_{G_i} )} = \log P(D_i \ | \ \hat{\theta}_{i | G_i} )
-$$ which simply shows that the goodness-of-the-fit terms agree between
-the criteria. Next, we consider the difference of regrets in
+\frac{P(D_{i, G_i} \ | \ \hat{\theta}_{i, G_i} )}{P(D_{G_i} \ | \ \hat{\theta}_{G_i} )} = P(D_i \ | \ \hat{\theta}_{i | G_i} ).
+$$
+\end{lemma}
+
+\begin{proof}
+We can write the terms in the left hand side of the equation as
+$$
+P(D_{i, G_i} \ | \ \hat{\theta}_{i, G_i}) = \prod_{j,k} \Le \frac{N_{ijk}}{N}  \Ri^{N_{ijk}}
+$$and
+$$
+P(D_{G_i} \ | \ \hat{\theta}_{G_i} ) =  \prod_{j} \Le \frac{N_{ij}}{N}  \Ri^{N_{ij}}.
+$$Here, $N_{ijk}$ denotes the number of times we observe $X_i$ taking value $k$ when its parents are in $j$:th configuration in our data matrix $D$.
+Also, $N_{ij} = \sum_k N_{ijk}$ (and $\sum_{k,j}N_{ijk} = N$ for all $i$).
+
+Now,
+\begin{align*}
+\frac{P(D_{i, G_i} \ | \ \hat{\theta}_{i, G_i} )}{P(D_{G_i} \ | \ \hat{\theta}_{G_i} )} &= \frac{ \prod_{j,k} \Le \frac{N_{ijk}}{N}  \Ri^{N_{ijk}}}{\prod_{j} \Le \frac{N_{ij}}{N}  \Ri^{N_{ij}}} \\
+&= \frac{ \prod_{j,k} \Le \frac{N_{ijk}}{N}  \Ri^{N_{ijk}}}{\prod_{j} \Le \frac{N_{ij}}{N}  \Ri^{\sum_k N_{ijk}}} \\
+&= \frac{ \prod_{j,k} \Le \frac{N_{ijk}}{N}  \Ri^{N_{ijk}}}{\prod_{j}\prod_{k} \Le \frac{N_{ij}}{N}  \Ri^{N_{ijk}}}\\
+&= \prod_{j,k} \Le \frac{N_{ijk}}{N_{ij}} \Ri^{N_{ijk}} \\
+&= P(D_i \ | \ \hat{\theta}_{i | G_i} ).
+\end{align*}
+
+\end{proof}
+
+Next, we consider the difference of regrets in
 (\ref{qnmlDef2}) which corresponds to the penalty term of BIC. The
-following theorem states that these two are asymptotically equal.
+following lemma states that these two are asymptotically equal:
 
-\begin{theorem}
-Let $r_i$ and $q_i$ denote the number of possible values for variable
-$X_i$ and the possible configurations of parents $X_{G_i}$,
-respectively. As $N \to \infty$,
+\begin{lemma}\label{penaltyLemma}
+As $N \to \infty$,
 $$reg(N,q_i r_i) - reg(N,q_i) = \frac{q_i(r_i - 1)}{2}\log N + O(1).$$   
-\end{theorem}   
+\end{lemma}
 \begin{proof}
-By considering only the leading term in (\ref{eqn:szp}), the regret
+The regret
 for a single multinomial variable with $m$ categories can be written
 asymptotically as
-$$
+\begin{equation}\label{regretAsymp}
 reg(N,m) = \frac{m-1}{2}\log N + O(1).
-$$ Using this, we have
+\end{equation}
+For the more precise statement with the underlying assumptions (which are fulfilled in the multinomial case) and for the proof, we refer to \cite{Riss96a, Grun07}. Using this, we have
 \begin{align*}
@@
 \end{proof}
-This suffices in order to show that qNML is asymptotically equivalent
-to BIC and, hence, consistent. The outline of this proof follows the
+This concludes our proof since lemmas \ref{MLLemma} and \ref{penaltyLemma} imply the Theorem \ref{consistency}. The outline of this proof follows the
 similar pattern as found in \cite{SilanderIJAR10} where the

### 2017-03-26 bn.tex
something on BDeu irregularity

@@
 
+Recently, Suzuki~\cite{Suzuki2017} discussed the theoretical properties of the BDeu of score and showed that in certain settings BDeu has tendency to add more and more parent variables for a child node even though the empirical conditional entropy of the child given the parents has already reached zero.
+More in detail, assume that in our data $D$, the values of $X_i$ are completely determined by variables in set $Z$, so that the entropy $H(X | Z) = 0$.
+Now, if we can further find one or more variables, denoted by $Y$, whose values are determined completely by the variables in $Z$, then BDeu will prefer the set $Z\cup Y$ over mere $Z$ as the parents of $X_i$.
+Suzuki argues that this kind of behaviour violates regularity in model selection as the more complex model is preferred over a simpler one even though it does not fit to data any better.
+The phenomena seems to stem from the way the hyperparameters for the Dirichlet distribution are chosen in BDeu as the  BD score based on the Jeffreys prior does not suffer from this anomaly.
+
 A natural solution to avoid parameter sensitivity of BDeu would be to

### 2017-03-26 bn.tex
Change regret approximation

@@
 computation of the regret is still prohibitive, it can be approximated
-efficiently using a so called Szpankowski
-approximation~\cite{cosco.aistat03}:
+efficiently using an approximation derived by Szpankowski and
+Weinberger~\cite{Szpankowski2012}:
 \begin{eqnarray}
-\label{eqn:szp}
-\lefteqn{reg(N,r) \approx \frac{\sqrt{2} r \Gamma{\left(\frac{r}{2} \right)}}
-                               {3 \sqrt{N} \Gamma{\left(\frac{r-1}{2}  \right)}}} \\
-&&+ \left(\frac{r-1}{2}\right) \log{\left (\frac{N}{2} \right )}
-- \log \Gamma{\left(\frac{r}{2} \right)} + \frac{1}{2} \log{\left (\pi \right )}\nonumber\\
-&&- \frac{r^{2} \Gamma^{2}{\left(\frac{r }{2} \right)}}
-         {9N \Gamma^{2}{\left(\frac{r-1}{2}\right)}}
-+ \frac{2r^3-3r^2-2r+3}{36N}\nonumber.
+    reg(N, r) & \approx & N\left(\log{\alpha} + (\alpha + 2)C_\alpha -
+                \frac{1}{C_\alpha}\right)\nonumber \\
+    && - \frac{1}{2} \log{\left(C_\alpha + \frac{2}{\alpha}\right)},
 \end{eqnarray}
+where $\alpha = \frac{r}{N}$ and
+\begin{equation}
+    C_\alpha = \frac{1}{2} + \frac{1}{2} \sqrt{1 + \frac{4}{\alpha}}.
+\end{equation}
 fNML solves the parameter sensitivity problem and yields predictive

### 2017-03-27 qnml.tex
Typos Please enter the commit message for your changes. Lines starting

@@
 \end{eqnarray}
-  due to equations
-  $\{A\}\cup G_A  = G_B$,
-  $\{B\}\cup G'_B = G'_A$,
-  $\{B\}\cup G_B  = \{A\} \cup G'_A$,
-  and $G_A        = G'_B$.
+  due to equations $\{A\}\cup G_A = G_B$, $\{B\}\cup G'_B = G'_A$,
+  $\{B\}\cup G_B = \{A\} \cup G'_A$, and $G_A = G'_B$ that follow
+  easily from the definition of the covered arc.
 \end{proof}
@@
 such a prefix appears in a $N\times n$ datamatrix $D$ by $n_D(x^{:i})$,
-($n_D(x^0)=N)$, we have
+($n_D(x^{:0})=N)$, we have
 \begin{eqnarray}
-  P(D;\hat\theta(D),G) &=& \prod_{j=1}^{N}P(d_j;\theta(D),G) \nonumber \\
+  P(D;\hat\theta(D),G) &=& \prod_{j=1}^{N}P(d_j;\hat\theta(D),G) \nonumber \\
   &=& \prod_{j=1}^{N} \prod_{i=1}^{n} \frac{n_D(d_j^{:i})}{n_D(d_j^{:i-1})} \nonumber\\
@@
 
-Collapsing many variables into a single one causes $r$ in the
-equation~(\ref{eqn:szp}) to be exponentially large (in our experiments
-around $2^{20}$).  This causes the $\Gamma^2$ terms to grow very large
-calling for care when computing the ratios of such huge numbers. More
-specifically, scoring the $i^\text{th}$ data column, i.e., computing
-$s^{qNML}_i$, requires the difference $dreg(N,q_i,r_i)=reg(N,q_i) -
-reg(N,r_i q_i)$, where $q$ denotes the number of values in the
+REWRITE THIS
+
+$dreg(N,q_i,r_i)=reg(N,q_i) -reg(N,r_i q_i)$, where $q$ denotes the number of values in the
 collapsed parent variable.  A numerically stable approximation is:

### 2017-03-27 bn.tex
Fix typo in approximation
A bit more on approximations

--- 2017-03-26 bn.tex
@@
 vector with $r$ different categorical values.  While the naive
-computation of the regret is still prohibitive, it can be approximated
-efficiently using a so called Szpankowski
+computation of the regret is still prohibitive, Silander et al.\
+approximate it efficiently using a so-called Szpankowski
 approximation~\cite{cosco.aistat03}:
 \begin{eqnarray}
-\label{eqn:szp}
+\label{eqn:szp1}
 \lefteqn{reg(N,r) \approx \frac{\sqrt{2} r \Gamma{\left(\frac{r}{2} \right)}}
@@
 \end{eqnarray}
+
+However, equation (\ref{eqn:szp1}) is derived only for the case
+where $r$ is constant and $N$ grows. While with
+fNML it is typical that $N$ is large compared to $r$, an
+approximation for all ranges of $N$ and $r$ derived by
+Szpankowski and Weinberger~\cite{Szpankowski2012} can also be used:
+\begin{eqnarray}
+\label{eqn:szp2}
+    reg(N, r) & \approx & N\left(\log{\alpha} + (\alpha + 2) \log{C_\alpha}
+                - \frac{1}{C_\alpha}\right)\nonumber \\
+    && - \frac{1}{2} \log{\left(C_\alpha + \frac{2}{\alpha}\right)},
+\end{eqnarray}
+where $\alpha = \frac{r}{N}$ and
+\begin{equation}
+    C_\alpha = \frac{1}{2} + \frac{1}{2} \sqrt{1 + \frac{4}{\alpha}}.
+\end{equation}
+These approximations are compared in Table \ref{tbl:regrets}
+to the exact regret for various values of $N$ and $r$.
+For a constant $N$, equation (\ref{eqn:szp1}) provides
+a progressively worse approximation as $r$ grows. Equation
+(\ref{eqn:szp2}) on the other hand is a good approximation of
+the regret regardless of the ratio of $N$ and $r$.
+
 fNML solves the parameter sensitivity problem and yields predictive
@@
 still sometimes appear surprisingly complicated.
+
+\begin{table}
+\centering
+\begin{tabular}{crrrr}
+\toprule
+N & r & eq. (\ref{eqn:szp1}) & eq. (\ref{eqn:szp2}) & exact \\
+\midrule
+\multirow{4}{*}{50} & 10 & 13.24 & 13.26 & 13.24 \\
+& 100 & 62.00 & 60.01 & 60.00 \\
+& 1000 & 491.63 & 153.28 & 153.28 \\
+& 10000 & 25635.15 & 265.28 & 265.28 \\
+\midrule
+\multirow{4}{*}{500} & 10 & 22.67 & 22.69 & 22.67 \\
+& 100 & 144.10 & 144.03 & 144.03 \\
+& 1000 & 624.35 & 603.93 & 603.93 \\
+& 10000 & 4927.24 & 1533.38 & 1533.38 \\
+\midrule
+\multirow{4}{*}{5000} & 10 & 32.74 & 32.76 & 32.74 \\
+& 100 & 247.97 & 247.97 & 247.97 \\
+& 1000 & 1452.51 & 1451.78 & 1451.78 \\
+& 10000 & 6247.83 & 6043.16 & 6043.16 \\
+\bottomrule
+\end{tabular}
+\caption{Regret values for various values of $N$ and $r$.}
+\label{tbl:regrets}
+\end{table}

### 2017-03-28 qnml.tex
Minor editing

@@
 
-\subsection{qNML equals NML for exponentially many models}
+\subsection{qNML equals NML for many models}
 The fNML criterion equals NML criterion just for the Bayesian network
@@
 
+\begin{theorem}
+If $G$ consists of $C$ connected components $(G^1,\ldots,G^C)$ with
+variable sets $(V^1,\ldots,V^C)$, then $\log P_{NML}(D;G) = s^{qNML}(D;G)$
+for all data sets $D$.
+\end{theorem}
+\begin{proof}
 We first note that the NML-criterion for the Bayesian networks
-decomposes by the connected components. Namely, if $G$ consists of $C$
-connected components $(G^1,\ldots,G^C)$ with variable sets $(V^1,\ldots,V^C)$,
-the maximum likelihood for the data $D$ decomposes, and we can write
+decomposes by the connected components.
+
+The maximum likelihood for the data $D$ decomposes, and we can write
 \begin{eqnarray}
@@
 of these distributions.
+\end{proof}
 
@@
 $$
-
-HERE SOME TEXT ABOUT GROWTH RATE
-
-\subsection{On implementation}
-
-REWRITE THIS
-
-$dreg(N,q_i,r_i)=reg(N,q_i) -reg(N,r_i q_i)$, where $q$ denotes the number of values in the
-collapsed parent variable.  A numerically stable approximation is:
-\begin{eqnarray}
-\lefteqn{dreg(N,q,r)\approx\frac{1-r^2}{18 N} q^{2}} \\
-&&+ \frac{\left(q - 1\right) \sqrt{4 q - 2} + \left(1 - q r\right) \sqrt{4 q r - 2}}{6 \sqrt{N}} \nonumber\\
-&&+\frac{q\left(3 N \left(r \log{r}  - \left(r - 1\right) \left(\log{N} + 1\right)\right) + r\right)}{6 N} \nonumber\\
-&&+ \frac{1}{12}(\log{(r+\frac{1-r}{q+1})}- 7\log{r}).\nonumber
-\end{eqnarray}
+For example, $a(4)=\frac{4!}{1!}+\frac{4!}{1!1!}+\frac{4!}{2!}+\frac{4!}{1!2!}+\frac{4!}{4!}=73$.

### 2017-03-30 bn.tex
some new stuff, mainly results

@@
 
-\subsection{Likelihood}
+\subsection{LIKELIHOOD}
 \label{ssec:likelihood}
@@
 
-\subsection{Bayesian structure learning}
+\subsection{BAYESIAN STRUCTURE LEARNING}
 
@@
 
-\subsection {Problems, solutions and problems}
+\subsection {PROBLEMS, SOLUTIONS AND PROBLEMS}
 
@@
 
-Recently, Suzuki~\cite{Suzuki2017} discussed the theoretical properties of the BDeu of score and showed that in certain settings BDeu has tendency to add more and more parent variables for a child node even though the empirical conditional entropy of the child given the parents has already reached zero.
-More in detail, assume that in our data $D$, the values of $X_i$ are completely determined by variables in set $Z$, so that the entropy $H(X | Z) = 0$.
-Now, if we can further find one or more variables, denoted by $Y$, whose values are determined completely by the variables in $Z$, then BDeu will prefer the set $Z\cup Y$ over mere $Z$ as the parents of $X_i$.
-Suzuki argues that this kind of behaviour violates regularity in model selection as the more complex model is preferred over a simpler one even though it does not fit to data any better.
-The phenomena seems to stem from the way the hyperparameters for the Dirichlet distribution are chosen in BDeu as the  BD score based on the Jeffreys prior does not suffer from this anomaly.
+Recently, Suzuki~\cite{Suzuki2017} discussed the theoretical
+properties of the BDeu of score and showed that in certain settings
+BDeu has tendency to add more and more parent variables for a child
+node even though the empirical conditional entropy of the child given
+the parents has already reached zero. More in detail, assume that in
+our data $D$, the values of $X_i$ are completely determined by
+variables in set $Z$, so that the entropy $H(X | Z) = 0$. Now, if we
+can further find one or more variables, denoted by $Y$, whose values
+are determined completely by the variables in $Z$, then BDeu will
+prefer the set $Z\cup Y$ over mere $Z$ as the parents of $X_i$. Suzuki
+argues that this kind of behaviour violates regularity in model
+selection as the more complex model is preferred over a simpler one
+even though it does not fit to data any better. The phenomena seems to
+stem from the way the hyperparameters for the Dirichlet distribution
+are chosen in BDeu as the BD score based on the Jeffreys prior does
+not suffer from this anomaly.
 
@@
 the normalizing constant renders the computation
-infeasible. Consequently, Silander et al.~\cite{cosco.pgm08a}
-suggested solving the parameter sensitivity problem by adopting the
+infeasible.
+
+Consequently, Silander et al.~\cite{cosco.pgm08a}
+suggested solving the BDeu parameter sensitivity problem by adopting the
 NML-code to the column partitions, i.e., changing the Bayesian mixture

### 2017-03-30 experiments.tex
some new stuff, mainly results

@@
 \section{EXPERIMENTAL RESULTS}
+
+We will next empirically compare the capacity of the qNML to that of
+BIC, BDeu and fNML in identifying the data generating structures,
+and producing models that are predictive and parsimonious.
+It seems that none of the criteria outperform others for all
+these requirements.
+
+\subsection{FINDING GENERATING STRUCTURE}
+
+We first compared the ability of different scoring criteria to
+discover the data generating structure. For this purpose we generated
+100 different 5-node Bayesian network structures with 4 edges and
+another 100 structures with 7 edges.  The variables were randomly
+assigned to have 2 – 4 values ($r_i \in \{2, 3, 4\}$). For each
+network, we generated parameters by two different schemes. The first
+scheme exactly matched the assumptions of the BDeu score with $\alpha
+= 1$, i.e., the parameters were distributed by $\theta_{ij} \sim
+Dir(\frac{1}{r_iq_i},\ldots,\frac{r_i}{q_i})$. The other scheme was
+to generate the parameters independently from a Dirichlet distribution
+$\theta_{ij} \sim Dir(\frac{1}{2},\ldots, \frac{1}{2})$ that is the
+Jeffreys' prior for multinomial model. This distribution was selected
+instead of the uniform distribution in order to make the generating
+structure more identifiable.  For each network (structure +
+parameters), we generated 100 data sets of 1000 data vectors, and
+studied how different scoring criteria ranked the structure of the
+generating network among all the 5-node networks as a function of
+(sub)sample size.
+
+Not surprisingly, the results indicate that when the parameter generation
+mechanism matches the assumptions of the BDeu-score, the BDeu usually
+also ranks the generating structure higher than the other scores
+(Figure X).  On the other hand, when the paramaters are drawn from the
+Jeffreys's prior, the fNML score appears to rank the generting structure
+highest. This too is not surprising, since the NML distribution for the
+multinomial data is known to closely match the marginal distribution of
+the data when the Jefrreys' prior is used (CITE).
+
+The underfitting tendency of BIC can also be clearly detected both for
+relatively sparse networks (4 arcs) and the dense ones (7 arcs). The
+ranking ability the qNML criterion appears to be between fNML and BDeu
+in 3 out of for settings which is good considering that one of these
+criteria is always the winner. For the dense networks with BDeu prior,
+the qNML appears to perform worse than BDeu and fNML but still much
+better than BIC.
+
+\subsection{PREDICTION AND PARSIMONY}
 
@@
 conducted using a dynamic programming based exact structure
-learning\cite{cosco.uai06} which limited the number $n$ of variables
+learning~\cite{cosco.uai06} which limited the number $n$ of variables
 to less than 20.

### 2017-03-30 intro.tex
some new stuff, mainly results

@@
 comparisons have not been favorable for these criteria in terms of
-structural stability and/or predictive
-performance~\cite{cosco.pgm08a}.  The most popular criterion has been
-the marginal likelihood (usually called BDeu for reasons explained
-later), but recent studies~\cite{cosco.uai07,Steck08} have found this
-criterion to be very sensitive to hyper parameters and to yield
-undesirably complex models for small sample sizes.
+identifying the appropriate structure and/or predictive
+performance~\cite{cosco.pgm08a}. Traditionally, the most popular
+criterion has been the marginal likelihood (usually called BDeu for
+reasons explained later), but recent studies~\cite{cosco.uai07,Steck08}
+have found this criterion to be very sensitive to hyper parameters
+and to yield undesirably complex models for small sample sizes.
 
@@
 criterion~\cite{Shta87,Riss96a} would otherwise be an ideal candidate
-for a good criterion.  but its exact calculation is likely to be
+for a good criterion,  but its exact calculation is likely to be
 prohibitively demanding. In 2008, Silander et al. introduced a
@@
 summarize the results for 20 datasets to back up our claim of qNML
-yielding parsimonious models with good predictive capabilities.
+yielding parsimonious models with good predictive capabilities. The
+experiments with artificial data demonstrate the capability to give
+high scores for the data generating model with relatively small sample
+sizes.

### 2017-03-30 gendata.sh

@@ -1,0 +1,13 @@
+for bnmfn in artnets/*.bnm; do
+    bbnm=`basename $bnmfn .bnm`
+    g=`echo $bbnm | cut -d'_' -f2`
+    idtdn=artdats/$bbnm
+    mkdir -p $idtdn
+    vdfn=artdats/data_$g.vd
+    echo $bbnm
+    for j in `seq 0 99`; do
+	idtfn=$idtdn/$j.idt
+	python -m bn.model.bn_gendata -o $idtfn $vdfn $bnmfn 1000
+    done
+done
+

### 2017-03-31 abstract.tex
proofreading + images etc.

@@
 Learning the dependency structure of a multivariate distribution from
-the observational data is an important task since it allows us to
+observational data is an important task since it allows us to
 speculate about the underlying causal mechanisms that induce
@@
 selection problem, but recent studies have revealed that the popular
-Bayesian model selection criterion is not satisfactory. We review some
-of the information theoretic alternatives and introduce a new one
-called a quotient normalized maximum likelihood criterion.
+Bayesian model selection criteria are not satisfactory. We review some
+of the information theoretic alternatives and introduce a new criterion
+called the quotient normalized maximum likelihood criterion.
 \end{abstract}

### 2017-03-31 bn.tex
proofreading + images etc.

@@
 Despite denoting the values with small integers, the model will treat
-the components of $X$ categorical.
+the components of $X$ as categorical variables.
 
@@
 
-Bayesian network $B=(G,\theta)$ defines a probability distribution for
+A Bayesian network $B=(G,\theta)$ defines a probability distribution for
 $X$. The component $G$ defines the structure of the model as a
@@
 Given a realization $x$ of $X$, we denote the sub\nobreakdash-vector
-of $x$ that consists of the values of the parents of $X_{i}$ in $x$ as
+of $x$ that consists of the values of the parents of $X_{i}$ in $x$ by
 $G_{i}(x)$. It is customary to enumerate all the possible
@@
 \begin{equation}
-P(x|G,\theta)=\prod_{i=1}^{n}P(x_{i}\mid G_{i}(x);\theta_{i})=\prod_{i=1}^{n}\theta_{iG_{i}(x)x_{i}}.
+P(x|G,\theta)=\prod_{i=1}^{n}P(x_{i}\mid
+G_{i}(x);\theta_{i})=\prod_{i=1}^{n}\theta_{iG_{i}(x)x_{i}}.
 \end{equation}
@@
 \label{eqn:bayesmix}
-P(D|G,\alpha) & = & \prod_{i=1}^{n}\prod_{j=1}^{q_i}\int
+P(D|G,\alpha) & = & \prod_{i=1}^{n}\prod_{j=1}^{q_i}
 P(D[G_i=j,i]|\alpha)\\ & = & \prod_{i=1}^{n}\prod_{j=1}^{q_i}\int
@@
 
-Finding the satisfactory Dirichlet hyperparameters for the Bayesian
+Finding satisfactory Dirichlet hyperparameters for the Bayesian
 mixture above has, however, turned out to be problematic. Early on,
 one of the desiderata for a good model selection criterion was that it
-would yield equal scores for essentially equivalent
-models~\cite{Verm90}.  For example, the score for the structure
-$X_1\rightarrow X_2$ should be the same as the score for the model
-$X_2 \rightarrow X_1$ since they both correspond to the hypothesis
-that variables $X_1$ and $X_2$ are statistically dependent on each
-other.  It can be shown~\cite{Heck95} that to achieve this, not all
-the hyperparameters $\alpha$ are possible and for practical reasons
-Buntine~\cite{Bunt91} suggested a so called BDeu score with just one
-hyperparameter $\alpha\in R_{++}$ so that $\theta_{ij\cdot}\sim
-Dir(\frac{\alpha}{q_i r_i},\ldots,\frac{\alpha}{q_i r_i})$.  However,
-it soon turned out that BDeu score was very sensitive to the selection
-of this hyperparameter~\cite{cosco.uai07} and that for small sample
-sizes this method detects spurious correlations~\cite{Steck08} leading
-to models with suspiciously many parameters.
+is score-equivalent, i.e., it would yield equal scores for
+essentially equivalent models~\cite{Verm90}.  For example, the score
+for the structure $X_1\rightarrow X_2$ should be the same as the score
+for the model $X_2 \rightarrow X_1$ since they both correspond to the
+hypothesis that variables $X_1$ and $X_2$ are statistically dependent
+on each other.  It can be shown~\cite{Heck95} that to achieve this,
+not all the hyperparameters $\alpha$ are possible and for practical
+reasons Buntine~\cite{Bunt91} suggested a so called BDeu score with
+just one hyperparameter $\alpha\in R_{++}$ so that
+$\theta_{ij\cdot}\sim Dir(\frac{\alpha}{q_i
+  r_i},\ldots,\frac{\alpha}{q_i r_i})$.  However, it soon turned out
+that the BDeu score was very sensitive to the selection of this
+hyperparameter~\cite{cosco.uai07} and that for small sample sizes this
+method detects spurious correlations~\cite{Steck08} leading to models
+with suspiciously many parameters.
 
 Recently, Suzuki~\cite{Suzuki2017} discussed the theoretical
-properties of the BDeu of score and showed that in certain settings
-BDeu has tendency to add more and more parent variables for a child
-node even though the empirical conditional entropy of the child given
-the parents has already reached zero. More in detail, assume that in
-our data $D$, the values of $X_i$ are completely determined by
-variables in set $Z$, so that the entropy $H(X | Z) = 0$. Now, if we
-can further find one or more variables, denoted by $Y$, whose values
-are determined completely by the variables in $Z$, then BDeu will
-prefer the set $Z\cup Y$ over mere $Z$ as the parents of $X_i$. Suzuki
+properties of the BDeu score and showed that in certain settings BDeu
+has tendency to add more and more parent variables for a child node
+even though the empirical conditional entropy of the child given the
+parents has already reached zero. In more detail, assume that in our
+data $D$, the values of $X_i$ are completely determined by variables
+in set $Z$, so that the entropy $H(X | Z) = 0$. Now, if we can further
+find one or more variables, denoted by $Y$, whose values are
+determined completely by the variables in $Z$, then BDeu will prefer
+the set $Z\cup Y$ over $Z$ alone as the parents of $X_i$. Suzuki
 argues that this kind of behaviour violates regularity in model
 selection as the more complex model is preferred over a simpler one
-even though it does not fit to data any better. The phenomena seems to
-stem from the way the hyperparameters for the Dirichlet distribution
-are chosen in BDeu as the BD score based on the Jeffreys prior does
-not suffer from this anomaly.
+even though it does not fit the data any better. The phenomenon seems
+to stem from the way the hyperparameters for the Dirichlet
+distribution are chosen in BDeu as using Jeffreys' prior,
+$\theta_{ijk}\sim Dir(\frac{1}{2},ldots,\frac{1}{2})$, does not suffer
+from this anomaly. However, using Jeffreys' prior causes marginal
+likelihood score not to be score-equivalent.
 
@@
 Consequently, Silander et al.~\cite{cosco.pgm08a}
-suggested solving the BDeu parameter sensitivity problem by adopting the
-NML-code to the column partitions, i.e., changing the Bayesian mixture
+suggested solving the BDeu parameter sensitivity problem by using the
+NML-code for the column partitions, i.e., changing the Bayesian mixture
 in equation~(\ref{eqn:bayesmix}) to
@@
 $P^1_{NML}$ depends only on the length $N$ of the categorical data
-vector with $r$ different categorical values.  While the naive
+vector with $r$ different categorical values,
+\begin{equation}
+reg(N,r)=\log \sum_{D\in \{1,\ldots,r\}^N} P(D|\hat\theta(D)).
+\end{equation}
+While the naive
 computation of the regret is still prohibitive, Silander et al.\
@@
 \end{equation}
-These approximations are compared in Table \ref{tbl:regrets}
-to the exact regret for various values of $N$ and $r$.
-For a constant $N$, equation (\ref{eqn:szp1}) provides
-a progressively worse approximation as $r$ grows. Equation
-(\ref{eqn:szp2}) on the other hand is a good approximation of
-the regret regardless of the ratio of $N$ and $r$.
+These approximations are compared in Table \ref{tbl:regrets} to the
+exact regret for various values of $N$ and $r$.  For a constant $N$,
+equation (\ref{eqn:szp1}) provides a progressively worse approximation
+as $r$ grows. Equation (\ref{eqn:szp2}) on the other hand is a good
+approximation of the regret regardless of the ratio of $N$ and $r$.
+In our experiments, we will use this approximation for 
+implementation of the qNML criterion.
 
@@
 models superior to BDeu.  However, the criterion does not satisfy the
-property of giving the same score for the models that correspond to
+property of giving the same score for models that correspond to
 the same dependence statements. Furthermore, the learned structures
-still sometimes appear surprisingly complicated.
+are often rather complex which hampers their interpretation.
 

### 2017-03-31 experiments.tex
proofreading + images etc.

--- 2017-03-16 experiments.tex
@@
 \section{EXPERIMENTAL RESULTS}
-
-To empirically compare the model selection criteria we took 20 UCI
-data sets~\cite{Lichman:2013} and run 1000 train and test experiments
-for 20 different sample sizes for all of them. The training was
-conducted using a dynamic programming based exact structure
-learning\cite{cosco.uai06} which limited the number $n$ of variables
-to less than 20.
-
-Figure~\ref{fig:bcmean} shows an example of the behavior of the
-studied selection criteria.  For small sample sizes fNML and qNML
-yield equally accurate predictions while BDeu requires more samples to
-obtain competitive performance. The curves are mean curves from 1000
-independent train and test splits.
 
@@
 \centering
-\includegraphics[width=8cm,height=5cm]{breast_cancer_mean.pdf}
-\caption{Predictive log-loss in a breast cancer data as a function of
-  sample size for different model selection criteria.}
-\label{fig:bcmean}
+\includegraphics[width=8cm,height=5cm]{qNML_images/art4_mean.pdf}
+\caption{Finding generating models of 4 arcs.}
+\label{fig:4arcs}
 \end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=8cm,height=5cm]{qNML_images/art7_mean.pdf}
+\caption{Finding generating models of 7 arcs.}
+\label{fig:4arcs}
+\end{figure}
+
+We empirically compare the capacity of qNML to that of BIC, BDeu and
+fNML in identifying the data generating structures, and producing
+models that are predictive and parsimonious.  It seems that none of
+the criteria uniformly outperform the others in all these desirable
+aspects of model selection criteria.
+
+\subsection{FINDING GENERATING STRUCTURE}
+
+We first compare the ability of the different scoring criteria to
+discover the structure that generated the data. For this purpose we
+create 100 random 5-node Bayesian network structures with 4
+edges and another 100 structures with 7 edges.  The variables were
+randomly assigned to have 2 -- 4 values ($r_i \in \{2, 3, 4\}$). For
+each network, we generated parameters by two different schemes. The
+first scheme exactly matched the assumptions of the BDeu score with
+$\alpha = 1$, i.e., the parameters were distributed by $\theta_{ij}
+\sim Dir(\frac{1}{r_iq_i},\ldots,\frac{1}{r_iq_i})$. The other scheme
+was to generate the parameters independently from a Dirichlet
+distribution $\theta_{ij} \sim Dir(\frac{1}{2},\ldots, \frac{1}{2})$
+that is the Jeffreys' prior for the multinomial model. This distribution
+was selected instead of the uniform distribution in order to make the
+generating structure more identifiable.  For each network (structure +
+parameters), we generated 100 data sets of 1000 data vectors, and
+studied how different scoring criteria ranked the structure of the
+generating network among all the 5-node networks as a function of
+(sub)sample size.
+
+Not surprisingly, the results indicate that when the parameter generation
+mechanism matches the assumptions of the BDeu-score, BDeu usually
+also ranks the generating structure higher than the other scores
+(Figure X).  On the other hand, when the parameters are drawn from the
+Jeffreys' prior, the fNML score appears to rank the generating structure
+highest. This too is not surprising, since the NML distribution for
+multinomial data is known to closely match the marginal distribution of
+the data when the Jefrreys' prior is used.
+
+The underfitting tendency of BIC can also be clearly detected both for
+relatively sparse networks (4 arcs) and for dense ones (7 arcs). The
+ranking ability the qNML criterion appears to be between fNML and BDeu
+in 3 out of 4 settings which is good considering that one of these
+criteria is always the winner. For the dense networks with BDeu prior,
+qNML appears to perform worse than BDeu and fNML but still much
+better than BIC.
+
+\subsection{PREDICTION AND PARSIMONY}
+
+To empirically compare the model selection criteria we took 20 UCI
+data sets~\cite{Lichman:2013} and ran 1000 train and test experiments
+for all of them. In each experiment, random 10\% of the original
+dataset was selected as training data and the rest 90\% as test data.
+The training was conducted using a dynamic programming based exact structure
+learning~\cite{cosco.uai06} which limited the number $n$ of variables
+to less than 20.
+
+When predicting with structures learned by the BDeu score, we used the
+Bayesian expected parameter values $\theta_{ijk} \propto
+N_{ijk}+\frac{1}{r_iq_i}$.  In the spirit of keeping the scores
+hyperparameter free, for structures learned by the other model
+selection criteria, we used the predictive conditional NML
+parametrization $\theta_{ijk}\propto e(n_{ijk})(n_{ijk}+1)$, where
+$e(n)=(\frac{n+1}{n})^n$ as suggested in~\cite{Riss07b}.
+
+\begin{figure}
+\centering
+\includegraphics[width=8cm,height=5cm]{qNML_images/breast_cancer_npmean.pdf}
+\caption{Number of parameters in a breast cancer model as a function
+  of sample size for different model selection criteria.}
+\label{fig:bcnpmean}
+\end{figure}
+
+\begin{table}
+\caption{Predictive log losses for small sample sizes for different model selection criteria in 20 different datasets.}
+\label{tbl:preds}
+\begin{center}
+ \begin{tabular}{crrrrr}
+    Data &     N &              BDeu &               BIC &               fNML &               qNML \\
+\midrule
+    Iris &    15 &     \textit{3.61} &              3.56 &   \underline{3.50} &      \textbf{3.47} \\
+ PostOpe &    18 &    \textit{10.52} &     \textbf{7.45} &   \underline{8.35} &               8.39 \\
+   Ecoli &    34 &     \textit{6.27} &              6.16 &      \textbf{5.56} &   \underline{5.63} \\
+   Liver &    35 &     \textit{4.22} &     \textbf{4.01} &               4.10 &   \underline{4.03} \\
+    Wine &    36 &    \textit{16.41} &    \textbf{11.27} &              12.07 &  \underline{11.86} \\
+   Glass &    44 &     \textit{7.48} &              6.65 &      \textbf{6.37} &   \underline{6.43} \\
+ Thyroid &    44 &              2.92 &     \textit{2.93} &      \textbf{2.81} &   \underline{2.83} \\
+ HeartSt &    54 &    \textit{14.37} &    \textbf{10.96} &              11.98 &  \underline{11.65} \\
+ BreastC &    58 &    \textit{11.21} &     \textbf{9.94} &  \underline{10.34} &              10.37 \\
+ HeartHu &    60 &     \textit{9.21} &     \textbf{8.38} &               8.80 &   \underline{8.61} \\
+ HeartCl &    62 &    \textit{14.88} &    \textbf{11.76} &              12.48 &  \underline{12.18} \\
+ BcWisco &    70 &     \textit{6.41} &     \textbf{5.09} &               5.29 &   \underline{5.28} \\
+ Diabete &    77 &     \textit{5.11} &  \underline{4.99} &               5.09 &      \textbf{4.98} \\
+ TicTacT &    96 &             10.99 &    \textbf{10.35} &  \underline{10.72} &     \textit{11.04} \\
+ Balance &   126 &              7.45 &  \underline{7.42} &      \textbf{7.27} &      \textit{7.73} \\
+   Yeast &   149 &              5.67 &     \textit{5.80} &      \textbf{5.48} &   \underline{5.49} \\
+ Abalone &   418 &  \underline{3.93} &     \textit{4.00} &      \textbf{3.88} &               3.96 \\
+ PageBlo &   548 &  \underline{2.36} &     \textit{2.40} &      \textbf{2.32} &               2.38 \\
+ Shuttle &  5800 &  \underline{1.69} &     \textit{1.72} &      \textbf{1.69} &               1.71 \\
+   Adult &  6513 &             10.16 &    \textit{10.25} &     \textbf{10.05} &  \underline{10.08} \\
+\end{tabular}
+\end{center}
+\end{table}
+
+Table~\ref{tbl:preds} features predictive losses for the small sample
+sizes.  (The results for large sample sizes converge for all sensible
+criteria.) BIC's bias for simplicity makes it often win (written bold
+in the table) with small sample sizes, but it performs worst (in
+italics) for the large sample sizes. the qNML seems to achieve a lot
+of runner-up positions (underlined) making it a rather safe choice.
 
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=8cm,height=5cm]{breast_cancer_np_mean.pdf}
-\caption{Number of parameters in a breast cancer model as a function
-  of sample size for different model selection criteria.}
-\label{fig:bcnpmean}
-\end{figure}
+Looking at the number of parents for our 20 datasets and sample
+sizes (Table~\ref{tbl:nofparams}) again features BIC's preference for
+simple models. qNML usually (18/20) beats fNML that performs worst in
+14 out of 20 data sets.
 
 \begin{table}
-\centering
-\begin{tabular}{crrrr}
-\toprule
-           Data &     N &   BDeu &            fNML &            qNML \\
+  \caption{Average number of parents per node 
+    for different model selection criteria in 20 different datasets.}
+\label{tbl:nofparams}
+\begin{center}
+\begin{tabular}{crrrrr}
+    Data &     N &              BDeu &               BIC &           fNML &              qNML \\
 \midrule
-           Iris &    16 &   3.59 &            3.47 &   \textbf{3.43} \\
-  PostOperative &    20 &  10.24 &   \textbf{8.27} &            8.32 \\
-           Wine &    27 &  18.11 &           12.68 &  \textbf{12.51} \\
- HeartHungarian &    45 &  10.15 &            9.10 &   \textbf{8.90} \\
-   BreastCancer &    45 &  12.01 &  \textbf{10.57} &           10.66 \\
- HeartCleveland &    48 &  16.36 &           12.81 &  \textbf{12.54} \\
-          Ecoli &    51 &   5.77 &   \textbf{5.30} &            5.37 \\
-          Liver &    54 &   4.07 &            3.97 &   \textbf{3.91} \\
-          Glass &    55 &   7.16 &   \textbf{6.22} &            6.28 \\
-        Thyroid &    55 &   2.83 &   \textbf{2.75} &            2.77 \\
-   HeartStatlog &    56 &  14.15 &           11.92 &  \textbf{11.60} \\
-      TicTacToe &    96 &  10.99 &  \textbf{10.73} &           11.05 \\
-        Balance &    96 &   7.48 &   \textbf{7.39} &            7.78 \\
-    BcWisconsin &   105 &   5.81 &            5.12 &   \textbf{5.08} \\
-       Diabetes &   117 &   4.95 &            4.95 &   \textbf{4.89} \\
-          Yeast &   225 &   5.54 &   \textbf{5.40} &            5.42 \\
-        Abalone &   418 &   3.93 &   \textbf{3.88} &            3.96 \\
-     PageBlocks &   548 &   2.36 &   \textbf{2.32} &            2.38 \\
-        Shuttle &  2900 &   1.70 &   \textbf{1.69} &            1.72 \\
-          Adult &  4885 &  10.22 &  \textbf{10.08} &           10.12 \\
-\bottomrule
+    Iris &    15 &              0.84 &     \textbf{0.63} &  \textit{0.87} &  \underline{0.82} \\
+ PostOpe &    18 &              1.35 &     \textbf{0.11} &  \textit{1.66} &  \underline{1.29} \\
+   Ecoli &    34 &              0.83 &     \textbf{0.30} &  \textit{0.96} &  \underline{0.77} \\
+   Liver &    35 &              0.60 &     \textbf{0.10} &  \textit{0.88} &  \underline{0.43} \\
+    Wine &    36 &     \textit{2.11} &     \textbf{0.75} &           1.86 &  \underline{1.44} \\
+   Glass &    44 &              1.32 &     \textbf{0.38} &  \textit{1.41} &  \underline{0.85} \\
+ Thyroid &    44 &              0.84 &     \textbf{0.48} &  \textit{0.97} &  \underline{0.66} \\
+ HeartSt &    54 &              1.81 &     \textbf{0.56} &  \textit{2.06} &  \underline{1.50} \\
+ BreastC &    58 &  \underline{1.27} &     \textbf{0.42} &  \textit{1.71} &              1.39 \\
+ HeartHu &    60 &  \underline{0.93} &     \textbf{0.56} &  \textit{1.55} &              1.04 \\
+ HeartCl &    62 &              1.69 &     \textbf{0.41} &  \textit{1.96} &  \underline{1.33} \\
+ BcWisco &    70 &     \textit{2.02} &     \textbf{0.76} &           1.96 &  \underline{1.14} \\
+ Diabete &    77 &  \underline{0.55} &     \textbf{0.23} &  \textit{1.04} &              0.58 \\
+ TicTacT &    96 &  \underline{1.60} &     \textbf{0.21} &           2.32 &     \textit{2.47} \\
+ Balance &   126 &     \textbf{0.06} &  \underline{0.15} &           0.71 &     \textit{1.18} \\
+   Yeast &   149 &              0.54 &     \textbf{0.14} &  \textit{0.78} &  \underline{0.45} \\
+ Abalone &   418 &              1.20 &     \textbf{0.78} &  \textit{1.49} &  \underline{0.93} \\
+ PageBlo &   548 &     \textit{1.67} &     \textbf{0.38} &           1.17 &  \underline{0.49} \\
+ Shuttle &  5800 &     \textit{2.02} &     \textbf{1.05} &           1.84 &  \underline{1.14} \\
+   Adult &  6513 &  \underline{1.34} &     \textbf{1.11} &  \textit{1.59} &              1.36 \\
 \end{tabular}
-\caption{Predictive log losses for small sample sizes for different model selection criteria in 20 different datasets.}
-\label{tbl:preds}
+\end{center}
 \end{table}
-
-Table~\ref{tbl:preds} features predictive losses for the small sample
-sizes.  (The results for large sample sizes converge for all sensible
-criteria.)  As we can see, fNML still usually (12/20) performs best
-but qNML is not much worse while BDeu may sometimes perform much
-worse.
-
-Looking at the number of parameters for the same datasets and sample
-sizes (Table~\ref{tbl:nofparams}) reveals that qNML usually (18/20)
-yields simplest models, and sometimes, like in the BreastCancer data,
-the differences are of orders of magnitude.
-
-\begin{table}
-\centering
-\begin{tabular}{crrrr}
-\toprule
-           Data &     N &          BDeu &  fNML &          qNML \\
-\midrule
-           Iris &    16 &            36 &    33 &   \textbf{28} \\
-  PostOperative &    20 &          1304 &   464 &  \textbf{115} \\
-           Wine &    27 &         11559 &   780 &  \textbf{214} \\
- HeartHungarian &    45 &          4706 &   838 &  \textbf{100} \\
-   BreastCancer &    45 &         26774 &  2710 &  \textbf{246} \\
- HeartCleveland &    48 &         29346 &  1407 &  \textbf{204} \\
-          Ecoli &    51 &           113 &   167 &   \textbf{80} \\
-          Liver &    54 &            40 &    60 &   \textbf{24} \\
-          Glass &    55 &          1488 &   572 &  \textbf{100} \\
-        Thyroid &    55 &            37 &    70 &   \textbf{28} \\
-   HeartStatlog &    56 &         18414 &  1099 &  \textbf{159} \\
-      TicTacToe &    96 &         13645 &  2438 &  \textbf{696} \\
-        Balance &    96 &   \textbf{20} &   237 &           168 \\
-    BcWisconsin &   105 &          5921 &   816 &   \textbf{88} \\
-       Diabetes &   117 &            35 &   216 &   \textbf{34} \\
-          Yeast &   225 &   \textbf{78} &   307 &            80 \\
-        Abalone &   418 &            90 &   156 &   \textbf{62} \\
-     PageBlocks &   548 &           699 &   377 &   \textbf{56} \\
-        Shuttle &  2900 &           380 &   594 &  \textbf{111} \\
-          Adult &  4885 &  \textbf{599} &  1431 &           876 \\
-\bottomrule
-\end{tabular}
-\caption{Number of model parameters for small sample sizes for different model selection criteria in 20 different datasets.}
-\label{tbl:nofparams}
-\end{table}

### 2017-03-31 intro.tex
proofreading + images etc.

@@
 underlying causal mechanisms.  Techniques for learning the structure
-of Bayesian networks from the observational data has therefore been
-used for many tasks such as discovering cell signaling pathways from
-the protein activity data~\cite{bn4sigpath02}, revealing the business
-process structures~\cite{bn4bpmining} from the transaction logs and
-modeling brain region connectivity using fMRI
+of Bayesian networks from observational data have therefore been used
+for many tasks such as discovering cell signaling pathways from
+protein activity data~\cite{bn4sigpath02}, revealing the business
+process structures~\cite{bn4bpmining} from transaction logs and
+modeling brain-region connectivity using fMRI
 data~\cite{bn4brainconnect}.
@@
 Learning the structure of statistical dependencies can be seen as a
-model selection task where each model is a different hypothesis about
+model-selection task where each model is a different hypothesis about
 the conditional dependencies between sets of variables. Traditional
@@
 criterion has been the marginal likelihood (usually called BDeu for
-reasons explained later), but recent studies~\cite{cosco.uai07,Steck08}
+reasons explained later), but studies~\cite{cosco.uai07,Steck08}
 have found this criterion to be very sensitive to hyper parameters
@@
 for a good criterion, but its exact calculation is likely to be
-prohibitively demanding. In 2008, Silander et al. introduced a
-hyper-parameter free, NML inspired criterion called a factorized NML
+prohibitively expensive. In 2008, Silander et al. introduced a
+hyperparameter free, NML inspired criterion called the factorized NML
 (fNML)~\cite{cosco.pgm08a} that was shown to yield good predictive
-models without sensitivity problems.  However, from the structure
+models without such sensitivity problems. However, from the structure
 learning point of view, the fNML still sometimes appears to yield
 overly complex models. In this paper we introduce another NML related
-criterion, a quotient NML (qNML) that yields simpler models without
-sacrificing predictive accuracy. Furthermore, unlike the fNML, the
-qNML gives equal scores to the structures that encode the same
+criterion, the \textit{quotient NML} (qNML) that yields simpler models
+without sacrificing predictive accuracy. Furthermore, unlike the fNML,
+the qNML gives equal scores to structures that encode the same
 independence statements. Like other common model selection criteria,
-the qNML is also consistent.
+qNML is also consistent.
 
-We will next briefly introduce Bayesian networks and then review the
-BDeu and fNML criteria and introduce the qNML criterion.  We will also
-summarize the results for 20 datasets to back up our claim of qNML
-yielding parsimonious models with good predictive capabilities. The
+We next briefly introduce Bayesian networks and review the BDeu and
+fNML criteria and then introduce the qNML criterion.  We also
+summarize the results for 20 datasets to back up our claim that qNML
+yields parsimonious models with good predictive capabilities. The
 experiments with artificial data demonstrate the capability to give

### 2017-03-31 qnml.tex
some new stuff, mainly results
proofreading + images etc.

@@
 criterion for learning Bayesian network structures.  While equally
-efficient to compute than BDeu and fNML, it is free from
+efficient to compute as BDeu and fNML, it is free from
 hyperparameters, and it can be proven to give equal scores to
@@
 with $\prod_{X_i\in S} r_i$ different values which can then be modeled
-with a one-dimensional $P^1_{NML}$ code.  The $s^{qNML}$ does not
-necessarily define a distribution for $D$, but it is easy to verify
-that it coincides with the $P_{NML}(D;G)$ for all the networks that
-are composed of fully connected components.  The number of such
-networks equals the number of nonempty partitions of a set of $n$
-elements, i.e., the $n^\text{th}$ Bell number.
+with a one-dimensional $P^1_{NML}$ code.  The $s^{qNML}$-score does
+not necessarily define a distribution for $D$, but it is easy to
+verify that it coincides with the $\log P_{NML}(D;G)$ for all networks
+that are composed of fully connected components.  The number of such
+networks is lower bounded by the number of nonempty partitions of a
+set of $n$ elements, i.e., the $n^\text{th}$ Bell number.
 
-\subsection {qNML is score equivalent}
+We are now ready to prove some important properties of the qNML-score.
 
-The qNML yields equal scores for network structures that encode same
-sets of independencies. Verma and Pearl~\cite{Verm90} showed that the
-equivalent networks are exactly those who a) are same when directed
-arcs are substituted by undirected ones and b) who have same
-V-structures, i.e.  the variable triplets $(A,B,C)$ where both $A$ and
-$B$ are parents of $C$, but there is no arc between $A$ and $B$ (in
-neither direction).  Later, Chickering~\cite{Chick95} showed that all
-the equivalent network structures, and only them, can be reached from
-each other by reversing, one by one, so called covered arcs, i.e. the
-arcs from node $A$ to $B$, for which $B$'s parents other than $A$ are
-exactly the $A$'s parents ($G_B=\{A\}\cup G_A$).
+\subsection {qNML IS SCORE EQUIVALENT}
+
+qNML yields equal scores for network structures that encode same sets
+of independencies. Verma and Pearl~\cite{Verm90} showed that the
+equivalent networks are exactly those which a) are same when directed
+arcs are substituted by undirected ones and b) which have same
+\textit{V-structures}, i.e. the variable triplets $(A,B,C)$ where both
+$A$ and $B$ are parents of $C$, but there is no arc between $A$ and
+$B$ (in either direction).  Later, Chickering~\cite{Chick95} showed
+that all the equivalent network structures, and only those structures,
+can be reached from each other by reversing, one by one, the so called
+\textit{covered arcs}, i.e. the arcs from node $A$ to $B$, for which
+$B$'s parents other than $A$ are exactly  $A$'s parents
+($G_B=\{A\}\cup G_A$).
 
@@
 \begin{proof}
-  Now the scores for structures can be decomposed as the
+  Now the scores for structures can be decomposed as
   $s^{qNML}(D;G)=\sum_{i=1}^{n}s_i^{qNML}(D;G)$ and
@@
   terms corresponding to the variables $A$ and $B$ in these sums are
-  different, it is enough to show that sum of these two terms are
+  different, it is enough to show that the sum of these two terms are
   equal for $G$ and $G'$. Since we can assume the data to be fixed we
-  lighten up the notations and write
+  lighten up the notation and write
   $P^1_{NML}(i,G_i) := P^1_{NML}(D[\cdot,(i,G_i)];G)$ and
@@
 \end{eqnarray}
-  due to equations $\{A\}\cup G_A = G_B$, $\{B\}\cup G'_B = G'_A$,
-  $\{B\}\cup G_B = \{A\} \cup G'_A$, and $G_A = G'_B$ that follow
-  easily from the definition of the covered arc.
+  using the equations $\{A\}\cup G_A = G_B$, $\{B\}\cup G'_B = G'_A$,
+  $\{B\}\cup G_B = \{A\} \cup G'_A$, and $G_A = G'_B$ which follow
+  easily from the definition of covered arcs.
 \end{proof}
 
-\subsection{qNML is consistent}
+\subsection{qNML IS CONSISTENT}
 
@@
 BIC criterion which is known to be consistent \cite{Schw78, Haug88}.
+The outline of this proof follows a similar pattern to that in
+\cite{SilanderIJAR10} where the consistency of fNML was proved.
 
@@
 \begin{equation}\label{BIC}
-\textnormal{BIC}(D,G) = \sum_{i = 1}^n \log P(D_i \ | \ \hat{\theta}_{i | G_i} ) - \frac{q_i(r_i - 1)}{2} \log N,
+\textnormal{BIC}(D;G) = \sum_{i = 1}^n \log P(D_i \ | \ \hat{\theta}_{i | G_i} ) - \frac{q_i(r_i - 1)}{2} \log N,
 \end{equation}
@@
 
-Since the both BIC and qNML scores are decomposable, we can focus
-on studying the local scores. We will next show that, asymptotically, the local qNML score equals the local BIC score. This is formulated in the following theorem:
+Since both the BIC and qNML scores are decomposable, we can focus on
+studying the local scores. We will next show that, asymptotically, the
+local qNML score equals the local BIC score. This is formulated in the
+following theorem:
 
@@
 Let $r_i$ and $q_i$ denote the number of possible values for variable
-$X_i$ and the possible configurations of parents $X_{G_i}$,
+$X_i$ and its possible configurations of parents $X_{G_i}$,
 respectively. As $N \to \infty$
@@
 
-By comparing the equations (\ref{BIC}) and (\ref{qnmlDef2}), we see that proving our clam boils down to showing two things: 1) the terms involving the maximized likelihoods are equal and 2) the penalty terms are asymptotically equivalent.
-We will formulate these into two lemmas.
+By comparing the equations (\ref{BIC}) and (\ref{qnmlDef2}), we see
+that proving our clam boils down to showing two things: 1) the terms
+involving the maximized likelihoods are equal and 2) the penalty terms
+are asymptotically equivalent. We will formulate these as two
+lemmas.
 
@@
 \begin{proof}
-We can write the terms in the left hand side of the equation as
+We can write the terms on the left side of the equation as
 $$
@@
 P(D_{G_i} \ | \ \hat{\theta}_{G_i} ) =  \prod_{j} \Le \frac{N_{ij}}{N}  \Ri^{N_{ij}}.
-$$Here, $N_{ijk}$ denotes the number of times we observe $X_i$ taking value $k$ when its parents are in $j$:th configuration in our data matrix $D$.
+$$Here, $N_{ijk}$ denotes the number of times we observe $X_i$ taking value $k$ when its parents are in $j^\text{th}$ configuration in our data matrix $D$.
 Also, $N_{ij} = \sum_k N_{ijk}$ (and $\sum_{k,j}N_{ijk} = N$ for all $i$).
 
-Now,
+Therefore,
 \begin{align*}
@@
 \end{proof}
-This concludes our proof since lemmas \ref{MLLemma} and \ref{penaltyLemma} imply the Theorem \ref{consistency}. The outline of this proof follows the
-similar pattern as found in \cite{SilanderIJAR10} where the
-consistency of fNML was proved.
+This concludes our proof since Lemmas \ref{MLLemma} and
+\ref{penaltyLemma} imply Theorem \ref{consistency}.
 
-\subsection{qNML equals NML for many models}
-The fNML criterion equals NML criterion just for the Bayesian network
-with no arcs.  We will next show that the qNML equals NML when the
-connected components of the network are tournaments (complete directed
-acyclic subgraphs of G).
+\subsection{qNML EQUALS NML FOR MANY MODELS}
+The fNML criterion can be seen as a computationally feasible
+approximation of the more desirable NML criterion.  However, the fNML
+criterion equals the NML criterion only for the Bayesian network
+structure with no arcs.  We will next show that the qNML criterion
+equals the NML criterion for all the networks $G$ whose connected
+components tournaments (i.e., complete directed acyclic subgraphs of
+$G$).
 
@@
 \begin{proof}
-We first note that the NML-criterion for the Bayesian networks
+We first show that the NML-criterion for a Bayesian network
 decomposes by the connected components.
 
-The maximum likelihood for the data $D$ decomposes, and we can write
+Because the maximum likelihood for the data $D$ decomposes,
+we can write
 \begin{eqnarray}
@@
             && = \frac{\prod_{c=1}^C P(D_{V^c};\hat\theta(D_{V^c}),G)} 
-            {\prod_{c=1}^C\sum_D^cP(D_{V^c};\hat\theta(D_{V^c}),G)}.\nonumber \\
+            {\prod_{c=1}^C\sum_{D_{V^c}}P(D_{V^c};\hat\theta(D_{V^c}),G)}.\nonumber \\
             && = \prod_{c=1}^CP_{NML}(D_{V^c};G).
@@
 then for any data $D$, $s^{qNML}(D;G)=\log P_{NML}(D;G)$.  Due to the
-score equivalence of both the NML criterion and the qNML criterion, we
-may pick such a tournament $G$ that linear ordering defined by $G$
-matches the ordering of the data columns, i.e., $i<j$ implies $G_i
-\subset G_j$. Now from the definition~(\ref{eqn:qnmldef}) of the qNML
+score equivalence of the NML criterion and the qNML criterion, we may
+pick a tournament $G$ such that the linear ordering defined by $G$ matches
+the ordering of the data columns, i.e., $i<j$ implies $G_i \subset
+G_j$. Now from the definition~(\ref{eqn:qnmldef}) of the qNML
 criterion we see that for the tournament $G$, the sum telescopes
 leaving us with $s^{qNML}(D;G) = \log P^1_{NML}(D[\cdot,G];G)$, thus
-it is enough to show that $P^1_{NML}(D;G)=P_{NML}(D;G)$.
-This follows, since for any data vector $x$ in data $D$, the
-$P^1(x;\hat\theta(D),G) = P(x;\hat\theta(D),G)$,
-where we have denoted by $P^1$ the model that takes $n$-dimensional
-vectors to be values of the single (collapsed) categorical variable.
-Denoting prefixes of data vector $x$ by $x^{:i}$, and number of times
-such a prefix appears in a $N\times n$ datamatrix $D$ by $n_D(x^{:i})$,
-($n_D(x^{:0})=N)$, we have
+it is enough to show that $P^1_{NML}(D;G)=P_{NML}(D;G)$.  This
+follows, since for any data vector $x$ in data $D$, we have
+$P^1(x;\hat\theta(D),G) = P(x;\hat\theta(D),G)$, where $P^1$ denotes
+the model that takes $n$-dimensional vectors to be values of the
+single (collapsed) categorical variable.  Denoting prefixes of data
+vector $x$ by $x^{:i}$, and the number of times such a prefix appears
+in a $N\times n$ data matrix $D=[d_1,\ldots,d_N]$ by $N_D(x^{:i})$,
+(so that $N_D(x^{:0})=N)$, we have
 \begin{eqnarray}
   P(D;\hat\theta(D),G) &=& \prod_{j=1}^{N}P(d_j;\hat\theta(D),G) \nonumber \\
-  &=& \prod_{j=1}^{N} \prod_{i=1}^{n} \frac{n_D(d_j^{:i})}{n_D(d_j^{:i-1})} \nonumber\\
-  &=& \prod_{j=1}^{N} \frac{n_D(d_j^{:n})}{N} \nonumber \\
+  &=& \prod_{j=1}^{N} \prod_{i=1}^{n} \frac{N_D(d_j^{:i})}{N_D(d_j^{:i-1})} \nonumber\\
+  &=& \prod_{j=1}^{N} \frac{N_D(d_j^{:n})}{N} \nonumber \\
   &=& P^1(D;\hat\theta^1(D),G).
@@
 Equality established, we would like to still state the number of
-different networks connected components of which are tournaments.  We
+different networks whose connected components are tournaments.  We
 start by generating all the $p(n)$ integer partitions i.e. ways to
@@
 one that corresponds just to the empty network.  We count number of
-networks for one such partition size profile (and then later sum these
+networks for one such partition size-profile (and then later sum these
 counts up).  For any such partition profile $(p_1,\ldots,p_k)$ we can
-(think we can) count the ways we can pick the nodes to different parts
+count the ways we can assign the nodes to different parts
 and then order each part. This leads to the product $n\choose
 p_1$$p_1$!${n-p_1}\choose p_2$$p_2!$${n-p_1-p2}\choose
-p_3$$p3!$$\ldots$${n-\sum_{j=1}^{k-1}p_j}\choose p_k$$p_k!$. However,
+p_3$$p_3!$$\ldots$${n-\sum_{j=1}^{k-1}p_j}\choose p_k$$p_k!$. However,
 the order of different parts of the same size does not matter, so for
@@
 
-With $p(n) =$ the number of integer partitions of $n$, 
-     $d(i)$ = the number of different parts of the $i$-th partition of $n$, $m(i,j) =$ multiplicity of the $j$-th part of the $i$-th partition of $n$,  one has:
-
+\textit{
+With $p(n) =$ the number of integer partitions of $n$, $d(i)$ = the
+number of different parts of the $i^\text{th}$ partition of $n$,
+$m(i,j) =$ multiplicity of the $j^\text{th}$ part of the $i^\text{th}$
+partition of $n$, one has:
+}
 $$

### 2017-03-31 conclusion.tex
added a bit to the conclusion

@@
 predictive terms, it often yields significantly simpler models than
-other common model selection criteria.  The computational cost of qNML
-equals the cost of earlier criteria.  The criterion also gives equal
-scores for models that encode same independence hypotheses about the
-joint probability distribution.  qNML also coincides with NML
-criterion for exponentially many models.
+other common model selection criteria other than BIC that has a very strong
+bias for simplicity. The computational cost of qNML
+equals the cost of the current state-of-the-art criteria. The
+criterion also gives equal scores for models that encode same
+independence hypotheses about the joint probability distribution.
+qNML also coincides with NML criterion for many models.
+In our experiments, the qNML criterion appears as a safe choice for a
+model selection criterion that balances
+parsimony, predictive capability and the ability to quickly converge to the
+generating model.

### 2017-03-31 bn.tex
finetuning

@@
 parameters and the sum in the denominator goes over all the possible
-$N\times n$ data matrices. While it is easy to see that this criterion
-satisfies the requirement of giving equal scores to equal structures,
-the normalizing constant renders the computation
-infeasible.
+$N\times n$ data matrices. This information theoretic NML criterion
+can be justified from the minimum description length point of view,
+\cite{Riss78,Grun07}. It has been shown to be robust with respect to
+different data generating mechanisms where a good choice of prior
+is challenging, see~\cite{eggeling2014robust,maatta16}. While it is
+easy to see that the NML criterion satisfies the requirement of giving
+equal scores to equal structures, the normalizing constant renders the
+computation infeasible.
 
@@
 models superior to BDeu.  However, the criterion does not satisfy the
-property of giving the same score for models that correspond to
-the same dependence statements. Furthermore, the learned structures
-are often rather complex which hampers their interpretation.
+property of giving the same score for models that correspond to the
+same dependence statements. Furthermore, the learned structures are
+often rather complex (see Figure~\ref{fig:bcnpmean}) which also
+hampers their interpretation. The quest for a model selection
+criterion that would yield more parsimonious, easier to interpret, but
+still predictive Bayesian networks structures is one of the main
+motivations for this work.
+
+\begin{figure}
+\centering
+\includegraphics[width=8cm,height=5cm]{qNML_images/breast_cancer_npmean.pdf}
+\caption{Number of parameters in a breast cancer model as a function
+  of sample size for different model selection criteria.}
+\label{fig:bcnpmean}
+\end{figure}
 
 \begin{table}
-\centering
+\caption{Regret values for various values of $N$ and $r$.}
+\label{tbl:regrets}
+\begin{center}
 \begin{tabular}{crrrr}
-\toprule
 N & r & eq. (\ref{eqn:szp1}) & eq. (\ref{eqn:szp2}) & exact \\
@@
 \end{tabular}
-\caption{Regret values for various values of $N$ and $r$.}
-\label{tbl:regrets}
+\end{center}
 \end{table}

### 2017-03-31 qnml.tex
finetuning

@@
 We can write the terms on the left side of the equation as
-$$
-P(D_{i, G_i} \ | \ \hat{\theta}_{i, G_i}) = \prod_{j,k} \Le \frac{N_{ijk}}{N}  \Ri^{N_{ijk}}
-$$and
-$$
-P(D_{G_i} \ | \ \hat{\theta}_{G_i} ) =  \prod_{j} \Le \frac{N_{ij}}{N}  \Ri^{N_{ij}}.
-$$Here, $N_{ijk}$ denotes the number of times we observe $X_i$ taking value $k$ when its parents are in $j^\text{th}$ configuration in our data matrix $D$.
+\begin{eqnarray*}
+P(D_{i, G_i} \ | \ \hat{\theta}_{i, G_i}) &=& \prod_{j,k} \Le \frac{N_{ijk}}{N}  \Ri^{N_{ijk}}, \text{ and }\\
+P(D_{G_i} \ | \ \hat{\theta}_{G_i} ) &=&  \prod_{j} \Le \frac{N_{ij}}{N}  \Ri^{N_{ij}}.
+\end{eqnarray*}
+Here, $N_{ijk}$ denotes the number of times we observe $X_i$ taking value $k$ when its parents are in $j^\text{th}$ configuration in our data matrix $D$.
 Also, $N_{ij} = \sum_k N_{ijk}$ (and $\sum_{k,j}N_{ijk} = N$ for all $i$).
-
 Therefore,
@@
 &= \frac{ \prod_{j,k} \Le \frac{N_{ijk}}{N}  \Ri^{N_{ijk}}}{\prod_{j}\prod_{k} \Le \frac{N_{ij}}{N}  \Ri^{N_{ijk}}}\\
-&= \prod_{j,k} \Le \frac{N_{ijk}}{N_{ij}} \Ri^{N_{ijk}} \\
+%% &= \prod_{j,k} \Le \frac{N_{ijk}}{N_{ij}} \Ri^{N_{ijk}} \\
 &= P(D_i \ | \ \hat{\theta}_{i | G_i} ).
@@
 \end{proof}
-
 Next, we consider the difference of regrets in
@@
 
-Equality established, we would like to still state the number of
-different networks whose connected components are tournaments.  We
-start by generating all the $p(n)$ integer partitions i.e. ways to
-partition $n$ labelled into parts with different size-profiles.  For
-example, for $n=4$, we have $p(4)=5$ partition profiles $[[4], [3, 1],
-  [2, 2], [2, 1, 1], [1, 1, 1, 1]]$. Each of these partition size
-profiles corresponds to many different networks, apart from the last
-one that corresponds just to the empty network.  We count number of
-networks for one such partition size-profile (and then later sum these
-counts up).  For any such partition profile $(p_1,\ldots,p_k)$ we can
-count the ways we can assign the nodes to different parts
-and then order each part. This leads to the product $n\choose
-p_1$$p_1$!${n-p_1}\choose p_2$$p_2!$${n-p_1-p2}\choose
+Equality established, we would like to still state the number $a(n)$
+of different $n$-node networks whose connected components are
+tournaments.  We start by generating all the $p(n)$ integer partitions
+i.e. ways to partition $n$ labelled into parts with different
+size-profiles.  For example, for $n=4$, we have $p(4)=5$ partition
+profiles $[[4], [3, 1], [2, 2], [2, 1, 1], [1, 1, 1, 1]]$. Each of
+these partition size profiles corresponds to many different networks,
+apart from the last one that corresponds just to the empty network.
+We count number of networks for one such partition size-profile (and
+then later sum these counts up).  For any such partition profile
+$(p_1,\ldots,p_k)$ we can count the ways we can assign the nodes to
+different parts and then order each part. This leads to the product
+$n\choose p_1$$p_1$!${n-p_1}\choose p_2$$p_2!$${n-p_1-p2}\choose
 p_3$$p_3!$$\ldots$${n-\sum_{j=1}^{k-1}p_j}\choose p_k$$p_k!$. However,
@@
 that the product above telescopes, leaving us a formula for OEIS
-sequence A000262 as described by Thomas Wieder:
+sequence A000262\footnote{https://oeis.org/A000108} as
+described by Thomas Wieder:
 
@@
 $$
-For example, $a(4)=\frac{4!}{1!}+\frac{4!}{1!1!}+\frac{4!}{2!}+\frac{4!}{1!2!}+\frac{4!}{4!}=73$.
+For example, $a(4)=\frac{4!}{1!}+\frac{4!}{1!1!}+\frac{4!}{2!}+\frac{4!}{1!2!}+\frac{4!}{4!}=73$. In general this sequence grows rapidly; $1, 1, 3, 13, 73, 501, 4051, 37633, 394353, 4596553, \ldots$.

### 2017-03-31 bn.tex
Typos

@@
 posterior probability $P(G|D,\alpha)$, where $\alpha$ denotes the
-hyper\nobreakdash-parameters for the model parameters $\theta$, and
+hyperparameters for the model parameters $\theta$, and
 the $D$ is a collection of $N$ $n$\nobreakdash-dimensional i.i.d. data
@@
 one of the desiderata for a good model selection criterion was that it
-is score-equivalent, i.e., it would yield equal scores for
+is score equivalent, i.e., it would yield equal scores for
 essentially equivalent models~\cite{Verm90}.  For example, the score
@@
 not all the hyperparameters $\alpha$ are possible and for practical
-reasons Buntine~\cite{Bunt91} suggested a so called BDeu score with
+reasons Buntine~\cite{Bunt91} suggested a so-called BDeu score with
 just one hyperparameter $\alpha\in R_{++}$ so that
@@
 distribution are chosen in BDeu as using Jeffreys' prior,
-$\theta_{ijk}\sim Dir(\frac{1}{2},ldots,\frac{1}{2})$, does not suffer
+$\theta_{ijk}\sim Dir(\frac{1}{2},\ldots,\frac{1}{2})$, does not suffer
 from this anomaly. However, using Jeffreys' prior causes marginal
-likelihood score not to be score-equivalent.
+likelihood score not to be score equivalent.
 
@@
 parameters and the sum in the denominator goes over all the possible
-$N\times n$ data matrices. This information theoretic NML criterion
+$N\times n$ data matrices. This information-theoretic NML criterion
 can be justified from the minimum description length point of view,

### 2017-03-31 experiments.tex
finetuning
Typos

@@
-\section{EXPERIMENTAL RESULTS}
-
-\begin{figure}
+\begin{figure}[h]
 \centering
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=8cm,height=5cm]{qNML_images/art7_mean.pdf}
-\caption{Finding generating models of 7 arcs.}
-\label{fig:4arcs}
-\end{figure}
+\section{EXPERIMENTAL RESULTS}
 
@@
 was to generate the parameters independently from a Dirichlet
-distribution $\theta_{ij} \sim Dir(\frac{1}{2},\ldots, \frac{1}{2})$
-that is the Jeffreys' prior for the multinomial model. This distribution
+distribution $\theta_{ij} \sim Dir(\frac{1}{2},\ldots, \frac{1}{2})$,
+that is, the Jeffreys' prior for the multinomial model. This distribution
 was selected instead of the uniform distribution in order to make the
@@
 (sub)sample size.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=8cm,height=5cm]{qNML_images/art7_mean.pdf}
+\caption{Finding generating models of 7 arcs.}
+\label{fig:4arcs}
+\end{figure}
 
@@
 multinomial data is known to closely match the marginal distribution of
-the data when the Jefrreys' prior is used.
+the data when the Jeffreys' prior is used.
 
@@
 dataset was selected as training data and the rest 90\% as test data.
-The training was conducted using a dynamic programming based exact structure
+The training was conducted using dynamic programming based exact structure
 learning~\cite{cosco.uai06} which limited the number $n$ of variables
@@
 N_{ijk}+\frac{1}{r_iq_i}$.  In the spirit of keeping the scores
-hyperparameter free, for structures learned by the other model
+hyperparameter-free, for structures learned by the other model
 selection criteria, we used the predictive conditional NML
@@
 $e(n)=(\frac{n+1}{n})^n$ as suggested in~\cite{Riss07b}.
-
-\begin{figure}
-\centering
-\includegraphics[width=8cm,height=5cm]{qNML_images/breast_cancer_npmean.pdf}
-\caption{Number of parameters in a breast cancer model as a function
-  of sample size for different model selection criteria.}
-\label{fig:bcnpmean}
-\end{figure}
 

### 2017-03-31 intro.tex
Typos

@@
 Learning the structure of statistical dependencies can be seen as a
-model-selection task where each model is a different hypothesis about
+model selection task where each model is a different hypothesis about
 the conditional dependencies between sets of variables. Traditional
@@
 reasons explained later), but studies~\cite{cosco.uai07,Steck08}
-have found this criterion to be very sensitive to hyper parameters
+have found this criterion to be very sensitive to hyperparameters
 and to yield undesirably complex models for small sample sizes.
 
-The information theoretic normalized maximum likelihood (NML)
+The information-theoretic normalized maximum likelihood (NML)
 criterion~\cite{Shta87,Riss96a} would otherwise be an ideal candidate

### 2017-03-31 qnml.tex
Typos

@@
 that all the equivalent network structures, and only those structures,
-can be reached from each other by reversing, one by one, the so called
+can be reached from each other by reversing, one by one, the so-called
 \textit{covered arcs}, i.e. the arcs from node $A$ to $B$, for which
@@
 with a probability tending to one as $N$ increases. We will show this
-by first proving that qNML is asymptotically equivalent to widely used
+by first proving that qNML is asymptotically equivalent to the widely used
 BIC criterion which is known to be consistent \cite{Schw78, Haug88}.

### 2017-03-31 experiments.tex
Minor fixes

@@
 was to generate the parameters independently from a Dirichlet
-distribution $\theta_{ij} \sim Dir(\frac{1}{2},\ldots, \frac{1}{2})$,
-that is, the Jeffreys' prior for the multinomial model. This distribution
+distribution $\theta_{ij} \sim Dir(\frac{1}{2},\ldots, \frac{1}{2})$
+that is the Jeffreys' prior for the multinomial model. This distribution
 was selected instead of the uniform distribution in order to make the
@@
 for all of them. In each experiment, random 10\% of the original
-dataset was selected as training data and the rest 90\% as test data.
+data set was selected as training data and the rest 90\% as test data.
 The training was conducted using dynamic programming based exact structure
@@
 \begin{table}
-\caption{Predictive log losses for small sample sizes for different model selection criteria in 20 different datasets.}
+\caption{Predictive log losses for small sample sizes for different model selection criteria in 20 different data sets.}
 \label{tbl:preds}
@@
 
-Looking at the number of parents for our 20 datasets and sample
+Looking at the number of parents for our 20 data sets and sample
 sizes (Table~\ref{tbl:nofparams}) again features BIC's preference for
@@
   \caption{Average number of parents per node 
-    for different model selection criteria in 20 different datasets.}
+    for different model selection criteria in 20 different data sets.}
 \label{tbl:nofparams}

### 2017-03-31 intro.tex
Minor fixes

@@
 The information-theoretic normalized maximum likelihood (NML)
-criterion~\cite{Shta87,Riss96a} would otherwise be an ideal candidate
+criterion~\cite{Shta87,Riss96a} would otherwise be a potential candidate
 for a good criterion, but its exact calculation is likely to be
@@
 fNML criteria and then introduce the qNML criterion.  We also
-summarize the results for 20 datasets to back up our claim that qNML
+summarize the results for 20 data sets to back up our claim that qNML
 yields parsimonious models with good predictive capabilities. The

### 2017-03-31 qnml.tex
Minor fixes

@@
 of independencies. Verma and Pearl~\cite{Verm90} showed that the
-equivalent networks are exactly those which a) are same when directed
-arcs are substituted by undirected ones and b) which have same
+equivalent networks are exactly those which a) are the same when directed
+arcs are substituted by undirected ones and b) which have the same
 \textit{V-structures}, i.e. the variable triplets $(A,B,C)$ where both

### 2017-04-01 abstract.tex
Typos
finetuning + Teemu's comments

--- 2017-03-16 abstract.tex
@@
 \begin{abstract}
-Learning the dependency structure of a multivariate distribution from
-the observational data is an important task since it allows us to
-speculate about the underlying causal mechanisms that induce
-dependencies. This learning task can be approached as a model
-selection problem, but recent studies have revealed that the popular
-Bayesian model selection criterion is not satisfactory. We review some
-of the information theoretic alternatives and introduce a new one
-called a quotient normalized maximum likelihood criterion.
+We introduce an information theoretic criterion for Bayesian network
+structure learning which we call quotient normalized maximum
+likelihood (qNML). In contrast to the closely related factorized
+normalized maximum likelihood criterion, qNML satisfies the property
+of likelihood equivalence. It is also decomposable and completely free
+of adjustable hyperparameters. For practical computations, we identify
+a remarkably accurate approximation proposed earlier by Szpankowski
+and Weinberger. Experiments on both simulated and real data
+demonstrate that the new criterion leads to parsimonious models with
+good predictive accuracy.
 \end{abstract}

### 2017-04-01 experiments.tex
finetuning + Teemu's comments

@@
 When predicting with structures learned by the BDeu score, we used the
-Bayesian expected parameter values $\theta_{ijk} \propto
+Bayesian predictive parameter values (BPP) $\theta_{ijk} \propto
 N_{ijk}+\frac{1}{r_iq_i}$.  In the spirit of keeping the scores
 hyperparameter-free, for structures learned by the other model
-selection criteria, we used the predictive conditional NML
+selection criteria, we used the sequential predictive NML (sNML)
 parametrization $\theta_{ijk}\propto e(n_{ijk})(n_{ijk}+1)$, where
@@
 \begin{tabular}{crrrrr}
-    Data &     N &              BDeu &               BIC &               fNML &               qNML \\
+       Data
+    & N
+    & \multicolumn{1}{p{0.7cm}}{\centering BDeu \\ BPP}
+    & \multicolumn{1}{p{0.8cm}}{\centering BIC \\ sNML}
+    & \multicolumn{1}{p{0.9cm}}{\centering fNML \\ sNML}
+    & \multicolumn{1}{p{0.9cm}}{\centering qNML \\ sNML}\\
        \midrule
@@
 \begin{tabular}{crrrrr}
-    Data &     N &              BDeu &               BIC &           fNML &              qNML \\
+%%    Data &     N &              BDeu &               BIC &           fNML &              qNML \\
+       Data
+    & N
+    & \multicolumn{1}{p{0.7cm}}{\centering BDeu\\ }
+    & \multicolumn{1}{p{0.8cm}}{\centering BIC\\ }
+    & \multicolumn{1}{p{0.9cm}}{\centering fNML\\}
+    & \multicolumn{1}{p{0.9cm}}{\centering qNML\\}\\
 \midrule

### 2017-04-01 intro.tex
finetuning + Teemu's comments

@@
 model selection criteria such as the Akaike information
-criterion~\cite{Akai73} and the Bayesian information
-criterion~\cite{Schw78} have also been used for the task, but recent
-comparisons have not been favorable for these criteria in terms of
-identifying the appropriate structure and/or predictive
-performance~\cite{cosco.pgm08a}. Traditionally, the most popular
+criterion (AIC)~\cite{Akai73} and the Bayesian information
+criterion (BIC)~\cite{Schw78} have also been used for the task, but recent
+comparisons have not been favorable for AIC, and BIC appears to
+require large sample sizes in order to identify appropriate
+structures~\cite{cosco.pgm08a,Liu2012}. Traditionally, the most popular
 criterion has been the marginal likelihood (usually called BDeu for
-reasons explained later), but studies~\cite{cosco.uai07,Steck08}
-have found this criterion to be very sensitive to hyperparameters
-and to yield undesirably complex models for small sample sizes.
+reasons explained later), but studies~\cite{cosco.uai07,Steck08} have
+found this criterion to be very sensitive to hyperparameters and to
+yield undesirably complex models for small sample sizes.
 

### 2017-04-01 experiments.tex
Finetuning

@@
 \caption{Finding generating models of 7 arcs.}
-\label{fig:4arcs}
+\label{fig:7arcs}
 \end{figure}
@@
 mechanism matches the assumptions of the BDeu-score, BDeu usually
-also ranks the generating structure higher than the other scores
-(Figure X).  On the other hand, when the parameters are drawn from the
+also ranks the generating structure higher than the other scores (see Figure
+~\ref{fig:4arcs} and \ref{fig:7arcs}).  On the other hand, when the parameters are drawn from the
 Jeffreys' prior, the fNML score appears to rank the generating structure
@@
 relatively sparse networks (4 arcs) and for dense ones (7 arcs). The
-ranking ability the qNML criterion appears to be between fNML and BDeu
+ranking ability of the qNML criterion appears to be between fNML and BDeu
 in 3 out of 4 settings which is good considering that one of these
-criteria is always the winner. For the dense networks with BDeu prior,
+criteria is always the winner. For dense networks with the BDeu prior,
 qNML appears to perform worse than BDeu and fNML but still much

### 2017-04-01 intro.tex
Finetuning

@@
 the conditional dependencies between sets of variables. Traditional
-model selection criteria such as the Akaike information
-criterion (AIC)~\cite{Akai73} and the Bayesian information
-criterion (BIC)~\cite{Schw78} have also been used for the task, but recent
+model selection criteria such as the Akaike information criterion
+(AIC)~\cite{Akai73} and the Bayesian information criterion
+(BIC)~\cite{Schw78} have also been used for the task, but recent
 comparisons have not been favorable for AIC, and BIC appears to
 require large sample sizes in order to identify appropriate
-structures~\cite{cosco.pgm08a,Liu2012}. Traditionally, the most popular
-criterion has been the marginal likelihood (usually called BDeu for
-reasons explained later), but studies~\cite{cosco.uai07,Steck08} have
+structures~\cite{cosco.pgm08a,Liu2012}. Traditionally, the most
+popular criterion has been the Bayesian marginal
+likelihood~\cite{Heck95b} and its BDeu variant (see
+section~\ref{sec:bns}), but studies~\cite{cosco.uai07,Steck08} have
 found this criterion to be very sensitive to hyperparameters and to
@@
 The information-theoretic normalized maximum likelihood (NML)
-criterion~\cite{Shta87,Riss96a} would otherwise be a potential candidate
-for a good criterion, but its exact calculation is likely to be
-prohibitively expensive. In 2008, Silander et al. introduced a
+criterion~\cite{Shta87,Riss96a} would otherwise be a potential
+candidate for a good criterion, but its exact calculation is likely to
+be prohibitively expensive. In 2008, Silander et al. introduced a
 hyperparameter free, NML inspired criterion called the factorized NML
@@
 models without such sensitivity problems. However, from the structure
-learning point of view, the fNML still sometimes appears to yield
+learning point of view, fNML still sometimes appears to yield
 overly complex models. In this paper we introduce another NML related
 criterion, the \textit{quotient NML} (qNML) that yields simpler models
-without sacrificing predictive accuracy. Furthermore, unlike the fNML,
-the qNML gives equal scores to structures that encode the same
-independence statements. Like other common model selection criteria,
-qNML is also consistent.
+without sacrificing predictive accuracy. Furthermore, unlike fNML,
+qNML is \textit{score equivalent}, i.e., it gives equal scores to
+structures that encode the same independence and dependence
+statements. Like other common model selection criteria, qNML is also
+consistent.
 

### 2017-04-01 qnml.tex
finetuning + Teemu's comments
Finetuning

@@
   && = \frac{P(D;\hat\theta(D),G)}
-            {\sum_{D_{V_1}}\ldots \sum_{D_{V_C}}\prod_{c=1}^C P(D_{V^c};\hat\theta(D_{V^c}),G)} \nonumber \\
+            {\sum_{D'_{V_1}}\ldots \sum_{D'_{V_C}}\prod_{c=1}^C P(D'_{V^c};\hat\theta(D'_{V^c}),G)} \nonumber \\
             && = \frac{\prod_{c=1}^C P(D_{V^c};\hat\theta(D_{V^c}),G)} 
-            {\prod_{c=1}^C\sum_{D_{V^c}}P(D_{V^c};\hat\theta(D_{V^c}),G)}.\nonumber \\
+            {\prod_{c=1}^C\sum_{D'_{V^c}}P(D'_{V^c};\hat\theta(D'_{V^c}),G)}.\nonumber \\
             && = \prod_{c=1}^CP_{NML}(D_{V^c};G).
@@
 
-\textit{
-With $p(n) =$ the number of integer partitions of $n$, $d(i)$ = the
+\textit{With $p(n) =$ the number of integer partitions of $n$, $d(i)$ = the
 number of different parts of the $i^\text{th}$ partition of $n$,
 $m(i,j) =$ multiplicity of the $j^\text{th}$ part of the $i^\text{th}$
-partition of $n$, one has:
-}
+partition of $n$, one has:}
 $$

### 2017-04-01 bn.tex
finetuning + Teemu's comments
Finetuning
small fixes

@@
 $X$. The component $G$ defines the structure of the model as a
-directed acyclic graph that has exactly one node for each component of
+directed acyclic graph (DAG) that has exactly one node for each component of
 $X$. The structure $G=(G_{1},\ldots,G_{n})$ defines for each
@@
 \label{eqn:bayesmix}
-P(D|G,\alpha) & = & \prod_{i=1}^{n}\prod_{j=1}^{q_i}
-P(D[G_i=j,i]|\alpha)\\ & = & \prod_{i=1}^{n}\prod_{j=1}^{q_i}\int
-P([G_i=j,i]|\theta)P(\theta|\alpha) d\theta.\nonumber
+\lefteqn{P(D|G,\alpha)}\nonumber\\
+&&=\prod_{i=1}^{n}\prod_{j=1}^{q_i}P(D[G_i=j,i]|\alpha)\nonumber\\
+&&=\prod_{i=1}^{n}\prod_{j=1}^{q_i}\int P([G_i=j,i]|\theta)P(\theta|\alpha) d\theta.
 \end{eqnarray}
@@
 data $D$, the values of $X_i$ are completely determined by variables
-in set $Z$, so that the entropy $H(X | Z) = 0$. Now, if we can further
+in set $Z$, so that the empirical entropy $H(X_i | Z) = 0$. Now, if we can further
 find one or more variables, denoted by $Y$, whose values are
@@
 \begin{equation}
-P^1_{NML}(D[G_i=j,i];G)=\frac{P(D|\hat\theta(D[G_i=j,i];G))}{\sum_{D'}{P(D';\hat\theta(D'|G))}},
+P^1_{NML}(D[G_i=j,i];G)=\frac{P(D|\hat\theta(D[G_i=j,i];G))}{\sum_{D'}{P(D';\hat\theta(D';G))}},
 \end{equation}
@@
 property of giving the same score for models that correspond to the
-same dependence statements. Furthermore, the learned structures are
+same dependence statements.
+The score equivalence is usually viewed desirable when DAGs are considered only as models for conditional independence, without any causal interpretation.
+Furthermore, the learned structures are
 often rather complex (see Figure~\ref{fig:bcnpmean}) which also

### 2017-04-01 qnml.tex
small fixes

@@
 Let $r_i$ and $q_i$ denote the number of possible values for variable
-$X_i$ and its possible configurations of parents $X_{G_i}$,
+$X_i$ and its possible configurations of parents $G_i$,
 respectively. As $N \to \infty$

### 2017-04-01 bn.tex
last tunings

@@
 $G_{i}(x)$. It is customary to enumerate all the possible
-sub\nobreakdash-vectors $G_{i}(X)$ from $1$ to $q_{i}=\prod_{h\in
+sub\nobreakdash-vectors $G_{i}(x)$ from $1$ to $q_{i}=\prod_{h\in
   G_{i}}r_{h}.$ In case $G_{i}$ is empty, we define $q_{i}=1$ and
@@
 For each variable $X_{i}$ there is a $q_{i}\times r_{i}$ table
-$\theta_{i}$ of parameters whose $j^{\textnormal{th}}$ row and
-$k^{\textnormal{th}}$ column defines the conditional probability
-$P(X_{i}=k\mid G_{i}(X)=j;\theta)=\theta_{ijk}$.  With structure $G$
-and parameters $\theta$, we can now express the likelihood function of
-the model as
+$\theta_{i}$ of parameters whose $k^{\textnormal{th}}$ column
+on the $j^{\textnormal{th}}$ row $\theta_{ij}$ defines the conditional
+probability $P(X_{i}=k\mid G_{i}(X)=j;\theta)=\theta_{ijk}$.  With
+structure $G$ and parameters $\theta$, we can now express the
+likelihood function of the model as
 \begin{equation}
@@
 posterior probability $P(G|D,\alpha)$, where $\alpha$ denotes the
-hyperparameters for the model parameters $\theta$, and
-the $D$ is a collection of $N$ $n$\nobreakdash-dimensional i.i.d. data
-vectors collected to a $N\times n$ design matrix.  We use structure
-$G$ to extend the common matrix indexing notation $D[i,j]$, and allow
-more general row selectors.  Notably the notation $D[G_i=j,i]$ is used
-to denote those rows of the column $i$ in which the parents $G_i$
-contain the value configuration number $j\in\{1,\ldots,q_i\}$.
+hyperparameters for the model parameters $\theta$, and the $D$ is a
+collection of $N$ $n$\nobreakdash-dimensional i.i.d. data vectors
+collected to a $N\times n$ design matrix. We use the notation $D_i$ to
+denote the $i^\text{th}$ column of the data matrix and notation $D_V$ to denote
+the columns that correspond to the variable subset $V$. We also write
+$D_{i,G_i}$ for $D_{\{i\}\cup G_i}$ and denote the entries of the
+column $i$ on the rows on which the parents $G_i$ contain the value
+configuration number by $D_{i,G_i=j}$, $j\in\{1,\ldots,q_i\}$.
 
@@
 \lefteqn{P(D|G,\alpha)}\nonumber\\
-&&=\prod_{i=1}^{n}\prod_{j=1}^{q_i}P(D[G_i=j,i]|\alpha)\nonumber\\
-&&=\prod_{i=1}^{n}\prod_{j=1}^{q_i}\int P([G_i=j,i]|\theta)P(\theta|\alpha) d\theta.
+&&=\prod_{i=1}^{n}\prod_{j=1}^{q_i}P(D_{i,G_i=j};\alpha)\nonumber\\
+&&=\prod_{i=1}^{n}\prod_{j=1}^{q_i}\int P(D_{i,G_i=j}|\theta_{ij})P(\theta_{ij};\alpha) d\theta_{ij}.
 \end{eqnarray}
-In coding terms this means that each data column $D[\cdot,i]$ is first partitioned based on the
-values in columns $G_i$, and each part is then coded using a Bayesian mixture that can be expressed
-in a closed form~\cite{Bunt91, Heck95}.
+In coding terms this means that each data column $D_i$ is first
+partitioned based on the values in columns $G_i$, and each part is
+then coded using a Bayesian mixture that can be expressed in a closed
+form~\cite{Bunt91, Heck95}.
+
+\begin{figure}
+\centering
+\includegraphics[width=8cm,height=5cm]{qNML_images/breast_cancer_npmean.pdf}
+\caption{Number of parameters in a breast cancer model as a function
+  of sample size for different model selection criteria.}
+\label{fig:bcnpmean}
+\end{figure}
 
@@
 \begin{equation}
-P^1_{NML}(D[G_i=j,i];G)=\frac{P(D|\hat\theta(D[G_i=j,i];G))}{\sum_{D'}{P(D';\hat\theta(D';G))}},
+P^1_{NML}(D_{i,G_i=j};G)=\frac{P(D|\hat\theta(D_{i,G_i=j};G))}{\sum_{D'}{P(D';\hat\theta(D';G))}},
 \end{equation}
-where $D'\in{\{1,\ldots,r_i\}}^{|D[G_i=j,i]|}$.  The logarithm of the
+where $D'\in{\{1,\ldots,r_i\}}^{|D_{i,G_i=j}|}$.  The logarithm of the
 denominator is often called a regret, since it indicates the extra
@@
 where $\alpha = \frac{r}{N}$ and
-\begin{equation}
-    C_\alpha = \frac{1}{2} + \frac{1}{2} \sqrt{1 + \frac{4}{\alpha}}.
-\end{equation}
+%\begin{equation}
+    $C_\alpha = \frac{1}{2} + \frac{1}{2} \sqrt{1 + \frac{4}{\alpha}}$.
+%\end{equation}
 These approximations are compared in Table \ref{tbl:regrets} to the
@@
 motivations for this work.
-
-\begin{figure}
-\centering
-\includegraphics[width=8cm,height=5cm]{qNML_images/breast_cancer_npmean.pdf}
-\caption{Number of parameters in a breast cancer model as a function
-  of sample size for different model selection criteria.}
-\label{fig:bcnpmean}
-\end{figure}
 

### 2017-04-01 experiments.tex
last tunings

@@
 parameters), we generated 100 data sets of 1000 data vectors, and
-studied how different scoring criteria ranked the structure of the
+studied how different scoring criteria ranked, on average, the structure of the
 generating network among all the 5-node networks as a function of
@@
 relatively sparse networks (4 arcs) and for dense ones (7 arcs). The
-ranking ability of the qNML criterion appears to be between fNML and BDeu
-in 3 out of 4 settings which is good considering that one of these
-criteria is always the winner. For dense networks with the BDeu prior,
-qNML appears to perform worse than BDeu and fNML but still much
-better than BIC.
+ranking ability of the qNML criterion appears to be between
+fNML and BDeu in 3 out of 4 settings which is good considering that
+one of these criteria is always the winner. For dense networks with
+the BDeu prior, qNML appears to perform worse than BDeu and fNML but
+still much better than BIC.
 
@@
 
-Table~\ref{tbl:preds} features predictive losses for the small sample
-sizes.  (The results for large sample sizes converge for all sensible
-criteria.) BIC's bias for simplicity makes it often win (written bold
+Table~\ref{tbl:preds} features the average predictive losses for different
+data sets.  BIC's bias for simplicity makes it often win (written bold
 in the table) with small sample sizes, but it performs worst (in
@@
 
-Looking at the number of parents for our 20 data sets and sample
-sizes (Table~\ref{tbl:nofparams}) again features BIC's preference for
-simple models. qNML usually (18/20) beats fNML that performs worst in
-14 out of 20 data sets.
+Looking at the number of parents for the same 20 data sets again
+features BIC's preference for simple models
+(Table~\ref{tbl:nofparams}).  qNML usually (18/20) yields more
+parsimonious models than fNML that selects the most complex model for 14
+out of 20 data sets.
 

### 2017-04-01 qnml.tex
last tunings

@@
 s^{qNML}(D;G) & := & \sum_{i=1}^n s^{qNML}_i(D;G)\\
-& := & \sum_{i=1}^n \log \frac{P^1_{NML}(D[\cdot,(i,G_i)];G)}
-                             {P^1_{NML}(D[\cdot,G_i];G)}.\nonumber
+& := & \sum_{i=1}^n \log \frac{P^1_{NML}(D_{i,G_i};G)}
+                             {P^1_{NML}(D_{G_i};G)}.\nonumber
 \end{eqnarray}
@@
   lighten up the notation and write
-  $P^1_{NML}(i,G_i) := P^1_{NML}(D[\cdot,(i,G_i)];G)$ and
-  $P^1_{NML}(G_i)   := P^1_{NML}(D[\cdot,G_i];G)$.
+  $P^1_{NML}(i,G_i) := P^1_{NML}(D_{i,G_i};G)$ and
+  $P^1_{NML}(G_i)   := P^1_{NML}(D_{G_i};G)$.
   \begin{eqnarray}
@@
 \end{equation}
-where $D_i = D[\cdot, i]$ stands for the data on variable $i$,
-$\hat{\theta}_{i | G_i}$ denotes the maximum likelihood parameters of
+where $\hat{\theta}_{i | G_i}$ denotes the maximum likelihood parameters of
 the conditional distribution of variable $i$ given its parents in
@@
 score equivalence of the NML criterion and the qNML criterion, we may
-pick a tournament $G$ such that the linear ordering defined by $G$ matches
-the ordering of the data columns, i.e., $i<j$ implies $G_i \subset
-G_j$. Now from the definition~(\ref{eqn:qnmldef}) of the qNML
+pick a tournament $G$ such that the linear ordering defined by $G$
+matches the ordering of the data columns, i.e., $i<j$ implies $G_i
+\subset G_j$. Now from the definition~(\ref{eqn:qnmldef}) of the qNML
 criterion we see that for the tournament $G$, the sum telescopes
-leaving us with $s^{qNML}(D;G) = \log P^1_{NML}(D[\cdot,G];G)$, thus
+leaving us with $s^{qNML}(D;G) = \log P^1_{NML}(D_G;G)$, thus
 it is enough to show that $P^1_{NML}(D;G)=P_{NML}(D;G)$.  This
@@
 vector $x$ by $x^{:i}$, and the number of times such a prefix appears
-in a $N\times n$ data matrix $D=[d_1,\ldots,d_N]$ by $N_D(x^{:i})$,
-(so that $N_D(x^{:0})=N)$, we have
+on $N$ rows $[d_1,\ldots,d_N]$ of the $N\times n$ data matrix $D$ by
+$N_D(x^{:i})$, (so that $N_D(x^{:0})=N)$, we have
 \begin{eqnarray}

### 2017-09-26 regularity.tex
draft for regularity proof

\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{bm}
\usepackage[round]{natbib}
\usepackage{bm}
\newcommand{\Le}{\left(}
\newcommand{\Ri}{\right)}
\newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{lemma}{Lemma}

\begin{document}

\section{Problem}

The score function $Q_n(X \mid Y)$, where $X$ refers to one random variable and $Y$ is some proposed parent set for that variable, is \textit{regular} if 
$$H_n(X \mid U) \leq H_n(X \mid U \cup V) \quad \Rightarrow \quad  Q_n(X\mid U) \geq Q_n(X \mid U \cup V), $$ 
where $H_n(\cdot\mid \cdot)$ refers to conditional empirical entropy and $U$ and $V$ are some sets of random variables.

Let $X$ be a categorical random variable with $r$ possible values. Let $U$ denote a possible parent set with $q$ different combinations of values for the variables and $V$ a parent set with $m$ different configurations. Assume that we have observed $n$ samples of $(X,U,V)$ and $H_n(X \mid U) \leq H_n(X \mid U \cup V)$ holds.

We want to show that
$$
\quad  Q^{qnml}_n(X\mid U) \geq Q^{qnml}_n(X \mid U \cup V).
$$
The assumption about entropy implies that the maximized likelihood terms of the qnml-score are equal. In order to prove the claim it suffices to study the penalty terms, and we want to show that
\begin{align*}
-[reg(n,rq) - reg(n,q) ] \ &\geq \  -[reg(n,rqm) - reg(n,qm) ] \\
reg(n,rq) - reg(n,q) \ &\leq \  reg(n,rqm) - reg(n,qm) \\
 \frac{C(n,rq)}{C(n,q)} \ &\leq \frac{C(n,rqm)}{C(n,qm)},
\end{align*} where $C(n,k)$ is the stochastic complexity for multinomial random variable with $k$ categories. Since, trivially, $q \leq qm$, we can prove the claim by showing that the function $k \mapsto C(n,rk)/C(n,k)$ is increasing for arbitrary $r \geq 2$ and $n \geq 1$.

\section{Stochastic complexity}
We can represent $C(n,k)$ via recursion $$C(n,k) = C(n,k-1) + C(n,k-2)n/(k-2)$$
or by using the following formula
\begin{align*}
C(n,k) &= \sum_{l=0}^n\frac{n^{\underline{l} } \ (k-1)^{\overline{l}}}{n^l \ l!} \\
&= \sum_{l=0}^{n-1}\frac{(n-1)^{\underline{l}}k^{\overline{l + 1}}}{n^{l+1} \ l!} 
\end{align*}where $x^{\underline{l}}$ and $x^{\overline{l}}$ denote falling and rising factorials, respectively.

\section{Case $n = 2$}
\begin{align*}
\frac{C(2,rk)}{C(2,k)} &= \frac{1 + (rk-1) + (rk-1)rk/4}{1 + (k-1) + (k-1)k/4}
= \frac{r^2k^2 + 3rk}{k^2 + 3k},
\end{align*}and since
$$
\frac{d}{dk} \Le\frac{C(2,rk)}{C(2,k)} \Ri =\frac{3r(r-1)k^2}{(k^2+3k)^2} = \frac{3r(r-1)}{(k+3)^2} > 0
$$for every $r \geq 2$ and $k \geq 1$, the claim holds.

\section{Case $n = 3$}
\begin{align*}
\frac{C(3,rk)}{C(3,k)} &= \frac{r^3k^2 + 9r^2k + 17r}{k^2 + 9k + 17} \textnormal{ and } \\ \frac{d}{dk} \Le\frac{C(3,rk)}{C(3,k)}\Ri &= \frac{(r-1)r(153 + 34 (r+1) k + 9 r k^2)}{(17 + 9 k + k^2)^2} > 0
\end{align*}

\section{Case $n = 4$}
\begin{align*}
\frac{C(4,rk)}{C(4,k)} = \frac{142 r + 95 r^2 k + 18 r^3 k^2 + r^4 k^3}{142 + 95 k + 18 k^2 + k^3}
\end{align*}

It seems that the denominator of the derivative contains "always" a polynomial of $k$ with positive coefficients, and since the nominator is some polynomial squared, the resulting quotient is always positive for arbitrary $r$.  
%\section{Case general}
%Using the first sum form from Sec. 2 and denoting $b_l = \frac{n^{\underline{l}}}{n^l l!}$, we have
%\begin{align*}
%\frac{C(n,rk)}{C(n,k)} = \frac{\sum_{l=0}^n b_l(rk-1)^{\overline{l}} } {\sum_{l=0}^n b_l(k-1)^{\overline{l}}}.
%\end{align*}
%
%Differentiating the denominator and nominator  w.r.t. $k$ gives
%\begin{align*}
%\frac{d}{dk}\Le\sum_{l=0}^n b_l(rk-1)^{\overline{l}}\Ri &=  \sum_{l=0}^nr b_l[\psi(rk-i+l)-\psi(rk-1)](rk-1)^{\overline{l}} \\
%&= r  \sum_{l=1}^n b_l\Le\sum_{i=0}^{l-1}\frac{(rk-1)^{\overline{l}}}{rk-1+i}\Ri
%\end{align*}and
%\begin{align*}
%\frac{d}{dk}\Le\sum_{l=0}^n b_l(k-1)^{\overline{l}}\Ri= \sum_{l=1}^n b_l\Le\sum_{i=0}^{l-1}\frac{(k-1)^{\overline{l}}}{k-1+i}\Ri
%\end{align*}
%Differentiating the quotient results in the following expression
%\begin{align*}
%\frac{d}{dk}\Le\frac{C(n,rk)}{C(n,k)}\Ri &= \\
%&r\Le   \sum_{l=1}^n b_l\Le\sum_{i=0}^{l-1}\frac{(rk-1)^{\overline{l}}}{rk-1+i}\Ri \Ri \Le \sum_{l=0}^n b_l(k-1)^{\overline{l}} \Ri \\ 
%-& \Le \sum_{l=1}^n b_l\Le\sum_{i=0}^{l-1}\frac{(k-1)^{\overline{l}}}{k-1+i}\Ri \Ri \Le \sum_{l=0}^n b_l(rk-1)^{\overline{l}}  \Ri 
%\end{align*}Hard to see anything from this..
\section{Case general}
We first derive a representation for $C(n,k)$ as a polynomial of $k$. We utilize the fact that the rising factorial can be represented as polynomial using unsigned Stirling numbers of the first kind
\begin{align*}
C(n,k) &=  \sum_{l=0}^{n-1}\frac{(n-1)^{\underline{l}}k^{\overline{l + 1}}}{n^{l+1} \ l!} \\
&= \sum_{l=0}^{n-1} b_l \ k^{\overline{l+1}} \\
&= \sum_{l=0}^{n-1} b_l \Le \sum_{j=1}^{l+1}|s(l + 1,j)| \ k^j \Ri \\
&= \sum_{l=0}^{n-1} \Le \sum_{j=1}^{n}b_l \ |s(l + 1,j)| \ k^j \Ri \\
&= \sum_{j=1}^{n} \Le \sum_{l=0}^{n-1} b_l \ |s(l + 1,j)| \ k^j \Ri \\
&= \sum_{j=1}^{n} \Le \sum_{l=0}^{n-1} b_l \ |s(l + 1,j)| \Ri k^j \\
&= \sum_{j=1}^{n} a_j  k^j,
\end{align*}where $s(x,y)$ denotes the Stirling number of the first kind and
$$
a_j = \Le \sum_{l=0}^{n-1} \frac{(n-1)^{\underline{l}}}{n^{l+1}l!} \ |s(l + 1,j)| \Ri,
$$ $a_j \geq 0$ for all $j$ (also it seems that $\sum_{j=1}^n a_j = 1 $). On the row 4, we used the property of Stirling numbers: $s(i,j) = 0$ for all $j > i$. Similarly,
$$
C(n,rk) = \sum_{j=1}^{n} a_j  r^jk^j
$$Derivatives are obtained easily from this form
\begin{align*}
\frac{d}{dk}C(n,k) &= \sum_{j=1}^{n}ja_jk^{j-1} \\
&= \sum_{j=0}^{n-1}(j+1) a_{j+1}k^{j}
\end{align*}and
\begin{align*}
\frac{d}{dk}C(n,rk) &= \sum_{j=1}^{n}ja_jr^jk^{j-1} \\
&= \sum_{j=0}^{n-1}(j+1)a_{j+1}r^{j+1}k^{j}. 
\end{align*}Consider next the products found in the derivative of the quotient. We obtain
\begin{align*}
\Le\frac{d}{dk}C(n,rk)\Ri C(n,k) &= \Le \sum_{j=0}^{n-1}(j+1)a_{j+1}r^{j+1}k^{j} \Ri \Le \sum_{l=1}^{n} a_l  k^l \Ri \\
&= \sum_{i = 1}^{2n-1} \Le \sum_{j+l = i} (j+1)a_{j+1}r^{j+1}a_l  \Ri k^i
\end{align*}and
%\begin{align*}
%\Le\frac{d}{dk}C(n,k)\Ri C(n,rk) &= \Le \sum_{j=0}^{n-1}(j+1)a_{j+1}k^{j} \Ri \Le %\sum_{l=1}^{n} a_l  r^lk^l  \Ri \\
%&= \sum_{i = 1}^{2n-1} \Le \sum_{m=0}^i (m+1)a_{m+1}a_{i-m}r^{i-m}   \Ri k^i
%\end{align*}
\begin{align*}
\Le\frac{d}{dk}C(n,k)\Ri C(n,rk) &= \Le \sum_{j=0}^{n-1}(j+1)a_{j+1}k^{j} \Ri \Le \sum_{l=1}^{n} a_l  r^lk^l  \Ri \\
&= \sum_{i = 1}^{2n-1} \Le \sum_{j+l = i}(j+1)a_{j+1}a_l  r^l   \Ri k^i.
\end{align*}
Subtracting these two expression yields
\begin{align*}
&\Le\frac{d}{dk}C(n,rk)\Ri C(n,k)-\Le\frac{d}{dk}C(n,k)\Ri C(n,rk) \\&= \sum_{i = 1}^{2n-1} \Le \sum_{j+l = i} (j+1)a_{j+1}r^{j+1}a_l  \Ri k^i - \sum_{i = 1}^{2n-1} \Le \sum_{j+l = i}(j+1)a_{j+1}a_l  r^l   \Ri k^i \\
&= \sum_{i = 1}^{2n-1} \Le \sum_{j+l = i}(j+1)a_{j+1}a_l  (r^{j+1}- r^l)   \Ri k^i
\end{align*}which is the polynomial in the denominator of the derivative of $C(n,rk)/C(n,k)$.
Next, we study the coefficient of $k^i$, if $i \leq n$
\begin{align*}
\sum_{j+l = i}(j+1)a_{j+1}a_l  (r^{j+1}- r^l) &= \sum_{l=1}^{i}(i-l+1)a_{i-l+1}a_l  (r^{i-l+1}- r^l) \\
& = \sum_{l=1}^{i}(i-l+1)c_l \\
&= \sum_{k =1}^{\lfloor i / 2 \rfloor}(i-k+1)c_k + (i-(i-k + 1)+1)c_{i-k+1} \\
&=  \sum_{k =1}^{\lfloor i / 2 \rfloor}(i-k+1)c_k + kc_{i-k+1} \\
&=  \sum_{k =1}^{\lfloor i / 2 \rfloor}(i-k+1)c_k - kc_{k} \\
&= \sum_{k =1}^{\lfloor i / 2 \rfloor}(i-2k+1)c_k.
\end{align*} On the first row, we re-wrote sum using only one running index. On the second row we denoted $c_l =a_{i-l+1}a_l  (r^{i-l+1}- r^l)$. On the third row, we re-arranged the sum so that we are summing over pairs of terms of the original sum: the first and the last term, the second and the second to last, and so on.  This resulting sum has $\lfloor i / 2 \rfloor$ terms. We have to use the floor-function since if $i$ is odd, there exists an index $l'$ in the original sum such that $r^{i-l'+1}-r^{l'} = 0$. On the fifth row, we make use of the identity $c_k = -c_{i-k+1}$ which is straightforward to verify. From the last row, we can observe that every term of the sum is positive since $i-2k+1$ and $r^{i-k+1}- r^k$ are both positive if $k \leq (i+1)/2$ which holds since $k$ ranges from $1$ to $\lfloor i / 2 \rfloor$.

Let us now consider the situation where $n < i \leq 2n-1$. We start with the special case where $i = 2n-1$. Then, we have only one term in the sum
\begin{align*}
\sum_{j+l = i}(j+1)a_{j+1}a_l  (r^{j+1}- r^l) &= \sum_{l=n}^{n}(2n-1-l+1)a_{2n-1-l+1}a_l  (r^{2n-1-l+1}- r^l) \\
&= na_{n}a_n  (r^{n}- r^n) \\
& = 0.
\end{align*}Now, let $n < i < 2n-1$, we follow a similar procedure as before to manipulate the sum
\begin{align*}
\sum_{j+l = i}(j+1)a_{j+1}a_l  (r^{j+1}- r^l) &= \sum_{l=i-n + 1}^{n}(i-l+1)a_{i-l+1}a_l  (r^{i-l+1}- r^l) \\
& = \sum_{l=i-n + 1}^{n}(i-l+1)c_l \\
& = \sum_{k=1}^{2n-i}(n-k+1)c_{i-n+k} \\
& = \sum_{k=1}^{\lfloor n - i/2 \rfloor}(n-k+1)c_{i-n+k}\\
&+  (n-(2n-i-k+1)+1)c_{i-n+(2n-i-k+1)} \\
&= \sum_{k=1}^{\lfloor n - i/2 \rfloor}(n-k+1)c_{i-n+k} +  (i-n+k)c_{n-k+1} \\
&= \sum_{k=1}^{\lfloor n - i/2 \rfloor}(n-k+1)c_{i-n+k} -  (i-n+k)c_{i-n+k} \\
&= \sum_{k=1}^{\lfloor n - i/2 \rfloor}(n-k+1-(i-n+k))c_{i-n+k} \\
&= \sum_{k=1}^{\lfloor n - i/2 \rfloor}(2n-i-2k+1))c_{i-n+k}.
\end{align*}It is now easy to verify that $(2n-i-2k+1)$ and $c_{i-n+k}$ are positive if $k \leq n - (i+2)/2$ which holds since $k$ ranges from $1$ to $\lfloor n - i/2 \rfloor$. The floor function is again used when we sum over pairs of terms since if $i$ is odd there is zero-term. 
Since all the coefficients are non-negative and the $k \geq 2$, the derivative is positive. This implies that the original function is increasing.

%$$
%C(n,rk) = \sum_{l=0}^{n-1} b_l \Le \sum_{j=1}^{l+1}|s(l + 1,j)| \ (rk)^j \Ri.
%$$ Derivatives are obtained easily from this form
%$$&=  \sum_{k=i-n + 1}^{\lfloor n - i/2 \rfloor}(i-l+1)c_k + (i-(n-k + 1)+1)c_{n-k+1} 
%\frac{d}{dk}C(n,k) = \sum_{l=0}^{n-1} b_l \Le\sum_{j=1}^{l+1}j \ |s(l + 1,j)| \ k^{j-1}\Ri
%$$and
%$$
%\frac{d}{dk}C(n,rk) = \sum_{l=0}^{n-1} b_l \Le \sum_{j=1}^{l+1}j \ r^{j} \ |s(l + 1,j)| \ k^{j-1}\Ri.
%$$
\end{document}

### 2017-09-26 regularity.tex
typos

@@
 &= \sum_{k=1}^{\lfloor n - i/2 \rfloor}(n-k+1-(i-n+k))c_{i-n+k} \\
-&= \sum_{k=1}^{\lfloor n - i/2 \rfloor}(2n-i-2k+1))c_{i-n+k}.
-\end{align*}It is now easy to verify that $(2n-i-2k+1)$ and $c_{i-n+k}$ are positive if $k \leq n - (i+2)/2$ which holds since $k$ ranges from $1$ to $\lfloor n - i/2 \rfloor$.
+&= \sum_{k=1}^{\lfloor n - i/2 \rfloor}(2n-i-2k+1)c_{i-n+k}.
+\end{align*}It is now easy to verify that $(2n-i-2k+1)$ and $c_{i-n+k}$ are positive if $k \leq n - (i-1)/2$ which holds since $k$ ranges from $1$ to $\lfloor n - i/2 \rfloor$.
 The floor function is again used when we sum over pairs of terms since if $i$ is odd there is zero-term.
@@
 %$$
+\section{Terms $a_j$ sum to one}
+\begin{align*}
+\sum_{j=1}^{n}a_j &=\sum_{j=1}^{n} \Le \sum_{l=0}^{n-1} \frac{(n-1)^{\underline{l}}}{n^{l+1}l!} \ |s(l + 1,j)| \Ri \\
+&= \sum_{l=0}^{n-1} \frac{(n-1)^{\underline{l}}}{n^{l+1}l!} \Le \sum_{j=1}^{n} |s(l + 1,j)| \Ri \\
+&= \sum_{l=0}^{n-1} \frac{(n-1)^{\underline{l}}}{n^{l+1}l!} \Le \sum_{j=0}^{l-1} |s(l + 1,j)| \Ri \\
+&= \sum_{l=0}^{n-1} \frac{(n-1)^{\underline{l}}}{n^{l+1}l!} \ (l+1)! \\
+& =  \sum_{l=0}^{n-1} \frac{(n-1)^{\underline{l}}}{n^{l+1}}(l+1)
+\end{align*}
 \end{document}

### 2017-10-03 qnml.tex
subsection on regularity

@@
 \frac{P(D_{i, G_i} \ | \ \hat{\theta}_{i, G_i} )}{P(D_{G_i} \ | \ \hat{\theta}_{G_i} )} &= \frac{ \prod_{j,k} \Le \frac{N_{ijk}}{N}  \Ri^{N_{ijk}}}{\prod_{j} \Le \frac{N_{ij}}{N}  \Ri^{N_{ij}}} \\
-&= \frac{ \prod_{j,k} \Le \frac{N_{ijk}}{N}  \Ri^{N_{ijk}}}{\prod_{j} \Le \frac{N_{ij}}{N}  \Ri^{\sum_k N_{ijk}}} \\
 &= \frac{ \prod_{j,k} \Le \frac{N_{ijk}}{N}  \Ri^{N_{ijk}}}{\prod_{j}\prod_{k} \Le \frac{N_{ij}}{N}  \Ri^{N_{ijk}}}\\
@@
 For example, $a(4)=\frac{4!}{1!}+\frac{4!}{1!1!}+\frac{4!}{2!}+\frac{4!}{1!2!}+\frac{4!}{4!}=73$. In general this sequence grows rapidly; $1, 1, 3, 13, 73, 501, 4051, 37633, 394353, 4596553, \ldots$.
+
+\subsection{QNML IS REGULAR}
+
+\cite{Suzuki2017} defines regularity for a scoring function $Q_n(X \mid Y)$ as follows:
+\begin{definition}
+Assume $H_N(X \mid Y') \leq H_N(X \mid Y)$, where $Y' \subset Y.$ We say that $Q_N(\cdot \mid \cdot)$ is regular if $Q_n(X \mid Y') \geq Q_N(X \mid Y)$.
+\end{definition}
+In the definition, $N$ denotes the sample size, $X$ is some random variable, $Y$ denotes the proposed parent set for $X$, and $H_N(\cdot \mid \cdot)$ refers to the empirical conditional entropy.
+\cite{Suzuki2017} shows that BDeu violates this principle and demonstrates that this causes the score to prefer more complex networks even though data do not support this.
+In \cite{Suzuki2017_2}, the authors also suggest that regular scores are computationally more efficient when applied with branch-and-bound type algorithms for Bayesian network structure learning.
+
+By analyzing the penalty term of the qNML scoring function, one can prove the following statement:
+\begin{theorem}
+qNML score is regular.
+\end{theorem}
+\begin{proof}
+The proof is given in the Appendix A in Supplementary Material.
+\end{proof}

### 2017-10-03 qnml.tex
typoss

@@
 In the definition, $N$ denotes the sample size, $X$ is some random variable, $Y$ denotes the proposed parent set for $X$, and $H_N(\cdot \mid \cdot)$ refers to the empirical conditional entropy.
-\cite{Suzuki2017} shows that BDeu violates this principle and demonstrates that this causes the score to prefer more complex networks even though data do not support this.
+\cite{Suzuki2017} shows that BDeu violates this principle and demonstrates that this can cause the score to prefer more complex networks even though data do not support this.
 In \cite{Suzuki2017_2}, the authors also suggest that regular scores are computationally more efficient when applied with branch-and-bound type algorithms for Bayesian network structure learning.

### 2017-10-03 supp_mat.tex
supplementary material

\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{bm}
\usepackage[round]{natbib}
\usepackage{bm}
\newcommand{\Le}{\left(}
\newcommand{\Ri}{\right)}
\newcommand{\N}{\mathbb{N}}
\newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{corollary}{Corollary}
\newtheorem{lemma}{Lemma}

\date{}
\title{Supplementary Materials for the article \textit{Quotient Normalized Maximum Likelihood Criterion for Learning Bayesian Network Structures} }
\begin{document}
\maketitle

\appendix
\section{Regularity proof}

\subsection{Preliminaries}
We start by recalling the definition of regularity \citep{Suzuki2017}:

\begin{definition}
Assume $H_N(X \mid Y') \leq H_N(X \mid Y)$, where $Y' \subset Y.$ We say that the scoring function $Q_N(\cdot \mid \cdot)$ is regular if $Q_n(X \mid Y') \geq Q_N(X \mid Y)$.
\end{definition}
In the definition, $N$ denotes the sample size, $X$ is some random variable, $Y$ denotes the proposed parent set for $X$, and $H_N(X \mid Y)$ refers to the empirical conditional entropy based on $N$ samples of variables $X$ and $Y$.

Let $X$ be a categorical random variable with $r$ possible values. Let $U$ denote a possible parent set with $q$ different combinations of values for the variables, and $V$ a parent set with $m$ different configurations. Assume that we have observed $N$ samples of $(X,U,V)$ (denoted by $x_N,u_N$ and $v_N$) and $H_N(X \mid U) \leq H_N(X \mid U \cup V)$ holds.

Recall the definition of the qNML score:
\begin{align*}
Q^{qnml}_N(X\mid U) &= \log P(x_n \mid \hat{\theta}_{X\mid U} \ ) - \Le reg(N,rq) - reg(N,q) \Ri \\
&= \log P(x_N \mid \hat{\theta}_{X\mid U} \ ) - \log \frac{C(N,rq)}{C(N,q)},
\end{align*}where $C(N,r)$ is the normalizing constant the of the NML distribution for a categorical variable with $r$ possible values and sample size $N$ and $ \hat{\theta}_{X\mid U}$ denotes the maximum likelihood parameters of the conditional distribution of $X$ given $U$ which are computed from the data $(x_N,u_N)$.

In order to prove the regularity, we need the following three lemmas:
\begin{lemma}\label{Cpolynom} We can write $C(N,k)$ as a polynomial of $k$, formally
$$
C(N,k) = \sum_{j=1}^N a_j k^j,
$$where $a_j > 0$. 
\end{lemma}
\begin{lemma}\label{MLterms}
Assume $H_N(X \mid Y') \leq H_N(X \mid Y)$, where $Y' \subset Y$. Now $\log P(x_N \mid \hat{\theta}_{X\mid Y} \ ) = \log P(x_N \mid \hat{\theta}_{X\mid Y'} \ )$.
\end{lemma}
\begin{lemma}\label{increasing}
Let $r \in \N, r \geq 2$.  The function $k \mapsto \frac{C(N,rk)}{C(N,k)}$ is increasing for every $k \geq 2$.
\end{lemma}
\noindent We present the proofs of these lemmas in Section \ref{lemmaproofs}.

\subsection{The main proof}
\begin{theorem}
qNML score is regular.
\end{theorem}
\begin{proof}
We want to show that
$$
Q^{qnml}_N(X\mid U) \geq Q^{qnml}_N(X \mid U \cup V).
$$assuming $H_N(X \mid U) \leq H_N(X \mid U \cup V)$. Using the entropy assumption and Lemma \ref{MLterms} implies that the maximized likelihood terms are equal. In order to prove the claim, it suffices to study the penalty terms, and we want to show that
\begin{align*}
- \Le reg(N,rq) - reg(N,q) \Ri \ &\geq \  -\Le reg(N,rqm) - reg(N,qm) \Ri \\
\log \frac{C(N,rq)}{C(N,q)} \ &\leq \log \frac{C(N,rqm)}{C(N,qm)}.
\end{align*}This holds, since logarithm is an increasing function, and $q \leq qm$, so we can apply Lemma \ref{increasing} to conclude the proof.

\end{proof}

\subsection{Proofs of lemmas}\label{lemmaproofs}
%We want to show that
%$$
%\quad  Q^{qnml}_N(X\mid U) \geq Q^{qnml}_N(X \mid U \cup V)
%$$which, by using the definition of qNML, is equivalent to 
%\begin{equation}\label{statement}
%\log P(X \mid )
%\end{equation}
%The assumption about entropy implies that the maximized likelihood terms of the qnml-score are equal. 
\setcounter{lemma}{0}
\begin{lemma} $C(N,k)$ can be written as a polynomial of $k$, formally
$$
C(N,k) = \sum_{j=1}^N a_j k^j,
$$where $a_j > 0$. 
\end{lemma}
\begin{proof} \cite{cosco.itsl08} derive the following representation for the normalizing constant
\begin{align*}  
C(N,k) = \sum_{l=0}^{N-1}\frac{(N-1)^{\underline{l}}k^{\overline{l + 1}}}{N^{l+1} \ l!} ,
\end{align*}where $x^{\underline{l}}$ and $x^{\overline{l}}$ denote falling and rising factorials, respectively.

We utilize the fact that the rising factorial can be represented as polynomial using unsigned Stirling numbers of the first kind (see \cite{Adamchik1997}, for instance)
\begin{align*}
C(N,k) &=  \sum_{l=0}^{N-1}\frac{(N-1)^{\underline{l}}k^{\overline{l + 1}}}{N^{l+1} \ l!} \\
&= \sum_{l=0}^{N-1} b_l \ k^{\overline{l+1}} \\
&= \sum_{l=0}^{N-1} b_l \Le \sum_{j=1}^{l+1}|s(l + 1,j)| \ k^j \Ri \\
&= \sum_{l=0}^{N-1} \Le \sum_{j=1}^{N}b_l \ |s(l + 1,j)| \ k^j \Ri \\
&= \sum_{j=1}^{N} \Le \sum_{l=0}^{N-1} b_l \ |s(l + 1,j)| \ k^j \Ri \\
&= \sum_{j=1}^{N} \Le \sum_{l=0}^{N-1} b_l \ |s(l + 1,j)| \Ri k^j \\
&= \sum_{j=1}^{N} a_j  k^j,
\end{align*}where $s(i,j)$ denotes the (signed) Stirling number of the first kind and
$$
a_j = \Le \sum_{l=0}^{N-1} \frac{(N-1)^{\underline{l}}}{N^{l+1}l!} \ |s(l + 1,j)| \Ri,
$$ $a_j > 0$ for all $j$. On the row 4, we used the property of Stirling numbers: $s(i,j) = 0$ for all $j > i$. 
\end{proof}
\begin{lemma}
Assume $H_N(X \mid Y') \leq H_N(X \mid Y)$, where $Y' \subset Y$. Now $\log P(x_N \mid \hat{\theta}_{X\mid Y} \ ) = \log P(x_N \mid \hat{\theta}_{X\mid Y'})$.
\end{lemma}
\begin{proof}
We can write the logarithm of the maximized likelihood, \\ $\log P(x_N \mid \hat{\theta}_{X\mid Y} \ )$, as follows \citep{KOLLER}
\begin{align*}
\log P(x_N \mid \hat{\theta}_{X\mid Y} \ ) & = -N \Le H_N(X) - I_N(X;Y) \Ri \\
&= -N \  H_N(X\mid Y),
\end{align*}where $I_N(\cdot;\cdot)$ is the empirical mutual information. This implies that the assumption
$$
H_N(X \mid Y') \leq H_N(X \mid Y)
$$ is equivalent to
$$
\log P(x_N \mid \hat{\theta}_{X\mid Y} \ ) \geq \log P(x_N \mid \hat{\theta}_{X\mid Y'} \ ) .
$$Actually we must have the equality holding in the above expression, since
$$
H_N(X \mid Y') < H_N(X \mid Y)
$$ would imply that
$$
I_N(X ; Z \mid Y ) < 0,
$$where $Z = Y \setminus Y'$, which is impossible.
\end{proof}
\begin{lemma}
Let $r \in \N, r \geq 2$.  The function $k \mapsto \frac{C(N,rk)}{C(N,k)}$ is increasing for every $k \geq 2$.
\end{lemma}

\begin{proof}
Lemma \ref{Cpolynom} lets us to write
\begin{equation}\label{C1}
C(N,k) = \sum_{j=1}^{N} a_j k^j
\end{equation}and, similarly,
\begin{equation}\label{C2}
C(N,rk) = \sum_{j=1}^{N} a_j  r^jk^j.
\end{equation}

From this, it it easy see that the derivative of the quotient, \\ $d/dk (C(N,rk)/C(N,k))$, will be a ratio of two polynomials of $k$. Our goal is to show that the polynomial in the numerator has positive coefficients, which will guarantee the positivity of derivative for every $k > 0$, and thus imply that the original function is increasing (polynomial in the denominator is squared and non-zero for $k > 0$, so it can be ignored).

Derivatives of (\ref{C1}) and (\ref{C2}) are obtained easily:
\begin{align*}
\frac{d}{dk}C(N,k) &= \sum_{j=1}^{N}ja_jk^{j-1} \\
&= \sum_{j=0}^{N-1}(j+1) a_{j+1}k^{j}
\end{align*}and
\begin{align*}
\frac{d}{dk}C(N,rk) &= \sum_{j=1}^{N}ja_jr^jk^{j-1} \\
&= \sum_{j=0}^{N-1}(j+1)a_{j+1}r^{j+1}k^{j}. 
\end{align*}Consider next the products found in the derivative of the quotient. We obtain
\begin{align*}
\Le\frac{d}{dk}C(N,rk)\Ri C(N,k) &= \Le \sum_{j=0}^{N-1}(j+1)a_{j+1}r^{j+1}k^{j} \Ri \Le \sum_{l=1}^{N} a_l  k^l \Ri \\
&= \sum_{i = 1}^{2N-1} \Le \sum_{j+l = i} (j+1)a_{j+1}r^{j+1}a_l  \Ri k^i
\end{align*}and
%\begin{align*}
%\Le\frac{d}{dk}C(n,k)\Ri C(n,rk) &= \Le \sum_{j=0}^{n-1}(j+1)a_{j+1}k^{j} \Ri \Le %\sum_{l=1}^{n} a_l  r^lk^l  \Ri \\
%&= \sum_{i = 1}^{2n-1} \Le \sum_{m=0}^i (m+1)a_{m+1}a_{i-m}r^{i-m}   \Ri k^i
%\end{align*}
\begin{align*}
\Le\frac{d}{dk}C(N,k)\Ri C(N,rk) &= \Le \sum_{j=0}^{N-1}(j+1)a_{j+1}k^{j} \Ri \Le \sum_{l=1}^{N} a_l  r^lk^l  \Ri \\
&= \sum_{i = 1}^{2N-1} \Le \sum_{j+l = i}(j+1)a_{j+1}a_l  r^l   \Ri k^i.
\end{align*}
Subtracting these two expression yields
\begin{align*}
&\Le\frac{d}{dk}C(N,rk)\Ri C(N,k)-\Le\frac{d}{dk}C(N,k)\Ri C(N,rk) \\&= \sum_{i = 1}^{2N-1} \Le \sum_{j+l = i} (j+1)a_{j+1}r^{j+1}a_l  \Ri k^i - \sum_{i = 1}^{2N-1} \Le \sum_{j+l = i}(j+1)a_{j+1}a_l  r^l   \Ri k^i \\
&= \sum_{i = 1}^{2N-1} \Le \sum_{j+l = i}(j+1)a_{j+1}a_l  (r^{j+1}- r^l)   \Ri k^i
\end{align*}which is the polynomial in the denominator of the derivative of $C(N,rk)/C(N,k)$.
Next, we study the coefficient of $k^i$, if $i \leq N$
\begin{align*}
\sum_{j+l = i}(j+1)a_{j+1}a_l  (r^{j+1}- r^l) &= \sum_{l=1}^{i}(i-l+1)a_{i-l+1}a_l  (r^{i-l+1}- r^l) \\
& = \sum_{l=1}^{i}(i-l+1)c_l \\
&= \sum_{k =1}^{\lfloor i / 2 \rfloor}(i-k+1)c_k + (i-(i-k + 1)+1)c_{i-k+1} \\
&=  \sum_{k =1}^{\lfloor i / 2 \rfloor}(i-k+1)c_k + kc_{i-k+1} \\
&=  \sum_{k =1}^{\lfloor i / 2 \rfloor}(i-k+1)c_k - kc_{k} \\
&= \sum_{k =1}^{\lfloor i / 2 \rfloor}(i-2k+1)c_k.
\end{align*} On the first row, we re-wrote sum using only one running index. On the second row we denoted $c_l =a_{i-l+1}a_l  (r^{i-l+1}- r^l)$. On the third row, we re-arranged the sum so that we are summing over pairs of terms of the original sum: the first and the last term, the second and the second to last, and so on.  This resulting sum has $\lfloor i / 2 \rfloor$ terms. We have to use the floor-function since if $i$ is odd, there exists an index $l'$ in the original sum such that $r^{i-l'+1}-r^{l'} = 0$. On the fifth row, we make use of the identity $c_k = -c_{i-k+1}$ which is straightforward to verify. From the last row, we can observe that every term of the sum is positive since $i-2k+1$ and $r^{i-k+1}- r^k$ are both positive if $k \leq (i+1)/2$ which holds since $k$ ranges from $1$ to $\lfloor i / 2 \rfloor$.

Let us now consider the situation where $n < i \leq 2N-1$. We start with the special case where $i = 2N-1$. Then, we have only one term in the sum
\begin{align*}
\sum_{j+l = i}(j+1)a_{j+1}a_l  (r^{j+1}- r^l) &= \sum_{l=N}^{N}(2N-1-l+1)a_{2N-1-l+1}a_l  (r^{2N-1-l+1}- r^l) \\
&= Na_{N}a_N  (r^{N}- r^N) \\
& = 0.
\end{align*}Now, let $N < i < 2N-1$, we follow a similar procedure as before to manipulate the sum
\begin{align*}
\sum_{j+l = i}(j+1)a_{j+1}a_l  (r^{j+1}- r^l) &= \sum_{l=i-N + 1}^{N}(i-l+1)a_{i-l+1}a_l  (r^{i-l+1}- r^l) \\
& = \sum_{l=i-N + 1}^{N}(i-l+1)c_l \\
& = \sum_{k=1}^{2N-i}(N-k+1)c_{i-N+k} \\
& = \sum_{k=1}^{\lfloor N - i/2 \rfloor}(N-k+1)c_{i-N+k}\\
&+  (N-(2N-i-k+1)+1)c_{i-N+(2N-i-k+1)} \\
&= \sum_{k=1}^{\lfloor N - i/2 \rfloor}(N-k+1)c_{i-N+k} +  (i-N+k)c_{N-k+1} \\
&= \sum_{k=1}^{\lfloor N - i/2 \rfloor}(N-k+1)c_{i-N+k} -  (i-N+k)c_{i-N+k} \\
&= \sum_{k=1}^{\lfloor N - i/2 \rfloor}(N-k+1-(i-N+k))c_{i-N+k} \\
&= \sum_{k=1}^{\lfloor N - i/2 \rfloor}(2N-i-2k+1)c_{i-N+k}.
\end{align*}It is now easy to verify that $(2N-i-2k+1)$ and $c_{i-N+k}$ are positive if $k \leq N - (i-1)/2$ which holds since $k$ ranges from $1$ to $\lfloor N - i/2 \rfloor$. The floor function is again used when we sum over pairs of terms since if $i$ is odd there is zero-term. 
Since all the coefficients are non-negative and the $k \geq 2$, the derivative is positive. This implies that the original function is increasing.
\end{proof}
\bibliographystyle{apalike}
\bibliography{cosco}
\end{document}

### 2017-10-03 regularity.tex
some fixes the initial version

@@
 \end{align*}
+\section{ML terms}
+Assume we have $n$ samples of $X,Y$ and $Z$. We can write the logarithm of the maximized likelihood, $\hat{P}_n(X \mid Y)$, as follows
+\begin{align*}
+\log \hat{P}_n(X \mid Y) & = -n \Le H_n(X) - I_n(X;Y) \Ri \\
+&= -n \  H_n(X\mid Y),
+\end{align*}where $I_n(\cdot;\cdot)$ is the empirical mutual information. This implies that the assumption
+$$
+H_n(X \mid Y) \leq H_n(X \mid Y \cup Z)
+$$ is equivalent to
+$$
+\log \hat{P}_n(X \mid Y) \geq \log \hat{P}_n(X \mid Y \cup Z). 
+$$Actually we must have the equality holding in the above expression, since
+$$
+H_n(X \mid Y) < H_n(X \mid Y \cup Z)
+$$ would imply that
+$$
+I_n(X ; Z \mid Y ) < 0
+$$which is impossible.
 \end{document}

### 2017-10-11 experiments.tex
shd experiment added

@@
-\begin{figure}[h]
-\centering
-\includegraphics[width=8cm,height=5cm]{qNML_images/art4_mean.pdf}
-\caption{Finding generating models of 4 arcs.}
-\label{fig:4arcs}
-\end{figure}
-
+%\begin{figure}[h]
+%\centering
+%\includegraphics[width=8cm,height=5cm]{qNML_images/art4_mean.pdf}
+%\caption{Finding generating models of 4 arcs.}
+%\label{fig:4arcs}
+%\end{figure}
 \section{EXPERIMENTAL RESULTS}
@@
 
-We first compare the ability of the different scoring criteria to
-discover the structure that generated the data. For this purpose we
-create 100 random 5-node Bayesian network structures with 4
-edges and another 100 structures with 7 edges.  The variables were
-randomly assigned to have 2 -- 4 values ($r_i \in \{2, 3, 4\}$). For
-each network, we generated parameters by two different schemes. The
-first scheme exactly matched the assumptions of the BDeu score with
-$\alpha = 1$, i.e., the parameters were distributed by $\theta_{ij}
-\sim Dir(\frac{1}{r_iq_i},\ldots,\frac{1}{r_iq_i})$. The other scheme
-was to generate the parameters independently from a Dirichlet
-distribution $\theta_{ij} \sim Dir(\frac{1}{2},\ldots, \frac{1}{2})$
-that is the Jeffreys' prior for the multinomial model. This distribution
-was selected instead of the uniform distribution in order to make the
-generating structure more identifiable.  For each network (structure +
-parameters), we generated 100 data sets of 1000 data vectors, and
-studied how different scoring criteria ranked, on average, the structure of the
-generating network among all the 5-node networks as a function of
-(sub)sample size.
+In our first experiment, we took five widely used benchmark Bayesian networks\footnote{Bayesian Network Repository: \url{http://www.bnlearn.com/bnrepository/}} (Asia, Cancer, Earthquake, Sachs and Survey), sampled data from them, and tried to learn the generating structure with the different scoring functions using various sample sizes.
+These networks were picked in order to use the dynamic programming based exact structure learning~\cite{cosco.uai06} which limited the number $n$ of variables
+to less than 20. We measured the quality of the learned structures using structural Hamming distance (SHD, \cite{Tsamardinos2006}).
+
+Figure \ref{fig:all_shd} shows SHDs for all the scoring criteria for each network.
+Sample sizes range from $10$ to $10000$ and the shown results are averages computed from $1000$ repetitions.
+None of the scores dominates in all settings considered.
+BIC fares well when the sample size is low, as it tends to produce nearly empty graph which is a good answer in terms of SHD when the generating networks are relatively sparse.
+qNML obtains strong results in 'Earthquake' and 'Asia' networks, being the best or the second to best with all the sample sizes considered.
 
@@
 \centering
-\includegraphics[width=8cm,height=5cm]{qNML_images/art7_mean.pdf}
-\caption{Finding generating models of 7 arcs.}
-\label{fig:7arcs}
+\includegraphics[width=\columnwidth]{qNML_images/shd_all.pdf}
+\caption{Sample size versus SHD with data generated from real world DAGs.}
+\label{fig:all_shd}
 \end{figure}
 
-Not surprisingly, the results indicate that when the parameter generation
-mechanism matches the assumptions of the BDeu-score, BDeu usually
-also ranks the generating structure higher than the other scores (see Figure
-~\ref{fig:4arcs} and \ref{fig:7arcs}).  On the other hand, when the parameters are drawn from the
-Jeffreys' prior, the fNML score appears to rank the generating structure
-highest. This too is not surprising, since the NML distribution for
-multinomial data is known to closely match the marginal distribution of
-the data when the Jeffreys' prior is used.
+Figure \ref{fig:shd_ranks} summarizes the SHD results in all networks by showing the average rank for each score.
+The ranking was done by giving the score with the least SHD rank $1$ and the worst one rank $4$.
+The shown results are averages computed from $5000$ values ($5$ networks, $1000$ repetitions).
+From this, we can see that qNML has never the worst average ranking, and it has the best ranking with sample sizes greater than $300$.
+This suggests that qNML is overall a safe choice in structure learning, especially with moderate sample sizes.
 
-The underfitting tendency of BIC can also be clearly detected both for
-relatively sparse networks (4 arcs) and for dense ones (7 arcs). The
-ranking ability of the qNML criterion appears to be between
-fNML and BDeu in 3 out of 4 settings which is good considering that
-one of these criteria is always the winner. For dense networks with
-the BDeu prior, qNML appears to perform worse than BDeu and fNML but
-still much better than BIC.
+%We first compare the ability of the different scoring criteria to
+%discover the structure that generated the data. For this purpose we
+%create 100 random 5-node Bayesian network structures with 4
+%edges and another 100 structures with 7 edges.  The variables were
+%randomly assigned to have 2 -- 4 values ($r_i \in \{2, 3, 4\}$). For
+%each network, we generated parameters by two different schemes. The
+%first scheme exactly matched the assumptions of the BDeu score with
+%$\alpha = 1$, i.e., the parameters were distributed by $\theta_{ij}
+%\sim Dir(\frac{1}{r_iq_i},\ldots,\frac{1}{r_iq_i})$. The other scheme
+%was to generate the parameters independently from a Dirichlet
+%distribution $\theta_{ij} \sim Dir(\frac{1}{2},\ldots, \frac{1}{2})$
+%that is the Jeffreys' prior for the multinomial model. This distribution
+%was selected instead of the uniform distribution in order to make the
+%generating structure more identifiable.  For each network (structure +
+%parameters), we generated 100 data sets of 1000 data vectors, and
+%studied how different scoring criteria ranked, on average, the structure of the
+%generating network among all the 5-node networks as a function of
+%(sub)sample size.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=8cm,height=5cm]{qNML_images/shd_rank_all.pdf}
+\caption{Average ranks for the scoring functions in structure learning experiments.}
+\label{fig:shd_ranks}
+\end{figure}
+
+%Not surprisingly, the results indicate that when the parameter generation
+%mechanism matches the assumptions of the BDeu-score, BDeu usually
+%also ranks the generating structure higher than the other scores (see Figure
+%~\ref{fig:4arcs} and \ref{fig:7arcs}).  On the other hand, when the parameters are drawn from the
+%Jeffreys' prior, the fNML score appears to rank the generating structure
+%highest. This too is not surprising, since the NML distribution for
+%multinomial data is known to closely match the marginal distribution of
+%the data when the Jeffreys' prior is used.
+%
+%The underfitting tendency of BIC can also be clearly detected both for
+%relatively sparse networks (4 arcs) and for dense ones (7 arcs). The
+%ranking ability of the qNML criterion appears to be between
+%fNML and BDeu in 3 out of 4 settings which is good considering that
+%one of these criteria is always the winner. For dense networks with
+%the BDeu prior, qNML appears to perform worse than BDeu and fNML but
+%still much better than BIC.
 
@@
 for all of them. In each experiment, random 10\% of the original
-data set was selected as training data and the rest 90\% as test data.
-The training was conducted using dynamic programming based exact structure
-learning~\cite{cosco.uai06} which limited the number $n$ of variables
-to less than 20.
+data set was selected as training data and the rest 90\% as test data. The training was conducted using the same dynamic programming based exact structure learning algorithm.
 

### 2017-10-11 main.tex
consistency proof reformulated a bit
A bit more on approximations
finetuning
definition environment added
use aistats2018 package

@@
-\documentclass[letterpaper]{article}
-\usepackage{proceed2e}
-\usepackage[margin=1in]{geometry}
-\usepackage{amsmath,amsthm,epsfig,booktabs}
+%\documentclass[letterpaper]{article}
+%\usepackage{proceed2e}
+%\usepackage[margin=1in]{geometry}
 
+\documentclass[twoside]{article}
+\usepackage{aistats2018}
+\usepackage{amsmath,amsthm,epsfig,booktabs,multirow}
+\usepackage{hyperref}
 \newtheorem{lemma}{Lemma}
 \newtheorem{theorem}{Theorem}
+\newtheorem{definition}{Definition}
+\newcommand{\Le}{\left(}
+\newcommand{\Ri}{\right)}
 
 % Set the typeface to Times Roman
-\usepackage{times}
+%\usepackage{times}
 
-\title{Quotient Normalized Maximum Likelihood Criterion for Learning Bayesian Network Structures}
+%\title{Quotient Normalized Maximum Likelihood Criterion for Learning Bayesian Network Structures}
 
@@
 
-\maketitle
+\twocolumn[
+
+\aistatstitle{Quotient Normalized Maximum Likelihood Criterion for Learning Bayesian Network Structures}
+
+\aistatsauthor{ Author 1 \And Author 2 \And  Author 3 }
+
+\aistatsaddress{ Institution 1 \And  Institution 2 \And Institution 3 } ]
+
+%\maketitle
 
@@
 
-\subsubsection*{Acknowledgements}
+%%\subsubsection*{Acknowledgements}
 
-Use unnumbered third level headings for the acknowledgements title.
-All acknowledgements go at the end of the paper.
+%%Use unnumbered third level headings for the acknowledgements title.
+%%All acknowledgements go at the end of the paper.
 

### 2017-10-11 supp_mat.tex
typos

@@
 \date{}
-\title{Supplementary Materials for the article \textit{Quotient Normalized Maximum Likelihood Criterion for Learning Bayesian Network Structures} }
+\title{Supplementary Material for the article \textit{Quotient Normalized Maximum Likelihood Criterion for Learning Bayesian Network Structures} }
 \begin{document}
@@
 a_j = \Le \sum_{l=0}^{N-1} \frac{(N-1)^{\underline{l}}}{N^{l+1}l!} \ |s(l + 1,j)| \Ri,
-$$ $a_j > 0$ for all $j$. On the row 4, we used the property of Stirling numbers: $s(i,j) = 0$ for all $j > i$. 
+$$ $a_j > 0$ for all $j$. On the second row, we denoted $b_l = (N-1)^{\underline{l}}/(N^{l+1} l!)$. On the row 4, we used the property of Stirling numbers: $s(i,j) = 0$ for all $j > i$. 
 \end{proof}
@@
 &= \sum_{i = 1}^{2N-1} \Le \sum_{j+l = i}(j+1)a_{j+1}a_l  (r^{j+1}- r^l)   \Ri k^i
-\end{align*}which is the polynomial in the denominator of the derivative of $C(N,rk)/C(N,k)$.
+\end{align*}which is the polynomial in the numerator of the derivative of $C(N,rk)/C(N,k)$.
 Next, we study the coefficient of $k^i$, if $i \leq N$

### 2017-10-11 qnml.tex
citation style

--- 2017-10-03 qnml.tex
@@
 
-\cite{Suzuki2017} defines regularity for a scoring function $Q_n(X \mid Y)$ as follows:
+Suzuki \cite{Suzuki2017} defines regularity for a scoring function $Q_n(X \mid Y)$ as follows:
 \begin{definition}
@@
 In the definition, $N$ denotes the sample size, $X$ is some random variable, $Y$ denotes the proposed parent set for $X$, and $H_N(\cdot \mid \cdot)$ refers to the empirical conditional entropy.
-\cite{Suzuki2017} shows that BDeu violates this principle and demonstrates that this causes the score to prefer more complex networks even though data do not support this.
-In \cite{Suzuki2017_2}, the authors also suggest that regular scores are computationally more efficient when applied with branch-and-bound type algorithms for Bayesian network structure learning.
+Suzuki \cite{Suzuki2017} shows that BDeu violates this principle and demonstrates that this can cause the score to prefer more complex networks even though data do not support this.
+Regular scores are also argued to be computationally more efficient when applied with branch-and-bound type algorithms for Bayesian network structure learning \cite{Suzuki2017_2}.
 

### 2017-10-11 shd_plots.py
shd results + plot script

@@ -1,0 +1,262 @@
+#!/usr/bin/env python3
+# -*- coding: utf-8 -*-
+import numpy as np
+import seaborn as sns
+import matplotlib.pyplot as plt
+
+
+def subplots(folder = "results/real_dag_shd/", save = False):
+    
+    names = ["Survey", "Earthquake", "Sachs","Cancer","Asia"]
+    
+    filee = open(folder + names[0].lower() + ".txt", mode = 'r')
+    line1 = filee.readline()
+    filee.close()
+    
+    line1 = line1.rstrip("\n")
+    header = line1.split(" ")
+    
+    methods = header.copy()
+    methods.pop(0)    
+
+    methods = ["BIC", "BDeu", "fNML", "qNML"] 
+    
+    fig,axs = plt.subplots(nrows = 3, ncols = 2,sharex = True, figsize = (10,11) )
+    
+    axs = axs.flatten()
+    linestyles = ['--', ':', '-.', '-']
+
+    jj = 0
+    
+    for dag_name in names: 
+    
+        x = np.loadtxt(folder + dag_name.lower() + ".txt", skiprows = 1)
+        
+        ns = x[:,0]
+        shds = x[:,1::]
+
+
+        sns.set_style("ticks")
+        sns.set_palette("colorblind")
+                
+        ax = axs[jj]
+
+        ax.set(xscale = "log")
+
+        for ii in range(0,len(methods)):
+            ax.plot(ns,shds[:,ii],linestyles[ii],label = methods[ii])
+
+        jj += 1
+        
+        ax.set_title(dag_name)
+        sns.set_context("poster", font_scale = 2, rc={"lines.linewidth": 7})
+        sns.despine()
+        
+        
+    ax = axs[len(names)]
+   
+    for i in range(0,len(methods)):
+        lab = methods[i]
+        ax.plot([],[],linestyles[i],label = lab)
+        
+    ax.legend(loc="center")
+    
+    ax.set_axis_off() 
+    
+    
+    for jj in range(0,len(names)):
+        ax = axs[jj]
+    
+        
+    
+    #plt.tight_layout()
+    plt.show()
+    
+    
+    # tweak the y labels
+    for jj in range(0,len(names)):
+        ax = axs[jj]
+    
+        labels = ax.get_yticklabels()
+        
+        #print(labels[2])
+        for kkk in range(0,len(labels)):
+            if kkk % 2 == 1:
+                labels[kkk].set_text("")
[…]

### 2017-10-12 supp_mat_B.tex
Moved the qNML=NML proof to appendix

--- 2017-10-03 qnml.tex
@@
-\section{QUOTIENT NML SCORE}
+\documentclass[12pt]{article}
+\usepackage{amsmath}
+\usepackage{amsthm}
+\usepackage{amssymb}
+\usepackage{bm}
+\usepackage[round]{natbib}
+\usepackage{bm}
+\newcommand{\Le}{\left(}
+\newcommand{\Ri}{\right)}
+\newcommand{\N}{\mathbb{N}}
+\newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
+\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
+\newtheorem{theorem}{Theorem}
+\newtheorem{definition}{Definition}
+\newtheorem{corollary}{Corollary}
+\newtheorem{lemma}{Lemma}
 
-We will now introduce a quotient normalized maximum likelihood (qNML)
-criterion for learning Bayesian network structures.  While equally
-efficient to compute as BDeu and fNML, it is free from
-hyperparameters, and it can be proven to give equal scores to
-equivalent models. Furthermore, it coincides with the actual NML score
-for exponentially many models. In our empirical tests it produces
-models featuring good predictive performance with significantly
-simpler structures than BDeu and fNML.
+\date{}
+\title{Supplementary Material for the article \textit{Quotient Normalized Maximum Likelihood Criterion for Learning Bayesian Network Structures} }
+\begin{document}
+\maketitle
 
-Like BDeu and fNML, qNML can be expressed as a product of $n$ terms,
-one for each variable, but unlike the other two, it is not based on
-further partitioning the corresponding data column
-\begin{eqnarray}
-\label{eqn:qnmldef}
-s^{qNML}(D;G) & := & \sum_{i=1}^n s^{qNML}_i(D;G)\\
-& := & \sum_{i=1}^n \log \frac{P^1_{NML}(D_{i,G_i};G)}
-                             {P^1_{NML}(D_{G_i};G)}.\nonumber
-\end{eqnarray}
-The trick here is to model a subset of columns as though there were no
-conditional independencies among the corresponding variables $S
-\subset X$.  In this case, we can collapse the $\prod_{X_i\in S} r_i$
-value configurations and consider them as values of a single variable
-with $\prod_{X_i\in S} r_i$ different values which can then be modeled
-with a one-dimensional $P^1_{NML}$ code.  The $s^{qNML}$-score does
-not necessarily define a distribution for $D$, but it is easy to
-verify that it coincides with the $\log P_{NML}(D;G)$ for all networks
-that are composed of fully connected components.  The number of such
-networks is lower bounded by the number of nonempty partitions of a
-set of $n$ elements, i.e., the $n^\text{th}$ Bell number.
-
-We are now ready to prove some important properties of the qNML-score.
-
-\subsection {qNML IS SCORE EQUIVALENT}
-
-qNML yields equal scores for network structures that encode same sets
-of independencies. Verma and Pearl~\cite{Verm90} showed that the
-equivalent networks are exactly those which a) are the same when directed
-arcs are substituted by undirected ones and b) which have the same
-\textit{V-structures}, i.e. the variable triplets $(A,B,C)$ where both
-$A$ and $B$ are parents of $C$, but there is no arc between $A$ and
-$B$ (in either direction).  Later, Chickering~\cite{Chick95} showed
-that all the equivalent network structures, and only those structures,
-can be reached from each other by reversing, one by one, the so-called
-\textit{covered arcs}, i.e. the arcs from node $A$ to $B$, for which
-$B$'s parents other than $A$ are exactly  $A$'s parents
-($G_B=\{A\}\cup G_A$).
-
-We will next state this as a
-theorem and sketch a proof for it. A more detailed proof appears in the
-supplementary material.
-\begin{theorem}
-  \label{thm:scoreqv}
-  Let $G$ and $G'$ be two Bayesian network structures that differ only
-  by a single covered arc reversal, i.e., the arc from $A$ to $B$ in $G$
-  has been reversed in $G'$ to point from $B$ to $A$, then
-  $$s^{qNML}(D;G)=s^{qNML}(D;G').$$
-\end{theorem}
-\begin{proof}
-  Now the scores for structures can be decomposed as
-  $s^{qNML}(D;G)=\sum_{i=1}^{n}s_i^{qNML}(D;G)$ and
-  $s^{qNML}(D;G')=\sum_{i=1}^{n}s_i^{qNML}(D;G')$.  Since only the
-  terms corresponding to the variables $A$ and $B$ in these sums are
-  different, it is enough to show that the sum of these two terms are
-  equal for $G$ and $G'$. Since we can assume the data to be fixed we
-  lighten up the notation and write
-  $P^1_{NML}(i,G_i) := P^1_{NML}(D_{i,G_i};G)$ and
-  $P^1_{NML}(G_i)   := P^1_{NML}(D_{G_i};G)$.
-  \begin{eqnarray}
-    \lefteqn{s_A^{qNML}(D;G)+s_B^{qNML}(D;G)} \nonumber\\
-    && =\log\frac{P^1_{NML}(A,G_{A})}{P^1_{NML}(G_{A})}
-            \frac{P^1_{NML}(B,G_{B})}{P^1_{NML}(G_{B})}\nonumber\\
-    && =\log 1\cdot\frac{P^1_{NML}(B,G_{B})}{P^1_{NML}(G_{A})}\nonumber\\
-    && =\log \frac{P^1_{NML}(B,G'_{B})}{P^1_{NML}(G'_{A})}
-             \frac{P^1_{NML}(A,G'_{A})}{P^1_{NML}(G'_{B})}\nonumber\\
- && =s_A^{qNML}(D;G')+s_B^{qNML}(D;G'),\nonumber
-\end{eqnarray}
-  using the equations $\{A\}\cup G_A = G_B$, $\{B\}\cup G'_B = G'_A$,
-  $\{B\}\cup G_B = \{A\} \cup G'_A$, and $G_A = G'_B$ which follow
-  easily from the definition of covered arcs.
-\end{proof}
-
-\subsection{qNML IS CONSISTENT}
-
-One important property possessed by nearly every model selection
-criterion is consistency. In our context, consistency means that given
-a data matrix with $N$ samples coming from a distribution faithful to
-some DAG $G$, the qNML will give the highest score to true graph $G$
-with a probability tending to one as $N$ increases. We will show this
-by first proving that qNML is asymptotically equivalent to the widely used
-BIC criterion which is known to be consistent \cite{Schw78, Haug88}.
-The outline of this proof follows a similar pattern to that in
-\cite{SilanderIJAR10} where the consistency of fNML was proved.
-
-The BIC criterion can be written as
-\begin{equation}\label{BIC}
-\textnormal{BIC}(D;G) = \sum_{i = 1}^n \log P(D_i \ | \ \hat{\theta}_{i | G_i} ) - \frac{q_i(r_i - 1)}{2} \log N,
-\end{equation}
-where $\hat{\theta}_{i | G_i}$ denotes the maximum likelihood parameters of
-the conditional distribution of variable $i$ given its parents in
-$G$.
-
-Since both the BIC and qNML scores are decomposable, we can focus on
-studying the local scores. We will next show that, asymptotically, the
-local qNML score equals the local BIC score. This is formulated in the
-following theorem:
-
-\begin{theorem}\label{consistency}
-Let $r_i$ and $q_i$ denote the number of possible values for variable
-$X_i$ and its possible configurations of parents $G_i$,
-respectively. As $N \to \infty$
-$$
-s^{qNML}_i(D;G) =  \log P(D_i \ | \ \hat{\theta}_{i | G_i} )  - \frac{q_i(r_i - 1)}{2} \log N.
-$$
-\end{theorem}
-
-In order to prove this, we start with the definition of qNML and write
-\begin{align}\label{qnmlDef2}
-s^{qNML}_i(D;G) &= \log \frac{P(D_{i, G_i} \ | \ \hat{\theta}_{i, G_i}
-  )}{P(D_{G_i} \ | \ \hat{\theta}_{G_i} )} \notag \\ & -(reg(N,q_i
-r_i) - reg(N,q_i)).
-\end{align}
-
-By comparing the equations (\ref{BIC}) and (\ref{qnmlDef2}), we see
-that proving our clam boils down to showing two things: 1) the terms
-involving the maximized likelihoods are equal and 2) the penalty terms
-are asymptotically equivalent. We will formulate these as two
-lemmas.
-
-\begin{lemma}\label{MLLemma} The maximized likelihood terms in equations (\ref{BIC}) and (\ref{qnmlDef2}) are equal:    
-$$
-\frac{P(D_{i, G_i} \ | \ \hat{\theta}_{i, G_i} )}{P(D_{G_i} \ | \ \hat{\theta}_{G_i} )} = P(D_i \ | \ \hat{\theta}_{i | G_i} ).
-$$
-\end{lemma}
-
-\begin{proof}
-We can write the terms on the left side of the equation as
-\begin{eqnarray*}
-P(D_{i, G_i} \ | \ \hat{\theta}_{i, G_i}) &=& \prod_{j,k} \Le \frac{N_{ijk}}{N}  \Ri^{N_{ijk}}, \text{ and }\\
-P(D_{G_i} \ | \ \hat{\theta}_{G_i} ) &=&  \prod_{j} \Le \frac{N_{ij}}{N}  \Ri^{N_{ij}}.
-\end{eqnarray*}
-Here, $N_{ijk}$ denotes the number of times we observe $X_i$ taking value $k$ when its parents are in $j^\text{th}$ configuration in our data matrix $D$.
-Also, $N_{ij} = \sum_k N_{ijk}$ (and $\sum_{k,j}N_{ijk} = N$ for all $i$).
-Therefore,
-\begin{align*}
-\frac{P(D_{i, G_i} \ | \ \hat{\theta}_{i, G_i} )}{P(D_{G_i} \ | \ \hat{\theta}_{G_i} )} &= \frac{ \prod_{j,k} \Le \frac{N_{ijk}}{N}  \Ri^{N_{ijk}}}{\prod_{j} \Le \frac{N_{ij}}{N}  \Ri^{N_{ij}}} \\
-&= \frac{ \prod_{j,k} \Le \frac{N_{ijk}}{N}  \Ri^{N_{ijk}}}{\prod_{j}\prod_{k} \Le \frac{N_{ij}}{N}  \Ri^{N_{ijk}}}\\
-%% &= \prod_{j,k} \Le \frac{N_{ijk}}{N_{ij}} \Ri^{N_{ijk}} \\
-&= P(D_i \ | \ \hat{\theta}_{i | G_i} ).
-\end{align*}
-
-\end{proof}
-Next, we consider the difference of regrets in
-(\ref{qnmlDef2}) which corresponds to the penalty term of BIC. The
-following lemma states that these two are asymptotically equal:
-
-\begin{lemma}\label{penaltyLemma}
-As $N \to \infty$,
-$$reg(N,q_i r_i) - reg(N,q_i) = \frac{q_i(r_i - 1)}{2}\log N + O(1).$$   
-\end{lemma}
-\begin{proof}
-The regret
-for a single multinomial variable with $m$ categories can be written
-asymptotically as
-\begin{equation}\label{regretAsymp}
-reg(N,m) = \frac{m-1}{2}\log N + O(1).
-\end{equation}
-For the more precise statement with the underlying assumptions (which are fulfilled in the multinomial case) and for the proof, we refer to \cite{Riss96a, Grun07}. Using this, we have
-\begin{align*}
-reg(N,q_i r_i)& -  reg(N,q_i) \\ 
-&= \frac{q_ir_i-1}{2}\log N-\frac{q_i-1}{2}\log N + O(1) \\
-&= \frac{q_ir_i-1-q_i + 1}{2}\log N + O(1) \\
-&= \frac{q_i(r_i - 1)}{2} \log N + O(1).
-\end{align*}
-\end{proof}
-This concludes our proof since Lemmas \ref{MLLemma} and
-\ref{penaltyLemma} imply Theorem \ref{consistency}.
+\appendix
+\section{qNML coincides with NML for many models}
 
@@
 
-\subsection{QNML IS REGULAR}
-
-\cite{Suzuki2017} defines regularity for a scoring function $Q_n(X \mid Y)$ as follows:
-\begin{definition}
-Assume $H_N(X \mid Y') \leq H_N(X \mid Y)$, where $Y' \subset Y.$ We say that $Q_N(\cdot \mid \cdot)$ is regular if $Q_n(X \mid Y') \geq Q_N(X \mid Y)$.
-\end{definition}
-In the definition, $N$ denotes the sample size, $X$ is some random variable, $Y$ denotes the proposed parent set for $X$, and $H_N(\cdot \mid \cdot)$ refers to the empirical conditional entropy.
-\cite{Suzuki2017} shows that BDeu violates this principle and demonstrates that this causes the score to prefer more complex networks even though data do not support this.
-In \cite{Suzuki2017_2}, the authors also suggest that regular scores are computationally more efficient when applied with branch-and-bound type algorithms for Bayesian network structure learning.
-
-By analyzing the penalty term of the qNML scoring function, one can prove the following statement:
-\begin{theorem}
-qNML score is regular.
-\end{theorem}
-\begin{proof}
-The proof is given in the Appendix A in Supplementary Material.
-\end{proof}
+\bibliographystyle{apalike}
+\bibliography{cosco}
+\end{document}

### 2017-10-12 experiments.tex
Edited experiments too

@@
 
-\subsection{FINDING GENERATING STRUCTURE}
+\subsection{Finding Generating Structure}
 
@@
 
-\subsection{PREDICTION AND PARSIMONY}
+\subsection{Prediction and Parsimony}
 
 To empirically compare the model selection criteria we took 20 UCI
-data sets~\cite{Lichman:2013} and ran 1000 train and test experiments
-for all of them. In each experiment, random 10\% of the original
-data set was selected as training data and the rest 90\% as test data. The training was conducted using the same dynamic programming based exact structure learning algorithm.
+data sets~\cite{Lichman:2013} and ran train and test experiments for
+all of them. To better compare the performance over different sample
+sizes, we picked different fractions ($10\%, 20\%, \ldots, 90\%$) of
+the data sets as training data and the rest as the test data.
+This was done for 1000 different permutations of each dataset.
+The training was conducted using the dynamic
+programming based exact structure learning algorithm.
 
-When predicting with structures learned by the BDeu score, we used the
+When predicting $P(d_{test}|D_{train},S,\theta)$,
+with structures $S$ learned by the BDeu score, we used the
 Bayesian predictive parameter values (BPP) $\theta_{ijk} \propto
@@
 \begin{table}
-\caption{Predictive log losses for small sample sizes for different model selection criteria in 20 different data sets.}
+\caption{Average predictive performance rank over different sample sizes for different model selection criteria in 20 different data sets.}
 \label{tbl:preds}
@@
        \midrule
-    Iris &    15 &     \textit{3.61} &              3.56 &   \underline{3.50} &      \textbf{3.47} \\
- PostOpe &    18 &    \textit{10.52} &     \textbf{7.45} &   \underline{8.35} &               8.39 \\
-   Ecoli &    34 &     \textit{6.27} &              6.16 &      \textbf{5.56} &   \underline{5.63} \\
-   Liver &    35 &     \textit{4.22} &     \textbf{4.01} &               4.10 &   \underline{4.03} \\
-    Wine &    36 &    \textit{16.41} &    \textbf{11.27} &              12.07 &  \underline{11.86} \\
-   Glass &    44 &     \textit{7.48} &              6.65 &      \textbf{6.37} &   \underline{6.43} \\
- Thyroid &    44 &              2.92 &     \textit{2.93} &      \textbf{2.81} &   \underline{2.83} \\
- HeartSt &    54 &    \textit{14.37} &    \textbf{10.96} &              11.98 &  \underline{11.65} \\
- BreastC &    58 &    \textit{11.21} &     \textbf{9.94} &  \underline{10.34} &              10.37 \\
- HeartHu &    60 &     \textit{9.21} &     \textbf{8.38} &               8.80 &   \underline{8.61} \\
- HeartCl &    62 &    \textit{14.88} &    \textbf{11.76} &              12.48 &  \underline{12.18} \\
- BcWisco &    70 &     \textit{6.41} &     \textbf{5.09} &               5.29 &   \underline{5.28} \\
- Diabete &    77 &     \textit{5.11} &  \underline{4.99} &               5.09 &      \textbf{4.98} \\
- TicTacT &    96 &             10.99 &    \textbf{10.35} &  \underline{10.72} &     \textit{11.04} \\
- Balance &   126 &              7.45 &  \underline{7.42} &      \textbf{7.27} &      \textit{7.73} \\
-   Yeast &   149 &              5.67 &     \textit{5.80} &      \textbf{5.48} &   \underline{5.49} \\
- Abalone &   418 &  \underline{3.93} &     \textit{4.00} &      \textbf{3.88} &               3.96 \\
- PageBlo &   548 &  \underline{2.36} &     \textit{2.40} &      \textbf{2.32} &               2.38 \\
- Shuttle &  5800 &  \underline{1.69} &     \textit{1.72} &      \textbf{1.69} &               1.71 \\
-   Adult &  6513 &             10.16 &    \textit{10.25} &     \textbf{10.05} &  \underline{10.08} \\
+ PostOpe &     90 &              2.79 &     \textbf{1.20} &  \underline{3.06} &           2.94 \\
+    Iris &    150 &  \underline{2.82} &              2.37 &     \textbf{2.27} &           2.54 \\
+    Wine &    178 &  \underline{3.23} &     \textbf{1.88} &              2.67 &           2.22 \\
+   Glass &    214 &  \underline{3.61} &              3.09 &     \textbf{1.42} &           1.88 \\
+ Thyroid &    215 &              2.55 &  \underline{3.21} &     \textbf{1.80} &           2.44 \\
+ HeartSt &    270 &  \underline{3.12} &     \textbf{1.39} &              3.12 &           2.37 \\
+ BreastC &    286 &  \underline{3.09} &     \textbf{1.41} &              2.97 &           2.53 \\
+ HeartHu &    294 &  \underline{3.18} &     \textbf{1.66} &              2.90 &           2.27 \\
+ HeartCl &    303 &  \underline{3.46} &     \textbf{1.38} &              2.99 &           2.17 \\
+   Ecoli &    336 &              3.20 &  \underline{3.53} &     \textbf{1.24} &           2.04 \\
+   Liver &    345 &  \underline{3.17} &              2.39 &              2.69 &  \textbf{1.75} \\
+ Balance &    625 &  \underline{3.35} &              1.91 &     \textbf{1.59} &           3.16 \\
+ BcWisco &    699 &  \underline{3.06} &              2.03 &              2.89 &  \textbf{2.02} \\
+ Diabete &    768 &  \underline{2.91} &              2.70 &              2.68 &  \textbf{1.71} \\
+ TicTacT &    958 &  \underline{3.44} &              2.71 &     \textbf{1.31} &           2.53 \\
+   Yeast &   1484 &              2.60 &  \underline{3.76} &     \textbf{1.55} &           2.10 \\
+ Abalone &   4177 &              2.60 &  \underline{3.64} &     \textbf{1.04} &           2.72 \\
+ PageBlo &   5473 &              2.24 &  \underline{3.61} &     \textbf{1.31} &           2.83 \\
+   Adult &  32561 &              3.23 &  \underline{3.77} &     \textbf{1.00} &           2.00 \\
+ Shuttle &  58000 &     \textbf{1.44} &  \underline{3.78} &              1.56 &           3.22 \\
 \end{tabular}
@@
 
-Table~\ref{tbl:preds} features the average predictive losses for different
-data sets.  BIC's bias for simplicity makes it often win (written bold
-in the table) with small sample sizes, but it performs worst (in
-italics) for the large sample sizes. the qNML seems to achieve a lot
-of runner-up positions (underlined) making it a rather safe choice.
+For each train/test sample, we ranked the predictive performance of
+the models learned by different four structure learning criteria (rank
+1 being the best and 4 the worst). Table~\ref{tbl:preds} features the
+average rank for different data sets, the average being taken over
+1000 different train test samples for each 9 sample sizes for each
+data.  BIC's bias for simplicity makes it often win (written bold in
+the table) with small sample sizes, but it often performs worst
+(underlined) for the larger sample sizes (for the same reason), while
+fNML seems to be good for large sample sizes. The striking feature
+about the qNML is its robustness.  It is usually between BIC and fNML
+for all the sample sizes making it a ``safe choice''. This can be
+quantified if we further average the columns of the
+table~\ref{tbl:preds}.  This yields the average ranks of $2.95, 2.57,
+2.10$, and $2.37$, with standard deviations $0.49, 0.90, 0.76$, and
+$0.43$.  While fNML achieves on average the best rank, the runner-up
+qNML has the lowest standard deviation.
 
@@
 parameters in the model. qNML, instead, appears to yield more
-parsimonious models.
+parsimonious models. To study the concern of fNML producing too
+complex models for small sample sizes, we studied the number of
+parameters in models produced by different model selection criteria
+when using 10\% of each data set for structure learning.
 
-Looking at the number of parents for the same 20 data sets again
+Looking at the number of parameters for the same 20 data sets again
 features BIC's preference for simple models
-(Table~\ref{tbl:nofparams}).  qNML usually (18/20) yields more
-parsimonious models than fNML that selects the most complex model for 14
-out of 20 data sets.
+(Table~\ref{tbl:nofparams}).  qNML usually (19/20) yields more
+parsimonious models than fNML that selects the most complex model for
+7 out of 20 data sets.
+
+The graphs for different sample sizes for both predictive accuracy and
+the number of parameters can be found in the Supplementary Material.
 
 \begin{table}
-  \caption{Average number of parents per node 
+  \caption{Average number of parameters in models 
     for different model selection criteria in 20 different data sets.}
@@
 \midrule
-    Iris &    15 &              0.84 &     \textbf{0.63} &  \textit{0.87} &  \underline{0.82} \\
- PostOpe &    18 &              1.35 &     \textbf{0.11} &  \textit{1.66} &  \underline{1.29} \\
-   Ecoli &    34 &              0.83 &     \textbf{0.30} &  \textit{0.96} &  \underline{0.77} \\
-   Liver &    35 &              0.60 &     \textbf{0.10} &  \textit{0.88} &  \underline{0.43} \\
-    Wine &    36 &     \textit{2.11} &     \textbf{0.75} &           1.86 &  \underline{1.44} \\
-   Glass &    44 &              1.32 &     \textbf{0.38} &  \textit{1.41} &  \underline{0.85} \\
- Thyroid &    44 &              0.84 &     \textbf{0.48} &  \textit{0.97} &  \underline{0.66} \\
- HeartSt &    54 &              1.81 &     \textbf{0.56} &  \textit{2.06} &  \underline{1.50} \\
- BreastC &    58 &  \underline{1.27} &     \textbf{0.42} &  \textit{1.71} &              1.39 \\
- HeartHu &    60 &  \underline{0.93} &     \textbf{0.56} &  \textit{1.55} &              1.04 \\
- HeartCl &    62 &              1.69 &     \textbf{0.41} &  \textit{1.96} &  \underline{1.33} \\
- BcWisco &    70 &     \textit{2.02} &     \textbf{0.76} &           1.96 &  \underline{1.14} \\
- Diabete &    77 &  \underline{0.55} &     \textbf{0.23} &  \textit{1.04} &              0.58 \\
- TicTacT &    96 &  \underline{1.60} &     \textbf{0.21} &           2.32 &     \textit{2.47} \\
- Balance &   126 &     \textbf{0.06} &  \underline{0.15} &           0.71 &     \textit{1.18} \\
-   Yeast &   149 &              0.54 &     \textbf{0.14} &  \textit{0.78} &  \underline{0.45} \\
- Abalone &   418 &              1.20 &     \textbf{0.78} &  \textit{1.49} &  \underline{0.93} \\
- PageBlo &   548 &     \textit{1.67} &     \textbf{0.38} &           1.17 &  \underline{0.49} \\
- Shuttle &  5800 &     \textit{2.02} &     \textbf{1.05} &           1.84 &  \underline{1.14} \\
-   Adult &  6513 &  \underline{1.34} &     \textbf{1.11} &  \textit{1.59} &              1.36 \\
+    Iris &    15 &     \underline{37} &   \textbf{23} &                33 &               29 \\
+ PostOpe &    18 &   \underline{1217} &   \textbf{19} &               397 &              146 \\
+   Ecoli &    34 &    \underline{182} &   \textbf{31} &               162 &               77 \\
+   Liver &    35 &                 45 &   \textbf{15} &    \underline{61} &               24 \\
+    Wine &    36 &  \underline{16521} &   \textbf{70} &               807 &              205 \\
+   Glass &    44 &   \underline{1677} &   \textbf{48} &               506 &               97 \\
+ Thyroid &    44 &                 40 &   \textbf{23} &    \underline{66} &               28 \\
+ HeartSt &    54 &  \underline{16861} &   \textbf{44} &              1110 &              256 \\
+ BreastC &    58 &  \underline{25797} &   \textbf{49} &              3767 &              844 \\
+ HeartHu &    60 &   \underline{1634} &   \textbf{43} &               792 &               90 \\
+ HeartCl &    62 &  \underline{34381} &   \textbf{47} &              1433 &              404 \\
+ BcWisco &    70 &   \underline{4630} &   \textbf{42} &               603 &               89 \\
+ Diabete &    77 &                 39 &   \textbf{22} &   \underline{216} &               34 \\
+ TicTacT &    96 &  \underline{13701} &   \textbf{25} &              1969 &              767 \\
+ Balance &   126 &        \textbf{20} &            24 &                49 &  \underline{611} \\
+   Yeast &   149 &                 71 &   \textbf{31} &   \underline{265} &               75 \\
+ Abalone &   418 &                 91 &   \textbf{46} &   \underline{150} &               63 \\
+ PageBlo &   548 &    \underline{703} &   \textbf{45} &               380 &               56 \\
+ Shuttle &  5800 &                535 &   \textbf{99} &   \underline{717} &              130 \\
+   Adult &  6513 &                699 &  \textbf{479} &  \underline{1555} &              945 \\
 \end{tabular}

### 2017-10-12 qnml.tex
Edited experiments too

@@
 
-\subsection {qNML IS SCORE EQUIVALENT}
+\subsection {qNML Is Score Equivalent}
 
@@
 
-\subsection{qNML IS CONSISTENT}
+\subsection{qNML Is Consistent}
 
@@
 
-\subsection{qNML EQUALS NML FOR MANY MODELS}
+\subsection{qNML Equals NML for Many Models}
 The fNML criterion can be seen as a computationally feasible
@@
 criterion equals the NML criterion only for the Bayesian network
-structure with no arcs.  We will next show that the qNML criterion
+structure with no arcs.  It can be shown that the qNML criterion
 equals the NML criterion for all the networks $G$ whose connected
-components tournaments (i.e., complete directed acyclic subgraphs of
-$G$).
+components are tournaments (i.e., complete directed acyclic subgraphs of
+$G$). These networks include the empty network, the fully connected
+one and many networks in between having different complexity. While
+the generating network is unlikely to be composed of tournament
+components, the result increases the plausibility that qNML is a
+reasonable approximation for NML in general\footnote{A claim that is
+  naturally subject for further study.}.
 
@@
 \begin{proof}
-We first show that the NML-criterion for a Bayesian network
-decomposes by the connected components.
-
-Because the maximum likelihood for the data $D$ decomposes,
-we can write
-\begin{eqnarray}
-  \lefteqn{P_{NML}(D;G)} \nonumber \\
-  && = \frac{P(D;\hat\theta(D),G)}
-            {\sum_{D'_{V_1}}\ldots \sum_{D'_{V_C}}\prod_{c=1}^C P(D'_{V^c};\hat\theta(D'_{V^c}),G)} \nonumber \\
-            && = \frac{\prod_{c=1}^C P(D_{V^c};\hat\theta(D_{V^c}),G)} 
-            {\prod_{c=1}^C\sum_{D'_{V^c}}P(D'_{V^c};\hat\theta(D'_{V^c}),G)}.\nonumber \\
-            && = \prod_{c=1}^CP_{NML}(D_{V^c};G).
-\end{eqnarray}
-
-Clearly, the qNML score also decomposes by the connected components,
-so it remains to show that if the (sub)network $G$ is a tournament,
-then for any data $D$, $s^{qNML}(D;G)=\log P_{NML}(D;G)$.  Due to the
-score equivalence of the NML criterion and the qNML criterion, we may
-pick a tournament $G$ such that the linear ordering defined by $G$
-matches the ordering of the data columns, i.e., $i<j$ implies $G_i
-\subset G_j$. Now from the definition~(\ref{eqn:qnmldef}) of the qNML
-criterion we see that for the tournament $G$, the sum telescopes
-leaving us with $s^{qNML}(D;G) = \log P^1_{NML}(D_G;G)$, thus
-it is enough to show that $P^1_{NML}(D;G)=P_{NML}(D;G)$.  This
-follows, since for any data vector $x$ in data $D$, we have
-$P^1(x;\hat\theta(D),G) = P(x;\hat\theta(D),G)$, where $P^1$ denotes
-the model that takes $n$-dimensional vectors to be values of the
-single (collapsed) categorical variable.  Denoting prefixes of data
-vector $x$ by $x^{:i}$, and the number of times such a prefix appears
-on $N$ rows $[d_1,\ldots,d_N]$ of the $N\times n$ data matrix $D$ by
-$N_D(x^{:i})$, (so that $N_D(x^{:0})=N)$, we have
-\begin{eqnarray}
-  P(D;\hat\theta(D),G) &=& \prod_{j=1}^{N}P(d_j;\hat\theta(D),G) \nonumber \\
-  &=& \prod_{j=1}^{N} \prod_{i=1}^{n} \frac{N_D(d_j^{:i})}{N_D(d_j^{:i-1})} \nonumber\\
-  &=& \prod_{j=1}^{N} \frac{N_D(d_j^{:n})}{N} \nonumber \\
-  &=& P^1(D;\hat\theta^1(D),G).
-\end{eqnarray}
-Since both $P^1_{NML}$ and $P_{NML}$ are defined in terms of the
-maximum likelihood probabilities, the equality above implies the equality
-of these distributions.
+The full proof can be found in Appendix B in the Supplementary
+Material.  The proof first shows that NML decomposes for these
+particular structures, so it is enough to show the equivalence for
+fully connected graphs.
+It further derives the number $a(n)$ of
+different $n$-node networks whose connected components are
+tournaments, which turns out to be the formula for OEIS sequence
+A000262\footnote{https://oeis.org/A000108}.
+In general this sequence grows rapidly; $1, 1, 3, 13, 73, 501, 4051,
+37633, 394353, 4596553, \ldots$.
 \end{proof}
 
-Equality established, we would like to still state the number $a(n)$
-of different $n$-node networks whose connected components are
-tournaments.  We start by generating all the $p(n)$ integer partitions
-i.e. ways to partition $n$ labelled into parts with different
-size-profiles.  For example, for $n=4$, we have $p(4)=5$ partition
-profiles $[[4], [3, 1], [2, 2], [2, 1, 1], [1, 1, 1, 1]]$. Each of
-these partition size profiles corresponds to many different networks,
-apart from the last one that corresponds just to the empty network.
-We count number of networks for one such partition size-profile (and
-then later sum these counts up).  For any such partition profile
-$(p_1,\ldots,p_k)$ we can count the ways we can assign the nodes to
-different parts and then order each part. This leads to the product
-$n\choose p_1$$p_1$!${n-p_1}\choose p_2$$p_2!$${n-p_1-p2}\choose
-p_3$$p_3!$$\ldots$${n-\sum_{j=1}^{k-1}p_j}\choose p_k$$p_k!$. However,
-the order of different parts of the same size does not matter, so for
-all groups of parts having the same size, we have to divide the
-product above by the factorial of the size of such group. Notice also,
-that the product above telescopes, leaving us a formula for OEIS
-sequence A000262\footnote{https://oeis.org/A000108} as
-described by Thomas Wieder:
-
-\textit{With $p(n) =$ the number of integer partitions of $n$, $d(i)$ = the
-number of different parts of the $i^\text{th}$ partition of $n$,
-$m(i,j) =$ multiplicity of the $j^\text{th}$ part of the $i^\text{th}$
-partition of $n$, one has:}
-$$
-a(n) = \sum_{i=1}^{p(n)} \frac{n!}{\prod_{j=1}^{d(i)} m(i,j)!}.
-$$
-For example, $a(4)=\frac{4!}{1!}+\frac{4!}{1!1!}+\frac{4!}{2!}+\frac{4!}{1!2!}+\frac{4!}{4!}=73$. In general this sequence grows rapidly; $1, 1, 3, 13, 73, 501, 4051, 37633, 394353, 4596553, \ldots$.
-
-\subsection{QNML IS REGULAR}
+\subsection{qNML Is Regular}
 

### 2017-10-12 supp_mat.tex
organizing supp mat

@@
 \appendix
-\section{Regularity proof}
+\input{supp_mat_A.tex}\newpage
+\input{supp_mat_B.tex}\newpage
+\input{supp_mat_C.tex}
 
-\subsection{Preliminaries}
-We start by recalling the definition of regularity \citep{Suzuki2017}:
-
-\begin{definition}
-Assume $H_N(X \mid Y') \leq H_N(X \mid Y)$, where $Y' \subset Y.$ We say that the scoring function $Q_N(\cdot \mid \cdot)$ is regular if $Q_n(X \mid Y') \geq Q_N(X \mid Y)$.
-\end{definition}
-In the definition, $N$ denotes the sample size, $X$ is some random variable, $Y$ denotes the proposed parent set for $X$, and $H_N(X \mid Y)$ refers to the empirical conditional entropy based on $N$ samples of variables $X$ and $Y$.
-
-Let $X$ be a categorical random variable with $r$ possible values.
-Let $U$ denote a possible parent set with $q$ different combinations of values for the variables, and $V$ a parent set with $m$ different configurations.
-Assume that we have observed $N$ samples of $(X,U,V)$ (denoted by $x_N,u_N$ and $v_N$) and $H_N(X \mid U) \leq H_N(X \mid U \cup V)$ holds.
-
-Recall the definition of the qNML score:
-\begin{align*}
-Q^{qnml}_N(X\mid U) &= \log P(x_n \mid \hat{\theta}_{X\mid U} \ ) - \Le reg(N,rq) - reg(N,q) \Ri \\
-&= \log P(x_N \mid \hat{\theta}_{X\mid U} \ ) - \log \frac{C(N,rq)}{C(N,q)},
-\end{align*}where $C(N,r)$ is the normalizing constant the of the NML distribution for a categorical variable with $r$ possible values and sample size $N$ and $ \hat{\theta}_{X\mid U}$ denotes the maximum likelihood parameters of the conditional distribution of $X$ given $U$ which are computed from the data $(x_N,u_N)$.
-
-In order to prove the regularity, we need the following three lemmas:
-\begin{lemma}\label{Cpolynom} We can write $C(N,k)$ as a polynomial of $k$, formally
-$$
-C(N,k) = \sum_{j=1}^N a_j k^j,
-$$where $a_j > 0$. 
-\end{lemma}
-\begin{lemma}\label{MLterms}
-Assume $H_N(X \mid Y') \leq H_N(X \mid Y)$, where $Y' \subset Y$. Now $\log P(x_N \mid \hat{\theta}_{X\mid Y} \ ) = \log P(x_N \mid \hat{\theta}_{X\mid Y'} \ )$.
-\end{lemma}
-\begin{lemma}\label{increasing}
-Let $r \in \N, r \geq 2$.  The function $k \mapsto \frac{C(N,rk)}{C(N,k)}$ is increasing for every $k \geq 2$.
-\end{lemma}
-\noindent We present the proofs of these lemmas in Section \ref{lemmaproofs}.
-
-\subsection{The main proof}
-\begin{theorem}
-qNML score is regular.
-\end{theorem}
-\begin{proof}
-We want to show that
-$$
-Q^{qnml}_N(X\mid U) \geq Q^{qnml}_N(X \mid U \cup V).
-$$assuming $H_N(X \mid U) \leq H_N(X \mid U \cup V)$.
-Using the entropy assumption and Lemma \ref{MLterms} implies that the maximized likelihood terms are equal.
-In order to prove the claim, it suffices to study the penalty terms, and we want to show that
-\begin{align*}
-- \Le reg(N,rq) - reg(N,q) \Ri \ &\geq \  -\Le reg(N,rqm) - reg(N,qm) \Ri \\
-\log \frac{C(N,rq)}{C(N,q)} \ &\leq \log \frac{C(N,rqm)}{C(N,qm)}.
-\end{align*}This holds, since logarithm is an increasing function, and $q \leq qm$, so we can apply Lemma \ref{increasing} to conclude the proof.
-
-\end{proof}
-
-\subsection{Proofs of lemmas}\label{lemmaproofs}
-%We want to show that
-%$$
-%\quad  Q^{qnml}_N(X\mid U) \geq Q^{qnml}_N(X \mid U \cup V)
-%$$which, by using the definition of qNML, is equivalent to 
-%\begin{equation}\label{statement}
-%\log P(X \mid )
-%\end{equation}
-%The assumption about entropy implies that the maximized likelihood terms of the qnml-score are equal. 
-\setcounter{lemma}{0}
-\begin{lemma} $C(N,k)$ can be written as a polynomial of $k$, formally
-$$
-C(N,k) = \sum_{j=1}^N a_j k^j,
-$$where $a_j > 0$. 
-\end{lemma}
-\begin{proof} \cite{cosco.itsl08} derive the following representation for the normalizing constant
-\begin{align*}  
-C(N,k) = \sum_{l=0}^{N-1}\frac{(N-1)^{\underline{l}}k^{\overline{l + 1}}}{N^{l+1} \ l!} ,
-\end{align*}where $x^{\underline{l}}$ and $x^{\overline{l}}$ denote falling and rising factorials, respectively.
-
-We utilize the fact that the rising factorial can be represented as polynomial using unsigned Stirling numbers of the first kind (see \cite{Adamchik1997}, for instance)
-\begin{align*}
-C(N,k) &=  \sum_{l=0}^{N-1}\frac{(N-1)^{\underline{l}}k^{\overline{l + 1}}}{N^{l+1} \ l!} \\
-&= \sum_{l=0}^{N-1} b_l \ k^{\overline{l+1}} \\
-&= \sum_{l=0}^{N-1} b_l \Le \sum_{j=1}^{l+1}|s(l + 1,j)| \ k^j \Ri \\
-&= \sum_{l=0}^{N-1} \Le \sum_{j=1}^{N}b_l \ |s(l + 1,j)| \ k^j \Ri \\
-&= \sum_{j=1}^{N} \Le \sum_{l=0}^{N-1} b_l \ |s(l + 1,j)| \ k^j \Ri \\
-&= \sum_{j=1}^{N} \Le \sum_{l=0}^{N-1} b_l \ |s(l + 1,j)| \Ri k^j \\
-&= \sum_{j=1}^{N} a_j  k^j,
-\end{align*}where $s(i,j)$ denotes the (signed) Stirling number of the first kind and
-$$
-a_j = \Le \sum_{l=0}^{N-1} \frac{(N-1)^{\underline{l}}}{N^{l+1}l!} \ |s(l + 1,j)| \Ri,
-$$ $a_j > 0$ for all $j$. On the second row, we denoted $b_l = (N-1)^{\underline{l}}/(N^{l+1} l!)$. On the row 4, we used the property of Stirling numbers: $s(i,j) = 0$ for all $j > i$. 
-\end{proof}
-\begin{lemma}
-Assume $H_N(X \mid Y') \leq H_N(X \mid Y)$, where $Y' \subset Y$. Now $\log P(x_N \mid \hat{\theta}_{X\mid Y} \ ) = \log P(x_N \mid \hat{\theta}_{X\mid Y'})$.
-\end{lemma}
-\begin{proof}
-We can write the logarithm of the maximized likelihood, \\ $\log P(x_N \mid \hat{\theta}_{X\mid Y} \ )$, as follows \citep{KOLLER}
-\begin{align*}
-\log P(x_N \mid \hat{\theta}_{X\mid Y} \ ) & = -N \Le H_N(X) - I_N(X;Y) \Ri \\
-&= -N \  H_N(X\mid Y),
-\end{align*}where $I_N(\cdot;\cdot)$ is the empirical mutual information. This implies that the assumption
-$$
-H_N(X \mid Y') \leq H_N(X \mid Y)
-$$ is equivalent to
-$$
-\log P(x_N \mid \hat{\theta}_{X\mid Y} \ ) \geq \log P(x_N \mid \hat{\theta}_{X\mid Y'} \ ) .
-$$Actually we must have the equality holding in the above expression, since
-$$
-H_N(X \mid Y') < H_N(X \mid Y)
-$$ would imply that
-$$
-I_N(X ; Z \mid Y ) < 0,
-$$where $Z = Y \setminus Y'$, which is impossible.
-\end{proof}
-\begin{lemma}
-Let $r \in \N, r \geq 2$.  The function $k \mapsto \frac{C(N,rk)}{C(N,k)}$ is increasing for every $k \geq 2$.
-\end{lemma}
-
-\begin{proof}
-Lemma \ref{Cpolynom} lets us to write
-\begin{equation}\label{C1}
-C(N,k) = \sum_{j=1}^{N} a_j k^j
-\end{equation}and, similarly,
-\begin{equation}\label{C2}
-C(N,rk) = \sum_{j=1}^{N} a_j  r^jk^j.
-\end{equation}
-
-From this, it it easy see that the derivative of the quotient, \\ $d/dk (C(N,rk)/C(N,k))$, will be a ratio of two polynomials of $k$.
-Our goal is to show that the polynomial in the numerator has positive coefficients, which will guarantee the positivity of derivative for every $k > 0$, and thus imply that the original function is increasing (polynomial in the denominator is squared and non-zero for $k > 0$, so it can be ignored).
-
-Derivatives of (\ref{C1}) and (\ref{C2}) are obtained easily:
-\begin{align*}
-\frac{d}{dk}C(N,k) &= \sum_{j=1}^{N}ja_jk^{j-1} \\
-&= \sum_{j=0}^{N-1}(j+1) a_{j+1}k^{j}
-\end{align*}and
-\begin{align*}
-\frac{d}{dk}C(N,rk) &= \sum_{j=1}^{N}ja_jr^jk^{j-1} \\
-&= \sum_{j=0}^{N-1}(j+1)a_{j+1}r^{j+1}k^{j}. 
-\end{align*}Consider next the products found in the derivative of the quotient. We obtain
-\begin{align*}
-\Le\frac{d}{dk}C(N,rk)\Ri C(N,k) &= \Le \sum_{j=0}^{N-1}(j+1)a_{j+1}r^{j+1}k^{j} \Ri \Le \sum_{l=1}^{N} a_l  k^l \Ri \\
-&= \sum_{i = 1}^{2N-1} \Le \sum_{j+l = i} (j+1)a_{j+1}r^{j+1}a_l  \Ri k^i
-\end{align*}and
-%\begin{align*}
-%\Le\frac{d}{dk}C(n,k)\Ri C(n,rk) &= \Le \sum_{j=0}^{n-1}(j+1)a_{j+1}k^{j} \Ri \Le %\sum_{l=1}^{n} a_l  r^lk^l  \Ri \\
-%&= \sum_{i = 1}^{2n-1} \Le \sum_{m=0}^i (m+1)a_{m+1}a_{i-m}r^{i-m}   \Ri k^i
-%\end{align*}
-\begin{align*}
-\Le\frac{d}{dk}C(N,k)\Ri C(N,rk) &= \Le \sum_{j=0}^{N-1}(j+1)a_{j+1}k^{j} \Ri \Le \sum_{l=1}^{N} a_l  r^lk^l  \Ri \\
-&= \sum_{i = 1}^{2N-1} \Le \sum_{j+l = i}(j+1)a_{j+1}a_l  r^l   \Ri k^i.
-\end{align*}
-Subtracting these two expression yields
-\begin{align*}
-&\Le\frac{d}{dk}C(N,rk)\Ri C(N,k)-\Le\frac{d}{dk}C(N,k)\Ri C(N,rk) \\&= \sum_{i = 1}^{2N-1} \Le \sum_{j+l = i} (j+1)a_{j+1}r^{j+1}a_l  \Ri k^i - \sum_{i = 1}^{2N-1} \Le \sum_{j+l = i}(j+1)a_{j+1}a_l  r^l   \Ri k^i \\
-&= \sum_{i = 1}^{2N-1} \Le \sum_{j+l = i}(j+1)a_{j+1}a_l  (r^{j+1}- r^l)   \Ri k^i
-\end{align*}which is the polynomial in the numerator of the derivative of $C(N,rk)/C(N,k)$.
-Next, we study the coefficient of $k^i$, if $i \leq N$
-\begin{align*}
-\sum_{j+l = i}(j+1)a_{j+1}a_l  (r^{j+1}- r^l) &= \sum_{l=1}^{i}(i-l+1)a_{i-l+1}a_l  (r^{i-l+1}- r^l) \\
-& = \sum_{l=1}^{i}(i-l+1)c_l \\
-&= \sum_{k =1}^{\lfloor i / 2 \rfloor}(i-k+1)c_k + (i-(i-k + 1)+1)c_{i-k+1} \\
-&=  \sum_{k =1}^{\lfloor i / 2 \rfloor}(i-k+1)c_k + kc_{i-k+1} \\
-&=  \sum_{k =1}^{\lfloor i / 2 \rfloor}(i-k+1)c_k - kc_{k} \\
-&= \sum_{k =1}^{\lfloor i / 2 \rfloor}(i-2k+1)c_k.
-\end{align*} On the first row, we re-wrote sum using only one running index.
-On the second row we denoted $c_l =a_{i-l+1}a_l  (r^{i-l+1}- r^l)$.
-On the third row, we re-arranged the sum so that we are summing over pairs of terms of the original sum: the first and the last term, the second and the second to last, and so on.
-This resulting sum has $\lfloor i / 2 \rfloor$ terms.
-We have to use the floor-function since if $i$ is odd, there exists an index $l'$ in the original sum such that $r^{i-l'+1}-r^{l'} = 0$.
-On the fifth row, we make use of the identity $c_k = -c_{i-k+1}$ which is straightforward to verify.
-From the last row, we can observe that every term of the sum is positive since $i-2k+1$ and $r^{i-k+1}- r^k$ are both positive if $k \leq (i+1)/2$ which holds since $k$ ranges from $1$ to $\lfloor i / 2 \rfloor$.
-
-Let us now consider the situation where $n < i \leq 2N-1$. We start with the special case where $i = 2N-1$. Then, we have only one term in the sum
-\begin{align*}
-\sum_{j+l = i}(j+1)a_{j+1}a_l  (r^{j+1}- r^l) &= \sum_{l=N}^{N}(2N-1-l+1)a_{2N-1-l+1}a_l  (r^{2N-1-l+1}- r^l) \\
-&= Na_{N}a_N  (r^{N}- r^N) \\
-& = 0.
-\end{align*}Now, let $N < i < 2N-1$, we follow a similar procedure as before to manipulate the sum
-\begin{align*}
-\sum_{j+l = i}(j+1)a_{j+1}a_l  (r^{j+1}- r^l) &= \sum_{l=i-N + 1}^{N}(i-l+1)a_{i-l+1}a_l  (r^{i-l+1}- r^l) \\
-& = \sum_{l=i-N + 1}^{N}(i-l+1)c_l \\
-& = \sum_{k=1}^{2N-i}(N-k+1)c_{i-N+k} \\
-& = \sum_{k=1}^{\lfloor N - i/2 \rfloor}(N-k+1)c_{i-N+k}\\
-&+  (N-(2N-i-k+1)+1)c_{i-N+(2N-i-k+1)} \\
-&= \sum_{k=1}^{\lfloor N - i/2 \rfloor}(N-k+1)c_{i-N+k} +  (i-N+k)c_{N-k+1} \\
-&= \sum_{k=1}^{\lfloor N - i/2 \rfloor}(N-k+1)c_{i-N+k} -  (i-N+k)c_{i-N+k} \\
-&= \sum_{k=1}^{\lfloor N - i/2 \rfloor}(N-k+1-(i-N+k))c_{i-N+k} \\
-&= \sum_{k=1}^{\lfloor N - i/2 \rfloor}(2N-i-2k+1)c_{i-N+k}.
-\end{align*}It is now easy to verify that $(2N-i-2k+1)$ and $c_{i-N+k}$ are positive if $k \leq N - (i-1)/2$ which holds since $k$ ranges from $1$ to $\lfloor N - i/2 \rfloor$.
-The floor function is again used when we sum over pairs of terms since if $i$ is odd there is zero-term.
-Since all the coefficients are non-negative and the $k \geq 2$, the derivative is positive. This implies that the original function is increasing.
-\end{proof}
 \bibliographystyle{apalike}

### 2017-10-12 supp_mat_A.tex
organizing supp mat

\section{Regularity proof}

\subsection{Preliminaries}
We start by recalling the definition of regularity \citep{Suzuki2017}:

\begin{definition}
Assume $H_N(X \mid Y') \leq H_N(X \mid Y)$, where $Y' \subset Y.$ We say that the scoring function $Q_N(\cdot \mid \cdot)$ is regular if $Q_n(X \mid Y') \geq Q_N(X \mid Y)$.
\end{definition}
In the definition, $N$ denotes the sample size, $X$ is some random variable, $Y$ denotes the proposed parent set for $X$, and $H_N(X \mid Y)$ refers to the empirical conditional entropy based on $N$ samples of variables $X$ and $Y$.

Let $X$ be a categorical random variable with $r$ possible values. Let $U$ denote a possible parent set with $q$ different combinations of values for the variables, and $V$ a parent set with $m$ different configurations. Assume that we have observed $N$ samples of $(X,U,V)$ (denoted by $x_N,u_N$ and $v_N$) and $H_N(X \mid U) \leq H_N(X \mid U \cup V)$ holds.

Recall the definition of the qNML score:
\begin{align*}
Q^{qnml}_N(X\mid U) &= \log P(x_n \mid \hat{\theta}_{X\mid U} \ ) - \Le reg(N,rq) - reg(N,q) \Ri \\
&= \log P(x_N \mid \hat{\theta}_{X\mid U} \ ) - \log \frac{C(N,rq)}{C(N,q)},
\end{align*}where $C(N,r)$ is the normalizing constant the of the NML distribution for a categorical variable with $r$ possible values and sample size $N$ and $ \hat{\theta}_{X\mid U}$ denotes the maximum likelihood parameters of the conditional distribution of $X$ given $U$ which are computed from the data $(x_N,u_N)$.

In order to prove the regularity, we need the following three lemmas:
\begin{lemma}\label{Cpolynom} We can write $C(N,k)$ as a polynomial of $k$, formally
$$
C(N,k) = \sum_{j=1}^N a_j k^j,
$$where $a_j > 0$. 
\end{lemma}
\begin{lemma}\label{MLterms}
Assume $H_N(X \mid Y') \leq H_N(X \mid Y)$, where $Y' \subset Y$. Now $\log P(x_N \mid \hat{\theta}_{X\mid Y} \ ) = \log P(x_N \mid \hat{\theta}_{X\mid Y'} \ )$.
\end{lemma}
\begin{lemma}\label{increasing}
Let $r \in \N, r \geq 2$.  The function $k \mapsto \frac{C(N,rk)}{C(N,k)}$ is increasing for every $k \geq 2$.
\end{lemma}
\noindent We present the proofs of these lemmas in Section \ref{lemmaproofs}.

\subsection{The main proof}
\begin{theorem}
qNML score is regular.
\end{theorem}
\begin{proof}
We want to show that
$$
Q^{qnml}_N(X\mid U) \geq Q^{qnml}_N(X \mid U \cup V).
$$assuming $H_N(X \mid U) \leq H_N(X \mid U \cup V)$. Using the entropy assumption and Lemma \ref{MLterms} implies that the maximized likelihood terms are equal. In order to prove the claim, it suffices to study the penalty terms, and we want to show that
\begin{align*}
- \Le reg(N,rq) - reg(N,q) \Ri \ &\geq \  -\Le reg(N,rqm) - reg(N,qm) \Ri \\
\log \frac{C(N,rq)}{C(N,q)} \ &\leq \log \frac{C(N,rqm)}{C(N,qm)}.
\end{align*}This holds, since logarithm is an increasing function, and $q \leq qm$, so we can apply Lemma \ref{increasing} to conclude the proof.

\end{proof}

\subsection{Proofs of lemmas}\label{lemmaproofs}
%We want to show that
%$$
%\quad  Q^{qnml}_N(X\mid U) \geq Q^{qnml}_N(X \mid U \cup V)
%$$which, by using the definition of qNML, is equivalent to 
%\begin{equation}\label{statement}
%\log P(X \mid )
%\end{equation}
%The assumption about entropy implies that the maximized likelihood terms of the qnml-score are equal. 
\setcounter{lemma}{0}
\begin{lemma} $C(N,k)$ can be written as a polynomial of $k$, formally
$$
C(N,k) = \sum_{j=1}^N a_j k^j,
$$where $a_j > 0$. 
\end{lemma}
\begin{proof} \cite{cosco.itsl08} derive the following representation for the normalizing constant
\begin{align*}  
C(N,k) = \sum_{l=0}^{N-1}\frac{(N-1)^{\underline{l}}k^{\overline{l + 1}}}{N^{l+1} \ l!} ,
\end{align*}where $x^{\underline{l}}$ and $x^{\overline{l}}$ denote falling and rising factorials, respectively.

We utilize the fact that the rising factorial can be represented as polynomial using unsigned Stirling numbers of the first kind (see \cite{Adamchik1997}, for instance)
\begin{align*}
C(N,k) &=  \sum_{l=0}^{N-1}\frac{(N-1)^{\underline{l}}k^{\overline{l + 1}}}{N^{l+1} \ l!} \\
&= \sum_{l=0}^{N-1} b_l \ k^{\overline{l+1}} \\
&= \sum_{l=0}^{N-1} b_l \Le \sum_{j=1}^{l+1}|s(l + 1,j)| \ k^j \Ri \\
&= \sum_{l=0}^{N-1} \Le \sum_{j=1}^{N}b_l \ |s(l + 1,j)| \ k^j \Ri \\
&= \sum_{j=1}^{N} \Le \sum_{l=0}^{N-1} b_l \ |s(l + 1,j)| \ k^j \Ri \\
&= \sum_{j=1}^{N} \Le \sum_{l=0}^{N-1} b_l \ |s(l + 1,j)| \Ri k^j \\
&= \sum_{j=1}^{N} a_j  k^j,
\end{align*}where $s(i,j)$ denotes the (signed) Stirling number of the first kind and
$$
a_j = \Le \sum_{l=0}^{N-1} \frac{(N-1)^{\underline{l}}}{N^{l+1}l!} \ |s(l + 1,j)| \Ri,
$$ $a_j > 0$ for all $j$. On the second row, we denoted $b_l = (N-1)^{\underline{l}}/(N^{l+1} l!)$. On the row 4, we used the property of Stirling numbers: $s(i,j) = 0$ for all $j > i$. 
\end{proof}
\begin{lemma}
Assume $H_N(X \mid Y') \leq H_N(X \mid Y)$, where $Y' \subset Y$. Now $\log P(x_N \mid \hat{\theta}_{X\mid Y} \ ) = \log P(x_N \mid \hat{\theta}_{X\mid Y'})$.
\end{lemma}
\begin{proof}
We can write the logarithm of the maximized likelihood, \\ $\log P(x_N \mid \hat{\theta}_{X\mid Y} \ )$, as follows \citep{KOLLER}
\begin{align*}
\log P(x_N \mid \hat{\theta}_{X\mid Y} \ ) & = -N \Le H_N(X) - I_N(X;Y) \Ri \\
&= -N \  H_N(X\mid Y),
\end{align*}where $I_N(\cdot;\cdot)$ is the empirical mutual information. This implies that the assumption
$$
H_N(X \mid Y') \leq H_N(X \mid Y)
$$ is equivalent to
$$
\log P(x_N \mid \hat{\theta}_{X\mid Y} \ ) \geq \log P(x_N \mid \hat{\theta}_{X\mid Y'} \ ) .
$$Actually we must have the equality holding in the above expression, since
$$
H_N(X \mid Y') < H_N(X \mid Y)
$$ would imply that
$$
I_N(X ; Z \mid Y ) < 0,
$$where $Z = Y \setminus Y'$, which is impossible.
\end{proof}
\begin{lemma}
Let $r \in \N, r \geq 2$.  The function $k \mapsto \frac{C(N,rk)}{C(N,k)}$ is increasing for every $k \geq 2$.
\end{lemma}

\begin{proof}
Lemma \ref{Cpolynom} lets us to write
\begin{equation}\label{C1}
C(N,k) = \sum_{j=1}^{N} a_j k^j
\end{equation}and, similarly,
\begin{equation}\label{C2}
C(N,rk) = \sum_{j=1}^{N} a_j  r^jk^j.
\end{equation}

From this, it it easy see that the derivative of the quotient, \\ $d/dk (C(N,rk)/C(N,k))$, will be a ratio of two polynomials of $k$. Our goal is to show that the polynomial in the numerator has positive coefficients, which will guarantee the positivity of derivative for every $k > 0$, and thus imply that the original function is increasing (polynomial in the denominator is squared and non-zero for $k > 0$, so it can be ignored).

Derivatives of (\ref{C1}) and (\ref{C2}) are obtained easily:
\begin{align*}
\frac{d}{dk}C(N,k) &= \sum_{j=1}^{N}ja_jk^{j-1} \\
&= \sum_{j=0}^{N-1}(j+1) a_{j+1}k^{j}
\end{align*}and
\begin{align*}
\frac{d}{dk}C(N,rk) &= \sum_{j=1}^{N}ja_jr^jk^{j-1} \\
&= \sum_{j=0}^{N-1}(j+1)a_{j+1}r^{j+1}k^{j}. 
\end{align*}Consider next the products found in the derivative of the quotient. We obtain
\begin{align*}
\Le\frac{d}{dk}C(N,rk)\Ri C(N,k) &= \Le \sum_{j=0}^{N-1}(j+1)a_{j+1}r^{j+1}k^{j} \Ri \Le \sum_{l=1}^{N} a_l  k^l \Ri \\
&= \sum_{i = 1}^{2N-1} \Le \sum_{j+l = i} (j+1)a_{j+1}r^{j+1}a_l  \Ri k^i
\end{align*}and
%\begin{align*}
%\Le\frac{d}{dk}C(n,k)\Ri C(n,rk) &= \Le \sum_{j=0}^{n-1}(j+1)a_{j+1}k^{j} \Ri \Le %\sum_{l=1}^{n} a_l  r^lk^l  \Ri \\
%&= \sum_{i = 1}^{2n-1} \Le \sum_{m=0}^i (m+1)a_{m+1}a_{i-m}r^{i-m}   \Ri k^i
%\end{align*}
\begin{align*}
\Le\frac{d}{dk}C(N,k)\Ri C(N,rk) &= \Le \sum_{j=0}^{N-1}(j+1)a_{j+1}k^{j} \Ri \Le \sum_{l=1}^{N} a_l  r^lk^l  \Ri \\
&= \sum_{i = 1}^{2N-1} \Le \sum_{j+l = i}(j+1)a_{j+1}a_l  r^l   \Ri k^i.
\end{align*}
Subtracting these two expression yields
\begin{align*}
&\Le\frac{d}{dk}C(N,rk)\Ri C(N,k)-\Le\frac{d}{dk}C(N,k)\Ri C(N,rk) \\&= \sum_{i = 1}^{2N-1} \Le \sum_{j+l = i} (j+1)a_{j+1}r^{j+1}a_l  \Ri k^i - \sum_{i = 1}^{2N-1} \Le \sum_{j+l = i}(j+1)a_{j+1}a_l  r^l   \Ri k^i \\
&= \sum_{i = 1}^{2N-1} \Le \sum_{j+l = i}(j+1)a_{j+1}a_l  (r^{j+1}- r^l)   \Ri k^i
\end{align*}which is the polynomial in the numerator of the derivative of $C(N,rk)/C(N,k)$.
Next, we study the coefficient of $k^i$, if $i \leq N$
\begin{align*}
\sum_{j+l = i}(j+1)a_{j+1}a_l  (r^{j+1}- r^l) &= \sum_{l=1}^{i}(i-l+1)a_{i-l+1}a_l  (r^{i-l+1}- r^l) \\
& = \sum_{l=1}^{i}(i-l+1)c_l \\
&= \sum_{k =1}^{\lfloor i / 2 \rfloor}(i-k+1)c_k + (i-(i-k + 1)+1)c_{i-k+1} \\
&=  \sum_{k =1}^{\lfloor i / 2 \rfloor}(i-k+1)c_k + kc_{i-k+1} \\
&=  \sum_{k =1}^{\lfloor i / 2 \rfloor}(i-k+1)c_k - kc_{k} \\
&= \sum_{k =1}^{\lfloor i / 2 \rfloor}(i-2k+1)c_k.
\end{align*} On the first row, we re-wrote sum using only one running index. On the second row we denoted $c_l =a_{i-l+1}a_l  (r^{i-l+1}- r^l)$. On the third row, we re-arranged the sum so that we are summing over pairs of terms of the original sum: the first and the last term, the second and the second to last, and so on.  This resulting sum has $\lfloor i / 2 \rfloor$ terms. We have to use the floor-function since if $i$ is odd, there exists an index $l'$ in the original sum such that $r^{i-l'+1}-r^{l'} = 0$. On the fifth row, we make use of the identity $c_k = -c_{i-k+1}$ which is straightforward to verify. From the last row, we can observe that every term of the sum is positive since $i-2k+1$ and $r^{i-k+1}- r^k$ are both positive if $k \leq (i+1)/2$ which holds since $k$ ranges from $1$ to $\lfloor i / 2 \rfloor$.

Let us now consider the situation where $n < i \leq 2N-1$. We start with the special case where $i = 2N-1$. Then, we have only one term in the sum
\begin{align*}
\sum_{j+l = i}(j+1)a_{j+1}a_l  (r^{j+1}- r^l) &= \sum_{l=N}^{N}(2N-1-l+1)a_{2N-1-l+1}a_l  (r^{2N-1-l+1}- r^l) \\
&= Na_{N}a_N  (r^{N}- r^N) \\
& = 0.
\end{align*}Now, let $N < i < 2N-1$, we follow a similar procedure as before to manipulate the sum
\begin{align*}
\sum_{j+l = i}(j+1)a_{j+1}a_l  (r^{j+1}- r^l) &= \sum_{l=i-N + 1}^{N}(i-l+1)a_{i-l+1}a_l  (r^{i-l+1}- r^l) \\
& = \sum_{l=i-N + 1}^{N}(i-l+1)c_l \\
& = \sum_{k=1}^{2N-i}(N-k+1)c_{i-N+k} \\
& = \sum_{k=1}^{\lfloor N - i/2 \rfloor}(N-k+1)c_{i-N+k}\\
&+  (N-(2N-i-k+1)+1)c_{i-N+(2N-i-k+1)} \\
&= \sum_{k=1}^{\lfloor N - i/2 \rfloor}(N-k+1)c_{i-N+k} +  (i-N+k)c_{N-k+1} \\
&= \sum_{k=1}^{\lfloor N - i/2 \rfloor}(N-k+1)c_{i-N+k} -  (i-N+k)c_{i-N+k} \\
&= \sum_{k=1}^{\lfloor N - i/2 \rfloor}(N-k+1-(i-N+k))c_{i-N+k} \\
&= \sum_{k=1}^{\lfloor N - i/2 \rfloor}(2N-i-2k+1)c_{i-N+k}.
\end{align*}It is now easy to verify that $(2N-i-2k+1)$ and $c_{i-N+k}$ are positive if $k \leq N - (i-1)/2$ which holds since $k$ ranges from $1$ to $\lfloor N - i/2 \rfloor$. The floor function is again used when we sum over pairs of terms since if $i$ is odd there is zero-term. 
Since all the coefficients are non-negative and the $k \geq 2$, the derivative is positive. This implies that the original function is increasing.
\end{proof}

### 2017-10-12 supp_mat_C.tex
organizing supp mat

\section{Prediction and Parsimony}

We took 20 UCI data sets created 1000 permutations of each data set.
We then took the first x\% of each permuted data set data (x in
${10,20,30,\ldots,90}$ as training data and used the remaining part as
a test data. We used exact structure learning to learn the best
scoring model using four model selection criteria, BDeu ($\alpha=1$),
BIC, fNML and qNML, recording the number of parameters in the learnt
models and the average predictive log-probability of each test
vector. These numbers are then collected for 1000 different data
permutations average and variance of which appear on the following
pages, one page per data set.

### 2017-10-12 makesuppmat.sh

@@ -1,0 +1,5 @@
+#/bin/bash
+pdflatex supp_mat.tex
+bibtex supp_mat
+pdflatex supp_mat.tex; pdflatex supp_mat.tex
+pdfunite supp_mat.pdf all_qNML_images4.pdf all_sup_mat.pdf

### 2017-10-13 supp_mat_B.tex
organizing supp mat
eq. ref corrected

@@
-\documentclass[12pt]{article}
-\usepackage{amsmath}
-\usepackage{amsthm}
-\usepackage{amssymb}
-\usepackage{bm}
-\usepackage[round]{natbib}
-\usepackage{bm}
-\newcommand{\Le}{\left(}
-\newcommand{\Ri}{\right)}
-\newcommand{\N}{\mathbb{N}}
-\newcommand\independent{\protect\mathpalette{\protect\independenT}{\perp}}
-\def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}
-\newtheorem{theorem}{Theorem}
-\newtheorem{definition}{Definition}
-\newtheorem{corollary}{Corollary}
-\newtheorem{lemma}{Lemma}
-
-\date{}
-\title{Supplementary Material for the article \textit{Quotient Normalized Maximum Likelihood Criterion for Learning Bayesian Network Structures} }
-\begin{document}
-\maketitle
-
-\appendix
 \section{qNML coincides with NML for many models}
 
-\subsection{qNML EQUALS NML FOR MANY MODELS}
+\subsection{qNML Equals NML for Many Models}
 The fNML criterion can be seen as a computationally feasible
@@
 matches the ordering of the data columns, i.e., $i<j$ implies $G_i
-\subset G_j$. Now from the definition~(\ref{eqn:qnmldef}) of the qNML
-criterion we see that for the tournament $G$, the sum telescopes
+\subset G_j$. Now from the definition~(8) of the qNML
+criterion (in the main paper), we see that for the tournament $G$, the sum telescopes
 leaving us with $s^{qNML}(D;G) = \log P^1_{NML}(D_G;G)$, thus
@@
 For example, $a(4)=\frac{4!}{1!}+\frac{4!}{1!1!}+\frac{4!}{2!}+\frac{4!}{1!2!}+\frac{4!}{4!}=73$. In general this sequence grows rapidly; $1, 1, 3, 13, 73, 501, 4051, 37633, 394353, 4596553, \ldots$.
-
-\bibliographystyle{apalike}
-\bibliography{cosco}
-\end{document}

### 2017-10-13 supp_mat_A.tex
fNML regularity

@@
 \end{proof}
+
+\subsection{fNML is Regular}
+In this section, we will show that fNML score is regular. Recall first the definition of fNML local score
+\begin{equation}
+Q^{fnml}_N(X  \vert \ U) = \log P(x_n \mid \hat{\theta}_{X\mid U} \ ) - \sum_{j = 1}^{r_U} reg(N_j,r_X),
+\end{equation}where $r_X$ is the number of categories for $X$, $r_U$ denotes the number of possible configurations of variables in $U$, and $N_j$ is the number of times the $j$:th configuration is observed in our samples $u_N$.
+Note that $reg(r_X, N_j) = 0$ for any configuration not observed in our sample.
+fNML criterion differs from qNML only by how the penalty term is defined. We can follow a highly similar strategy in order to prove the regularity for fNML. To this end, we need the following Lemma:
+\begin{lemma}\label{regineq}
+Let $N = N_1 + N_2$ and $r \in\N, r > 2$. Now,
+$$
+reg(N,r) \leq reg(N_1,r) + reg(N_2,r).
+$$
+\end{lemma}
+\begin{proof}
+We start with the definition of $C(N_1,r) = \exp(reg(N_1,r))$,
+\begin{align*}
+C(N_1,r)C(N_2,r) &= \sum_{x_{N_1}}P(x_{N_1} \vert \hat{\theta}(x_{N_1} ))\sum_{x_{N_2}}P(x_{N_2} \vert \hat{\theta}(x_{N_2} ))\\ & \geq
+\sum_{x_{N_1}}P(x_{N_1} \vert \hat{\theta}(x_{N} ))\sum_{x_{N_2}}P(x_{N_2} \vert \hat{\theta}(x_{N} )) \\ &=
+\sum_{x_{N}}P(x_{N} \vert \hat{\theta}(x_{N} )) \\ & = C(N,r).
+\end{align*}Taking the logarithm on the both sides yields our claim.
+On the second row, we used the definition of maximum likelihood parameters: the probability of data vector $x_{N_1}$ is maximized under parameters $\hat{\theta}(x_{N_1})$, and if we use different parameters, $\hat{\theta}(x_{N_1}$, probability can only go down or stay the same.
+On the third row, we used the i.i.d, assumption, which allows us to take the single sum over data vectors of length $N = N_1 + N_2$.
+\end{proof}
+Now we can proceed to the actual proof.
+\begin{theorem}
+fNML score is regular.
+\end{theorem}
+\begin{proof}
+Using the entropy assumption and Lemma \ref{C2}, we can ignore the maximized likelihood terms and study the penalty terms. We want show that
+\begin{equation}\label{fnmlPenalty}
+\sum_{j = 1}^{r_{Y'}} reg(N'_j,r_X) \leq \sum_{l = 1}^{r_{Y}} reg(N_l,r_X),
+\end{equation}where we used $r_{Y'}$ and $r_Y$ to denote the numbers of possible configurations for variables in $Y'$ and $Y$, respectively.
+Also, $N_j$ refers to the number times the $j$:th configuration of variables $Y$ is observed in our sample, and $N'_l$ denotes the same for $Y'$.
+
+Now, since $r_{Y'} \leq r_Y$, and we know that $\sum_j N'_j = \sum_l N_l = N$, we can just apply Lemma (\ref{regineq}) multiple times to conclude our proof. 
+\end{proof}

### 2017-10-13 experiments.tex
info on used dags

@@
 
-In our first experiment, we took five widely used benchmark Bayesian networks\footnote{Bayesian Network Repository: \url{http://www.bnlearn.com/bnrepository/}} (Asia, Cancer, Earthquake, Sachs and Survey), sampled data from them, and tried to learn the generating structure with the different scoring functions using various sample sizes.
+In our first experiment, we took five widely used benchmark Bayesian networks\footnote{Bayesian Network Repository: \url{http://www.bnlearn.com/bnrepository/}}, sampled data from them, and tried to learn the generating structure with the different scoring functions using various sample sizes.
+We used the following networks: Asia ($n=5$, 8 arcs), Cancer ($n=5$, 4 arcs), Earthquake ($n=5$, 4 arcs), Sachs ($n=11$, 17 arcs) and Survey ($n=6$, 6 arcs).
 These networks were picked in order to use the dynamic programming based exact structure learning~\cite{cosco.uai06} which limited the number $n$ of variables

### 2017-10-13 supp_mat_A.tex
Add score equivalency proof to supplementary material

--- 2017-10-12 supp_mat_A.tex
@@
-\section{Regularity proof}
+\section{Score equivalency proof}
 
-\subsection{Preliminaries}
-We start by recalling the definition of regularity \citep{Suzuki2017}:
+Chickering showed that any equivalent structure can be reached from
+another by a series of arc reversal operations without leaving the
+equivalence class~\cite{Chick95}. This means that in order to show
+that equivalent structures leads to the same qNML score, we only need
+to prove that this is the case for equivalent structures that differ
+by a single arc reversal.
 
-\begin{definition}
-Assume $H_N(X \mid Y') \leq H_N(X \mid Y)$, where $Y' \subset Y.$ We say that the scoring function $Q_N(\cdot \mid \cdot)$ is regular if $Q_n(X \mid Y') \geq Q_N(X \mid Y)$.
-\end{definition}
-In the definition, $N$ denotes the sample size, $X$ is some random variable, $Y$ denotes the proposed parent set for $X$, and $H_N(X \mid Y)$ refers to the empirical conditional entropy based on $N$ samples of variables $X$ and $Y$.
+The network without any arcs is the sole member of its equivalence
+class. All the other networks have at least one arc. We will first
+presents some lemmas that characterize the relations of the parent sets
+before and after the arc reversal. Let $G$ be a network structure
+and $G'$ an equivalent network structure after the arc from $A$
+to $B$ has been reversed to point from $B$ to $A$. We
+will denote the parent sets of $A$ and $B$ in $G$ by $G_{A}$
+and $G_{B}$, and the parent sets of $A$ and $B$ in $G'$ by $G'_{A}$
+and $G'_{B}$. Note that not all the arc reversals lead to the equivalent
+structures. By saying that $G$ and $G'$ are equivalent we imply
+that our arc reversal does not destroy existing V-structures or create
+new ones. The crucial observation is that if reversing the arc that
+goes from $A$ to $B$ does not create or destroy V-structures, it
+must be that the parents of $B$ other than $A$ are exactly the same
+as parents of $A$ (statement 3 below).
 
-Let $X$ be a categorical random variable with $r$ possible values.
-Let $U$ denote a possible parent set with $q$ different combinations of values for the variables, and $V$ a parent set with $m$ different configurations.
-Assume that we have observed $N$ samples of $(X,U,V)$ (denoted by $x_N,u_N$ and $v_N$) and $H_N(X \mid U) \leq H_N(X \mid U \cup V)$ holds.
-
-Recall the definition of the qNML score:
-\begin{align*}
-Q^{qnml}_N(X\mid U) &= \log P(x_n \mid \hat{\theta}_{X\mid U} \ ) - \Le reg(N,rq) - reg(N,q) \Ri \\
-&= \log P(x_N \mid \hat{\theta}_{X\mid U} \ ) - \log \frac{C(N,rq)}{C(N,q)},
-\end{align*}where $C(N,r)$ is the normalizing constant the of the NML distribution for a categorical variable with $r$ possible values and sample size $N$ and $ \hat{\theta}_{X\mid U}$ denotes the maximum likelihood parameters of the conditional distribution of $X$ given $U$ which are computed from the data $(x_N,u_N)$.
-
-In order to prove the regularity, we need the following three lemmas:
-\begin{lemma}\label{Cpolynom} We can write $C(N,k)$ as a polynomial of $k$, formally
-$$
-C(N,k) = \sum_{j=1}^N a_j k^j,
-$$where $a_j > 0$. 
-\end{lemma}
-\begin{lemma}\label{MLterms}
-Assume $H_N(X \mid Y') \leq H_N(X \mid Y)$, where $Y' \subset Y$. Now $\log P(x_N \mid \hat{\theta}_{X\mid Y} \ ) = \log P(x_N \mid \hat{\theta}_{X\mid Y'} \ )$.
-\end{lemma}
-\begin{lemma}\label{increasing}
-Let $r \in \N, r \geq 2$.  The function $k \mapsto \frac{C(N,rk)}{C(N,k)}$ is increasing for every $k \geq 2$.
-\end{lemma}
-\noindent We present the proofs of these lemmas in Section \ref{lemmaproofs}.
-
-\subsection{The main proof}
-\begin{theorem}
-qNML score is regular.
-\end{theorem}
-\begin{proof}
-We want to show that
-$$
-Q^{qnml}_N(X\mid U) \geq Q^{qnml}_N(X \mid U \cup V).
-$$assuming $H_N(X \mid U) \leq H_N(X \mid U \cup V)$.
-Using the entropy assumption and Lemma \ref{MLterms} implies that the maximized likelihood terms are equal.
-In order to prove the claim, it suffices to study the penalty terms, and we want to show that
-\begin{align*}
-- \Le reg(N,rq) - reg(N,q) \Ri \ &\geq \  -\Le reg(N,rqm) - reg(N,qm) \Ri \\
-\log \frac{C(N,rq)}{C(N,q)} \ &\leq \log \frac{C(N,rqm)}{C(N,qm)}.
-\end{align*}This holds, since logarithm is an increasing function, and $q \leq qm$, so we can apply Lemma \ref{increasing} to conclude the proof.
-
-\end{proof}
-
-\subsection{Proofs of lemmas}\label{lemmaproofs}
-%We want to show that
-%$$
-%\quad  Q^{qnml}_N(X\mid U) \geq Q^{qnml}_N(X \mid U \cup V)
-%$$which, by using the definition of qNML, is equivalent to 
-%\begin{equation}\label{statement}
-%\log P(X \mid )
-%\end{equation}
-%The assumption about entropy implies that the maximized likelihood terms of the qnml-score are equal. 
-\setcounter{lemma}{0}
-\begin{lemma} $C(N,k)$ can be written as a polynomial of $k$, formally
-$$
-C(N,k) = \sum_{j=1}^N a_j k^j,
-$$where $a_j > 0$. 
-\end{lemma}
-\begin{proof} \cite{cosco.itsl08} derive the following representation for the normalizing constant
-\begin{align*}  
-C(N,k) = \sum_{l=0}^{N-1}\frac{(N-1)^{\underline{l}}k^{\overline{l + 1}}}{N^{l+1} \ l!} ,
-\end{align*}where $x^{\underline{l}}$ and $x^{\overline{l}}$ denote falling and rising factorials, respectively.
-
-We utilize the fact that the rising factorial can be represented as polynomial using unsigned Stirling numbers of the first kind (see \cite{Adamchik1997}, for instance)
-\begin{align*}
-C(N,k) &=  \sum_{l=0}^{N-1}\frac{(N-1)^{\underline{l}}k^{\overline{l + 1}}}{N^{l+1} \ l!} \\
-&= \sum_{l=0}^{N-1} b_l \ k^{\overline{l+1}} \\
-&= \sum_{l=0}^{N-1} b_l \Le \sum_{j=1}^{l+1}|s(l + 1,j)| \ k^j \Ri \\
-&= \sum_{l=0}^{N-1} \Le \sum_{j=1}^{N}b_l \ |s(l + 1,j)| \ k^j \Ri \\
-&= \sum_{j=1}^{N} \Le \sum_{l=0}^{N-1} b_l \ |s(l + 1,j)| \ k^j \Ri \\
-&= \sum_{j=1}^{N} \Le \sum_{l=0}^{N-1} b_l \ |s(l + 1,j)| \Ri k^j \\
-&= \sum_{j=1}^{N} a_j  k^j,
-\end{align*}where $s(i,j)$ denotes the (signed) Stirling number of the first kind and
-$$
-a_j = \Le \sum_{l=0}^{N-1} \frac{(N-1)^{\underline{l}}}{N^{l+1}l!} \ |s(l + 1,j)| \Ri,
-$$ $a_j > 0$ for all $j$. On the second row, we denoted $b_l = (N-1)^{\underline{l}}/(N^{l+1} l!)$. On the row 4, we used the property of Stirling numbers: $s(i,j) = 0$ for all $j > i$. 
-\end{proof}
-\begin{lemma}
-Assume $H_N(X \mid Y') \leq H_N(X \mid Y)$, where $Y' \subset Y$. Now $\log P(x_N \mid \hat{\theta}_{X\mid Y} \ ) = \log P(x_N \mid \hat{\theta}_{X\mid Y'})$.
-\end{lemma}
-\begin{proof}
-We can write the logarithm of the maximized likelihood, \\ $\log P(x_N \mid \hat{\theta}_{X\mid Y} \ )$, as follows \citep{KOLLER}
-\begin{align*}
-\log P(x_N \mid \hat{\theta}_{X\mid Y} \ ) & = -N \Le H_N(X) - I_N(X;Y) \Ri \\
-&= -N \  H_N(X\mid Y),
-\end{align*}where $I_N(\cdot;\cdot)$ is the empirical mutual information. This implies that the assumption
-$$
-H_N(X \mid Y') \leq H_N(X \mid Y)
-$$ is equivalent to
-$$
-\log P(x_N \mid \hat{\theta}_{X\mid Y} \ ) \geq \log P(x_N \mid \hat{\theta}_{X\mid Y'} \ ) .
-$$Actually we must have the equality holding in the above expression, since
-$$
-H_N(X \mid Y') < H_N(X \mid Y)
-$$ would imply that
-$$
-I_N(X ; Z \mid Y ) < 0,
-$$where $Z = Y \setminus Y'$, which is impossible.
-\end{proof}
-\begin{lemma}
-Let $r \in \N, r \geq 2$.  The function $k \mapsto \frac{C(N,rk)}{C(N,k)}$ is increasing for every $k \geq 2$.
+\begin{lemma}\label{thm:sameparents}
+\item $G_{B}=G_{A}\cup\{A\}$
 \end{lemma}
@@
 \begin{proof}
-Lemma \ref{Cpolynom} lets us to write
-\begin{equation}\label{C1}
-C(N,k) = \sum_{j=1}^{N} a_j k^j
-\end{equation}and, similarly,
-\begin{equation}\label{C2}
-C(N,rk) = \sum_{j=1}^{N} a_j  r^jk^j.
-\end{equation}
+We show the direction $G_{A}\cup\{A\}\subset G_{B}$ 
+by contradiction. First of all, a parent of $A$ cannot be a child
+of $B$ or otherwise there would be a loop in $G.$ If $A$ had a
+parent $Z$ that is not a parent of $B$, then $B$ and $Z$ were
+not adjacent and the reversal of the arc would create a new V-structure
+$(B,A,Z)$. Since this was forbidden by the equivalence of $G$ and
+$G'$, $Z$ must also be a parent of $B$.
 
-From this, it it easy see that the derivative of the quotient, \\ $d/dk (C(N,rk)/C(N,k))$, will be a ratio of two polynomials of $k$.
-Our goal is to show that the polynomial in the numerator has positive coefficients, which will guarantee the positivity of derivative for every $k > 0$, and thus imply that the original function is increasing (polynomial in the denominator is squared and non-zero for $k > 0$, so it can be ignored).
+Also, $B$ cannot have parents other than $A$ that are not also parents
+of $A$, i.e., $G_{B}\subset G_{A}\cup{A}$ . If there were such
+a parent $Z$ it should be anyway adjacent to $A$, otherwise the
+V-structure $(A,B,Z)$ would be destroyed in the reversal. Since the
+$Z$ was not a parent of $A$, the only possibility for adjacency
+is that $Z$ were a child of $A$. However, in this case arc reversal
+would lead to a loop $B,A,Z,B$.
+\end{proof}
 
-Derivatives of (\ref{C1}) and (\ref{C2}) are obtained easily:
+We wil next list some simple consequences of the equivalence preserving
+arc reversal:
+
+\begin{lemma}
+\label{thm:reveqs}
+\hangindent\leftmargini
+\hspace{1mm}
+\begin{enumerate}
+\item $G'_{B}=G_{B}\setminus\{A\}$
+\item $G'_{A}=G_{A}\cup\{B\}$ 
+\item $G'_{B}=G_{A}$
+\item $\{B\}\cup G_{B}=\{B\}\cup G_{A}\cup\{A\}=G'_{A}\cup\{A\}$
+\item $G'_{A}=G_{A}\cup\{B\}=G'_{B}\cup\{B\}$
+\end{enumerate}
+\end{lemma}
+
+\begin{proof}
+
+The statements 1 and 2 are trivial since they just state the fact
+that an arc reversal removes $A$ from the parents of $B$ and adds
+$B$ to the parents of $A$.
+
+The
+statement 3 follows directly by using the lemma \ref{thm:sameparents}
+to the statement 1. Combining these equalities we can generate more of them 
+such as statements 4 and 5. 
+\end{proof}
+
+Let us take structures $G$ and $G'$. Since we can assume the data to be fixed we lighten
+up the notations and write 
+$P^1_{NML}(i,G_i) := P^1_{NML}(D[\cdot,(i,G_i)];G)$ and
+$P^1_{NML}(G_i) := P^1_{NML}(D[\cdot,G_i];G)$.
+
+\begin{theorem}
+$  s^{qNML}(D;G)=s^{qNML}(D;G').$
+\end{theorem}
+
+\begin{proof}
+
+  Now the scores for structures can be decomposed as the
+  $s^{qNML}(D;G)=\sum_{i=1}^{n}s_i^{qNML}(D;G)$ and 
+  $s^{qNML}(D;G')=\sum_{i=1}^{n}s_i^{qNML}(D;G')$.
+
+Since only the terms corresponding to the variables $A$ and $B$
+in these sums are different, it is enough to show that
+
+$$
+s_A^{qNML}(D;G)+s_B^{qNML}(D;G) = s_A^{qNML}(D;G')+s_B^{qNML}(D;G')
+$$
+Now
+
 \begin{align*}
-\frac{d}{dk}C(N,k) &= \sum_{j=1}^{N}ja_jk^{j-1} \\
-&= \sum_{j=0}^{N-1}(j+1) a_{j+1}k^{j}
-\end{align*}and
-\begin{align*}
-\frac{d}{dk}C(N,rk) &= \sum_{j=1}^{N}ja_jr^jk^{j-1} \\
-&= \sum_{j=0}^{N-1}(j+1)a_{j+1}r^{j+1}k^{j}. 
-\end{align*}Consider next the products found in the derivative of the quotient. We obtain
-\begin{align*}
-\Le\frac{d}{dk}C(N,rk)\Ri C(N,k) &= \Le \sum_{j=0}^{N-1}(j+1)a_{j+1}r^{j+1}k^{j} \Ri \Le \sum_{l=1}^{N} a_l  k^l \Ri \\
-&= \sum_{i = 1}^{2N-1} \Le \sum_{j+l = i} (j+1)a_{j+1}r^{j+1}a_l  \Ri k^i
-\end{align*}and
-%\begin{align*}
-%\Le\frac{d}{dk}C(n,k)\Ri C(n,rk) &= \Le \sum_{j=0}^{n-1}(j+1)a_{j+1}k^{j} \Ri \Le %\sum_{l=1}^{n} a_l  r^lk^l  \Ri \\
-%&= \sum_{i = 1}^{2n-1} \Le \sum_{m=0}^i (m+1)a_{m+1}a_{i-m}r^{i-m}   \Ri k^i
-%\end{align*}
-\begin{align*}
-\Le\frac{d}{dk}C(N,k)\Ri C(N,rk) &= \Le \sum_{j=0}^{N-1}(j+1)a_{j+1}k^{j} \Ri \Le \sum_{l=1}^{N} a_l  r^lk^l  \Ri \\
-&= \sum_{i = 1}^{2N-1} \Le \sum_{j+l = i}(j+1)a_{j+1}a_l  r^l   \Ri k^i.
+s_A^{qNML}(D;G)+s_B^{qNML}(D;G)& =\log\frac{P^1_{NML}(A,G_{A})}{P^1_{NML}(G_{A})}\frac{P^1_{NML}(B,G_{B})}{P^1_{NML}(G_{B})}\\
+ & =\log 1\cdot\frac{P^1_{NML}(B,G_{B})}{P^1_{NML}(G_{A})}\\
+ & =\log \frac{P^1_{NML}(B,G'_{B})}{P^1_{NML}(G'_{A})}\frac{P^1_{NML}(A,G'_{A})}{P^1_{NML}(G'_{B})}\\
+ & =s_A^{qNML}(D;G')+s_B^{qNML}(D;G').
 \end{align*}
-Subtracting these two expression yields
-\begin{align*}
-&\Le\frac{d}{dk}C(N,rk)\Ri C(N,k)-\Le\frac{d}{dk}C(N,k)\Ri C(N,rk) \\&= \sum_{i = 1}^{2N-1} \Le \sum_{j+l = i} (j+1)a_{j+1}r^{j+1}a_l  \Ri k^i - \sum_{i = 1}^{2N-1} \Le \sum_{j+l = i}(j+1)a_{j+1}a_l  r^l   \Ri k^i \\
-&= \sum_{i = 1}^{2N-1} \Le \sum_{j+l = i}(j+1)a_{j+1}a_l  (r^{j+1}- r^l)   \Ri k^i
-\end{align*}which is the polynomial in the numerator of the derivative of $C(N,rk)/C(N,k)$.
-Next, we study the coefficient of $k^i$, if $i \leq N$
-\begin{align*}
-\sum_{j+l = i}(j+1)a_{j+1}a_l  (r^{j+1}- r^l) &= \sum_{l=1}^{i}(i-l+1)a_{i-l+1}a_l  (r^{i-l+1}- r^l) \\
-& = \sum_{l=1}^{i}(i-l+1)c_l \\
-&= \sum_{k =1}^{\lfloor i / 2 \rfloor}(i-k+1)c_k + (i-(i-k + 1)+1)c_{i-k+1} \\
-&=  \sum_{k =1}^{\lfloor i / 2 \rfloor}(i-k+1)c_k + kc_{i-k+1} \\
-&=  \sum_{k =1}^{\lfloor i / 2 \rfloor}(i-k+1)c_k - kc_{k} \\
-&= \sum_{k =1}^{\lfloor i / 2 \rfloor}(i-2k+1)c_k.
-\end{align*} On the first row, we re-wrote sum using only one running index.
-On the second row we denoted $c_l =a_{i-l+1}a_l  (r^{i-l+1}- r^l)$.
-On the third row, we re-arranged the sum so that we are summing over pairs of terms of the original sum: the first and the last term, the second and the second to last, and so on.
-This resulting sum has $\lfloor i / 2 \rfloor$ terms.
-We have to use the floor-function since if $i$ is odd, there exists an index $l'$ in the original sum such that $r^{i-l'+1}-r^{l'} = 0$.
-On the fifth row, we make use of the identity $c_k = -c_{i-k+1}$ which is straightforward to verify.
-From the last row, we can observe that every term of the sum is positive since $i-2k+1$ and $r^{i-k+1}- r^k$ are both positive if $k \leq (i+1)/2$ which holds since $k$ ranges from $1$ to $\lfloor i / 2 \rfloor$.
-
-Let us now consider the situation where $n < i \leq 2N-1$. We start with the special case where $i = 2N-1$. Then, we have only one term in the sum
-\begin{align*}
-\sum_{j+l = i}(j+1)a_{j+1}a_l  (r^{j+1}- r^l) &= \sum_{l=N}^{N}(2N-1-l+1)a_{2N-1-l+1}a_l  (r^{2N-1-l+1}- r^l) \\
-&= Na_{N}a_N  (r^{N}- r^N) \\
-& = 0.
-\end{align*}Now, let $N < i < 2N-1$, we follow a similar procedure as before to manipulate the sum
-\begin{align*}
-\sum_{j+l = i}(j+1)a_{j+1}a_l  (r^{j+1}- r^l) &= \sum_{l=i-N + 1}^{N}(i-l+1)a_{i-l+1}a_l  (r^{i-l+1}- r^l) \\
-& = \sum_{l=i-N + 1}^{N}(i-l+1)c_l \\
-& = \sum_{k=1}^{2N-i}(N-k+1)c_{i-N+k} \\
-& = \sum_{k=1}^{\lfloor N - i/2 \rfloor}(N-k+1)c_{i-N+k}\\
-&+  (N-(2N-i-k+1)+1)c_{i-N+(2N-i-k+1)} \\
-&= \sum_{k=1}^{\lfloor N - i/2 \rfloor}(N-k+1)c_{i-N+k} +  (i-N+k)c_{N-k+1} \\
-&= \sum_{k=1}^{\lfloor N - i/2 \rfloor}(N-k+1)c_{i-N+k} -  (i-N+k)c_{i-N+k} \\
-&= \sum_{k=1}^{\lfloor N - i/2 \rfloor}(N-k+1-(i-N+k))c_{i-N+k} \\
-&= \sum_{k=1}^{\lfloor N - i/2 \rfloor}(2N-i-2k+1)c_{i-N+k}.
-\end{align*}It is now easy to verify that $(2N-i-2k+1)$ and $c_{i-N+k}$ are positive if $k \leq N - (i-1)/2$ which holds since $k$ ranges from $1$ to $\lfloor N - i/2 \rfloor$.
-The floor function is again used when we sum over pairs of terms since if $i$ is odd there is zero-term.
-Since all the coefficients are non-negative and the $k \geq 2$, the derivative is positive. This implies that the original function is increasing.
+The second equation follows from the lemma \ref{thm:sameparents}, and the third from
+the statements 5 and 4 of the lemma \ref{thm:reveqs}.
 \end{proof}

### 2017-10-13 experiments.tex
Correct Appendix references & typos

--- 2017-10-12 experiments.tex
@@
 None of the scores dominates in all settings considered.
-BIC fares well when the sample size is low, as it tends to produce nearly empty graph which is a good answer in terms of SHD when the generating networks are relatively sparse.
-qNML obtains strong results in 'Earthquake' and 'Asia' networks, being the best or the second to best with all the sample sizes considered.
+BIC fares well when the sample size is small, as it tends to produce a nearly empty graph which is a good answer in terms of SHD when the generating networks are relatively sparse.
+qNML obtains strong results in 'Earthquake' and 'Asia' networks, being the best or the second best with all the sample sizes considered.
 
@@
 Figure \ref{fig:shd_ranks} summarizes the SHD results in all networks by showing the average rank for each score.
-The ranking was done by giving the score with the least SHD rank $1$ and the worst one rank $4$.
+The ranking was done by giving the score with the lowest SHD rank $1$ and the worst one rank $4$.
 The shown results are averages computed from $5000$ values ($5$ networks, $1000$ repetitions).
-From this, we can see that qNML has never the worst average ranking, and it has the best ranking with sample sizes greater than $300$.
-This suggests that qNML is overall a safe choice in structure learning, especially with moderate sample sizes.
+From this, we can see that qNML never has the worst average ranking, and it has the best ranking with sample sizes greater than $300$.
+This suggests that qNML is overall a safe choice in structure learning, especially with moderate and large sample sizes.
 
@@
 the data sets as training data and the rest as the test data.
-This was done for 1000 different permutations of each dataset.
+This was done for 1000 different permutations of each data set.
 The training was conducted using the dynamic
@@
 For each train/test sample, we ranked the predictive performance of
-the models learned by different four structure learning criteria (rank
+the models learned by four different structure learning criteria (rank
 1 being the best and 4 the worst). Table~\ref{tbl:preds} features the

### 2017-10-13 qnml.tex
fNML result mentioned
Correct Appendix references & typos

@@
 We will next state this as a
-theorem and sketch a proof for it. A more detailed proof appears in the
-supplementary material.
+theorem and sketch a proof for it. A more detailed proof appears in Appendix A the Supplementary Material.
 \begin{theorem}
@@
 \begin{proof}
-The full proof can be found in Appendix B in the Supplementary
+The full proof can be found in Appendix C in the Supplementary
 Material.  The proof first shows that NML decomposes for these
@@
 \begin{proof}
-The proof is given in the Appendix A in Supplementary Material.
+The proof is given in Appendix B in the Supplementary Material.
 \end{proof}

### 2017-10-13 bn.tex
Edited experiments too

@@
 
-\subsection{LIKELIHOOD}
+\subsection{Likelihood}
 \label{ssec:likelihood}
@@
 
-\subsection{BAYESIAN STRUCTURE LEARNING}
+\subsection{Bayesian Structure Learning}
 
@@
 
-\subsection {PROBLEMS, SOLUTIONS AND PROBLEMS}
+\subsection {Problems, Solutions and Problems}
 
@@
 data $D$, the values of $X_i$ are completely determined by variables
-in set $Z$, so that the empirical entropy $H(X_i | Z) = 0$. Now, if we can further
+in set $Z$, so that the empirical entropy $H_N(X_i | Z) = 0$. Now, if we can further
 find one or more variables, denoted by $Y$, whose values are
@@
 the set $Z\cup Y$ over $Z$ alone as the parents of $X_i$. Suzuki
-argues that this kind of behaviour violates regularity in model
+argues that this kind of behaviour violates \textit{regularity} in model
 selection as the more complex model is preferred over a simpler one
@@
 likelihood score not to be score equivalent.
+In Section \ref{sec:regularity}, we will give the formal definition of regularity and state that qNML is regular.
+In addition, we provide a proof of regularity for fNML criterion, which has not appeared in the literature before.
+The detailed proofs can be found in Appendix A in the Supplementary Material.
 

### 2017-10-13 intro.tex

@@
 protein activity data~\cite{bn4sigpath02}, revealing the business
-process structures~\cite{bn4bpmining} from transaction logs and
+process structures from transaction logs ~\cite{bn4bpmining} and
 modeling brain-region connectivity using fMRI
@@
 yields parsimonious models with good predictive capabilities. The
-experiments with artificial data demonstrate the capability to give
-high scores for the data generating model with relatively small sample
-sizes.
+experiments with artificial data generated from real world Bayesian networks demonstrate the capability of our score to quickly learn a structure close to the generating one.

### 2017-10-13 supp_mat_B.tex
small typos

@@
-\section{qNML coincides with NML for many models}
+\section{Regularity proof}
 
-\subsection{qNML Equals NML for Many Models}
-The fNML criterion can be seen as a computationally feasible
-approximation of the more desirable NML criterion.  However, the fNML
-criterion equals the NML criterion only for the Bayesian network
-structure with no arcs.  We will next show that the qNML criterion
-equals the NML criterion for all the networks $G$ whose connected
-components tournaments (i.e., complete directed acyclic subgraphs of
-$G$).
+\subsection{Preliminaries}
+We start by recalling the definition of regularity \citep{Suzuki2017}:
 
+\begin{definition}
+Assume $H_N(X \mid Y') \leq H_N(X \mid Y)$, where $Y' \subset Y.$ We say that the scoring function $Q_N(\cdot \mid \cdot)$ is regular if $Q_n(X \mid Y') \geq Q_N(X \mid Y)$.
+\end{definition}
+In the definition, $N$ denotes the sample size, $X$ is some random variable, $Y$ denotes the proposed parent set for $X$, and $H_N(X \mid Y)$ refers to the empirical conditional entropy based on $N$ samples of variables $X$ and $Y$.
+
+Let $X$ be a categorical random variable with $r$ possible values.
+Let $U$ denote a possible parent set with $q$ different combinations of values for the variables, and $V$ a parent set with $m$ different configurations.
+Assume that we have observed $N$ samples of $(X,U,V)$ (denoted by $x_N,u_N$ and $v_N$) and $H_N(X \mid U) \leq H_N(X \mid U \cup V)$ holds.
+
+Recall the definition of the qNML score:
+\begin{align*}
+Q^{qnml}_N(X\mid U) &= \log P(x_n \mid \hat{\theta}_{X\mid U} \ ) - \Le reg(N,rq) - reg(N,q) \Ri \\
+&= \log P(x_N \mid \hat{\theta}_{X\mid U} \ ) - \log \frac{C(N,rq)}{C(N,q)},
+\end{align*}where $C(N,r)$ is the normalizing constant the of the NML distribution for a categorical variable with $r$ possible values and sample size $N$ and $ \hat{\theta}_{X\mid U}$ denotes the maximum likelihood parameters of the conditional distribution of $X$ given $U$ which are computed from the data $(x_N,u_N)$.
+
+In order to prove the regularity, we need the following three lemmas:
+\begin{lemma}\label{Cpolynom} We can write $C(N,k)$ as a polynomial of $k$, formally
+$$
+C(N,k) = \sum_{j=1}^N a_j k^j,
+$$where $a_j > 0$. 
+\end{lemma}
+\begin{lemma}\label{MLterms}
+Assume $H_N(X \mid Y') \leq H_N(X \mid Y)$, where $Y' \subset Y$. Now $\log P(x_N \mid \hat{\theta}_{X\mid Y} \ ) = \log P(x_N \mid \hat{\theta}_{X\mid Y'} \ )$.
+\end{lemma}
+\begin{lemma}\label{increasing}
+Let $r \in \N, r \geq 2$.  The function $k \mapsto \frac{C(N,rk)}{C(N,k)}$ is increasing for every $k \geq 2$.
+\end{lemma}
+\noindent We present the proofs of these lemmas in Section \ref{lemmaproofs}.
+
+\subsection{The main proof}
 \begin{theorem}
-If $G$ consists of $C$ connected components $(G^1,\ldots,G^C)$ with
-variable sets $(V^1,\ldots,V^C)$, then $\log P_{NML}(D;G) = s^{qNML}(D;G)$
-for all data sets $D$.
+qNML score is regular.
 \end{theorem}
 \begin{proof}
-We first show that the NML-criterion for a Bayesian network
-decomposes by the connected components.
+We want to show that
+$$
+Q^{qnml}_N(X\mid U) \geq Q^{qnml}_N(X \mid U \cup V).
+$$assuming $H_N(X \mid U) \leq H_N(X \mid U \cup V)$.
+Using the entropy assumption and Lemma \ref{MLterms} implies that the maximized likelihood terms are equal.
+In order to prove the claim, it suffices to study the penalty terms, and we want to show that
+\begin{align*}
+- \Le reg(N,rq) - reg(N,q) \Ri \ &\geq \  -\Le reg(N,rqm) - reg(N,qm) \Ri \\
+\log \frac{C(N,rq)}{C(N,q)} \ &\leq \log \frac{C(N,rqm)}{C(N,qm)}.
+\end{align*}This holds, since logarithm is an increasing function, and $q \leq qm$, so we can apply Lemma \ref{increasing} to conclude the proof.
 
-Because the maximum likelihood for the data $D$ decomposes,
-we can write
-\begin{eqnarray}
-  \lefteqn{P_{NML}(D;G)} \nonumber \\
-  && = \frac{P(D;\hat\theta(D),G)}
-            {\sum_{D'_{V_1}}\ldots \sum_{D'_{V_C}}\prod_{c=1}^C P(D'_{V^c};\hat\theta(D'_{V^c}),G)} \nonumber \\
-            && = \frac{\prod_{c=1}^C P(D_{V^c};\hat\theta(D_{V^c}),G)} 
-            {\prod_{c=1}^C\sum_{D'_{V^c}}P(D'_{V^c};\hat\theta(D'_{V^c}),G)}.\nonumber \\
-            && = \prod_{c=1}^CP_{NML}(D_{V^c};G).
-\end{eqnarray}
-
-Clearly, the qNML score also decomposes by the connected components,
-so it remains to show that if the (sub)network $G$ is a tournament,
-then for any data $D$, $s^{qNML}(D;G)=\log P_{NML}(D;G)$.  Due to the
-score equivalence of the NML criterion and the qNML criterion, we may
-pick a tournament $G$ such that the linear ordering defined by $G$
-matches the ordering of the data columns, i.e., $i<j$ implies $G_i
-\subset G_j$. Now from the definition~(8) of the qNML
-criterion (in the main paper), we see that for the tournament $G$, the sum telescopes
-leaving us with $s^{qNML}(D;G) = \log P^1_{NML}(D_G;G)$, thus
-it is enough to show that $P^1_{NML}(D;G)=P_{NML}(D;G)$.  This
-follows, since for any data vector $x$ in data $D$, we have
-$P^1(x;\hat\theta(D),G) = P(x;\hat\theta(D),G)$, where $P^1$ denotes
-the model that takes $n$-dimensional vectors to be values of the
-single (collapsed) categorical variable.  Denoting prefixes of data
-vector $x$ by $x^{:i}$, and the number of times such a prefix appears
-on $N$ rows $[d_1,\ldots,d_N]$ of the $N\times n$ data matrix $D$ by
-$N_D(x^{:i})$, (so that $N_D(x^{:0})=N)$, we have
-\begin{eqnarray}
-  P(D;\hat\theta(D),G) &=& \prod_{j=1}^{N}P(d_j;\hat\theta(D),G) \nonumber \\
-  &=& \prod_{j=1}^{N} \prod_{i=1}^{n} \frac{N_D(d_j^{:i})}{N_D(d_j^{:i-1})} \nonumber\\
-  &=& \prod_{j=1}^{N} \frac{N_D(d_j^{:n})}{N} \nonumber \\
-  &=& P^1(D;\hat\theta^1(D),G).
-\end{eqnarray}
-Since both $P^1_{NML}$ and $P_{NML}$ are defined in terms of the
-maximum likelihood probabilities, the equality above implies the equality
-of these distributions.
 \end{proof}
 
-Equality established, we would like to still state the number $a(n)$
-of different $n$-node networks whose connected components are
-tournaments.  We start by generating all the $p(n)$ integer partitions
-i.e. ways to partition $n$ labelled into parts with different
-size-profiles.  For example, for $n=4$, we have $p(4)=5$ partition
-profiles $[[4], [3, 1], [2, 2], [2, 1, 1], [1, 1, 1, 1]]$. Each of
-these partition size profiles corresponds to many different networks,
-apart from the last one that corresponds just to the empty network.
-We count number of networks for one such partition size-profile (and
-then later sum these counts up).  For any such partition profile
-$(p_1,\ldots,p_k)$ we can count the ways we can assign the nodes to
-different parts and then order each part. This leads to the product
-$n\choose p_1$$p_1$!${n-p_1}\choose p_2$$p_2!$${n-p_1-p2}\choose
-p_3$$p_3!$$\ldots$${n-\sum_{j=1}^{k-1}p_j}\choose p_k$$p_k!$. However,
-the order of different parts of the same size does not matter, so for
-all groups of parts having the same size, we have to divide the
-product above by the factorial of the size of such group. Notice also,
-that the product above telescopes, leaving us a formula for OEIS
-sequence A000262\footnote{https://oeis.org/A000108} as
-described by Thomas Wieder:
+\subsection{Proofs of lemmas}\label{lemmaproofs}
+%We want to show that
+%$$
+%\quad  Q^{qnml}_N(X\mid U) \geq Q^{qnml}_N(X \mid U \cup V)
+%$$which, by using the definition of qNML, is equivalent to 
+%\begin{equation}\label{statement}
+%\log P(X \mid )
+%\end{equation}
+%The assumption about entropy implies that the maximized likelihood terms of the qnml-score are equal. 
+\setcounter{lemma}{0}
+\begin{lemma} $C(N,k)$ can be written as a polynomial of $k$, formally
+$$
+C(N,k) = \sum_{j=1}^N a_j k^j,
+$$where $a_j > 0$. 
+\end{lemma}
+\begin{proof} \cite{cosco.itsl08} derive the following representation for the normalizing constant
+\begin{align*}  
+C(N,k) = \sum_{l=0}^{N-1}\frac{(N-1)^{\underline{l}}k^{\overline{l + 1}}}{N^{l+1} \ l!} ,
+\end{align*}where $x^{\underline{l}}$ and $x^{\overline{l}}$ denote falling and rising factorials, respectively.
 
-\textit{With $p(n) =$ the number of integer partitions of $n$, $d(i)$ = the
-number of different parts of the $i^\text{th}$ partition of $n$,
-$m(i,j) =$ multiplicity of the $j^\text{th}$ part of the $i^\text{th}$
-partition of $n$, one has:}
+We utilize the fact that the rising factorial can be represented as polynomial using unsigned Stirling numbers of the first kind (see \cite{Adamchik1997}, for instance)
+\begin{align*}
+C(N,k) &=  \sum_{l=0}^{N-1}\frac{(N-1)^{\underline{l}}k^{\overline{l + 1}}}{N^{l+1} \ l!} \\
+&= \sum_{l=0}^{N-1} b_l \ k^{\overline{l+1}} \\
+&= \sum_{l=0}^{N-1} b_l \Le \sum_{j=1}^{l+1}|s(l + 1,j)| \ k^j \Ri \\
+&= \sum_{l=0}^{N-1} \Le \sum_{j=1}^{N}b_l \ |s(l + 1,j)| \ k^j \Ri \\
+&= \sum_{j=1}^{N} \Le \sum_{l=0}^{N-1} b_l \ |s(l + 1,j)| \ k^j \Ri \\
+&= \sum_{j=1}^{N} \Le \sum_{l=0}^{N-1} b_l \ |s(l + 1,j)| \Ri k^j \\
+&= \sum_{j=1}^{N} a_j  k^j,
+\end{align*}where $s(i,j)$ denotes the (signed) Stirling number of the first kind and
 $$
-a(n) = \sum_{i=1}^{p(n)} \frac{n!}{\prod_{j=1}^{d(i)} m(i,j)!}.
+a_j = \Le \sum_{l=0}^{N-1} \frac{(N-1)^{\underline{l}}}{N^{l+1}l!} \ |s(l + 1,j)| \Ri,
+$$ $a_j > 0$ for all $j$. On the second row, we denoted $b_l = (N-1)^{\underline{l}}/(N^{l+1} l!)$. On the row 4, we used the property of Stirling numbers: $s(i,j) = 0$ for all $j > i$. 
+\end{proof}
+\begin{lemma}
+Assume $H_N(X \mid Y') \leq H_N(X \mid Y)$, where $Y' \subset Y$. Now $\log P(x_N \mid \hat{\theta}_{X\mid Y} \ ) = \log P(x_N \mid \hat{\theta}_{X\mid Y'})$.
+\end{lemma}
+\begin{proof}
+We can write the logarithm of the maximized likelihood, \\ $\log P(x_N \mid \hat{\theta}_{X\mid Y} \ )$, as follows \citep{KOLLER}
+\begin{align*}
+\log P(x_N \mid \hat{\theta}_{X\mid Y} \ ) & = -N \Le H_N(X) - I_N(X;Y) \Ri \\
+&= -N \  H_N(X\mid Y),
+\end{align*}where $I_N(\cdot;\cdot)$ is the empirical mutual information. This implies that the assumption
 $$
-For example, $a(4)=\frac{4!}{1!}+\frac{4!}{1!1!}+\frac{4!}{2!}+\frac{4!}{1!2!}+\frac{4!}{4!}=73$. In general this sequence grows rapidly; $1, 1, 3, 13, 73, 501, 4051, 37633, 394353, 4596553, \ldots$.
+H_N(X \mid Y') \leq H_N(X \mid Y)
+$$ is equivalent to
+$$
+\log P(x_N \mid \hat{\theta}_{X\mid Y} \ ) \geq \log P(x_N \mid \hat{\theta}_{X\mid Y'} \ ) .
+$$Actually we must have the equality holding in the above expression, since
+$$
+H_N(X \mid Y') < H_N(X \mid Y)
+$$ would imply that
+$$
+I_N(X ; Z \mid Y ) < 0,
+$$where $Z = Y \setminus Y'$, which is impossible.
+\end{proof}
+\begin{lemma}
+Let $r \in \N, r \geq 2$.  The function $k \mapsto \frac{C(N,rk)}{C(N,k)}$ is increasing for every $k \geq 2$.
+\end{lemma}
+
+\begin{proof}
+Lemma \ref{Cpolynom} lets us to write
+\begin{equation}\label{C1}
+C(N,k) = \sum_{j=1}^{N} a_j k^j
+\end{equation}and, similarly,
+\begin{equation}\label{C2}
+C(N,rk) = \sum_{j=1}^{N} a_j  r^jk^j.
+\end{equation}
+
+From this, it it easy see that the derivative of the quotient, \\ $d/dk (C(N,rk)/C(N,k))$, will be a ratio of two polynomials of $k$.
+Our goal is to show that the polynomial in the numerator has positive coefficients, which will guarantee the positivity of derivative for every $k > 0$, and thus imply that the original function is increasing (polynomial in the denominator is squared and non-zero for $k > 0$, so it can be ignored).
+
+Derivatives of (\ref{C1}) and (\ref{C2}) are obtained easily:
+\begin{align*}
+\frac{d}{dk}C(N,k) &= \sum_{j=1}^{N}ja_jk^{j-1} \\
+&= \sum_{j=0}^{N-1}(j+1) a_{j+1}k^{j}
+\end{align*}and
+\begin{align*}
+\frac{d}{dk}C(N,rk) &= \sum_{j=1}^{N}ja_jr^jk^{j-1} \\
+&= \sum_{j=0}^{N-1}(j+1)a_{j+1}r^{j+1}k^{j}. 
+\end{align*}Consider next the products found in the derivative of the quotient. We obtain
+\begin{align*}
+\Le\frac{d}{dk}C(N,rk)\Ri C(N,k) &= \Le \sum_{j=0}^{N-1}(j+1)a_{j+1}r^{j+1}k^{j} \Ri \Le \sum_{l=1}^{N} a_l  k^l \Ri \\
+&= \sum_{i = 1}^{2N-1} \Le \sum_{j+l = i} (j+1)a_{j+1}r^{j+1}a_l  \Ri k^i
+\end{align*}and
+%\begin{align*}
+%\Le\frac{d}{dk}C(n,k)\Ri C(n,rk) &= \Le \sum_{j=0}^{n-1}(j+1)a_{j+1}k^{j} \Ri \Le %\sum_{l=1}^{n} a_l  r^lk^l  \Ri \\
+%&= \sum_{i = 1}^{2n-1} \Le \sum_{m=0}^i (m+1)a_{m+1}a_{i-m}r^{i-m}   \Ri k^i
+%\end{align*}
+\begin{align*}
+\Le\frac{d}{dk}C(N,k)\Ri C(N,rk) &= \Le \sum_{j=0}^{N-1}(j+1)a_{j+1}k^{j} \Ri \Le \sum_{l=1}^{N} a_l  r^lk^l  \Ri \\
+&= \sum_{i = 1}^{2N-1} \Le \sum_{j+l = i}(j+1)a_{j+1}a_l  r^l   \Ri k^i.
+\end{align*}
+Subtracting these two expression yields
+\begin{align*}
+&\Le\frac{d}{dk}C(N,rk)\Ri C(N,k)-\Le\frac{d}{dk}C(N,k)\Ri C(N,rk) \\&= \sum_{i = 1}^{2N-1} \Le \sum_{j+l = i} (j+1)a_{j+1}r^{j+1}a_l  \Ri k^i - \sum_{i = 1}^{2N-1} \Le \sum_{j+l = i}(j+1)a_{j+1}a_l  r^l   \Ri k^i \\
+&= \sum_{i = 1}^{2N-1} \Le \sum_{j+l = i}(j+1)a_{j+1}a_l  (r^{j+1}- r^l)   \Ri k^i
+\end{align*}which is the polynomial in the numerator of the derivative of $C(N,rk)/C(N,k)$.
+Next, we study the coefficient of $k^i$, if $i \leq N$
+\begin{align*}
+\sum_{j+l = i}(j+1)a_{j+1}a_l  (r^{j+1}- r^l) &= \sum_{l=1}^{i}(i-l+1)a_{i-l+1}a_l  (r^{i-l+1}- r^l) \\
+& = \sum_{l=1}^{i}(i-l+1)c_l \\
+&= \sum_{k =1}^{\lfloor i / 2 \rfloor}(i-k+1)c_k + (i-(i-k + 1)+1)c_{i-k+1} \\
+&=  \sum_{k =1}^{\lfloor i / 2 \rfloor}(i-k+1)c_k + kc_{i-k+1} \\
+&=  \sum_{k =1}^{\lfloor i / 2 \rfloor}(i-k+1)c_k - kc_{k} \\
+&= \sum_{k =1}^{\lfloor i / 2 \rfloor}(i-2k+1)c_k.
+\end{align*} On the first row, we re-wrote sum using only one running index.
+On the second row we denoted $c_l =a_{i-l+1}a_l  (r^{i-l+1}- r^l)$.
+On the third row, we re-arranged the sum so that we are summing over pairs of terms of the original sum: the first and the last term, the second and the second to last, and so on.
+This resulting sum has $\lfloor i / 2 \rfloor$ terms.
+We have to use the floor-function since if $i$ is odd, there exists an index $l'$ in the original sum such that $r^{i-l'+1}-r^{l'} = 0$.
+On the fifth row, we make use of the identity $c_k = -c_{i-k+1}$ which is straightforward to verify.
+From the last row, we can observe that every term of the sum is positive since $i-2k+1$ and $r^{i-k+1}- r^k$ are both positive if $k \leq (i+1)/2$ which holds since $k$ ranges from $1$ to $\lfloor i / 2 \rfloor$.
+
+Let us now consider the situation where $n < i \leq 2N-1$. We start with the special case where $i = 2N-1$. Then, we have only one term in the sum
+\begin{align*}
+\sum_{j+l = i}(j+1)a_{j+1}a_l  (r^{j+1}- r^l) &= \sum_{l=N}^{N}(2N-1-l+1)a_{2N-1-l+1}a_l  (r^{2N-1-l+1}- r^l) \\
+&= Na_{N}a_N  (r^{N}- r^N) \\
+& = 0.
+\end{align*}Now, let $N < i < 2N-1$, we follow a similar procedure as before to manipulate the sum
+\begin{align*}
+\sum_{j+l = i}(j+1)a_{j+1}a_l  (r^{j+1}- r^l) &= \sum_{l=i-N + 1}^{N}(i-l+1)a_{i-l+1}a_l  (r^{i-l+1}- r^l) \\
+& = \sum_{l=i-N + 1}^{N}(i-l+1)c_l \\
+& = \sum_{k=1}^{2N-i}(N-k+1)c_{i-N+k} \\
+& = \sum_{k=1}^{\lfloor N - i/2 \rfloor}(N-k+1)c_{i-N+k}\\
+&+  (N-(2N-i-k+1)+1)c_{i-N+(2N-i-k+1)} \\
+&= \sum_{k=1}^{\lfloor N - i/2 \rfloor}(N-k+1)c_{i-N+k} +  (i-N+k)c_{N-k+1} \\
+&= \sum_{k=1}^{\lfloor N - i/2 \rfloor}(N-k+1)c_{i-N+k} -  (i-N+k)c_{i-N+k} \\
+&= \sum_{k=1}^{\lfloor N - i/2 \rfloor}(N-k+1-(i-N+k))c_{i-N+k} \\
+&= \sum_{k=1}^{\lfloor N - i/2 \rfloor}(2N-i-2k+1)c_{i-N+k}.
+\end{align*}It is now easy to verify that $(2N-i-2k+1)$ and $c_{i-N+k}$ are positive if $k \leq N - (i-1)/2$ which holds since $k$ ranges from $1$ to $\lfloor N - i/2 \rfloor$.
+The floor function is again used when we sum over pairs of terms since if $i$ is odd there is zero-term.
+Since all the coefficients are non-negative and the $k \geq 2$, the derivative is positive. This implies that the original function is increasing.
+\end{proof}
+
+\subsection{fNML is Regular}
+In this section, we will show that fNML score is regular. Recall first the definition of fNML local score
+\begin{equation}
+Q^{fnml}_N(X  \vert \ U) = \log P(x_n \mid \hat{\theta}_{X\mid U} \ ) - \sum_{j = 1}^{r_U} reg(N_j,r_X),
+\end{equation}where $r_X$ is the number of categories for $X$, $r_U$ denotes the number of possible configurations of variables in $U$, and $N_j$ is the number of times the $j$:th configuration is observed in our samples $u_N$.
+Note that $reg(r_X, N_j) = 0$ for any configuration not observed in our sample.
+fNML criterion differs from qNML only by how the penalty term is defined. We can follow a highly similar strategy in order to prove the regularity for fNML. To this end, we need the following Lemma:
+\begin{lemma}\label{regineq}
+Let $N = N_1 + N_2$ and $r \in\N, r > 2$. Now,
+$$
+reg(N,r) \leq reg(N_1,r) + reg(N_2,r).
+$$
+\end{lemma}
+\begin{proof}
+We start with the definition of $C(N_1,r) = \exp(reg(N_1,r))$,
+\begin{align*}
+C(N_1,r)C(N_2,r) &= \sum_{x_{N_1}}P(x_{N_1} \vert \hat{\theta}(x_{N_1} ))\sum_{x_{N_2}}P(x_{N_2} \vert \hat{\theta}(x_{N_2} ))\\ & \geq
+\sum_{x_{N_1}}P(x_{N_1} \vert \hat{\theta}(x_{N} ))\sum_{x_{N_2}}P(x_{N_2} \vert \hat{\theta}(x_{N} )) \\ &=
+\sum_{x_{N}}P(x_{N} \vert \hat{\theta}(x_{N} )) \\ & = C(N,r).
+\end{align*}Taking the logarithm on the both sides yields our claim.
+On the second row, we used the definition of maximum likelihood parameters: the probability of data vector $x_{N_1}$ is maximized under parameters $\hat{\theta}(x_{N_1})$, and if we use different parameters, $\hat{\theta}(x_{N_1})$, probability can only go down or stay the same.
+On the third row, we used the i.i.d, assumption, which allows us to take the single sum over data vectors of length $N = N_1 + N_2$.
+\end{proof}
+Now we can proceed to the actual proof.
+\begin{theorem}
+fNML score is regular.
+\end{theorem}
+\begin{proof}
+Using the entropy assumption and Lemma \ref{C2}, we can ignore the maximized likelihood terms and study the penalty terms. We want show that
+\begin{equation}\label{fnmlPenalty}
+\sum_{j = 1}^{r_{Y'}} reg(N'_j,r_X) \leq \sum_{l = 1}^{r_{Y}} reg(N_l,r_X),
+\end{equation}where we used $r_{Y'}$ and $r_Y$ to denote the numbers of possible configurations for variables in $Y'$ and $Y$, respectively.
+Also, $N_j$ refers to the number times the $j$:th configuration of variables $Y$ is observed in our sample, and $N'_l$ denotes the same for $Y'$.
+
+Now, since $r_{Y'} \leq r_Y$, and we know that $\sum_j N'_j = \sum_l N_l = N$, we can just apply Lemma (\ref{regineq}) multiple times to conclude our proof. 
+\end{proof}

### 2017-10-13 experiments.tex
Add note on BDeu ESS

@@
 
-We empirically compare the capacity of qNML to that of BIC, BDeu and
+We empirically compare the capacity of qNML to that of BIC, BDeu ($\alpha = 1$) and
 fNML in identifying the data generating structures, and producing

### 2017-10-13 conclusion.tex
Typos

@@
 equals the cost of the current state-of-the-art criteria. The
-criterion also gives equal scores for models that encode same
+criterion also gives equal scores for models that encode the same
 independence hypotheses about the joint probability distribution.
-qNML also coincides with NML criterion for many models.
+qNML also coincides with the NML criterion for many models.
 In our experiments, the qNML criterion appears as a safe choice for a

### 2017-10-13 supp_mat_B.tex
Typos

@@
 \sum_{x_{N}}P(x_{N} \vert \hat{\theta}(x_{N} )) \\ & = C(N,r).
-\end{align*}Taking the logarithm on the both sides yields our claim.
-On the second row, we used the definition of maximum likelihood parameters: the probability of data vector $x_{N_1}$ is maximized under parameters $\hat{\theta}(x_{N_1})$, and if we use different parameters, $\hat{\theta}(x_{N_1})$, probability can only go down or stay the same.
-On the third row, we used the i.i.d, assumption, which allows us to take the single sum over data vectors of length $N = N_1 + N_2$.
+\end{align*}Taking the logarithm on both sides yields our claim.
+On the second row, we used the definition of maximum likelihood parameters: the probability of the data vector $x_{N_1}$ is maximized under parameters $\hat{\theta}(x_{N_1})$, and if we use different parameters, $\hat{\theta}(x_{N_1})$, the probability can only go down or stay the same.
+On the third row, we used the i.i.d., assumption, which allows us to take the single sum over data vectors of length $N = N_1 + N_2$.
 \end{proof}

### 2017-10-13 experiments.tex
typos

@@
 selection criteria, we used the sequential predictive NML (sNML)
-parametrization $\theta_{ijk}\propto e(n_{ijk})(n_{ijk}+1)$, where
+parametrization $\theta_{ijk}\propto e(N_{ijk})(N_{ijk}+1)$, where
 $e(n)=(\frac{n+1}{n})^n$ as suggested in~\cite{Riss07b}.
@@
 The graphs for different sample sizes for both predictive accuracy and
-the number of parameters can be found in the Supplementary Material.
+the number of parameters can be found in Appendix D in the Supplementary Material.
 

### 2017-10-13 qnml.tex
small typos
typos

@@
 We will next state this as a
-theorem and sketch a proof for it. A more detailed proof appears in Appendix A the Supplementary Material.
+theorem and sketch a proof for it. A more detailed proof appears in Appendix A in the Supplementary Material.
 \begin{theorem}
@@
 
-\subsection{qNML Is Consistent}
+\subsection{qNML is Consistent}
 
@@
 
-\subsection{qNML Is Regular}
+\subsection{qNML is Regular}\label{sec:regularity}
 
@@
 \end{proof}
+As fNML criterion differs from qNML only by how the penalty term is defined, we obtain the following result with little extra work:
+\begin{theorem}
+fNML score is regular.
+\end{theorem}
+\begin{proof}
+The proof is given in Appendix B in Supplementary Material.
+\end{proof}

### 2017-10-13 experiments.tex
Added a comment about BDq to supp mat.

@@
 The graphs for different sample sizes for both predictive accuracy and
-the number of parameters can be found in Appendix D in the Supplementary Material.
+the number of parameters can be found in Appendix E in the Supplementary Material.
 

### 2017-10-13 qnml.tex
added a reference for BDq

@@
 \end{proof}
+
+Suzuki \cite{Suzuki2017} independently introduces a Bayesian Dirichlet
+quotient (BDq) score, that can also be shown to be score equivalent
+and regular.  However, like BDeu, this score features a single
+hyperparameter $\alpha$, and our initial studies suggest that BDq is
+also very sensitive to this hyperparameter (see Appendix D in the
+Supplementarty Material), the issue that was one of the main
+motivations to develop a parameter-free model selection criterion like
+qNML.

### 2017-11-23 AnswerToReviewers.txt
answer to reviewers (draft)

We thank reviewers for their feedback. While there appears to be a big spread in overall ratings, we are happy to see that the reviews are unanimously positive about the theoretical properties of the proposed model selection criterion, qNML, that is decomposable (unlike NML), consistent, score equivalent (unlike fNML), parameter free (unlike BDeu or BDq), regular (unlike BDeu) and possibly a reasonable approximation to the NML criterion.

Our original motivation for looking for a new model selection criterion was, however, extremely practical. Most of our "clients" want to use Bayesian networks due to their intuitive interpretability, and for the insight Bayesian networks provide to the domain. Unfortunately, that is where existing methods too often appear to fail us. To practitioner's disappointment, BIC often yields an empty network that is hard to sell, say, to a social scientist that has spent months to painstakingly collect his/her (often not very big) data set. Unfortunately, our experiments using Structural Hamming Distance (SHD) does not fully reveal this situation, since the empty network can on average still be relatively competitive in terms of SHD. 
Naturally, providing very "false insight" is also undesirable so the models should also be reasonably good in prediction. To our experience, the  qNML is closest to strike a good balance for providing insight while still being predictively well performing, the fact that for example reviewer 1 appears to have appreciated. While probably chosen for communicative clarity, we feel that 
"it does not work in practice" by reviewer 2 is not a best possible summary of the practical value of qNML, which unlike other methods(!), never performs very poorly.

Clearly, none of the criteria perform the best for all sample sizes and purposes (prediction vs. SHD), which might even suggest picking different criteria for different situations and needs. We like, like reviewer 3, feel that in this situation, proposing a theoretically well behaved and empirically validated criterion is indeed an interesting contribution to the field.

Reviewer 1 gave some detailed comments that are easy to take into account in the final version of the paper. In the longer version of the paper, the comparison with the BDq, as suggested by reviewer 3, is a natural thing to do. The issue with BDq is that it features a hyper-parameter and our preliminary results indicate that BDq can be very sensitive to this parameter. Therefore, for more meaningful comparison, the not yet thoroughly studied parameter sensitivity issue must also be addressed. For including it to this conference paper might be to much of a squeeze.

### 2017-11-27 AnswerToReviewers.txt
pointed out that qNML is the best in SHD for bigger sample sizes.

@@
 While probably chosen for communicative clarity, we feel that
-"it does not work in practice" by reviewer 2 is not a best possible summary of the practical value of qNML, which unlike other methods(!), never performs very poorly.
+"it does not work in practice" by reviewer 2 is not a best possible summary of the practical value of qNML, since it appears as a best method in SHD (see Figure 3) from medium sample sizes up.
+Furthermore, unlike BIC and BDeu, it seldom performs very poorly, and, like pointed out by the reviewer 1, it has both theoretical and practical advantages over the other robust criterion, fNML.
 
@@
 SHD), which might even suggest picking different criteria for different situations and needs.
-We like, like reviewer 3, feel that in this situation, proposing a theoretically well behaved and empirically validated criterion is indeed an interesting contribution to the field.
+We like, like reviewer 3, feel that in this situation, proposing a theoretically well behaved and empirically validated criterion is indeed an interesting contribution to the field - a contribution that the Bayesian network learning community should know about.
 

### 2017-11-27 AnswerToReviewers.txt
some comments on empirical results, order of chapters changed

@@
 While there appears to be a big spread in overall ratings, we are happy to see that the reviews are unanimously positive about the theoretical properties of the proposed model selection criterion, qNML, that is decomposable (unlike NML), consistent, score equivalent (unlike fNML), parameter free (unlike BDeu or BDq), regular (unlike BDeu) and possibly a reasonable approximation to the NML criterion.
-
-Our original motivation for looking for a new model selection criterion was, however, extremely practical.
-Most of our "clients" want to use Bayesian networks due to their intuitive interpretability, and for the insight Bayesian networks provide to the domain.
-Unfortunately, that is where existing methods too often appear to fail us.
-To practitioner's disappointment, BIC often yields an empty network that is hard to sell, say, to a social scientist that has spent months to painstakingly collect his/her (often not very big) data set.
-Unfortunately, our experiments using Structural Hamming Distance (SHD) does not fully reveal this situation, since the empty network can on average still be relatively competitive in terms of SHD.
-Naturally, providing very "false insight" is also undesirable so the models should also be reasonably good in prediction.
-To our experience, the  qNML is closest to strike a good balance for providing insight while still being predictively well performing, the fact that for example reviewer 1 appears to have appreciated.
-While probably chosen for communicative clarity, we feel that
-"it does not work in practice" by reviewer 2 is not a best possible summary of the practical value of qNML, since it appears as a best method in SHD (see Figure 3) from medium sample sizes up.
-Furthermore, unlike BIC and BDeu, it seldom performs very poorly, and, like pointed out by the reviewer 1, it has both theoretical and practical advantages over the other robust criterion, fNML.
-
-Clearly, none of the criteria perform the best for all sample sizes and purposes (prediction vs.
-SHD), which might even suggest picking different criteria for different situations and needs.
-We like, like reviewer 3, feel that in this situation, proposing a theoretically well behaved and empirically validated criterion is indeed an interesting contribution to the field - a contribution that the Bayesian network learning community should know about.
 
@@
 Therefore, for more meaningful comparison, the not yet thoroughly studied parameter sensitivity issue must also be addressed.
-For including it to this conference paper might be to much of a squeeze.
+For including it to this conference paper might be too much of a squeeze.
+
+Our original motivation for looking for a new model selection criterion was extremely practical.
+Most of our "clients" want to use Bayesian networks due to their intuitive interpretability, and for the insight Bayesian networks provide to the domain.
+Unfortunately, that is where existing methods too often appear to fail us.
+To practitioner's disappointment, BIC often yields an empty network that is hard to sell, say, to a social scientist that has spent months to painstakingly collect his/her (often not very big) data set.
+Unfortunately, our experiments using Structural Hamming Distance (SHD) does not fully reveal this situation, since the empty network can on average still be relatively competitive in terms of SHD.
+
+Naturally, providing very "false insight" is also undesirable so the models should also be reasonably good in prediction.
+To our experience, the  qNML is closest to strike a good balance for providing insight while still being predictively well performing, the fact that for example reviewer 1 appears to have appreciated.
+While probably chosen for communicative clarity, we feel that
+"it does not work in practice" by reviewer 2 is not a best possible summary of the practical value of qNML, since it appears as a best method in SHD (see Figure 3) from medium sample sizes up.
+Furthermore, taking a closer look at the empirical results we can see the following:
+
+- based on the average rank, qNML is better in prediction than BIC which produces the worst predictions on large sample size settings.
+- qNML obtains always a better rank than fNML in structure learning experiments (see Figure 3).
+When averaged over all the sample sizes, fNML obtains the best rank in prediction with qNML being the second to best.
+However, the standard deviation of fNML's rank is 0.76 which is around 77% larger than qNML's (0.43).
+Also, as Table 3 suggests, qNML obtains this competitive performance in prediction usually with considerably simpler models when compared to fNML.
+- BDeu is bested by all the other methods nearly always in prediction.
+In structure learning, there are couple settings with moderate sample sizes where BDeu has a better average rank than qNML but otherwise qNML outperforms BDeu.
+
+Clearly, none of the criteria perform the best for all sample sizes and purposes (prediction vs.
+SHD), which might even suggest picking different criteria for different situations and needs.
+We, like reviewer 3, feel that in this situation, proposing a theoretically well behaved and empirically validated criterion is indeed an interesting contribution to the field - a contribution that the Bayesian network learning community should know about.

### 2017-11-28 AnswerToReviewers.txt
Teemu's input

--- 2017-11-23 AnswerToReviewers.txt
@@
 We thank reviewers for their feedback.
-While there appears to be a big spread in overall ratings, we are happy to see that the reviews are unanimously positive about the theoretical properties of the proposed model selection criterion, qNML, that is decomposable (unlike NML), consistent, score equivalent (unlike fNML), parameter free (unlike BDeu or BDq), regular (unlike BDeu) and possibly a reasonable approximation to the NML criterion.
+While there appears to be some spread in overall ratings, we are happy to see that the reviews are unanimously positive about the theoretical properties of the proposed model selection criterion, qNML, that is decomposable (unlike NML), consistent, score equivalent (unlike fNML), parameter free (unlike BDeu or BDq), regular (unlike BDeu) and possibly a reasonable approximation to the NML criterion.
 
-Our original motivation for looking for a new model selection criterion was, however, extremely practical.
-Most of our "clients" want to use Bayesian networks due to their intuitive interpretability, and for the insight Bayesian networks provide to the domain.
-Unfortunately, that is where existing methods too often appear to fail us.
-To practitioner's disappointment, BIC often yields an empty network that is hard to sell, say, to a social scientist that has spent months to painstakingly collect his/her (often not very big) data set.
-Unfortunately, our experiments using Structural Hamming Distance (SHD) does not fully reveal this situation, since the empty network can on average still be relatively competitive in terms of SHD.
-Naturally, providing very "false insight" is also undesirable so the models should also be reasonably good in prediction.
-To our experience, the  qNML is closest to strike a good balance for providing insight while still being predictively well performing, the fact that for example reviewer 1 appears to have appreciated.
-While probably chosen for communicative clarity, we feel that
-"it does not work in practice" by reviewer 2 is not a best possible summary of the practical value of qNML, which unlike other methods(!), never performs very poorly.
+Reviewer 1 gave some detailed comments that are easy to take into account in the final version of the paper.
+In a journal version of the paper, the comparison with the BDq, as suggested by reviewer 3, would be a natural thing to do.
+The issue with BDq is that it features a hyper-parameter and our preliminary results indicate that like BDeu, BDq too can be extremely sensitive to the choice of the hyperparameter.
+Therefore, for more meaningful comparison, the not yet thoroughly studied parameter sensitivity issue with BDq should also be addressed.
+
+About the practicality of the proposed method: Our original motivation for looking for a new model selection criterion was actually extremely practical.
+We have quite a lot of experience of working with users (researchers in applied fields), and they want to use Bayesian networks due to the insight Bayesian networks provide into their domain.
+Unfortunately, that is where the existing methods too often appear to fail.
+For example, to the practitioners' disappointment, BIC often yields an empty network.
+This is not useful at all, say, to a social scientist that has spent months to painstakingly collect his/her (often not very big) data set.
+Our experiments using Structural Hamming Distance (SHD) do not fully reveal this situation, since the empty network can actually be relatively competitive in terms of SHD.
+
+Obviously, providing more but misleading "insight" is even worse.
+So the models should also capture the underlying distribution, which we suggest to measure by prediction performance.
+What we are particularly happy about is that qNML finds a good balance by providing insight while still performing well in predictive tasks -- as pointed out by reviewer 1.
+
+Based on these points, "it does not work in practice" (reviewer 2) is perhaps not really an accurate summary of the practicality of qNML.
+Besides the above points about balanced performance, it is the best method in SHD (see Figure 3) from medium sample sizes up.
+
+Further remarks about the empirical results:
+
+- based on the average rank, qNML is better in prediction than BIC which produces the worst predictions on large sample size settings (Table 2)
+- qNML obtains always a better rank than fNML in structure learning experiments (Fig. 3).
+- when averaged over all the sample sizes, fNML obtains the best rank in prediction with qNML being the second to best.
+However, the standard deviation of fNML's rank is 0.76 which is larger than qNML's (0.43) (data from Table 2; averages and variances not shown explicitly in the paper)
+- qNML obtains this competitive performance in prediction with considerably simpler models when compared to fNML (Table 3).
+- BDeu is outperformed by all the other methods in almost all prediction experiments (Table 2).
+- In structure learning, qNML outperforms BDeu except in a few settings with moderate sample sizes (Fig. 3).
 
@@
 SHD), which might even suggest picking different criteria for different situations and needs.
-We like, like reviewer 3, feel that in this situation, proposing a theoretically well behaved and empirically validated criterion is indeed an interesting contribution to the field.
-
-Reviewer 1 gave some detailed comments that are easy to take into account in the final version of the paper.
-In the longer version of the paper, the comparison with the BDq, as suggested by reviewer 3, is a natural thing to do.
-The issue with BDq is that it features a hyper-parameter and our preliminary results indicate that BDq can be very sensitive to this parameter.
-Therefore, for more meaningful comparison, the not yet thoroughly studied parameter sensitivity issue must also be addressed.
-For including it to this conference paper might be to much of a squeeze.
+We, like reviewer 3, feel that in this situation, proposing a theoretically well behaved and empirically robust criterion is indeed an interesting contribution to the field - a contribution that the Bayesian network learning community is likely to be interested to know about.

### 2018-02-01 shd_plots.py
typo in axis label fixed

@@ -223,7 +223,7 @@ def rank_shd_from_single_tests(folder = "results/real_dag_shd/", save = False):
         ax.plot(ns,avg_ranks[:,ii],linestyles[ii],label = methods[ii], linewidth = 6)
             
       
-    ax.set_ylabel("Avgerage rank")
+    ax.set_ylabel("Average rank")
     ax.set_xlabel("Sample size")    
     
     plt.legend(loc = "best")
@@ -253,10 +253,10 @@ def score_to_rank(x):
     return(ranks)
     
 if __name__ == "__main__":
-    save = False
+    save = True
     folder = "results/real_dag_shd/"
-    subplots(folder = folder, save = save)
+    #subplots(folder = folder, save = save)
     #rank_shd(folder = folder, save = save)
-    #rank_shd_from_single_tests(folder = folder, save = save)
+    rank_shd_from_single_tests(folder = folder, save = save)
     
     

### 2018-02-01 poster/qnmlPoster.tex
add poster template

\documentclass[final]{beamer}

\mode<presentation>
{
  \usetheme{hiitposter}
}
%\graphicspath{{imgs/}}
\usepackage{amsmath,amsthm, amssymb,mathrsfs}
\usepackage{color}
\usepackage{times}
\usepackage{subfigure}
\usepackage{bm}
\usepackage{url}
%\usepackage[orientation=portrait,size=a0,scale=1.4,debug]{beamerposter}        
\usepackage[orientation=portrait,width = 91 ,height = 122 ,scale=1.39,debug]{beamerposter} 
%\setbeamerfont{itemize/enumerate body}{size={\fontsize{25}{31}}}
%\setbeamerfont{itemize/enumerate subbody}{size=\footnotesize}

\graphicspath{{../qNML_images/}}

\def\newblock{\hskip .11em plus .33em minus 3.07em}

\definecolor{DRed}{rgb}{0.8,0.6,0.6}
\definecolor{DGreen}{rgb}{0.6,0.7,0.6}
\definecolor{BRed}{rgb}{0.7,0,0}
\definecolor{BGreen}{rgb}{0,0.7,0}

\newcommand{\mechanism}{\mathcal{M}}
\newcommand{\dataset}{\mathcal{D}}

\newcommand{\myred}[1]{{\color{BRed}#1}}
\newcommand{\mygreen}[1]{{\color{BGreen}#1}}

\newcommand{\heading}[1]{\alert{\large #1}\\}
\definecolor{myPurple}{RGB}{174, 206, 195}

\usepackage{textpos}
\usepackage{fancybox}
\usepackage{tikz}
\usepackage{mathtools}
\usepackage{empheq}

\theoremstyle{plain}
\newtheorem{assumption}{Assumption}

\newcommand{\vOmega}{\bm{\Omega}}
\newcommand{\bBeta}{\bm{\beta}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\vA}{\bm{A}}
\newcommand{\vx}{\bm{x}}
\newcommand{\vy}{\bm{y}}
\definecolor{myblue}{rgb}{.85, .85, 1.0}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newlength\mytemplen
\newsavebox\mytempbox

\makeatletter
\newcommand\mybluebox{%
    \@ifnextchar[%]
       {\@mybluebox}%
       {\@mybluebox[0pt]}}

\def\@mybluebox[#1]{%
    \@ifnextchar[%]
       {\@@mybluebox[#1]}%
       {\@@mybluebox[#1][0pt]}}

\def\@@mybluebox[#1][#2]#3{
    \sbox\mytempbox{#3}%
    \mytemplen\ht\mytempbox
    \advance\mytemplen #1\relax
    \ht\mytempbox\mytemplen
    \mytemplen\dp\mytempbox
    \advance\mytemplen #2\relax
    \dp\mytempbox\mytemplen
    \colorbox{myblue}{\hspace{1em}\usebox{\mytempbox}\hspace{1em}}}

\makeatother
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\makeatletter
\newcommand\mathcircled[1]{%
  \mathpalette\@mathcircled{#1}%
}
\newcommand\@mathcircled[2]{%
  \tikz[baseline=(math.base)] \node[red,draw,circle,inner sep=1pt] (math) {$\m@th#1#2$};%
}
\makeatother

\title{Quotient Normalized Maximum Likelihood Criterion for Learning Bayesian Network Structures}
\author{Tomi Silander$^1$, Janne Lepp{\"a}-aho$^2$, Elias J{\"a}{\"a}saari$^2$, and Teemu Roos$^2$}
\institute{$^{1}$ NAVER LABS Europe, France \\ 
$^{2}$ HIIT / Department of Computer Science, University of Helsinki, Finland}

\begin{document}
\begin{frame}{}
\vskip-1.0ex

{
\setbeamercolor{block body}{bg=myPurple}
\begin{block}{Abstract}
	\large
  	We introduce an information theoretic criterion for Bayesian network
  	structure learning which we call quotient normalized maximum
	likelihood (qNML). In contrast to the closely related factorized
	normalized maximum likelihood criterion, qNML satisfies the property
	of score equivalence. It is also decomposable and completely free
	of adjustable hyperparameters. For practical computations, we identify
	a remarkably accurate approximation proposed earlier by Szpankowski
	and Weinberger. Experiments on both simulated and real data
	demonstrate that the new criterion leads to parsimonious models with
	good predictive accuracy.
\end{block}
}

\begin{block}{Section 1}
\begin{columns}[T]
   \begin{column}{0.3\textwidth} % first column {{{
     \heading{Subsection 1}

   \end{column}

   \begin{column}{0.005\textwidth}\linethickness{0.3ex} % separator {{{
      \color{myPurple} \line(0,1){400}
   \end{column} % }}}

   \begin{column}{0.3\textwidth} % second column {{{
     \heading{Subsection 2}
   \end{column}

   \begin{column}{0.005\textwidth}\linethickness{0.3ex} % separator {{{
      \color{myPurple} \line(0,1){400}
   \end{column} % }}}
   \begin{column}{0.3\textwidth}% third column {{{
     \heading{Subsection 3}

   \end{column} % end of first column }}}
\end{columns}
\end{block}

\begin{block}{Section 2}
\begin{columns}[T]
   \begin{column}{0.25\textwidth} % first column {{{
	\heading{Subsection 1}

   \end{column}

   \begin{column}{0.005\textwidth}\linethickness{0.3ex} % separator {{{
      \color{myPurple} \line(0,1){650}
   \end{column} % }}}

   \begin{column}{0.3\textwidth} % second column {{{
	\heading{Subsection 2}

   \end{column}

   \begin{column}{0.005\textwidth}\linethickness{0.3ex} % separator {{{
      \color{myPurple} \line(0,1){650}
   \end{column} % }}}

   \begin{column}{0.3\textwidth}% third column {{{

      \heading{Subsection 3}

   \end{column} % end of first column }}}
   % \begin{column}{0.005\textwidth}\linethickness{0.3ex} % separator {{{
   %    \color{myPurple} \line(0,1){600}
   % \end{column} % }}}
   % \begin{column}{0.3\textwidth} % third column {{{
   % \end{column} % end of third column }}}

\end{columns}
\end{block}

\begin{block}{Section 3}
  \begin{columns}[T]
    \begin{column}{0.35\textwidth}
      \heading{Subsection 1}

    \end{column}
   \begin{column}{0.005\textwidth}\linethickness{0.3ex}
      \color{myPurple} \line(0,1){800}
   \end{column} % }}}

    \begin{column}{0.23\textwidth}
    \heading{Subsection 2}

    \end{column}

    \begin{column}{0.005\textwidth}\linethickness{0.3ex} % separator {{{
      \color{myPurple} \line(0,1){800}
   \end{column} % }}}
    \begin{column}{0.3\textwidth}
    \heading{Subsection 3}

    \end{column}
  \end{columns}
\end{block}

\vskip-1.0ex
\begin{columns}[T]
  \begin{column}{0.63\paperwidth}
\begin{block}{References} %{{{
%\scriptsize
\tiny
{
[1] 
}
\end{block} %}}}
  \end{column}
  \begin{column}{0.33\paperwidth}
\begin{block}{Acknowledgements} %{{{
%\scriptsize
\tiny
{
..........
}
\end{block} %}}}
  \end{column}
\end{columns}
\end{frame}
\end{document}

### 2018-02-12 bn.tex
ref to right appendix
address minor comment

@@
 
-Bayesian learning of Bayesian network structures is based on the
+Score-based Bayesian learning of Bayesian network structures evaluates the goodness of different structures $G$ using their
 posterior probability $P(G|D,\alpha)$, where $\alpha$ denotes the
@@
 In addition, we provide a proof of regularity for fNML criterion, which has not appeared in the literature before.
-The detailed proofs can be found in Appendix A in the Supplementary Material.
+The detailed proofs can be found in Appendix B in the Supplementary Material.
 
@@
 \begin{equation}
-P^1_{NML}(D_{i,G_i=j};G)=\frac{P(D|\hat\theta(D_{i,G_i=j};G))}{\sum_{D'}{P(D';\hat\theta(D';G))}},
+P^1_{NML}(D_{i,G_i=j};G)=\frac{P(D|\hat\theta(D_{i,G_i=j};G))}{\sum_{D'}{P(D'|\hat\theta(D';G))}},
 \end{equation}

### 2018-02-12 supp_mat_B.tex
typos/polishing

@@
 \begin{definition}
-Assume $H_N(X \mid Y') \leq H_N(X \mid Y)$, where $Y' \subset Y.$ We say that the scoring function $Q_N(\cdot \mid \cdot)$ is regular if $Q_n(X \mid Y') \geq Q_N(X \mid Y)$.
+Assume $H_N(X \mid Y') \leq H_N(X \mid Y)$, where $Y' \subset Y.$ We say that the scoring function $Q_N(\cdot \mid \cdot)$ is regular if $Q_N(X \mid Y') \geq Q_N(X \mid Y)$.
 \end{definition}
@@
 Let $X$ be a categorical random variable with $r$ possible values.
-Let $U$ denote a possible parent set with $q$ different combinations of values for the variables, and $V$ a parent set with $m$ different configurations.
+Let $U$ denote a possible parent set with $q$ different combinations of values for the variables, and $V$ a set with $m$ different configurations.
 Assume that we have observed $N$ samples of $(X,U,V)$ (denoted by $x_N,u_N$ and $v_N$) and $H_N(X \mid U) \leq H_N(X \mid U \cup V)$ holds.
@@
 \begin{align*}
-Q^{qnml}_N(X\mid U) &= \log P(x_n \mid \hat{\theta}_{X\mid U} \ ) - \Le reg(N,rq) - reg(N,q) \Ri \\
+Q^{qnml}_N(X\mid U) &= \log P(x_N \mid \hat{\theta}_{X\mid U} \ ) - \Le reg(N,rq) - reg(N,q) \Ri \\
 &= \log P(x_N \mid \hat{\theta}_{X\mid U} \ ) - \log \frac{C(N,rq)}{C(N,q)},
@@
 $$
-\log P(x_N \mid \hat{\theta}_{X\mid Y} \ ) \geq \log P(x_N \mid \hat{\theta}_{X\mid Y'} \ ) .
+\log P(x_N \mid \hat{\theta}_{X\mid Y} \ ) \leq \log P(x_N \mid \hat{\theta}_{X\mid Y'} \ ) .
 $$Actually we must have the equality holding in the above expression, since
@@
 
-From this, it it easy see that the derivative of the quotient, \\ $d/dk (C(N,rk)/C(N,k))$, will be a ratio of two polynomials of $k$.
+In the following, we assume that $k$ is a real number.
+From (\ref{C1}) and (\ref{C2}), it is easy see that the derivative of the quotient, $d/dk (C(N,rk)/C(N,k))$, will be a ratio of two polynomials of $k$.
 Our goal is to show that the polynomial in the numerator has positive coefficients, which will guarantee the positivity of derivative for every $k > 0$, and thus imply that the original function is increasing (polynomial in the denominator is squared and non-zero for $k > 0$, so it can be ignored).
@@
 & = \sum_{l=1}^{i}(i-l+1)c_l \\
-&= \sum_{k =1}^{\lfloor i / 2 \rfloor}(i-k+1)c_k + (i-(i-k + 1)+1)c_{i-k+1} \\
-&=  \sum_{k =1}^{\lfloor i / 2 \rfloor}(i-k+1)c_k + kc_{i-k+1} \\
-&=  \sum_{k =1}^{\lfloor i / 2 \rfloor}(i-k+1)c_k - kc_{k} \\
-&= \sum_{k =1}^{\lfloor i / 2 \rfloor}(i-2k+1)c_k.
+&= \sum_{t =1}^{\lfloor i / 2 \rfloor}(i-t+1)c_t + (i-(i-t + 1)+1)c_{i-t+1} \\
+&=  \sum_{t =1}^{\lfloor i / 2 \rfloor}(i-t+1)c_t + tc_{i-t+1} \\
+&=  \sum_{t =1}^{\lfloor i / 2 \rfloor}(i-t+1)c_t - tc_{t} \\
+&= \sum_{t =1}^{\lfloor i / 2 \rfloor}(i-2t+1)c_t.
 \end{align*} On the first row, we re-wrote sum using only one running index.
@@
 We have to use the floor-function since if $i$ is odd, there exists an index $l'$ in the original sum such that $r^{i-l'+1}-r^{l'} = 0$.
-On the fifth row, we make use of the identity $c_k = -c_{i-k+1}$ which is straightforward to verify.
-From the last row, we can observe that every term of the sum is positive since $i-2k+1$ and $r^{i-k+1}- r^k$ are both positive if $k \leq (i+1)/2$ which holds since $k$ ranges from $1$ to $\lfloor i / 2 \rfloor$.
+On the fifth row, we make use of the identity $c_t = -c_{i-t+1}$ which is straightforward to verify.
+From the last row, we can observe that every term of the sum is positive since $i-2t+1$ and $r^{i-t+1}- r^t$ are both positive if $t \leq (i+1)/2$ which holds since $t$ ranges from $1$ to $\lfloor i / 2 \rfloor$.
 
@@
 & = \sum_{l=i-N + 1}^{N}(i-l+1)c_l \\
-& = \sum_{k=1}^{2N-i}(N-k+1)c_{i-N+k} \\
-& = \sum_{k=1}^{\lfloor N - i/2 \rfloor}(N-k+1)c_{i-N+k}\\
-&+  (N-(2N-i-k+1)+1)c_{i-N+(2N-i-k+1)} \\
-&= \sum_{k=1}^{\lfloor N - i/2 \rfloor}(N-k+1)c_{i-N+k} +  (i-N+k)c_{N-k+1} \\
-&= \sum_{k=1}^{\lfloor N - i/2 \rfloor}(N-k+1)c_{i-N+k} -  (i-N+k)c_{i-N+k} \\
-&= \sum_{k=1}^{\lfloor N - i/2 \rfloor}(N-k+1-(i-N+k))c_{i-N+k} \\
-&= \sum_{k=1}^{\lfloor N - i/2 \rfloor}(2N-i-2k+1)c_{i-N+k}.
-\end{align*}It is now easy to verify that $(2N-i-2k+1)$ and $c_{i-N+k}$ are positive if $k \leq N - (i-1)/2$ which holds since $k$ ranges from $1$ to $\lfloor N - i/2 \rfloor$.
-The floor function is again used when we sum over pairs of terms since if $i$ is odd there is zero-term.
+& = \sum_{t=1}^{2N-i}(N-t+1)c_{i-N+t} \\
+& = \sum_{t=1}^{\lfloor N - i/2 \rfloor}(N-t+1)c_{i-N+t}\\
+&+  (N-(2N-i-t+1)+1)c_{i-N+(2N-i-t+1)} \\
+&= \sum_{t=1}^{\lfloor N - i/2 \rfloor}(N-t+1)c_{i-N+t} +  (i-N+t)c_{N-t+1} \\
+&= \sum_{t=1}^{\lfloor N - i/2 \rfloor}(N-t+1)c_{i-N+t} -  (i-N+t)c_{i-N+t} \\
+&= \sum_{t=1}^{\lfloor N - i/2 \rfloor}(N-t+1-(i-N+t))c_{i-N+t} \\
+&= \sum_{t=1}^{\lfloor N - i/2 \rfloor}(2N-i-2t+1)c_{i-N+t}.
+\end{align*}It is now easy to verify that $(2N-i-2t+1)$ and $c_{i-N+t}$ are positive if $t \leq N - (i-1)/2$ which holds since $t$ ranges from $1$ to $\lfloor N - i/2 \rfloor$.
+The floor function is again used when we sum over pairs of terms since if $i$ is odd there is a zero-term.
 Since all the coefficients are non-negative and the $k \geq 2$, the derivative is positive. This implies that the original function is increasing.
@@
 \begin{equation}
-Q^{fnml}_N(X  \vert \ U) = \log P(x_n \mid \hat{\theta}_{X\mid U} \ ) - \sum_{j = 1}^{r_U} reg(N_j,r_X),
-\end{equation}where $r_X$ is the number of categories for $X$, $r_U$ denotes the number of possible configurations of variables in $U$, and $N_j$ is the number of times the $j$:th configuration is observed in our samples $u_N$.
-Note that $reg(r_X, N_j) = 0$ for any configuration not observed in our sample.
+Q^{fnml}_N(X  \vert \ U) = \log P(x_N \mid \hat{\theta}_{X\mid U} \ ) - \sum_{j = 1}^{r_U} reg(N_j,r_X),
+\end{equation}where $r_X$ is the number of categories for $X$, $r_U$ denotes the number of possible configurations of variables in $U$, and $N_j$ is the number of times the $j^\text{th}$ configuration is observed in our samples $u_N$.
+Note that $reg(N_j,r_X) = 0$ for any configuration not observed in our sample.
 fNML criterion differs from qNML only by how the penalty term is defined. We can follow a highly similar strategy in order to prove the regularity for fNML. To this end, we need the following Lemma:

### 2018-02-12 qnml.tex
typo

@@
 \begin{definition}
-Assume $H_N(X \mid Y') \leq H_N(X \mid Y)$, where $Y' \subset Y.$ We say that $Q_N(\cdot \mid \cdot)$ is regular if $Q_n(X \mid Y') \geq Q_N(X \mid Y)$.
+Assume $H_N(X \mid Y') \leq H_N(X \mid Y)$, where $Y' \subset Y.$ We say that $Q_N(\cdot \mid \cdot)$ is regular if $Q_N(X \mid Y') \geq Q_N(X \mid Y)$.
 \end{definition}

### 2018-02-14 bn.tex
Minor typos

@@
 variable/node $X_{i}$ its (possibly empty) parent set $G_{i}$, i.e.,
-the nodes from which there are a directed edges to the variable
+the nodes from which there is a directed edge to the variable
 $X_{i}$.
@@
 posterior probability $P(G|D,\alpha)$, where $\alpha$ denotes the
-hyperparameters for the model parameters $\theta$, and the $D$ is a
+hyperparameters for the model parameters $\theta$, and $D$ is a
 collection of $N$ $n$\nobreakdash-dimensional i.i.d. data vectors
 collected to a $N\times n$ design matrix. We use the notation $D_i$ to
-denote the $i^\text{th}$ column of the data matrix and notation $D_V$ to denote
+denote the $i^\text{th}$ column of the data matrix and the notation $D_V$ to denote
 the columns that correspond to the variable subset $V$. We also write
@@
 partitioned based on the values in columns $G_i$, and each part is
-then coded using a Bayesian mixture that can be expressed in a closed
+then coded using a Bayesian mixture that can be expressed in closed
 form~\cite{Bunt91, Heck95}.
@@
 properties of the BDeu score and showed that in certain settings BDeu
-has tendency to add more and more parent variables for a child node
+has the tendency to add more and more parent variables for a child node
 even though the empirical conditional entropy of the child given the
 parents has already reached zero. In more detail, assume that in our
-data $D$, the values of $X_i$ are completely determined by variables
-in set $Z$, so that the empirical entropy $H_N(X_i | Z) = 0$. Now, if we can further
+data $D$ the values of $X_i$ are completely determined by variables
+in set $Z$, so that the empirical entropy $H_N(X_i | Z)$ is zero. Now, if we can further
 find one or more variables, denoted by $Y$, whose values are
@@
 the set $Z\cup Y$ over $Z$ alone as the parents of $X_i$. Suzuki
-argues that this kind of behaviour violates \textit{regularity} in model
+argues that this kind of behavior violates \textit{regularity} in model
 selection as the more complex model is preferred over a simpler one
@@
 $N\times n$ data matrices. This information-theoretic NML criterion
-can be justified from the minimum description length point of view,
+can be justified from the minimum description length point of view
 \cite{Riss78,Grun07}. It has been shown to be robust with respect to
@@
 suggested solving the BDeu parameter sensitivity problem by using the
-NML-code for the column partitions, i.e., changing the Bayesian mixture
+NML code for the column partitions, i.e., changing the Bayesian mixture
 in equation~(\ref{eqn:bayesmix}) to

### 2018-02-14 experiments.tex
mention the SHD ranking in case of ties
Minor typos

@@
 These networks were picked in order to use the dynamic programming based exact structure learning~\cite{cosco.uai06} which limited the number $n$ of variables
-to less than 20. We measured the quality of the learned structures using structural Hamming distance (SHD, \cite{Tsamardinos2006}).
+to less than 20. We measured the quality of the learned structures using structural Hamming distance (SHD) \cite{Tsamardinos2006}.
 
@@
 None of the scores dominates in all settings considered.
-BIC fares well when the sample size is small, as it tends to produce a nearly empty graph which is a good answer in terms of SHD when the generating networks are relatively sparse.
-qNML obtains strong results in 'Earthquake' and 'Asia' networks, being the best or the second best with all the sample sizes considered.
+BIC fares well when the sample size is small as it tends to produce a nearly empty graph which is a good answer in terms of SHD when the generating networks are relatively sparse.
+qNML obtains strong results in the Earthquake and Asia networks, being the best or the second best with all the sample sizes considered.
 
@@
 The ranking was done by giving the score with the lowest SHD rank $1$ and the worst one rank $4$.
+In case of ties, the methods with the same SHD were given the same rank.
 The shown results are averages computed from $5000$ values ($5$ networks, $1000$ repetitions).
@@
 
-When predicting $P(d_{test}|D_{train},S,\theta)$,
+When predicting $P(d_{test}|D_{train},S,\theta)$
 with structures $S$ learned by the BDeu score, we used the

### 2018-02-14 intro.tex
Minor typos

@@
 likelihood~\cite{Heck95b} and its BDeu variant (see
-section~\ref{sec:bns}), but studies~\cite{cosco.uai07,Steck08} have
+Section~\ref{sec:bns}), but studies~\cite{cosco.uai07,Steck08} have
 found this criterion to be very sensitive to hyperparameters and to
@@
 be prohibitively expensive. In 2008, Silander et al. introduced a
-hyperparameter free, NML inspired criterion called the factorized NML
+hyperparameter-free, NML inspired criterion called the factorized NML
 (fNML)~\cite{cosco.pgm08a} that was shown to yield good predictive
@@
 yields parsimonious models with good predictive capabilities. The
-experiments with artificial data generated from real world Bayesian networks demonstrate the capability of our score to quickly learn a structure close to the generating one.
+experiments with artificial data generated from real-world Bayesian networks demonstrate the capability of our score to quickly learn a structure close to the generating one.

### 2018-02-14 qnml.tex
Minor typos

@@
 with $\prod_{X_i\in S} r_i$ different values which can then be modeled
-with a one-dimensional $P^1_{NML}$ code.  The $s^{qNML}$-score does
+with a one-dimensional $P^1_{NML}$ code.  The $s^{qNML}$ score does
 not necessarily define a distribution for $D$, but it is easy to
-verify that it coincides with the $\log P_{NML}(D;G)$ for all networks
+verify that it coincides with the NML score for all networks
 that are composed of fully connected components.  The number of such
@@
 
-We are now ready to prove some important properties of the qNML-score.
+We are now ready to prove some important properties of the qNML score.
 
@@
 
-qNML yields equal scores for network structures that encode same sets
+qNML yields equal scores for network structures that encode the same set
 of independencies. Verma and Pearl~\cite{Verm90} showed that the
@@
   $P^1_{NML}(i,G_i) := P^1_{NML}(D_{i,G_i};G)$ and
-  $P^1_{NML}(G_i)   := P^1_{NML}(D_{G_i};G)$.
+  $P^1_{NML}(G_i)   := P^1_{NML}(D_{G_i};G)$. Now
   \begin{eqnarray}
@@
 a data matrix with $N$ samples coming from a distribution faithful to
-some DAG $G$, the qNML will give the highest score to true graph $G$
+some DAG $G$, the qNML will give the highest score to the true graph $G$
 with a probability tending to one as $N$ increases. We will show this
@@
 $X_i$ and its possible configurations of parents $G_i$,
-respectively. As $N \to \infty$
+respectively. As $N \to \infty$,
 $$
@@
 tournaments, which turns out to be the formula for OEIS sequence
-A000262\footnote{https://oeis.org/A000108}.
+A000262\footnote{https://oeis.org/A000262}.
 In general this sequence grows rapidly; $1, 1, 3, 13, 73, 501, 4051,
@@
 In the definition, $N$ denotes the sample size, $X$ is some random variable, $Y$ denotes the proposed parent set for $X$, and $H_N(\cdot \mid \cdot)$ refers to the empirical conditional entropy.
-Suzuki \cite{Suzuki2017} shows that BDeu violates this principle and demonstrates that this can cause the score to prefer more complex networks even though data do not support this.
+Suzuki \cite{Suzuki2017} shows that BDeu violates this principle and demonstrates that this can cause the score to prefer more complex networks even though the data do not support this.
 Regular scores are also argued to be computationally more efficient when applied with branch-and-bound type algorithms for Bayesian network structure learning \cite{Suzuki2017_2}.
@@
 \begin{proof}
-The proof is given in Appendix B in Supplementary Material.
+The proof is given in Appendix B in the Supplementary Material.
 \end{proof}
@@
 Suzuki \cite{Suzuki2017} independently introduces a Bayesian Dirichlet
-quotient (BDq) score, that can also be shown to be score equivalent
+quotient (BDq) score that can also be shown to be score equivalent
 and regular.  However, like BDeu, this score features a single

### 2018-02-14 supp_mat_A.tex
Typos
Minor typos

@@
 class. All the other networks have at least one arc. We will first
-presents some lemmas that characterize the relations of the parent sets
+present some lemmas that characterize the relations of the parent sets
 before and after the arc reversal. Let $G$ be a network structure
@@
 
-We wil next list some simple consequences of the equivalence preserving
+We will next list some simple consequences of the equivalence preserving
 arc reversal:

### 2018-02-14 supp_mat_C.tex
Minor typos

@@
-\section{Prediction and Parsimony}
+\section{qNML coincides with NML for many models}
 
-We took 20 UCI data sets created 1000 permutations of each data set.
-We then took the first x\% of each permuted data set data (x in
-${10,20,30,\ldots,90}$ as training data and used the remaining part as
-a test data. We used exact structure learning to learn the best
-scoring model using four model selection criteria, BDeu ($\alpha=1$),
-BIC, fNML and qNML, recording the number of parameters in the learnt
-models and the average predictive log-probability of each test
-vector. These numbers are then collected for 1000 different data
-permutations average and variance of which appear on the following
-pages, one page per data set.
+\subsection{qNML Equals NML for Many Models}
+The fNML criterion can be seen as a computationally feasible
+approximation of the more desirable NML criterion.  However, the fNML
+criterion equals the NML criterion only for the Bayesian network
+structure with no arcs.  We will next show that the qNML criterion
+equals the NML criterion for all the networks $G$ whose connected
+components tournaments (i.e., complete directed acyclic subgraphs of
+$G$).
+
+\begin{theorem}
+If $G$ consists of $C$ connected components $(G^1,\ldots,G^C)$ with
+variable sets $(V^1,\ldots,V^C)$, then $\log P_{NML}(D;G) = s^{qNML}(D;G)$
+for all data sets $D$.
+\end{theorem}
+\begin{proof}
+We first show that the NML-criterion for a Bayesian network
+decomposes by the connected components.
+
+Because the maximum likelihood for the data $D$ decomposes,
+we can write
+\begin{eqnarray}
+  \lefteqn{P_{NML}(D;G)} \nonumber \\
+  && = \frac{P(D;\hat\theta(D),G)}
+            {\sum_{D'_{V_1}}\ldots \sum_{D'_{V_C}}\prod_{c=1}^C P(D'_{V^c};\hat\theta(D'_{V^c}),G)} \nonumber \\
+            && = \frac{\prod_{c=1}^C P(D_{V^c};\hat\theta(D_{V^c}),G)} 
+            {\prod_{c=1}^C\sum_{D'_{V^c}}P(D'_{V^c};\hat\theta(D'_{V^c}),G)}.\nonumber \\
+            && = \prod_{c=1}^CP_{NML}(D_{V^c};G).
+\end{eqnarray}
+
+Clearly, the qNML score also decomposes by the connected components,
+so it remains to show that if the (sub)network $G$ is a tournament,
+then for any data $D$, $s^{qNML}(D;G)=\log P_{NML}(D;G)$.  Due to the
+score equivalence of the NML criterion and the qNML criterion, we may
+pick a tournament $G$ such that the linear ordering defined by $G$
+matches the ordering of the data columns, i.e., $i<j$ implies $G_i
+\subset G_j$. Now from the definition~(8) of the qNML
+criterion (in the main paper), we see that for the tournament $G$, the sum telescopes
+leaving us with $s^{qNML}(D;G) = \log P^1_{NML}(D_G;G)$, thus
+it is enough to show that $P^1_{NML}(D;G)=P_{NML}(D;G)$.  This
+follows, since for any data vector $x$ in data $D$, we have
+$P^1(x;\hat\theta(D),G) = P(x;\hat\theta(D),G)$, where $P^1$ denotes
+the model that takes $n$-dimensional vectors to be values of the
+single (collapsed) categorical variable.  Denoting prefixes of data
+vector $x$ by $x^{:i}$, and the number of times such a prefix appears
+on $N$ rows $[d_1,\ldots,d_N]$ of the $N\times n$ data matrix $D$ by
+$N_D(x^{:i})$, (so that $N_D(x^{:0})=N)$, we have
+\begin{eqnarray}
+  P(D;\hat\theta(D),G) &=& \prod_{j=1}^{N}P(d_j;\hat\theta(D),G) \nonumber \\
+  &=& \prod_{j=1}^{N} \prod_{i=1}^{n} \frac{N_D(d_j^{:i})}{N_D(d_j^{:i-1})} \nonumber\\
+  &=& \prod_{j=1}^{N} \frac{N_D(d_j^{:n})}{N} \nonumber \\
+  &=& P^1(D;\hat\theta^1(D),G).
+\end{eqnarray}
+Since both $P^1_{NML}$ and $P_{NML}$ are defined in terms of the
+maximum likelihood probabilities, the equality above implies the equality
+of these distributions.
+\end{proof}
+
+Equality established, we would like to still state the number $a(n)$
+of different $n$-node networks whose connected components are
+tournaments.  We start by generating all the $p(n)$ integer partitions
+i.e. ways to partition $n$ labelled into parts with different
+size-profiles.  For example, for $n=4$, we have $p(4)=5$ partition
+profiles $[[4], [3, 1], [2, 2], [2, 1, 1], [1, 1, 1, 1]]$. Each of
+these partition size profiles corresponds to many different networks,
+apart from the last one that corresponds just to the empty network.
+We count number of networks for one such partition size-profile (and
+then later sum these counts up).  For any such partition profile
+$(p_1,\ldots,p_k)$ we can count the ways we can assign the nodes to
+different parts and then order each part. This leads to the product
+$n\choose p_1$$p_1$!${n-p_1}\choose p_2$$p_2!$${n-p_1-p2}\choose
+p_3$$p_3!$$\ldots$${n-\sum_{j=1}^{k-1}p_j}\choose p_k$$p_k!$. However,
+the order of different parts of the same size does not matter, so for
+all groups of parts having the same size, we have to divide the
+product above by the factorial of the size of such group. Notice also,
+that the product above telescopes, leaving us a formula for OEIS
+sequence A000262\footnote{https://oeis.org/A000262} as
+described by Thomas Wieder:
+
+\textit{With $p(n) =$ the number of integer partitions of $n$, $d(i)$ = the
+number of different parts of the $i^\text{th}$ partition of $n$,
+$m(i,j) =$ multiplicity of the $j^\text{th}$ part of the $i^\text{th}$
+partition of $n$, one has:}
+$$
+a(n) = \sum_{i=1}^{p(n)} \frac{n!}{\prod_{j=1}^{d(i)} m(i,j)!}.
+$$
+For example, $a(4)=\frac{4!}{1!}+\frac{4!}{1!1!}+\frac{4!}{2!}+\frac{4!}{1!2!}+\frac{4!}{4!}=73$. In general this sequence grows rapidly; $1, 1, 3, 13, 73, 501, 4051, 37633, 394353, 4596553, \ldots$.

### 2018-02-14 bn.tex
More minor typos

@@
 column $i$ on the rows on which the parents $G_i$ contain the value
-configuration number by $D_{i,G_i=j}$, $j\in\{1,\ldots,q_i\}$.
+configuration number $j$ by $D_{i,G_i=j}$, $j\in\{1,\ldots,q_i\}$.
 
@@
 data $D$ the values of $X_i$ are completely determined by variables
-in set $Z$, so that the empirical entropy $H_N(X_i | Z)$ is zero. Now, if we can further
+in a set $Z$, so that the empirical entropy $H_N(X_i | Z)$ is zero. Now, if we can further
 find one or more variables, denoted by $Y$, whose values are
@@
 where $D'\in{\{1,\ldots,r_i\}}^{|D_{i,G_i=j}|}$.  The logarithm of the
-denominator is often called a regret, since it indicates the extra
+denominator is often called the regret, since it indicates the extra
 code length needed compared to the code length obtained using the (a

### 2018-02-14 experiments.tex
More minor typos

@@
 
-To empirically compare the model selection criteria we took 20 UCI
+To empirically compare the model selection criteria, we took 20 UCI
 data sets~\cite{Lichman:2013} and ran train and test experiments for
@@
 for all the sample sizes making it a ``safe choice''. This can be
-quantified if we further average the columns of the
-table~\ref{tbl:preds}.  This yields the average ranks of $2.95, 2.57,
+quantified if we further average the columns of
+Table~\ref{tbl:preds}.  This yields the average ranks of $2.95, 2.57,
 2.10$, and $2.37$, with standard deviations $0.49, 0.90, 0.76$, and
@@
 Figure~\ref{fig:bcnpmean} shows how fNML still sometimes behaves
-strangely in terms of model complexity here measured by the number of
+strangely in terms of model complexity, here measured by the number of
 parameters in the model. qNML, instead, appears to yield more

### 2018-02-14 qnml.tex
More minor typos

@@
 also very sensitive to this hyperparameter (see Appendix D in the
-Supplementarty Material), the issue that was one of the main
+Supplementary Material), the issue that was one of the main
 motivations to develop a parameter-free model selection criterion like

### 2018-02-14 main.tex
[hidelinks] added to hyperref
[accepted] options + authot info

@@
 \documentclass[twoside]{article}
-\usepackage{aistats2018}
+\usepackage[accepted]{aistats2018}
 \usepackage{amsmath,amsthm,epsfig,booktabs,multirow}
-\usepackage{hyperref}
+\usepackage[hidelinks]{hyperref}
 \newtheorem{lemma}{Lemma}
@@
 
+\runningauthor{Tomi Silander, Janne Lepp\"{a}-aho, Elias J\"{a}\"{a}saari, Teemu Roos}
+
 \begin{document}
@@
 
-\aistatsauthor{ Author 1 \And Author 2 \And  Author 3 }
+\aistatsauthor{ Tomi Silander$^1$ \And Janne Lepp\"{a}-aho$^{2,4}$ \And Elias J\"{a}\"{a}saari$^{2,3}$ \And Teemu Roos$^{2,4}$   }
 
-\aistatsaddress{ Institution 1 \And  Institution 2 \And Institution 3 } ]
+\aistatsaddress{ $^1$NAVER LABS Europe \And $^2$Helsinki Institute for \\ Information Technology HIIT \And $^3$Department of CS,\\Aalto University \And $^4$Department of CS,\\University of Helsinki} ]
 

### 2018-02-15 supp_mat_BDq.tex
added missing supp_mat_BDq.tex

\section{Bayesian Dirichlet quotient score}

\cite{Suzuki2017} independently suggested a Bayesian Dirichlet
quotient score BDq, in which the NML-distributions
$P^1_{NML}(D_{i,G_i};G)$ and $P^1_{NML}(D_{G_i};G)$ of the qNML are
replaced by the Bayesian marginal likelihoods
$P^1(D_{i,G_i};G,\alpha)$ and $P^1_{NML}(D_{G_i};G,\alpha)$ using
parameter prior $\theta_{ij} \sim
Dirichlet(\alpha,\ldots,\alpha)$. Suzuki further suggested using the
Jeffreys' prior by setting $\alpha=\frac{1}{2}$. While this is a
popular choice in statistics, it is also possible to argue for other
values of $\alpha$ like $\alpha=\frac{1}{2}-\frac{\ln 2}{2\ln
  N}$~\citep{watanabe15a}. Our yet to be published paper will present
a case in which $\alpha=\frac{1}{3}$ could be preferable.

Our initial experiments indicate that model selection by BDq is still
highly sensitive to the hyperparameter $\alpha$. On next 5 pages we
present results for learning the posterior distribution of the network
structures for the four predictor variables (discretized to three
equal width bins) in the iris-data (N=150). The structure prior is
assumed to be uniform. Since BDq is score equivalent, we only include
one representative from each equivalence class. The images also show
the most probable structure and the predictive distribution given by
the most probable structure and the predictive distribution when
marginalizing over all the structures.

One can see how the structure posterior is quite sensitive to $\alpha$
and the selected network is different for $\alpha$-values $0.2$, $0.3$
and $0.5$.

\includepdf[pages={1-5},angle=90]{qsens.pdf}

### 2018-02-15 bn.tex
small edits to make intro fit to one page and to fix the beginning of the section 2.

@@
 Bayesian networks are a general way to describe the dependencies
-between the components of an $n$\nobreakdash-dimensional discrete data
-vector $X=(X_{1},\ldots,X_{n})$ in which the component $X_{i}$ may
-take any of the discrete values in a set $\{1,\ldots,r_{i}\}$.
-Despite denoting the values with small integers, the model will treat
-the components of $X$ as categorical variables.
+between the components of an $n$\nobreakdash-dimensional random data
+vector. In this paper we only address the case in which the component
+$X_{i}$ of the data vector $X=(X_{1},\ldots,X_{n})$ may take any of
+the discrete values in a set $\{1,\ldots,r_{i}\}$.  Despite denoting
+the values with small integers, the model will treat each $X_i$
+as a categorical variable.
 

### 2018-02-15 intro.tex
small edits to make intro fit to one page and to fix the beginning of the section 2.

@@
 likelihood~\cite{Heck95b} and its BDeu variant (see
-Section~\ref{sec:bns}), but studies~\cite{cosco.uai07,Steck08} have
-found this criterion to be very sensitive to hyperparameters and to
+Section~\ref{sec:bns}), but studies~\cite{cosco.uai07,Steck08} show
+this criterion to be sensitive to hyperparameters and to
 yield undesirably complex models for small sample sizes.

### 2018-02-17 AISTATS2018PaperPack/Style file AISTATS 2018/sample_paper.tex
Adding latest style files

\documentclass[twoside]{article}
%\usepackage{aistats2018}
 % If your paper is accepted, change the options for the package
% aistats2018 as follows:
%
\usepackage[accepted]{aistats2018}
%
% This option will print headings for the title of your paper and
% headings for the authors names, plus a copyright note at the end of
% the first column of the first page.

\begin{document}

% If your paper is accepted and the title of your paper is very long,
% the style will print as headings an error message. Use the following
% command to supply a shorter title of your paper so that it can be
% used as headings.
%
%\runningtitle{I use this title instead because the last one was very long}

% If your paper is accepted and the number of authors is large, the
% style will print as headings an error message. Use the following
% command to supply a shorter version of the authors names so that
% they can be used as headings (for example, use only the surnames)
%
%\runningauthor{Surname 1, Surname 2, Surname 3, ...., Surname n}

\twocolumn[

\aistatstitle{Instructions for paper submissions to AISTATS 2018}

\aistatsauthor{ Author 1 \And Author 2 \And  Author 3 }

\aistatsaddress{ Institution 1 \And  Institution 2 \And Institution 3 } ]

\begin{abstract}
  The Abstract paragraph should be indented 0.25 inch (1.5 picas) on
  both left and right-hand margins. Use 10~point type, with a vertical
  spacing of 11~points. The {\bf Abstract} heading must be centered,
  bold, and in point size 12. Two line spaces precede the
  Abstract. The Abstract must be limited to one paragraph.
\end{abstract}

\section{GENERAL FORMATTING INSTRUCTIONS}

The camera-ready versions of the accepted papers are 8 pages,
plus any additional pages needed for references.

Papers are in 2 columns with the overall line width of 6.75~inches (41~picas). Each column is 3.25~inches wide (19.5~picas).  The space
between the columns is .25~inches wide (1.5~picas).  The left margin is 1~inch (6~picas).  Use 10~point type with a vertical spacing of
11~points. Please use US Letter size paper instead of A4.

Paper title is 16~point, caps/lc, bold, centered between 2~horizontal rules.  Top rule is 4~points thick and bottom rule is 1~point thick.
Allow 1/4~inch space above and below title to rules.

Author descriptions are center-justified, initial caps.  The lead
author is to be listed first (left-most), and the Co-authors are set
to follow.  If up to three authors, use a single row of author
descriptions, each one center-justified, and all set side by side;
with more authors or unusually long names or institutions, use more
rows.

Use one-half line space between paragraphs, with no indent.

\section{FIRST LEVEL HEADINGS}

First level headings are all caps, flush left, bold, and in point size
12. Use one line space before the first level heading and one-half line space
after the first level heading.

\subsection{Second Level Heading}

Second level headings are initial caps, flush left, bold, and in point
size 10. Use one line space before the second level heading and one-half line
space after the second level heading.

\subsubsection{Third Level Heading}

Third level headings are flush left, initial caps, bold, and in point
size 10. Use one line space before the third level heading and one-half line
space after the third level heading.

\paragraph{Fourth Level Heading}

Fourth level headings must be flush left, initial caps, bold, and
Roman type.  Use one line space before the fourth level heading, and
place the section text immediately after the heading with no line
break, but an 11 point horizontal space.

\subsection{CITATIONS, FIGURES, REFERENCES}

\subsubsection{Citations in Text}

Citations within the text should include the author's last name and
year, e.g., (Cheesman, 1985). References should follow any style that
you are used to using, as long as their style is consistent throughout
the paper.  Be sure that the sentence reads correctly if the citation
is deleted: e.g., instead of ``As described by (Cheesman, 1985), we
first frobulate the widgets,'' write ``As described by Cheesman
(1985), we first frobulate the widgets.''  %Be sure to avoid
%accidentally disclosing author identities through citations.

\subsubsection{Footnotes}

Indicate footnotes with a number\footnote{Sample of the first
  footnote.} in the text. Use 8 point type for footnotes. Place the
footnotes at the bottom of the column in which their markers appear,
continuing to the next column if required. Precede the footnote
section of a column with a 0.5 point horizontal rule 1~inch (6~picas)
long.\footnote{Sample of the second footnote.}

\subsubsection{Figures}

All artwork must be centered, neat, clean, and legible.  All lines
should be very dark for purposes of reproduction, and art work should
not be hand-drawn.  Figures may appear at the top of a column, at the
top of a page spanning multiple columns, inline within a column, or
with text wrapped around them, but the figure number and caption
always appear immediately below the figure.  Leave 2 line spaces
between the figure and the caption. The figure caption is initial caps
and each figure should be numbered consecutively.

Make sure that the figure caption does not get separated from the
figure. Leave extra white space at the bottom of the page rather than
splitting the figure and figure caption.
\begin{figure}[h]
\vspace{.3in}
\centerline{\fbox{This figure intentionally left non-blank}}
\vspace{.3in}
\caption{Sample Figure Caption}
\end{figure}

\subsubsection{Tables}

All tables must be centered, neat, clean, and legible. Do not use hand-drawn tables. Table number and title always appear above the table.
See Table~\ref{sample-table}.

Use one line space before the table title, one line space after the table title, and one line space after the table. The table title must be
initial caps and each table numbered consecutively.

\begin{table}[h]
\caption{Sample Table Title} \label{sample-table}
\begin{center}
\begin{tabular}{ll}
{\bf PART}  &{\bf DESCRIPTION} \\
\hline \\
Dendrite         &Input terminal \\
Axon             &Output terminal \\
Soma             &Cell body (contains cell nucleus) \\
\end{tabular}
\end{center}
\end{table}

\section{SUPPLEMENTARY MATERIAL}

If you need to include additional appendices during submission, you
can include them in the supplementary material file.

\section{INSTRUCTIONS FOR CAMERA-READY PAPERS}

For the camera-ready paper, if you are using \LaTeX, please make sure
that you follow these instructions.  (If you are not using \LaTeX,
please make sure to achieve the same effect using your chosen
typesetting package.)

\begin{enumerate}
    \item Download \texttt{fancyhdr.sty} -- the
    \texttt{aistats2018.sty} file will make use of it.
    \item Begin your document with
    \begin{flushleft}
    \texttt{\textbackslash documentclass[twoside]\{article\}}\\
    \texttt{\textbackslash usepackage[accepted]\{aistats2018\}}
    \end{flushleft}
    The \texttt{twoside} option for the class article allows the
    package \texttt{fancyhdr.sty} to include headings for even and odd
    numbered pages. The option \texttt{accepted} for the package
    \texttt{aistats2018.sty} will write a copyright notice at the end of
    the first column of the first page. This option will also print
    headings for the paper.  For the \emph{even} pages, the title of
    the paper will be used as heading and for \emph{odd} pages the
    author names will be used as heading.  If the title of the paper
    is too long or the number of authors is too large, the style will
    print a warning message as heading. If this happens additional
    commands can be used to place as headings shorter versions of the
    title and the author names. This is explained in the next point.
    \item  If you get warning messages as described above, then
    immediately after $\texttt{\textbackslash
    begin\{document\}}$, write
    \begin{flushleft}
    \texttt{\textbackslash runningtitle\{Provide here an alternative shorter version of the title of your
    paper\}}\\
    \texttt{\textbackslash runningauthor\{Provide here the surnames of the authors of your paper, all separated by
    commas\}}
    \end{flushleft}
    Note that the text that appears as argument in \texttt{\textbackslash
      runningtitle} will be printed as a heading in the \emph{even}
    pages. The text that appears as argument in \texttt{\textbackslash
      runningauthor} will be printed as a heading in the \emph{odd}
    pages.  If even the author surnames do not fit, it is acceptable
    to give a subset of author names followed by ``et al.''

    \item Use the file sample\_paper.tex as an example.

    \item The camera-ready versions of the accepted papers are 8
      pages, plus any additional pages needed for references.

    \item If you need to include additional appendices,
      you can include them in the supplementary
      material file.

    \item Please, don't change the layout given by the above
      instructions and by the style file.

\end{enumerate}

\subsubsection*{Acknowledgements}

Use unnumbered third level headings for the acknowledgements.  All
acknowledgements go at the end of the paper.

\subsubsection*{References}

References follow the acknowledgements.  Use an unnumbered third level
heading for the references section.  Any choice of citation style is
acceptable as long as you are consistent.  Please use the same font
size for references as for the body of the paper---remember that
references do not count against your page length total.

J.~Alspector, B.~Gupta, and R.~B.~Allen (1989). Performance of a
stochastic learning microchip.  In D. S. Touretzky (ed.), {\it
  Advances in Neural Information Processing Systems 1}, 748-760.  San
Mateo, Calif.: Morgan Kaufmann.

F.~Rosenblatt (1962). {\it Principles of Neurodynamics.} Washington,
D.C.: Spartan Books.

G.~Tesauro (1989). Neurogammon wins computer Olympiad.  {\it Neural
  Computation} {\bf 1}(3):321-323.

\end{document}

### 2018-02-19 supp_mat_BDq.tex
Add reference to ALT paper

@@
 Dirichlet(\alpha,\ldots,\alpha)$. Suzuki further suggested using the
-Jeffreys' prior by setting $\alpha=\frac{1}{2}$. While this is a
+Jeffreys' prior with $\alpha=\frac{1}{2}$. While this is a
 popular choice in statistics, it is also possible to argue for other
 values of $\alpha$ like $\alpha=\frac{1}{2}-\frac{\ln 2}{2\ln
-  N}$~\citep{watanabe15a}. Our yet to be published paper will present
-a case in which $\alpha=\frac{1}{3}$ could be preferable.
+  N}$~\citep{watanabe15a} and $\alpha=\frac{1}{3}$~\citep{jaasaari18}.
 
 Our initial experiments indicate that model selection by BDq is still
-highly sensitive to the hyperparameter $\alpha$. On next 5 pages we
+highly sensitive to the hyperparameter $\alpha$. On the next 5 pages we
 present results for learning the posterior distribution of the network
 structures for the four predictor variables (discretized to three
-equal width bins) in the iris-data (N=150). The structure prior is
+equal width bins) in the iris data (N=150). The structure prior is
 assumed to be uniform. Since BDq is score equivalent, we only include

### 2018-02-19 supp_mat_C.tex
Add reference to ALT paper

@@
 all groups of parts having the same size, we have to divide the
-product above by the factorial of the size of such group. Notice also,
+product above by the factorial of the size of such group. Notice also
 that the product above telescopes, leaving us a formula for OEIS

### 2018-02-19 supp_mat_D.tex
Add reference to ALT paper

\section{Prediction and Parsimony}

We took 20 UCI data sets and created 1000 permutations of each data set.
We then took the first $x$\% of each permuted data set ($x$ in
${10,20,30,\ldots,90}$) as training data and used the remaining part as
test data. We used exact structure learning to learn the best
scoring model using BDeu ($\alpha=1$), BIC, fNML and qNML,
recording the number of parameters in the learnt
models and the average predictive log-probability of each test
vector. These numbers are collected for 1000 different data
permutations, the average and variance of which appear on the following
pages, one page per data set.

### 2018-02-19 makesuppmat.sh
including result images of supplementary material construction with includepdf instead of spearate bash script

@@ -1,5 +1,0 @@
-#/bin/bash
-pdflatex supp_mat.tex
-bibtex supp_mat
-pdflatex supp_mat.tex; pdflatex supp_mat.tex
-pdfunite supp_mat.pdf all_qNML_images4.pdf all_sup_mat.pdf

### 2018-02-19 experiments.tex
Add acknowledgements

@@
 For each train/test sample, we ranked the predictive performance of
-the models learned by four different structure learning criteria (rank
+the models learned by the four different scores (rank
 1 being the best and 4 the worst). Table~\ref{tbl:preds} features the
 average rank for different data sets, the average being taken over
-1000 different train test samples for each 9 sample sizes for each
-data.  BIC's bias for simplicity makes it often win (written bold in
-the table) with small sample sizes, but it often performs worst
+1000 different train/test samples for each 9 sample sizes.
+BIC's bias for simplicity makes it often win (written bold in
+the table) with small sample sizes, but it performs worst
 (underlined) for the larger sample sizes (for the same reason), while
@@
 quantified if we further average the columns of
-Table~\ref{tbl:preds}.  This yields the average ranks of $2.95, 2.57,
+Table~\ref{tbl:preds}, yielding the average ranks of $2.95, 2.57,
 2.10$, and $2.37$, with standard deviations $0.49, 0.90, 0.76$, and
@@
 Figure~\ref{fig:bcnpmean} shows how fNML still sometimes behaves
-strangely in terms of model complexity, here measured by the number of
+strangely in terms of model complexity as measured by the number of
 parameters in the model. qNML, instead, appears to yield more
@@
 complex models for small sample sizes, we studied the number of
-parameters in models produced by different model selection criteria
-when using 10\% of each data set for structure learning.
+parameters in models produced by different scores when using
+10\% of each data set for structure learning.
 
@@
   \caption{Average number of parameters in models 
-    for different model selection criteria in 20 different data sets.}
+    for different scores in 20 different data sets.}
 \label{tbl:nofparams}

### 2018-02-20 supp_mat_B.tex
small typo
one typo more..

@@
 \begin{lemma}\label{regineq}
-Let $N = N_1 + N_2$ and $r \in\N, r > 2$. Now,
+Let $N = N_1 + N_2$ and $r \in\N, r \geq 2$. Now,
 $$
@@
 \end{align*}Taking the logarithm on both sides yields our claim.
-On the second row, we used the definition of maximum likelihood parameters: the probability of the data vector $x_{N_1}$ is maximized under parameters $\hat{\theta}(x_{N_1})$, and if we use different parameters, $\hat{\theta}(x_{N_1})$, the probability can only go down or stay the same.
+On the second row, we used the definition of maximum likelihood parameters: the probability of the data vector $x_{N_1}$ is maximized under parameters $\hat{\theta}(x_{N_1})$, and if we use different parameters, $\hat{\theta}(x_{N})$, the probability can only go down or stay the same.
 On the third row, we used the i.i.d., assumption, which allows us to take the single sum over data vectors of length $N = N_1 + N_2$.
@@
 \end{equation}where we used $r_{Y'}$ and $r_Y$ to denote the numbers of possible configurations for variables in $Y'$ and $Y$, respectively.
-Also, $N_j$ refers to the number times the $j$:th configuration of variables $Y$ is observed in our sample, and $N'_l$ denotes the same for $Y'$.
+Also, $N_j$ refers to the number times the $j$:th configuration of variables $Y$ is observed in our sample, and $N'_j$ denotes the same for $Y'$.
 

### 2018-02-26 shd_plots.py
change fonts in shd figs

@@ -1,9 +1,11 @@
 #!/usr/bin/env python3
 # -*- coding: utf-8 -*-
 import numpy as np
-import seaborn as sns
 import matplotlib.pyplot as plt
-
+plt.rcParams['ps.useafm'] = True
+plt.rcParams['pdf.use14corefonts'] = True
+plt.rcParams['text.usetex'] = True
+import seaborn as sns
 
 def subplots(folder = "results/real_dag_shd/", save = False):
     
@@ -255,7 +257,7 @@ def score_to_rank(x):
 if __name__ == "__main__":
     save = True
     folder = "results/real_dag_shd/"
-    #subplots(folder = folder, save = save)
+    subplots(folder = folder, save = save)
     #rank_shd(folder = folder, save = save)
     rank_shd_from_single_tests(folder = folder, save = save)
     

### 2018-02-27 make_all_4.sh
work to get rid of the Type 3 fonts

@@ -1,0 +1,22 @@
+pdfunite \
+qNML_images/adult_mean.pdf qNML_images/adult_var.pdf qNML_images/adult_np_mean.pdf qNML_images/adult_np_var.pdf \
+qNML_images/wine_mean.pdf qNML_images/wine_var.pdf qNML_images/wine_np_mean.pdf qNML_images/wine_np_var.pdf \
+qNML_images/heart_statlog_mean.pdf qNML_images/heart_statlog_var.pdf qNML_images/heart_statlog_np_mean.pdf qNML_images/heart_statlog_np_var.pdf \
+qNML_images/heart_hungarian_mean.pdf qNML_images/heart_hungarian_var.pdf qNML_images/heart_hungarian_np_mean.pdf qNML_images/heart_hungarian_np_var.pdf \
+qNML_images/heart_cleveland_mean.pdf qNML_images/heart_cleveland_var.pdf qNML_images/heart_cleveland_np_mean.pdf qNML_images/heart_cleveland_np_var.pdf \
+qNML_images/page_blocks_mean.pdf qNML_images/page_blocks_var.pdf qNML_images/page_blocks_np_mean.pdf qNML_images/page_blocks_np_var.pdf \
+qNML_images/glass_mean.pdf qNML_images/glass_var.pdf qNML_images/glass_np_mean.pdf qNML_images/glass_np_var.pdf \
+qNML_images/bc_wisconsin_mean.pdf qNML_images/bc_wisconsin_var.pdf qNML_images/bc_wisconsin_np_mean.pdf qNML_images/bc_wisconsin_np_var.pdf \
+qNML_images/tic_tac_toe_mean.pdf qNML_images/tic_tac_toe_var.pdf qNML_images/tic_tac_toe_np_mean.pdf qNML_images/tic_tac_toe_np_var.pdf \
+qNML_images/shuttle_mean.pdf qNML_images/shuttle_var.pdf qNML_images/shuttle_np_mean.pdf qNML_images/shuttle_np_var.pdf \
+qNML_images/breast_cancer_mean.pdf qNML_images/breast_cancer_var.pdf qNML_images/breast_cancer_np_mean.pdf qNML_images/breast_cancer_np_var.pdf \
+qNML_images/yeast_mean.pdf qNML_images/yeast_var.pdf qNML_images/yeast_np_mean.pdf qNML_images/yeast_np_var.pdf \
[…]

### 2018-02-27 make_all_4.sh
Work to get rid of the Type 3 fonts

@@ -20,3 +20,4 @@ qNML_images/thyroid_mean.pdf qNML_images/thyroid_var.pdf qNML_images/thyroid_np_
 qNML_images/iris_mean.pdf qNML_images/iris_var.pdf qNML_images/iris_np_mean.pdf qNML_images/iris_np_var.pdf \
 qNML_images/balance_mean.pdf qNML_images/balance_var.pdf qNML_images/balance_np_mean.pdf qNML_images/balance_np_var.pdf \
 all4.pdf
+echo Then print all4.pdf to the pdf file with 4 pages per sheet in landscape

### 2018-03-16 poster/qnmlPoster.tex
starting with the poster

@@
 \usepackage{url}
-%\usepackage[orientation=portrait,size=a0,scale=1.4,debug]{beamerposter}        
-\usepackage[orientation=portrait,width = 91 ,height = 122 ,scale=1.39,debug]{beamerposter} 
+\usepackage[numbers]{natbib}
+\usepackage[orientation=portrait,size=a0,scale=1.4,debug]{beamerposter}        
+%\usepackage[orientation=portrait,width = 91 ,height = 122 ,scale=1.39,debug]{beamerposter} 
 %\setbeamerfont{itemize/enumerate body}{size={\fontsize{25}{31}}}
@@
 
-\newcommand{\mechanism}{\mathcal{M}}
-\newcommand{\dataset}{\mathcal{D}}
-
 \newcommand{\myred}[1]{{\color{BRed}#1}}
@@
 \definecolor{myPurple}{RGB}{174, 206, 195}
+%\definecolor{myPurple}{RGB}{165,206,190}
 
@@
 
-\newcommand{\vOmega}{\bm{\Omega}}
-\newcommand{\bBeta}{\bm{\beta}}
-\newcommand{\R}{\mathbb{R}}
-\newcommand{\vA}{\bm{A}}
-\newcommand{\vx}{\bm{x}}
-\newcommand{\vy}{\bm{y}}
 \definecolor{myblue}{rgb}{.85, .85, 1.0}
@@
 
-\begin{block}{Section 1}
+\begin{block}{Background: Structure Learning of Bayesian Networks}
 \begin{columns}[T]
    \begin{column}{0.3\textwidth} % first column {{{
-     \heading{Subsection 1}
-
+     \heading{Bayesian Networks}
+     \vspace*{12pt}
+     \begin{itemize}
+     \setlength\itemsep{1em}
+     \item Provide a compact way to represent a joint distribution over a random vector $X = (X_1, \ldots , X_d)$. 
+     %\item Here, each $X_i$ is a categorical random variable.
+     \item Consist of:
+     \begin{itemize}
+     \item[1.] A Directed acyclic graph $G$ which encodes the dependencies between the components of $X$.
+     \item[2.] Parameters $\theta = (\theta_1,\ldots,\theta_d)$, where $\theta_i$ denotes the parameters of the conditional distribution of $X_i$ given its parents $G_i$. 
+     \end{itemize}
+     \item Decomposition: $P(X \mid G, \theta) = \prod_{i=1}^dP(X_i\mid G_i,\theta_i)$
+     \end{itemize}     
    \end{column}
@@
    \begin{column}{0.005\textwidth}\linethickness{0.3ex} % separator {{{
-      \color{myPurple} \line(0,1){400}
+      \color{myPurple} \line(0,1){450}
    \end{column} % }}}
@@
    \begin{column}{0.3\textwidth} % second column {{{
-     \heading{Subsection 2}
+     \heading{Structure Learning}
+     \vspace*{12pt}
+     \begin{itemize}
+     \setlength\itemsep{1em}
+     \item[Data:] Each $X_i$ is a categorical variable. We observe $n$ independent samples of $X$ which are collected in a data matrix $D$ of size $n\times d$. 
+     \item[Goal:] We consider a \textit{score-based} approach and seek a graph $G$ that maximizes a scoring function which evaluates how well a given graph fits the observed data.
+     \item Some scoring functions: BIC, BDeu and fNML. 
+   	 \end{itemize}
    \end{column}
@@
    \begin{column}{0.005\textwidth}\linethickness{0.3ex} % separator {{{
-      \color{myPurple} \line(0,1){400}
+      \color{myPurple} \line(0,1){450}
    \end{column} % }}}
    \begin{column}{0.3\textwidth}% third column {{{
-     \heading{Subsection 3}
+     \heading{Scoring Functions}
+     \begin{itemize}
+
+	 \vspace*{12pt}
+	 \setlength\itemsep{1em}     
+     \item[BDeu] Bayesian marginal likelihood based on Dirichlet priors. Depends on a single 
+     hyperparameter $\alpha >0$ called equivalent sample size. 
+     \item[BIC] Maximized likelihood with penalty $\frac{k}{2}\log n$, where $k$ is the number of free parameters in the network. 
+     \item[fNML] Factorized Normalized Maximum likelihood. Maximized likelihood with penalty defined via \textit{regret} functions.
+     \end{itemize}
 
@@
 
-\begin{block}{Section 2}
+\begin{block}{Quotient Normalized Maximum Likelihood Criterion}
 \begin{columns}[T]
-   \begin{column}{0.25\textwidth} % first column {{{
-	\heading{Subsection 1}
+   \begin{column}{0.18\textwidth} % first column {{{
+   \heading{Motivation}
+   \vspace*{12pt}
+   	\textcolor{teal}{BDeu}
+	\begin{itemize}	    
+    \item Is very sensitive to the choice of the hyperparameter $\alpha$ \cite{cosco.uai07}.
+    \item Is not \textit{regular} \cite{Suzuki2017} (can be shown in certain situations to favour too complex models over simpler) 
+    \end{itemize}
+    \vspace*{12pt}
+    \textcolor{teal}{fNML}
+	\begin{itemize}	    
+    \item Is not score equivalent: the graphs expressing the same independence statements are not scored equally.
+    \item Learned structures are often rather complex, which hampers their interpretation.
+    \end{itemize}
+
+    %\vspace*{12pt} 
+	%\textcolor{HIITgreen}{Figure:} Number of parameters as a function of sample size for UCI Breast Cancer  data.
+
+    \end{column}
+
+   \begin{column}{0.18\textwidth}
+   \vphantom{\large Motivation}
+   \vspace*{12pt} 
+    \textcolor{teal}{BIC}
+    \begin{itemize}
+    \item Appears to
+require large sample sizes in order to identify appropriate
+structures \cite{cosco.pgm08a,Liu2012}. 
+	\end{itemize}
+
+	\begin{figure}
+  	\includegraphics[width=\textwidth]{breast_cancer_npmean.pdf}
+  	\caption{Number of parameters as a function of sample size for Breast Cancer (UCI) data.}
+  	\end{figure}
 
@@
    \begin{column}{0.3\textwidth} % second column {{{
-	\heading{Subsection 2}
+	\heading{qNML: Definition}
+	\vspace*{12pt}
+	\begin{itemize}
+	\item We would like to choose a model $G$ maximizing	
+	\begin{equation}
+	P_{NML}(D;G)=\frac{P(D \mid \hat\theta , G)}{\sum_{D'}{P(D' \mid \hat  \theta,G)}},
+	\end{equation}where $\hat\theta$ denotes the maximum likelihood parameters.
+	\end{itemize}
 
@@
 
-   \begin{column}{0.3\textwidth}% third column {{{
+   \begin{column}{0.25\textwidth}% third column {{{
 
-      \heading{Subsection 3}
+      \heading{qNML: Properties}
 
@@
 
-\begin{block}{Section 3}
+\begin{block}{Experimental Results}
   \begin{columns}[T]
     \begin{column}{0.35\textwidth}
-      \heading{Subsection 1}
+      \heading{Structure Learning}
 
@@
    \begin{column}{0.005\textwidth}\linethickness{0.3ex}
-      \color{myPurple} \line(0,1){800}
+      \color{myPurple} \line(0,1){700}
    \end{column} % }}}
@@
     \begin{column}{0.23\textwidth}
-    \heading{Subsection 2}
+    \heading{Prediction}
 
@@
     \begin{column}{0.005\textwidth}\linethickness{0.3ex} % separator {{{
-      \color{myPurple} \line(0,1){800}
+      \color{myPurple} \line(0,1){700}
    \end{column} % }}}
     \begin{column}{0.3\textwidth}
-    \heading{Subsection 3}
+    \heading{Conclusions}
 
@@
 {
-[1] 
+\bibliographystyle{apalike}
+\bibliography{../cosco}
 }
@@
 {
-..........
+J.L., E.J. and T.R. were supported by the Academy of Finland (COIN CoE and Project \textsc{TensorML}). J.L. was supported by the DoCS doctoral programme of the University of Helsinki.
 }

### 2018-03-19 poster/qnmlPoster.tex
more poster

@@
 \usepackage[orientation=portrait,size=a0,scale=1.4,debug]{beamerposter}
+\usepackage{blkarray}        
 %\usepackage[orientation=portrait,width = 91 ,height = 122 ,scale=1.39,debug]{beamerposter} 
@@
      \end{itemize}
-     \item Decomposition: $P(X \mid G, \theta) = \prod_{i=1}^dP(X_i\mid G_i,\theta_i)$
+     \item Decomposition: $P(X \mid G, \theta) = \prod_{i=1}^dP(X_i\mid X_{G_i},\theta_i)$
      \end{itemize}     
@@
      \begin{itemize}
-
 	 \vspace*{12pt}
@@
      hyperparameter $\alpha >0$ called equivalent sample size. 
-     \item[BIC] Maximized likelihood with penalty $\frac{k}{2}\log n$, where $k$ is the number of free parameters in the network. 
-     \item[fNML] Factorized Normalized Maximum likelihood. Maximized likelihood with penalty defined via \textit{regret} functions.
+     \item[BIC] Maximized log-likelihood with penalty $\frac{k}{2}\log n$, where $k$ is the number of free parameters in the network. 
+     \item[fNML] Factorized Normalized Maximum likelihood. Maximized log-likelihood with penalty defined via \textit{regret} functions.
      \end{itemize}
@@
    \begin{column}{0.3\textwidth} % second column {{{
-	\heading{qNML: Definition}
+	\heading{qNML: Idea}
 	\vspace*{12pt}
 	\begin{itemize}
-	\item We would like to choose a model $G$ maximizing	
+	\item Following the \textit{MDL-principle}, we would like to pick a model $G$ maximizing
+
 	\begin{equation}
-	P_{NML}(D;G)=\frac{P(D \mid \hat\theta , G)}{\sum_{D'}{P(D' \mid \hat  \theta,G)}},
-	\end{equation}where $\hat\theta$ denotes the maximum likelihood parameters.
+	\log P_{NML}(D;G) = \underbrace{\log P(D \mid \hat\theta , G)}_{\text{Maximized log-likelihood}} - \underbrace{\log \sum_{D'}P(D' \mid \hat  \theta,G)}_{:= \text{ regret.
+} \textbf{Intractable!}}, \notag
+	\end{equation}%where $\hat\theta$ denotes the maximum likelihood parameters.
+
+	\item However, for \textbf{a single categorical data vector}, regret and one dimensional $P^1_{NML}$-code can be computed.
+	%\item In $P^1_{NML}$, regret depends on $N$, the length of the categorical data vector $D_i$ and $r_i$, the number of possible values for corresponding variable $X_i$. 
+	%\item Justification from \textit{Minimum description length} principle.
+	\item \textbf{Trick}. Recall factorization w.r.t. graph $G$:
+	$$
+	P(D\mid G) = \prod_{i = 1}^{d}P(D_i \mid D_{G_i}) = \prod_{i = 1}^{d}\frac{P(D_{i,G_i})}{P(D_{G_i})}
+	$$
+	\item Assume there are \textbf{no independencies} among $G_i$. Now, $D_{i,G_i} = $
+	\vspace*{12pt}
 	\end{itemize}
+	\begin{minipage}{0.45\textwidth}
+	\begin{align*}
+	\begin{blockarray}{c|ccc}
+	D_i\phantom{s} & \phantom{..}  & D_{G_i} & \phantom{..}\\
+	\begin{block}{(c|ccc)}
+  	1 & 1 & 1 & 1  \\
+  	0 & 1 & 0 & 0 \\
+  	1 & 1 &0 & 0  \\
+  	1 & 1 & 1 & 1  \\
+	\end{block}
+	\end{blockarray} 
+	\ \to \ 
+	\begin{blockarray}{c}
+	\phantom{...}\\
+	\begin{block}{(c)}
+  	0   \\
+  	1  \\
+  	2  \\
+  	0   \\
+	\end{block}
+	\end{blockarray}
+ 	\end{align*}
+	\end{minipage}\quad
+	\begin{minipage}{0.45\textwidth}
+	\fbox{\parbox{\textwidth}{
+	\begin{itemize}
+	\item We can treat $D_{i,G_i}$ and $D_{G_i}$ as values of single variables!
+	\item $P(D_{i,G_i})$ and $P(D_{i,G_i})$ can be modeled using $P_{NML}^1$-code.
+	\end{itemize}}}  
+	\end{minipage}
 
@@
 
-      \heading{qNML: Properties}
+   	\heading{qNML: Definition \& Properties}
+   	Definition of qNML score:
+    	\begin{empheq}[box={\mybluebox[4pt][4pt]}]{equation}	    		
+	s^{qNML} := \sum_{i = 1}^d \log \frac{P^1_{NML}(D_{i,G_i})}{P^1_{NML}(D_{G_i})},\notag
+	\end{empheq}$\log P^1_{NML}(D) = \log P(D \mid \hat{\theta}) - reg(n,r).$
 
-   \end{column} % end of first column }}}
-   % \begin{column}{0.005\textwidth}\linethickness{0.3ex} % separator {{{
-   %    \color{myPurple} \line(0,1){600}
-   % \end{column} % }}}
-   % \begin{column}{0.3\textwidth} % third column {{{
-   % \end{column} % end of third column }}}
+	\begin{itemize}
+	 \item $D$ is understood as  $n$ observations on single categorical variable with $r$ possible values.
+	\item Accurate approximation for $reg(n,r)$ from \cite{Szpankowski2012}:
+	\begin{align*}
+    reg(n, r) &\approx  n\left(\log{\alpha} + (\alpha + 2) \log{C_\alpha} - \frac{1}{C_\alpha}\right) \\ 
+    &- \frac{1}{2} \log{\left(C_\alpha + \frac{2}{\alpha}\right)},
+	\end{align*}
+where $\alpha = \frac{r}{n}$ and $C_\alpha = \frac{1}{2} + \frac{1}{2} \sqrt{1 + \frac{4}{\alpha}}$.
+	\end{itemize}
+	\vspace*{12pt}
+	qNML score is:
+	\fbox{\parbox{0.95\textwidth}{
+	\begin{itemize}	
+	\item \textcolor{teal}{\textbf{hyper-parameter free}}
+	\item \textcolor{teal}{\textbf{consistent}}
+	\item \textcolor{teal}{\textbf{score equivalent}}
+	\item \textcolor{teal}{\textbf{regular}}
+	\item qNML agrees with NML for every network composed of fully connected sub-graphs.
+	\end{itemize}}}
+
+   \end{column}
 
@@
 
+	  \begin{minipage}{0.45\textwidth}
+	  \begin{figure}[h]
+
+		\includegraphics[width=\columnwidth]{shd_all.pdf}
+		\caption{SHD as a function of sample size.}
+		\label{fig:all_shd}
+		\end{figure}
+	  \end{minipage}
+	  \begin{minipage}{0.45\textwidth}
+	  \begin{itemize}
+	  \item Data generated from real world DAG structures.
+	  \item Exact DP structure learning algorithm.
+	  \end{itemize}
+	  \begin{figure}[h]
+		\includegraphics[width=\columnwidth]{shd_rank_all.pdf}
+		\caption{Average ranks for each score.}
+		\label{fig:all_shd}
+		\end{figure}
+	  \end{minipage}
+
+	\vspace*{12pt}
+	\textbf{qNML seems overall a safe choice in structure learning, especially with moderate and large sample sizes!} 
     \end{column}
    \begin{column}{0.005\textwidth}\linethickness{0.3ex}
-      \color{myPurple} \line(0,1){700}
+      \color{myPurple} \line(0,1){600}
    \end{column} % }}}
@@
     \begin{column}{0.005\textwidth}\linethickness{0.3ex} % separator {{{
-      \color{myPurple} \line(0,1){700}
+      \color{myPurple} \line(0,1){600}
    \end{column} % }}}

### 2018-03-20 poster/qnmlPoster.tex
poster updated

@@
 \usepackage{fancybox}
-\usepackage{tikz}
-\usepackage{mathtools}
+\setlength{\fboxsep}{12pt}
+%\usepackage{mathtools}
 \usepackage{empheq}
@@
 
-   \begin{column}{0.3\textwidth} % second column {{{
+   \begin{column}{0.25\textwidth} % second column {{{
      \heading{Structure Learning}
@@
 \begin{columns}[T]
-   \begin{column}{0.18\textwidth} % first column {{{
+
+	\begin{column}{0.34\textwidth} % first column {{{
    		\heading{Motivation}
+   		\begin{minipage}{0.49\textwidth}
    			\vspace*{12pt}
@@
 
+   		\end{minipage}
+   		\begin{minipage}{0.49\textwidth}		   
+		   \phantom{why vspace* does nothing??}
 		   %\vspace*{12pt} 
-	%\textcolor{HIITgreen}{Figure:} Number of parameters as a function of sample size for UCI Breast Cancer  data.
-
-    \end{column}
-
-   \begin{column}{0.18\textwidth}
-   \vphantom{\large Motivation}
-   \vspace*{12pt} 
 		   \textcolor{teal}{BIC}
@@
 
+   		\end{minipage}
+
+   		\vspace*{12pt}
+
+   		\shadowbox{
+		\parbox{0.97\textwidth}{
+		\vspace*{12pt}
+		The quest for a model selection criterion that would yield \textbf{more parsimonious}, \textbf{easier to interpret}, but still \textbf{predictive} Bayesian networks structures is one of the main motivations for this work.
+		\vspace*{12pt}
+		}}
+
    	\end{column}
@@
    \begin{column}{0.005\textwidth}\linethickness{0.3ex} % separator {{{
-      \color{myPurple} \line(0,1){650}
+      \color{myPurple} \line(0,1){810}
    \end{column} % }}}
@@
 	\begin{minipage}{0.45\textwidth}
-	\fbox{\parbox{\textwidth}{
+	\shadowbox{
+	\parbox{\textwidth}{
+	\vspace*{12pt}
 	\begin{itemize}
@@
    	\heading{qNML: Definition \& Properties}
+   	\vspace*{12pt}
    	Definition of qNML score:
@@
 	\end{itemize}
+
 	\vspace*{12pt}
-	qNML score is:
-	\fbox{\parbox{0.95\textwidth}{
+	\shadowbox{\parbox{0.85\textwidth}{
+	\vspace*{12pt}
+	\textbf{qNML score is:}
 	\begin{itemize}	
@@
 	\item \textcolor{teal}{\textbf{regular}}
-	\item qNML agrees with NML for every network composed of fully connected sub-graphs.
+	\item \textcolor{teal}{\textbf{it equals NML for many models.}}
+	%\item qNML agrees with NML for every network composed of fully connected sub-graphs.
 	\end{itemize}}}
@@
 	\vspace*{12pt}
-	\textbf{qNML seems overall a safe choice in structure learning, especially with moderate and large sample sizes!} 
+	%\textbf{qNML seems overall a safe choice in structure learning, especially with moderate and large sample sizes!} 
     \end{column}
    \begin{column}{0.005\textwidth}\linethickness{0.3ex}
-      \color{myPurple} \line(0,1){600}
+      \color{myPurple} \line(0,1){590}
    \end{column} % }}}
@@
     \begin{column}{0.005\textwidth}\linethickness{0.3ex} % separator {{{
-      \color{myPurple} \line(0,1){600}
+      \color{myPurple} \line(0,1){590}
    \end{column} % }}}

### 2018-03-20 poster/qnmlPoster.tex
first poster version ready

@@
 }
-%\graphicspath{{imgs/}}
+
 \usepackage{amsmath,amsthm, amssymb,mathrsfs}
@@
 \usepackage{subfigure}
+\usepackage{booktabs}
 \usepackage{bm}
@@
 \usepackage[numbers]{natbib}
-\usepackage[orientation=portrait,size=a0,scale=1.4,debug]{beamerposter}
+\usepackage[orientation=portrait,size=a0,scale=1.34,debug]{beamerposter}
 \usepackage{blkarray}        
@@
 
-\definecolor{DRed}{rgb}{0.8,0.6,0.6}
-\definecolor{DGreen}{rgb}{0.6,0.7,0.6}
-\definecolor{BRed}{rgb}{0.7,0,0}
-\definecolor{BGreen}{rgb}{0,0.7,0}
-
-\newcommand{\myred}[1]{{\color{BRed}#1}}
-\newcommand{\mygreen}[1]{{\color{BGreen}#1}}
-
 \newcommand{\heading}[1]{\alert{\large #1}\\}
 \definecolor{myPurple}{RGB}{174, 206, 195}
-%\definecolor{myPurple}{RGB}{165,206,190}
 
-\usepackage{textpos}
 \usepackage{fancybox}
 \setlength{\fboxsep}{12pt}
-%\usepackage{mathtools}
+
 \usepackage{empheq}
 
-\theoremstyle{plain}
-\newtheorem{assumption}{Assumption}
-
 \definecolor{myblue}{rgb}{.85, .85, 1.0}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\makeatletter
-\newcommand\mathcircled[1]{%
-  \mathpalette\@mathcircled{#1}%
-}
-\newcommand\@mathcircled[2]{%
-  \tikz[baseline=(math.base)] \node[red,draw,circle,inner sep=1pt] (math) {$\m@th#1#2$};%
-}
-\makeatother
 
@@
      \item Provide a compact way to represent a joint distribution over a random vector $X = (X_1, \ldots , X_d)$. 
-     %\item Here, each $X_i$ is a categorical random variable.
      \item Consist of:
@@
    \begin{column}{0.005\textwidth}\linethickness{0.3ex} % separator {{{
-      \color{myPurple} \line(0,1){450}
+      \color{myPurple} \line(0,1){430}
    \end{column} % }}}
@@
      \item[BIC] Maximized log-likelihood with penalty $\frac{k}{2}\log n$, where $k$ is the number of free parameters in the network. 
-     \item[fNML] Factorized Normalized Maximum likelihood. Maximized log-likelihood with penalty defined via \textit{regret} functions.
+     \item[fNML] Factorized Normalized Maximum Likelihood. Maximized log-likelihood with penalty defined via \textit{regret} functions.
      \end{itemize}
@@
    \begin{column}{0.005\textwidth}\linethickness{0.3ex} % separator {{{
-      \color{myPurple} \line(0,1){810}
+      \color{myPurple} \line(0,1){820}
    \end{column} % }}}
@@
 	\item However, for \textbf{a single categorical data vector}, regret and one dimensional $P^1_{NML}$-code can be computed.
-	%\item In $P^1_{NML}$, regret depends on $N$, the length of the categorical data vector $D_i$ and $r_i$, the number of possible values for corresponding variable $X_i$. 
-	%\item Justification from \textit{Minimum description length} principle.
 	\item \textbf{Trick}. Recall factorization w.r.t. graph $G$:
@@
 	\begin{align*}
-    reg(n, r) &\approx  n\left(\log{\alpha} + (\alpha + 2) \log{C_\alpha} - \frac{1}{C_\alpha}\right) \\ 
-    &- \frac{1}{2} \log{\left(C_\alpha + \frac{2}{\alpha}\right)},
+    reg(n, r) &\approx  n\left(\log{\beta} + (\beta + 2) \log{C_\beta} - \frac{1}{C_\beta}\right) \\ 
+    &- \frac{1}{2} \log{\left(C_\beta + \frac{2}{\beta}\right)},
 	\end{align*}
-where $\alpha = \frac{r}{n}$ and $C_\alpha = \frac{1}{2} + \frac{1}{2} \sqrt{1 + \frac{4}{\alpha}}$.
+where $\beta = \frac{r}{n}$ and $C_\beta = \frac{1}{2} + \frac{1}{2} \sqrt{1 + \frac{4}{\beta}}$.
 	\end{itemize}
@@
 	\begin{itemize}	
-	\item \textcolor{teal}{\textbf{hyper-parameter free}}
-	\item \textcolor{teal}{\textbf{consistent}}
-	\item \textcolor{teal}{\textbf{score equivalent}}
-	\item \textcolor{teal}{\textbf{regular}}
-	\item \textcolor{teal}{\textbf{it equals NML for many models.}}
-	%\item qNML agrees with NML for every network composed of fully connected sub-graphs.
+	\item {\textbf{hyper-parameter free}}
+	\item {\textbf{consistent}}
+	\item {\textbf{score equivalent}}
+	\item {\textbf{regular}}
+	\item {\textbf{it equals NML for many models.}}
 	\end{itemize}}}
@@
   \begin{columns}[T]
-    \begin{column}{0.35\textwidth}
+    \begin{column}{0.4\textwidth}
       \heading{Structure Learning}
 
-	  \begin{minipage}{0.45\textwidth}
+      \begin{minipage}{0.3\textwidth}
+
+      \begin{itemize}
+	  \item Data generated from real world DAG structures.
+	  \item $5$ networks, $d  = 6,5,11,5$ and $8$, \\ edges: $ 6,4,17,4$ and $8$. 
+	  \item Networks learned using exact DP-algorithm.
+	  \item We measure Structural Hamming Distance (SHD).
+	  %\item Average ranks (over $5$ networks and $1000$ tests) also shown.
+	  \end{itemize}
+
+	  \end{minipage}
+      \begin{minipage}{0.35\textwidth}
+      	\phantom{dsaasdasdasdasd}
+		\begin{itemize}
+	  	\item Average ranks (over $5$ networks and $1000$ tests) also shown.
+	  	\end{itemize}
+
 	  \begin{figure}[h]
-
 		\includegraphics[width=\columnwidth]{shd_all.pdf}
@@
 	  \end{minipage}
-	  \begin{minipage}{0.45\textwidth}
-	  \begin{itemize}
-	  \item Data generated from real world DAG structures.
-	  \item Exact DP structure learning algorithm.
-	  \end{itemize}
+	  \begin{minipage}{0.33\textwidth}
 	  \begin{figure}[h]
@@
 		\caption{Average ranks for each score.}
-		\label{fig:all_shd}
 		\end{figure}
+
+		\shadowbox{\parbox{0.95\textwidth}{
+		qNML:
+		\begin{itemize}	
+		\item obtains the best ranking when $n > 300$.
+		\item never has the worst average ranking.
+		\end{itemize}}}
+
 	  \end{minipage}
@@
 	\vspace*{12pt}
-	%\textbf{qNML seems overall a safe choice in structure learning, especially with moderate and large sample sizes!} 
     \end{column}
    \begin{column}{0.005\textwidth}\linethickness{0.3ex}
-      \color{myPurple} \line(0,1){590}
+      \color{myPurple} \line(0,1){630}
    \end{column} % }}}
 
-    \begin{column}{0.23\textwidth}
+    \begin{column}{0.55\textwidth}
     \heading{Prediction}
+
+    \vspace*{12pt}
+	\begin{minipage}{0.3\textwidth}
+	\vspace*{12pt}
+	\begin{itemize}
+	\item 20 UCI data sets split to train and test sets.
+	\item Used training fractions: $10\%, 20\% , \ldots, 90\%$.
+	%\item Graphs learned using the exact algorithm.
+	\item BIC, fNML and qNML use sequential predictive (sNML) parameters.
+	\item BDeu uses Bayesian predictive parameter (BPP) values with $\alpha = 1$. 
+	\item Scores are ranked from 1 to 4 according to test set log-loss.
+	\end{itemize}	
+	\end{minipage}
+	\begin{minipage}{0.35\textwidth}	
+		\centering
+		\tiny{
+		\textcolor{HIITgreen}{Table:} Average predictive performance rank over \\ different sample sizes for different scores. \\ \textbf{Best}, \underline{worst}.    
+	    \begin{table}
+		%\caption{Average predictive performance rank over different sample sizes for different model selection criteria in 20 different data sets.}
+		\label{tbl:preds}
+		\begin{center}
+		\begin{tabular}{crrrrr}
+		       Data
+		    & $n$
+		    & \multicolumn{1}{p{0.7cm}}{\centering BDeu}
+		    & \multicolumn{1}{p{0.8cm}}{\centering BIC}
+		    & \multicolumn{1}{p{0.9cm}}{\centering fNML}
+		    & \multicolumn{1}{p{0.9cm}}{\centering qNML}\\
+		       \midrule
+		 PostOpe &     90 &              2.79 &     \textbf{1.20} &  \underline{3.06} &           2.94 \\
+		    Iris &    150 &  \underline{2.82} &              2.37 &     \textbf{2.27} &           2.54 \\
+		    Wine &    178 &  \underline{3.23} &     \textbf{1.88} &              2.67 &           2.22 \\
+		   Glass &    214 &  \underline{3.61} &              3.09 &     \textbf{1.42} &           1.88 \\
+		 Thyroid &    215 &              2.55 &  \underline{3.21} &     \textbf{1.80} &           2.44 \\
+		 HeartSt &    270 &  \underline{3.12} &     \textbf{1.39} &              3.12 &           2.37 \\
+		 BreastC &    286 &  \underline{3.09} &     \textbf{1.41} &              2.97 &           2.53 \\
+		 HeartHu &    294 &  \underline{3.18} &     \textbf{1.66} &              2.90 &           2.27 \\
+		 HeartCl &    303 &  \underline{3.46} &     \textbf{1.38} &              2.99 &           2.17 \\
+		   Ecoli &    336 &              3.20 &  \underline{3.53} &     \textbf{1.24} &           2.04 \\
+		   Liver &    345 &  \underline{3.17} &              2.39 &              2.69 &  \textbf{1.75} \\
+		 Balance &    625 &  \underline{3.35} &              1.91 &     \textbf{1.59} &           3.16 \\
+		 BcWisco &    699 &  \underline{3.06} &              2.03 &              2.89 &  \textbf{2.02} \\
+		 Diabete &    768 &  \underline{2.91} &              2.70 &              2.68 &  \textbf{1.71} \\
+		 TicTacT &    958 &  \underline{3.44} &              2.71 &     \textbf{1.31} &           2.53 \\
+		   Yeast &   1484 &              2.60 &  \underline{3.76} &     \textbf{1.55} &           2.10 \\
+		 Abalone &   4177 &              2.60 &  \underline{3.64} &     \textbf{1.04} &           2.72 \\
+		 PageBlo &   5473 &              2.24 &  \underline{3.61} &     \textbf{1.31} &           2.83 \\
+		   Adult &  32561 &              3.23 &  \underline{3.77} &     \textbf{1.00} &           2.00 \\
+		 Shuttle &  58000 &     \textbf{1.44} &  \underline{3.78} &              1.56 &           3.22 \\
+		\midrule		 
+		 \textcolor{teal}{Mean} &  \phantom{} & \underline{2.95}  & 2.57 & \textbf{2.10}  &2.37 \\
+		 \textcolor{teal}{SD}  &  \phantom{} & 0.49 & \underline{0.90} &0.76 & \textbf{0.43} \\
+		\end{tabular}
+		\end{center}
+		\end{table}}	
+	\end{minipage}
+	\begin{minipage}{0.3\textwidth}
+
+	\shadowbox{\parbox{0.95\textwidth}{
+	\begin{itemize}
+	\item fNML obtains on average the best rank with qNML as the runner-up.	
+	\item qNML has the lowest standard deviation. 
+	\end{itemize}}}
+
+	We also studied the number of parameters in learned models when using $10\%$ samples for structure learning.
+	\vspace*{12pt}
+
+	\shadowbox{\parbox{0.95\textwidth}{
+	\begin{itemize}
+	\item qNML almost always yields more parsimonious models than fNML, but more complex than BIC.	
+	\end{itemize}}}
+	\end{minipage}
 
@@
 
-    \begin{column}{0.005\textwidth}\linethickness{0.3ex} % separator {{{
-      \color{myPurple} \line(0,1){590}
-   \end{column} % }}}
-    \begin{column}{0.3\textwidth}
-    \heading{Conclusions}
-
-    \end{column}
   \end{columns}

### 2018-03-26 poster/qnmlPoster.tex
polishing

@@
      \item[1.] A Directed acyclic graph $G$ which encodes the dependencies between the components of $X$.
-     \item[2.] Parameters $\theta = (\theta_1,\ldots,\theta_d)$, where $\theta_i$ denotes the parameters of the conditional distribution of $X_i$ given its parents $G_i$. 
+     \item[2.] Parameters $\theta = (\theta_1,\ldots,\theta_d)$, where $\theta_i$ denotes the parameters of the conditional distribution of $X_i$ given its parents $X_{G_i}$. 
      \end{itemize}
@@
 	$$
-	\item Assume there are \textbf{no independencies} among $G_i$. Now, $D_{i,G_i} = $
+	\item Assume there are \textbf{no independencies} among $X_{G_i}$. Now, $D_{i,G_i} = $
 	\vspace*{12pt}
@@
 %\scriptsize
-\tiny
-{
-\bibliographystyle{apalike}
-\bibliography{../cosco}
-}
+%\tiny
+%{
+%\bibliographystyle{apalike}
+%\bibliography{../cosco}
+%}
+
+\begin{tiny}
+\begin{thebibliography}{}
+\bibitem[Liu et~al., 2012]{Liu2012}
+Liu, Z. et~al. (2012).
+\newblock Empirical evaluation of scoring functions for {B}ayesian network
+  model selection.
+\newblock {\em BMC Bioinformatics}, 13(15):S14.
+
+\bibitem[Silander, T. et~al., 2007]{cosco.uai07}
+Silander et~al. (2007).
+\newblock On sensitivity of the {MAP} {Bayesian} network structure to the
+  equivalent sample size parameter.
+\newblock In {\em Proceedings of {UAI--07}}.
+
+\bibitem[Silander, T. et~al., 2008]{cosco.pgm08a}
+Silander et~al. (2008).
+\newblock Factorized normalized maximum likelihood criterion for learning
+  {B}ayesian network structures.
+\newblock In {\em Proceedings of {PGM-08}}.
+
+\bibitem[Suzuki, 2017]{Suzuki2017}
+Suzuki, J. (2017).
+\newblock A theoretical analysis of the {BDeu} scores in {B}ayesian network
+  structure learning.
+\newblock {\em Behaviormetrika}, 44(1):97--116.
+
+\bibitem[Szpankowski and Weinberger, 2012]{Szpankowski2012}
+Szpankowski, W. and Weinberger, M.~J. (2012).
+\newblock Minimax pointwise redundancy for memoryless models over large
+  alphabets.
+\newblock {\em IEEE Transactions on Information Theory}, 58(7):4094--4104.
+\end{thebibliography}
+\end{tiny}
+
 \end{block} %}}}

### 2024-08-27 gendata.sh
A version uploaded to arXiv

@@ -1,13 +1,0 @@
-for bnmfn in artnets/*.bnm; do
-    bbnm=`basename $bnmfn .bnm`
-    g=`echo $bbnm | cut -d'_' -f2`
-    idtdn=artdats/$bbnm
-    mkdir -p $idtdn
-    vdfn=artdats/data_$g.vd
-    echo $bbnm
-    for j in `seq 0 99`; do
-	idtfn=$idtdn/$j.idt
-	python -m bn.model.bn_gendata -o $idtfn $vdfn $bnmfn 1000
-    done
-done
-

### 2024-08-27 make_all_4.sh

@@ -1,23 +1,0 @@
-pdfunite \
-qNML_images/adult_mean.pdf qNML_images/adult_var.pdf qNML_images/adult_np_mean.pdf qNML_images/adult_np_var.pdf \
-qNML_images/wine_mean.pdf qNML_images/wine_var.pdf qNML_images/wine_np_mean.pdf qNML_images/wine_np_var.pdf \
-qNML_images/heart_statlog_mean.pdf qNML_images/heart_statlog_var.pdf qNML_images/heart_statlog_np_mean.pdf qNML_images/heart_statlog_np_var.pdf \
-qNML_images/heart_hungarian_mean.pdf qNML_images/heart_hungarian_var.pdf qNML_images/heart_hungarian_np_mean.pdf qNML_images/heart_hungarian_np_var.pdf \
-qNML_images/heart_cleveland_mean.pdf qNML_images/heart_cleveland_var.pdf qNML_images/heart_cleveland_np_mean.pdf qNML_images/heart_cleveland_np_var.pdf \
-qNML_images/page_blocks_mean.pdf qNML_images/page_blocks_var.pdf qNML_images/page_blocks_np_mean.pdf qNML_images/page_blocks_np_var.pdf \
-qNML_images/glass_mean.pdf qNML_images/glass_var.pdf qNML_images/glass_np_mean.pdf qNML_images/glass_np_var.pdf \
-qNML_images/bc_wisconsin_mean.pdf qNML_images/bc_wisconsin_var.pdf qNML_images/bc_wisconsin_np_mean.pdf qNML_images/bc_wisconsin_np_var.pdf \
-qNML_images/tic_tac_toe_mean.pdf qNML_images/tic_tac_toe_var.pdf qNML_images/tic_tac_toe_np_mean.pdf qNML_images/tic_tac_toe_np_var.pdf \
-qNML_images/shuttle_mean.pdf qNML_images/shuttle_var.pdf qNML_images/shuttle_np_mean.pdf qNML_images/shuttle_np_var.pdf \
-qNML_images/breast_cancer_mean.pdf qNML_images/breast_cancer_var.pdf qNML_images/breast_cancer_np_mean.pdf qNML_images/breast_cancer_np_var.pdf \
-qNML_images/yeast_mean.pdf qNML_images/yeast_var.pdf qNML_images/yeast_np_mean.pdf qNML_images/yeast_np_var.pdf \
[…]

### 2024-08-27 shd_plots.py

@@ -1,264 +1,0 @@
-#!/usr/bin/env python3
-# -*- coding: utf-8 -*-
-import numpy as np
-import matplotlib.pyplot as plt
-plt.rcParams['ps.useafm'] = True
-plt.rcParams['pdf.use14corefonts'] = True
-plt.rcParams['text.usetex'] = True
-import seaborn as sns
-
-def subplots(folder = "results/real_dag_shd/", save = False):
-    
-    names = ["Survey", "Earthquake", "Sachs","Cancer","Asia"]
-    
-    filee = open(folder + names[0].lower() + ".txt", mode = 'r')
-    line1 = filee.readline()
-    filee.close()
-    
-    line1 = line1.rstrip("\n")
-    header = line1.split(" ")
-    
-    methods = header.copy()
-    methods.pop(0)    
-
-    methods = ["BIC", "BDeu", "fNML", "qNML"] 
-    
-    fig,axs = plt.subplots(nrows = 3, ncols = 2,sharex = True, figsize = (10,11) )
-    
-    axs = axs.flatten()
-    linestyles = ['--', ':', '-.', '-']
-
-    jj = 0
-    
-    for dag_name in names: 
-    
-        x = np.loadtxt(folder + dag_name.lower() + ".txt", skiprows = 1)
-        
-        ns = x[:,0]
-        shds = x[:,1::]
-
-
-        sns.set_style("ticks")
-        sns.set_palette("colorblind")
-                
-        ax = axs[jj]
-
-        ax.set(xscale = "log")
-
-        for ii in range(0,len(methods)):
-            ax.plot(ns,shds[:,ii],linestyles[ii],label = methods[ii])
-
-        jj += 1
-        
-        ax.set_title(dag_name)
-        sns.set_context("poster", font_scale = 2, rc={"lines.linewidth": 7})
-        sns.despine()
-        
-        
-    ax = axs[len(names)]
-   
-    for i in range(0,len(methods)):
-        lab = methods[i]
-        ax.plot([],[],linestyles[i],label = lab)
-        
-    ax.legend(loc="center")
-    
-    ax.set_axis_off() 
-    
-    
-    for jj in range(0,len(names)):
-        ax = axs[jj]
-    
-        
-    
-    #plt.tight_layout()
-    plt.show()
-    
-    
-    # tweak the y labels
-    for jj in range(0,len(names)):
-        ax = axs[jj]
-    
-        labels = ax.get_yticklabels()
-        
-        #print(labels[2])
[…]
